<!-- ============================================================
INSTRUCTOR CUSTOMISATION — update before rendering for each cohort

  Search "PVA2"        — update if session numbering changes
  Search "Lab 3"       — verify Lab 3 filename is correct
  Search "icebreaker"  — verify Lab 2 residual plot renders correctly
  Line ~47             — ShelveLoc factor levels: Bad < Medium < Good
                         (must match Labs 2 and 3 recoding)

============================================================ -->
---
title: "Demo: Multiple Regression"
subtitle: "CRISP-DM Phase 4 — Modeling | Carseats Case Study"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en-US
bibliography: ../../lit/my_bib.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse)   # data wrangling and ggplot2
library(moderndive)  # get_regression_table(), get_regression_points()
library(broom)       # augment(), tidy()
library(gt)          # formatted tables
library(patchwork)   # combine multiple ggplots
library(plotly)      # interactive 3D plot
library(car)         # vif()
library(ISLR2)       # Carseats and Credit datasets

tbl_carseats <- tibble(Carseats) %>%
  mutate(ShelveLoc = factor(ShelveLoc,
                            levels = c("Bad", "Medium", "Good")))

# Shared colour palette — used in all ShelveLoc plots
col_shelf       <- c("Bad" = "#D50006", "Medium" = "#686868", "Good" = "#502479")
col_shelf_resid <- c("Bad" = "#D50006", "Medium" = "#b3b2b2", "Good" = "#502479")

# Models fitted live in the document body (mod_shelf, mod_int_shelf)
# Pre-fitted here only where needed before their section:
mod_baseline  <- lm(Sales ~ Price,                           data = tbl_carseats)
mod_adv       <- lm(Sales ~ Price + Advertising,             data = tbl_carseats)
mod_int_adv   <- lm(Sales ~ Price * Advertising,             data = tbl_carseats)
mod_full      <- lm(Sales ~ Price + Advertising + ShelveLoc, data = tbl_carseats)
```

:::{.callout-note collapse="true"}
## Purpose of this document

This document serves **two functions:**

1. **For the instructor:** Annotated step-by-step guide for the live demo in RStudio.
   `r fontawesome::fa("chalkboard-teacher")` **Instructor note** callouts signal suggested questions,
   transitions, and emphasis points.

2. **For students:** Complete reference document available after the session.

**Notation (descriptive setting):**

| Symbol | Meaning |
|--------|---------|
| $b_0, b_1, \ldots, b_k$ | OLS coefficients — descriptive, computed from *these* 400 stores |
| $e_i = y_i - \hat{y}_i$ | Residual — calculated from data |
| $\hat{y}_i$ | Fitted value on the regression line / plane |

In PVA2, when we treat the 400 stores as a *sample* from a larger population,
notation shifts to $\hat\beta_j$ estimating unknown $\beta_j$, and $e_i$
approximating unobservable $\varepsilon_i$.
:::

:::{.callout-note}
## Learning Objectives

After this session you will be able to:

- **Extend** the baseline equation with a categorical regressor and explain
  why R converts it into dummy variables automatically
- **Interpret** OLS coefficients as partial derivatives (ceteris paribus)
  in both the numeric and categorical case
- **Recognise** that `geom_smooth()` with aesthetics `color` implicitly fits an
  interaction model — and test empirically whether that complexity is warranted
- **Apply** Occam's Razor via adjusted R²: choose the parallel-slopes model
  when interaction adds no explanatory power
- **Identify** confounding variables and explain how controlling for them
  corrects Omitted Variable Bias in the Price coefficient
- **Explain** Simpson's Paradox: why an aggregate slope can reverse sign
  when a third variable is controlled for
- **Distinguish** the descriptive purpose of residual diagnostics
  (misspecification detection) from the inferential purpose (distributional checks, Session 3)
:::

:::{.callout-tip}
## Navigation

| Symbol | Box colour | Meaning | When to read |
|--------|-----------|---------|-------------|
| `r fontawesome::fa("chalkboard-teacher")` | <span style="color:#2f9e44">**Green — Instructor Note**</span> | Suggested questions, transitions, emphasis points | Before class |
| ⚠️ | <span style="color:#e02424">**Red — Action Required**</span> | Key concepts, model warnings | During demo |
| ℹ️ | <span style="color:#1c7ed6">**Blue, open — Note**</span> | Interpretation aids, technical clarifications | As needed |
| 📚 | <span style="color:#1c7ed6">**Blue, collapsed — Background Theory**</span> | Deeper derivations, optional theory | At home |

**Normal text** and code chunks are the **main content** — work through them in order.
**Background sections** are marked with a **collapsed blue box** and can be expanded at any time.

⏱ This demo runs approximately **60 minutes** in the live session.

:::

---

## Icebreaker: What Does the Baseline Miss?

### Where are we in CRISP-DM?

```{r crisp-overview}
#| echo: false
tibble::tribble(
  ~Phase, ~Name,                    ~Content,                                   ~Status,
  "1",    "Business Understanding", "Identify underperforming stores",           "Done (lecture)",
  "2",    "Data Understanding",     "EDA: distributions, correlations, OVB",     "Done (Lab 1–2)",
  "3",    "Data Preparation",       "ShelveLoc dummy coding, Advertising",        "Done (Lab 2)",
  "4",    "Modelling",              "Multiple regression: Sales ~ Price + ...",   "← This demo",
  "5",    "Evaluation",             "Residuals as performance score, adj. R²",    "Lab 3",
  "6",    "Deployment",             "Underperformer ranking, Managerial Memo",     "Lab 3"
) %>%
  gt() %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = Phase == "4")
  ) %>%
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(table.font.size = "13px")
```

### The residual tells the story

In Lab 2 we estimated the simplest possible model:

$$\widehat{\text{Sales}}_i = b_0 + b_1 \cdot \text{Price}_i$$

```{r baseline-recap}
#| code-fold: false
tb <- mod_baseline %>%
  get_regression_summaries() %>%
  select(r_squared, adj_r_squared, rmse) %>%
  mutate(Model = "Sales ~ Price") %>%
  relocate(Model) %>%
  gt()
```
```{r baseline-recap-fmt}
#| code-fold: true
#| code-summary: "Show code: table formatting"
tb %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse), decimals = 3) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header("Baseline — Lab 2")
```

The residual $e_i = \text{Sales}_i - \hat{y}_i$ is both a **diagnostic** and a **scoring tool:**

- **Diagnostic:** systematic patterns in $e_i$ reveal omitted variables
- **Scoring:** $e_i < 0$ = store sells less than its price-based benchmark → underperformer

```{r icebreaker-resid}
#| code-fold: false
#| fig-height: 3.2
gg <- augment(mod_baseline, data = tbl_carseats) %>%
  ggplot(aes(.fitted, .resid, color = ShelveLoc)) +
  geom_point(alpha = 0.45, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#333") +
  geom_smooth(aes(group = 1), method = "loess", se = FALSE,
              color = "black", linewidth = 0.9)
```
```{r icebreaker-resid-fmt}
#| code-fold: true
#| code-summary: "Show code: labels & theme"
#| fig-height: 3.2
gg +
  scale_color_manual(values = col_shelf_resid) +
  labs(title    = "Residual plot — baseline model, coloured by ShelveLoc",
       subtitle = "Purple (Good) above zero | Red (Bad) below zero | LOESS not flat → misspecification",
       x = "Fitted values", y = "Residual", color = "ShelveLoc") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

:::{.callout-tip}
## `r fontawesome::fa("chalkboard-teacher")` Instructor note — Icebreaker

Point to the colour separation:
*"Purple dots sit above zero, red dots below — systematically.
`ShelveLoc` explains part of the residual that our model ignores.
The obvious fix: add `ShelveLoc` to the equation."*
:::

---

## The General Model

For $k$ regressors, the regression model (descriptive model notation) is:

$$y_i = b_0 + b_1 x_{i1} + b_2 x_{i2} + \dots + b_k x_{ik} + e_i$$

OLS minimises $\sum_{i=1}^n e_i^2$.
Because the model is linear, each coefficient is the partial derivative of the
fitted surface — the slope in one direction while all others are held fixed:

$$b_j = \frac{\partial \hat{y}}{\partial x_j}
\quad \text{(ceteris paribus)}$$

This is a **mathematical** statement about the slope of the fitted surface.
It is **not** a causal claim.

### What combinations of regressors will we see?

Multiple regression handles regressors of different scale levels —
the combination determines both the geometry and the interpretation.
This demo works through two cases in order of increasing insight, then
points to what happens with more than two regressors.

```{r regressor-overview}
#| echo: false
tibble::tribble(
  ~`Case`,       ~`Regressors`,                    ~`Geometry`,           ~`This demo`,
  "Baseline",    "1 numeric (Price)",               "Line in 2D",          "Lab 2 — starting point",
  "Case A",      "1 numeric + 1 categorical\n(Price + ShelveLoc)", "Parallel lines in 2D\n(one per group)", "← Section 3",
  "Case B",      "2 numeric\n(Price + Advertising)", "Plane in 3D",        "← Section 4",
  "Full model",  "2 numeric + 1 categorical\n(Price + Adv + ShelveLoc)", "Parallel planes in 3D", "← Section 5",
  "General k>2", "k regressors (any mix)",           "Hyperplane — not\nvisualisable", "Same algebra, PVA2"
) %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#f0fff0"),
            locations = cells_body(rows = 2)) %>%
  tab_style(style = cell_fill(color = "#fff8e1"),
            locations = cells_body(rows = 3)) %>%
  tab_style(style = cell_fill(color = "#f5f0fa"),
            locations = cells_body(rows = 4)) %>%
  tab_style(style = cell_fill(color = "#f5f5f5"),
            locations = cells_body(rows = 5)) %>%
  tab_style(style = cell_fill(color = "#f5f0fa"),
            locations = cells_column_labels()) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(
    "Regressor combinations covered in this demo",
    subtitle = "Case A first — ShelveLoc explains nearly twice as much variance as Advertising"
  )
```

:::{.callout-note}
## Why Case A before Case B?

A quick empirical check before we start:

- `Sales ~ Price + ShelveLoc` → adj. R² ≈ **0.54**
- `Sales ~ Price + Advertising` → adj. R² ≈ **0.28**

The categorical variable `ShelveLoc` explains almost twice as much variance
as the numeric variable `Advertising`. The modelling sequence follows the data —
not textbook convention.
:::

::: {.callout-caution collapse="true"}
## Background: Matrix notation (optional)

For $n$ observations and $k$ regressors (including an intercept column of ones):

$$\mathbf{y} = \mathbf{X}\,\mathbf{b} + \mathbf{e}$$

OLS solution: $\mathbf{b} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{y}$
(requires $\mathbf{X}'\mathbf{X}$ invertible — no perfect multicollinearity).
Each element of $\mathbf{b}$ is the same partial derivative as above;
the matrix form makes computation tractable for arbitrary $k$.
:::

---

## Case A: What if We Just Add `ShelveLoc`?

This section covers **Case A** from the overview table: one numeric regressor
(`Price`) and one categorical regressor (`ShelveLoc`).
This is the most common starting point in applied regression —
one continuous measure of interest, one grouping variable that needs to be
controlled for. The key challenge: a categorical variable cannot enter a
regression equation as-is. It must first be converted into a numeric representation.

### The model equation — written out

The residual plot showed that `ShelveLoc` is the missing piece.
Adding it extends the baseline equation to:

$$\widehat{\text{Sales}}_i = b_0 + b_1 \cdot \text{Price}_i + b_2 \cdot \text{ShelveLoc}_i \tag{1}$$

But wait — `ShelveLoc` is not a number. As we have seen in **Lab 2**, `ShelveLoc` is a **factor** variable with three ordered levels ($k = 3$): Bad → Medium → Good.
Since a category cannot be multiplied with a coefficient, `r fontawesome::fa('r-project')` automatically converts factor objects into so called **dummy variables**: one binary indicator per category, minus one for the reference group. Hence, in our case ($k = 3$) `r fontawesome::fa('r-project')` creates $k - 1 = 2$ dummies for `ShelveLoc`:

::: {.callout-caution collapse="true"}
## Background: Dummy variables — binary, dichotomous, indicator

A **dummy variable** (also called **binary**, **dichotomous**, or **indicator** variable)
takes only two values:

| Value | Meaning |
|-------|---------|
| `1` | The observation belongs to this category |
| `0` | It does not |

For `ShelveLoc` with $k = 3$ categories, R creates $k - 1 = 2$ dummies:

| Store | ShelveLocMedium | ShelveLocGood |
|-------|:--------------:|:------------:|
| Bad shelf | 0 | 0 |
| Medium shelf | 1 | 0 |
| Good shelf | 0 | 1 |

**Why $k - 1$ and not $k$?**
If all three dummies were included, their columns would always sum to 1 —
a perfect linear combination of the intercept column.
The design matrix $\mathbf{X}$ would be singular ($\mathbf{X}'\mathbf{X}$ not invertible)
and OLS would have no unique solution.
Dropping one category — the **reference group** (`Bad`) — resolves this.
Its effect is absorbed into the intercept $b_0$.

**Synonyms in the literature:**

| Term | Common in |
|------|-----------|
| Dummy variable | Econometrics (Wooldridge, Greene) |
| Indicator variable | Statistics, R documentation |
| Binary variable | Machine learning |
| Dichotomous variable | Psychology, survey research |
| One-hot encoding | Machine learning (all $k$ dummies, no intercept) |
:::

```{r dummy-table}
#| echo: false
tibble(
  ShelveLoc         = c("Bad (reference)", "Medium", "Good"),
  `Absorbed into`   = c("b₀ (intercept)", "—", "—"),
  `ShelveLocGood`   = c(0L, 0L, 1L),
  `ShelveLocMedium` = c(0L, 1L, 0L)
) %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#fff8e1"),
            locations = cells_body(rows = 1)) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote(
    "Reference = Bad (first level — set explicitly in setup). Change with relevel(ShelveLoc, ref = 'Good') if needed."
  )
```

The full equation with dummies therefore replaces the placeholder $b_2 \cdot \text{ShelveLoc}_i$
from equation (1) with two explicit dummy terms:

$$\widehat{\text{Sales}}_i = b_0 + b_1 \cdot \text{Price}_i
  + b_2 \cdot \mathbf{1}_{\text{Good},\,i}
  + b_3 \cdot \mathbf{1}_{\text{Medium},\,i} \tag{2}$$

Each `ShelveLoc` group gets its own **intercept shift** ($b_2$ or $b_3$)
but the **same Price slope** ($b_1$) — that is the **parallel slopes** assumption.

:::{.callout-note}
## From equation to `lm()` — the direct mapping

| Equation (2) term | R formula term |
|---------------|---------------|
| $b_0$ (intercept) | always included by default |
| $b_1 \cdot \text{Price}_i$ | `Price` |
| $b_2 \cdot \mathbf{1}_{\text{Good}}$ | automatically from `ShelveLoc` |
| $b_3 \cdot \mathbf{1}_{\text{Medium}}$ | automatically from `ShelveLoc` |
| **Additive (`+`): same slope** | `lm(Sales ~ Price + ShelveLoc)` |
| **Interaction (`*`): different slopes** | `lm(Sales ~ Price * ShelveLoc)` |

The `+` operator means: same slope for all groups (additive / parallel slopes).
The `*` operator means: different slope per group (interaction term included).
:::

### Step 1 — `lm(Sales ~ Price + ShelveLoc)`: fit and inspect

Following equation (2), we translate directly into R:

```{r parallel-fit-first}
#| code-fold: false
mod_shelf <- lm(Sales ~ Price + ShelveLoc, data = tbl_carseats)

tb <- mod_shelf %>%
  get_regression_table() %>%
  select(term, estimate) %>%
  gt()
```
```{r parallel-fit-first-fmt}
#| code-fold: true
#| code-summary: "Show code: table formatting"
tb %>%
  fmt_number(columns = estimate, decimals = 4) %>%
  tab_style(style = cell_fill(color = "#f5f0fa"),
            locations = cells_body(rows = str_detect(term, "ShelveLoc"))) %>%
  tab_style(style = cell_fill(color = "#fff0f0"),
            locations = cells_body(rows = term == "Price")) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(
    paste0("Sales ~ Price + ShelveLoc  |  adj. R² = ",
           round(summary(mod_shelf)$adj.r.squared, 3)),
    subtitle = "Same Price slope (b₁) for all three groups — parallel slopes by construction"
  )
```

::: {.callout-caution collapse="true"}
## Background: Reading the parallel slopes table

The table has four rows, corresponding directly to equation (2):

- **`(Intercept)`** = $b_0$ — predicted Sales for a **Bad**-shelf store at Price = 0
  (not meaningful; see centering later)
- **`Price`** = $b_1$ — the Price effect, **identical for all three shelf groups**
- **`ShelveLocGood`** = $b_2$ — how much higher Good-shelf stores sell
  compared to Bad, at any given Price
- **`ShelveLocMedium`** = $b_3$ — same logic for Medium vs. Bad

The key constraint: there is **one single Price slope** $b_1$.
This is the parallel-slopes assumption — all three lines tilt at the same angle.
:::

:::{.callout-note}
## In descriptive regression we focus on the `estimate` column only

The full `get_regression_table()` output also shows `std_error`, `statistic`,
`p_value`, `lower_ci`, and `upper_ci`. These are **inferential** quantities —
meaningful only when treating the 400 stores as a *sample* from a larger population.

In the **descriptive** setting, all 400 stores *are* the population.
Standard errors and p-values are not defined for a population —
and the term **`estimate`** is itself slightly misleading here:
there is nothing to estimate when you have the full population.
The coefficients $b_0, b_1, \ldots$ are exact descriptive summaries, not estimates.

We return to these inferential statistics in Session 3.
:::

### Step 2 — Visualise: `geom_smooth()` vs. what our model implies

Our model (equation 2) implies parallel lines — same slope, different intercepts.
Let's plot it, then compare with the `geom_smooth()` picture from earlier:

```{r parallel-slopes-plot}
#| code-fold: true
#| code-summary: "Show code: parallel slopes plot"
#| fig-height: 3.8
tbl_carseats %>%
  ggplot(aes(x = Price, y = Sales, color = ShelveLoc)) +
  geom_point(alpha = 0.35, size = 1.8) +
  geom_parallel_slopes(se = FALSE, linewidth = 1.1) +
  scale_color_manual(values = col_shelf) +
  labs(title    = "lm(Sales ~ Price + ShelveLoc): parallel slopes",
       subtitle = "Same Price slope for all groups — one shift per group",
       x = "Price (USD)", y = "Sales (000 units)", color = "ShelveLoc") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

Now compare with what `geom_smooth()` produced when we visualised earlier:

```{r geom-smooth-plot}
#| code-fold: true
#| code-summary: "Show code: geom_smooth() plot"
#| fig-height: 3.8
tbl_carseats %>%
  ggplot(aes(x = Price, y = Sales, color = ShelveLoc)) +
  geom_point(alpha = 0.35, size = 1.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.1) +
  scale_color_manual(values = col_shelf) +
  labs(title    = "geom_smooth(): what model did this fit implicitly?",
       subtitle = "Slightly different slopes across groups — look closely at Good vs. Bad",
       x = "Price (USD)", y = "Sales (000 units)", color = "ShelveLoc") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

Looking at the `geom_smooth()` plot, the three lines appear nearly — but not perfectly —
parallel. A **close inspection** reveals **slightly different slopes** across shelf locations:
the price sensitivity seems marginally stronger for `Good` stores than for `Bad` ones.
But should we trust this visual impression, or is it an artefact of fitting
separate lines to noisy data?

:::{.callout-important}
## `geom_smooth()` implicitly fits the interaction model

When aesthetics `color` maps to a categorical variable, ggplot2 fits
**separate lines per group** — each with its own intercept *and* slope.
This corresponds to a different equation:

$$\widehat{\text{Sales}}_i = b_0 + b_1 \cdot \text{Price}_i
  + b_2 \cdot \mathbf{1}_{\text{Good}}
  + b_3 \cdot \mathbf{1}_{\text{Medium}}
  + b_4 \cdot (\text{Price}_i \times \mathbf{1}_{\text{Good}})
  + b_5 \cdot (\text{Price}_i \times \mathbf{1}_{\text{Medium}}) \tag{3}$$

i.e. `lm(Sales ~ Price * ShelveLoc)` — **not** `lm(Sales ~ Price + ShelveLoc)`.
:::

::: {.callout-caution collapse="true"}
## Background: `geom_parallel_slopes()` — why it is not in base ggplot2

`geom_parallel_slopes()` is a function from the **moderndive** package —
it is not available in base ggplot2.

ggplot2's `geom_smooth(method = "lm")` always fits the most general model
within each colour group — the interaction model (separate intercept *and* slope).
There is no built-in option to force equal slopes across groups.
`geom_parallel_slopes()` fills this gap: it fits `lm(y ~ x + group)` internally
and plots the resulting parallel lines.

**The practical rule:** whenever you use `geom_smooth()` with a categorical
colour aesthetic, you are visualising an interaction model —
whether you intend to or not.
:::

### Step 3 — `lm(Sales ~ Price * ShelveLoc)`: resolve the contradiction

`geom_smooth()` fitted equation (3) — the interaction model.
The interaction terms $b_4$ and $b_5$ capture whether the **effect of Price on Sales
differs depending on shelf location**. In other words: does a price increase hurt
sales more for a prominently placed product than for one tucked away on a bad shelf?
If so, Price and ShelveLoc *interact* — and the slopes would differ across groups.

```{r interaction-fit}
#| code-fold: false
mod_int_shelf <- lm(Sales ~ Price * ShelveLoc, data = tbl_carseats)

tb <- mod_int_shelf %>%
  get_regression_table() %>%
  select(term, estimate) %>%
  gt()
```
```{r interaction-fit-fmt}
#| code-fold: true
#| code-summary: "Show code: table formatting"
tb %>%
  fmt_number(columns = estimate, decimals = 4) %>%
  tab_style(style = cell_fill(color = "#f5f0fa"),
            locations = cells_body(rows = str_detect(term, "ShelveLoc") &
                                     !str_detect(term, ":"))) %>%
  tab_style(style = cell_fill(color = "#fff8e1"),
            locations = cells_body(rows = str_detect(term, ":"))) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(
    paste0("Sales ~ Price × ShelveLoc  |  adj. R² = ",
           round(summary(mod_int_shelf)$adj.r.squared, 3)),
    subtitle = "Purple = ShelveLoc intercept shifts  |  Yellow = slope offsets (interaction terms, eq. 3)"
  )
```

::: {.callout-caution collapse="true"}
## Background: Reading the interaction table

The table has **six rows** — two more than equation (2).
The yellow rows are the interaction terms: slope offsets relative to Bad.

**Reconstruct each group's line from the table:**

```{r interaction-lines}
#| echo: false
b <- coef(mod_int_shelf)
tibble(
  `ShelveLoc group` = c("Bad (reference)", "Good", "Medium"),
  `Intercept`       = c(round(b[1], 3),
                        round(b[1] + b["ShelveLocGood"], 3),
                        round(b[1] + b["ShelveLocMedium"], 3)),
  `Price slope`     = c(round(b["Price"], 4),
                        round(b["Price"] + b["Price:ShelveLocGood"], 4),
                        round(b["Price"] + b["Price:ShelveLocMedium"], 4)),
  `How computed`    = c(
    "b₀  /  b₁",
    "(b₀ + b_Good)  /  (b₁ + b_{Price:Good})",
    "(b₀ + b_Medium)  /  (b₁ + b_{Price:Medium})"
  )
) %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#fff8e1"),
            locations = cells_body(rows = 1)) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("Bad is the reference group — its values appear directly in b₀ and b₁")
```

The three Price slopes are
`r round(b["Price"], 4)`,
`r round(b["Price"] + b["Price:ShelveLocGood"], 4)`, and
`r round(b["Price"] + b["Price:ShelveLocMedium"], 4)`.
They differ by less than 0.005 — negligible.
:::

:::{.callout-tip}
## `r fontawesome::fa("chalkboard-teacher")` Instructor note — interaction tables

*"The table has 6 rows. None gives you the intercept or slope for Good or Medium
directly — you always add the offset to the reference-group value."*

*"Look at the slope offsets (yellow rows). They are tiny — less than 0.003.
That is the formal answer to the visual impression: the slopes differ, but only marginally.
Is that worth 2 extra parameters?"*
:::

### Occam's Razor: the simpler model wins

```{r occam-comparison}
#| code-fold: false
tb <- bind_rows(
  get_regression_summaries(mod_shelf) %>%
    mutate(Model = "Parallel slopes: Sales ~ Price + ShelveLoc", k = 4),
  get_regression_summaries(mod_int_shelf) %>%
    mutate(Model = "Interaction:     Sales ~ Price × ShelveLoc", k = 6)
) %>%
  select(Model, k, r_squared, adj_r_squared, rmse) %>%
  gt()
```
```{r occam-comparison-fmt}
#| code-fold: true
#| code-summary: "Show code: table formatting"
tb %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse), decimals = 4) %>%
  tab_style(style = cell_fill(color = "#f0fff0"),
            locations = cells_body(rows = 1)) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(
    "Parallel slopes vs. interaction: adj. R² is the referee",
    subtitle = "adj. R² penalises the 2 extra parameters of the interaction model"
  )
```

:::{.callout-note}
## Occam's Razor: prefer the simpler model when fit is equal

> *"Under otherwise equal conditions, simpler models are preferable to complex ones."*

adj. R² is **identical at 0.539** for both models (equations 2 and 3),
yet the interaction model uses 2 extra parameters. Zero marginal benefit →
equation (2), the parallel slopes model, wins.

The near-parallel lines in `geom_smooth()` were not misleading —
they reflected a genuine data property: the Price sensitivity is the same
across shelf locations.

See the background box below for a counter-example where interaction *is* warranted
and adj. R² confirms it.
:::

### Eyeball test: with vs. without interaction term

```{r eyeball-test}
#| code-fold: true
#| code-summary: "Show code: side-by-side comparison"
#| fig-height: 4
p_par <- tbl_carseats %>%
  ggplot(aes(x = Price, y = Sales, color = ShelveLoc)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_parallel_slopes(se = FALSE, linewidth = 1.1) +
  scale_color_manual(values = col_shelf) +
  labs(title    = "Without interaction term",
       subtitle = "geom_parallel_slopes()",
       x = "Price (USD)", y = "Sales (000 units)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")

p_int <- tbl_carseats %>%
  ggplot(aes(x = Price, y = Sales, color = ShelveLoc)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.1) +
  scale_color_manual(values = col_shelf) +
  labs(title    = "With interaction term",
       subtitle = "geom_smooth()",
       x = "Price (USD)", y = "Sales (000 units)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")

(p_par + p_int) +
  plot_annotation(
    title    = "The two models look nearly identical — because adj. R² is identical",
    subtitle = "Left: parallel slopes (eq. 2)  |  Right: interaction model (eq. 3)"
  )
```

### The parallel slopes model: interpret and verify

The model is already fitted above. Here we spell out what each coefficient means
and verify that the predictions match the observed group means.

**What each coefficient means:**

```{r parallel-interp}
#| echo: false
b_s <- coef(mod_shelf)
tibble(
  Term           = names(b_s),
  Estimate       = round(b_s, 4),
  Interpretation = c(
    paste0("Avg. Sales for a Bad-shelf store at Price = 0  (not meaningful here — see centering)"),
    paste0("Price effect: +1 USD → ", round(b_s[2], 4),
           " 000 units, holding ShelveLoc fixed — same for all three groups"),
    paste0("Good vs. Bad shelf: +", round(b_s[3], 3),
           " 000 units at any Price — parallel upward shift"),
    paste0("Medium vs. Bad shelf: +", round(b_s[4], 3),
           " 000 units at any Price — parallel upward shift")
  )
) %>%
  gt() %>%
  tab_style(style = cell_fill(color = "#f5f0fa"),
            locations = cells_body(rows = str_detect(Term, "ShelveLoc"))) %>%
  tab_options(table.font.size = "13px")
```

**Verify: predicted means match observed means**

```{r parallel-verify}
#| code-fold: false
b_s <- coef(mod_shelf)
mp  <- mean(tbl_carseats$Price)

cat("Predicted Sales at mean Price (", round(mp, 1), " USD):\n")
cat("  Bad:   ", round(b_s[1] + b_s[2] * mp, 2), "\n")
cat("  Good:  ", round(b_s[1] + b_s[2] * mp + b_s[3], 2), "\n")
cat("  Medium:", round(b_s[1] + b_s[2] * mp + b_s[4], 2), "\n\n")
cat("Observed mean Sales:\n")
tbl_carseats %>%
  group_by(ShelveLoc) %>%
  summarise(mean_sales = round(mean(Sales), 2)) %>%
  print()
```

The Carseats data give a clear verdict: parallel slopes.
But the same procedure on a different dataset can yield the opposite answer.
The background box below shows exactly that case.

::: {.callout-caution collapse="true"}
## Background: When interaction IS warranted — UT Austin (moderndive Ch. 6)

The moderndive textbook (`evals` dataset, $n = 463$) shows the opposite case:
here the interaction *is* justified.

```{r ut-austin}
#| message: false
evals_ch6 <- evals %>% select(score, age, gender)

mod_ut_par <- lm(score ~ age + gender,  data = evals_ch6)
mod_ut_int <- lm(score ~ age * gender,  data = evals_ch6)

bind_rows(
  get_regression_summaries(mod_ut_par) %>%
    mutate(Model = "Parallel slopes: score ~ age + gender", k = 3),
  get_regression_summaries(mod_ut_int) %>%
    mutate(Model = "Interaction:     score ~ age × gender", k = 4)
) %>%
  select(Model, k, r_squared, adj_r_squared, rmse) %>%
  gt() %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse), decimals = 3) %>%
  tab_style(style = cell_fill(color = "#f0fff0"),
            locations = cells_body(rows = 2)) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header("UT Austin: interaction IS warranted",
             subtitle = "adj. R² rises 0.035 → 0.045 (+29% relative) with 1 extra parameter")
```

```{r ut-austin-viz}
#| fig-height: 3.5
p_ut_int <- evals_ch6 %>%
  ggplot(aes(x = age, y = score, color = gender)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.1) +
  scale_color_manual(values = c("female" = "#D50006", "male" = "#502479")) +
  labs(title = "Interaction model",
       subtitle = "Lines cross (X-pattern) — different slopes justified",
       x = "Age", y = "Score") +
  theme_minimal(base_size = 9) + theme(legend.position = "bottom")

p_ut_par <- evals_ch6 %>%
  ggplot(aes(x = age, y = score, color = gender)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_parallel_slopes(se = FALSE, linewidth = 1.1) +
  scale_color_manual(values = c("female" = "#D50006", "male" = "#502479")) +
  labs(title = "Parallel slopes model",
       subtitle = "Forced equal slopes — misses the X-pattern",
       x = "Age", y = "Score") +
  theme_minimal(base_size = 9) + theme(legend.position = "bottom")

(p_ut_int + p_ut_par) +
  plot_annotation(
    title    = "Same procedure — different data answer",
    subtitle = "UT Austin: lines cross visually AND adj. R² improves → interaction warranted"
  )
```

| | Carseats (ShelveLoc) | UT Austin (gender) |
|--|--|--|
| Visual | Nearly parallel | Clear X-pattern |
| adj. R² gain | 0.000 | +0.010 (+29% rel.) |
| **Verdict** | **Parallel slopes** | **Interaction** |

**The procedure is identical — the data decide.**
:::

### ShelveLoc as a confounding variable

:::{.callout-tip}
## `r fontawesome::fa("chalkboard-teacher")` Instructor note — confounding

*"`ShelveLoc` is a **confounding variable**: it correlates with Price
(prime shelf locations support higher prices) AND with Sales
(better placement drives more units).
Without it in the model, the Price coefficient was biased — **Omitted Variable Bias**."*

*"Adding `ShelveLoc` as a **control variable** lets us read the Price effect
within comparable stores (same shelf location).
That is ceteris paribus — and it is why the Price coefficient changes
when we add ShelveLoc. We will quantify this in the Model Quality section."*
:::

---

## Case B: Adding a Second Numeric Regressor: `Advertising`

This section covers **Case B** from the overview table: two numeric regressors
(`Price` and `Advertising`). The key difference from Case A: with two numeric
variables, OLS no longer fits a line — it fits a **plane** in three-dimensional
space, and the ceteris-paribus interpretation becomes directly visible in the geometry.

`ShelveLoc` already jumped adj. R² from 0.15 to 0.54.
Can `Advertising` add anything on top?
Adding a second numeric regressor also changes the geometry:
where `Price + ShelveLoc` gave parallel lines, `Price + Advertising` gives a **plane**.

### From parallel lines to a plane in 3D

```{r viz-3d}
#| fig-height: 5
price_grid <- seq(min(tbl_carseats$Price),       max(tbl_carseats$Price),       length.out = 30)
adv_grid   <- seq(min(tbl_carseats$Advertising), max(tbl_carseats$Advertising), length.out = 30)

# plotly add_surface expects z[i,j] = f(y_grid[i], x_grid[j])
# outer() produces z[i,j] = f(x_grid[i], y_grid[j]) → transpose needed
z_mat <- outer(price_grid, adv_grid,
               function(p, a) coef(mod_adv)[1] +
                               coef(mod_adv)[2] * p +
                               coef(mod_adv)[3] * a)

plot_ly() %>%
  add_trace(
    data   = tbl_carseats,
    x = ~Price, y = ~Advertising, z = ~Sales,
    type = "scatter3d", mode = "markers",
    marker = list(size = 3, color = "#7d0a52", opacity = 0.6),
    name = "Stores"
  ) %>%
  add_surface(
    x = price_grid, y = adv_grid, z = t(z_mat),
    colorscale = list(c(0,"#f5f0fa"), c(0.5,"#7d0a52"), c(1,"#502479")),
    opacity = 0.55, showscale = FALSE,
    name = "Regression plane"
  ) %>%
  layout(
    title = list(
      text = "Sales ~ Price + Advertising: the regression plane",
      font = list(size = 13)
    ),
    scene = list(
      xaxis  = list(title = "Price (USD)"),
      yaxis  = list(title = "Advertising (USD 000)"),
      zaxis  = list(title = "Sales (000 units)"),
      camera = list(eye = list(x = 1.8, y = -1.5, z = 0.9))
    )
  )
```

:::{.callout-note}
## What the 3D plot shows

**Simple regression** ($k=1$): OLS fits a **line** — residual = vertical distance to line.

**Two numeric regressors** ($k=2$): OLS fits a **plane** — residual = vertical distance to plane.

**Partial effect from geometry:** $b_1$ is the slope of the plane *along the Price axis*
while the Advertising axis is held fixed. Tilt the plot — that is ceteris paribus made visible.

**Adding ShelveLoc** (categorical): instead of one plane,
OLS fits **parallel planes** (one per group, same slope) — exactly what we just estimated.
With an interaction term, the planes would be **tilted** relative to each other.

**$k > 2$ regressors:** geometrically unplottable, but
$\mathbf{b} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{y}$
and the partial-derivative interpretation remain identical.
:::

### Fit and interpret

```{r two-numeric-fit}
#| code-fold: false
tb <- mod_adv %>%
  get_regression_table() %>%
  select(term, estimate) %>%
  gt()
```
```{r two-numeric-fit-fmt}
#| code-fold: true
#| code-summary: "Show code: table formatting"
tb %>%
  fmt_number(columns = estimate, decimals = 4) %>%
  tab_style(style = cell_fill(color = "#f0fff8"),
            locations = cells_body(rows = term == "Advertising")) %>%
  tab_style(style = cell_fill(color = "#fff0f0"),
            locations = cells_body(rows = term == "Price")) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(paste0("Sales ~ Price + Advertising  |  R² = ",
                    round(summary(mod_adv)$r.squared, 3),
                    "  |  adj. R² = ",
                    round(summary(mod_adv)$adj.r.squared, 3)))
```

### Centering: making the intercept meaningful

$b_0 = $ `r round(coef(mod_adv)[1], 2)` is the predicted Sales when
Price = 0 *and* Advertising = 0 — both outside the observed range.
**Mean centering** replaces $x_j$ with $x_j^c = x_j - \bar{x}_j$:
then $b_0^c$ is the predicted Sales for a store with **average Price and
average Advertising** — a meaningful quantity.

```{r centering}
#| code-fold: false
tbl_c <- tbl_carseats %>%
  mutate(Price_c       = Price       - mean(Price),
         Advertising_c = Advertising - mean(Advertising))

mod_adv_c <- lm(Sales ~ Price_c + Advertising_c, data = tbl_c)

tb <- bind_rows(
  tidy(mod_adv)   %>% mutate(Model = "Original"),
  tidy(mod_adv_c) %>% mutate(Model = "Mean-centred")
) %>%
  select(Model, term, estimate) %>%
  pivot_wider(names_from = Model, values_from = estimate) %>%
  gt()
```
```{r centering-fmt}
#| code-fold: true
#| code-summary: "Show code: table formatting"
tb %>%
  fmt_number(columns = c(Original, `Mean-centred`), decimals = 4) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(
    "Centering: intercept changes, slopes unchanged",
    subtitle = paste0(
      "Centred intercept = predicted Sales at mean Price (",
      round(mean(tbl_carseats$Price), 1), " USD) and mean Advertising (",
      round(mean(tbl_carseats$Advertising), 1), " USD 000)")
  )
```

:::{.callout-note}
## What centering does and does not change

**Changes:** $b_0^c$ is now the predicted outcome at the mean of all regressors —
a meaningful number: *"A store with average price and average advertising
is expected to sell `r round(coef(mod_adv_c)[1], 2)` thousand units."*

**Does not change:** slopes $b_1, b_2$, R², residuals, fitted values.
Centering is a reparametrisation, not a different model.
Use it whenever $x = 0$ is not in the observed range.
:::

### Simpson's Paradox — when aggregate trends mislead

Adding a second numeric regressor can reverse the sign of a coefficient.
This is Simpson's Paradox: a trend visible in the aggregate disappears or reverses
when a third variable is controlled for.

```{r simpson-viz}
#| fig-height: 3.8
# Simple bivariate plots
p_price <- tbl_carseats %>%
  ggplot(aes(x = Price, y = Sales)) +
  geom_point(alpha = 0.3, color = "#7d0a52", size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#D50006", linewidth = 1.1) +
  annotate("text", x = 70, y = 15,
           label = paste0("b₁ = ", round(coef(mod_baseline)["Price"], 3)),
           color = "#D50006", size = 3.5, hjust = 0) +
  labs(title = "Price vs. Sales (bivariate)",
       x = "Price (USD)", y = "Sales") +
  theme_minimal(base_size = 9)

p_adv_simple <- tbl_carseats %>%
  ggplot(aes(x = Advertising, y = Sales)) +
  geom_point(alpha = 0.3, color = "#7d0a52", size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#502479", linewidth = 1.1) +
  annotate("text", x = 1, y = 15,
           label = paste0("b₁ = +", round(coef(lm(Sales~Advertising,tbl_carseats))["Advertising"], 3)),
           color = "#502479", size = 3.5, hjust = 0) +
  labs(title = "Advertising vs. Sales (bivariate)",
       x = "Advertising (USD 000)", y = "Sales") +
  theme_minimal(base_size = 9)

(p_price + p_adv_simple) +
  plot_annotation(
    title    = "Bivariate picture — each variable in isolation",
    subtitle = "Both relationships look straightforward here"
  )
```

```{r simpson-binned}
#| fig-height: 3.8
# ntile() avoids non-unique breaks caused by Advertising's many zero values
tbl_carseats %>%
  mutate(adv_q = factor(ntile(Advertising, 4),
                        labels = c("Q1 (low)", "Q2", "Q3", "Q4 (high)"))) %>%
  ggplot(aes(x = Price, y = Sales, color = adv_q)) +
  geom_point(alpha = 0.35, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.0) +
  scale_color_manual(values = c("Q1 (low)" = "#D50006",
                                "Q2"       = "#7d0a52",
                                "Q3"       = "#502479",
                                "Q4 (high)"= "#686868")) +
  labs(
    title    = "Price vs. Sales — disaggregated by Advertising quartile",
    subtitle = paste0(
      "Within each quartile: same negative Price slope  |  ",
      "Gap between lines = partial effect of Advertising  |  ",
      "Parallel ≈ no interaction (confirms adj. R² finding)"),
    x = "Price (USD)", y = "Sales (000 units)",
    color = "Advertising quartile"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

:::{.callout-note}
## Simpson's Paradox in the multiple regression context

When two regressors are correlated, the bivariate slope of one variable can differ
substantially from its partial slope in the multiple model.

**In Carseats:** Price and Advertising are only weakly correlated
(`r round(cor(tbl_carseats$Price, tbl_carseats$Advertising), 3)`),
so the sign reversal that characterises strong Simpson cases does not occur here.
The slopes within each Advertising quartile are nearly parallel — confirming the
parallel-slopes model and the absence of a meaningful interaction.

**The classic Simpson case** (moderndive Ch. 6, Credit data): Income has a *positive*
bivariate relationship with credit card debt, but a *negative* partial slope once
credit limit is controlled for — a full sign reversal caused by a strong confounder.
The background box below shows the full story.
:::

::: {.callout-caution collapse="true"}
## Background: Simpson's Paradox — the credit card example (moderndive Ch. 6)

```{r credit-setup}
#| message: false
credit_ch6 <- Credit %>%
  as_tibble() %>%
  select(debt         = Balance,
         credit_limit = Limit,
         income       = Income)
```

```{r credit-bivariate}
#| fig-height: 3.5
p_c1 <- credit_ch6 %>%
  ggplot(aes(x = credit_limit, y = debt)) +
  geom_point(alpha = 0.3, color = "#7d0a52", size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#D50006", linewidth = 1.1) +
  labs(title = "Debt ~ Credit limit",
       subtitle = paste0("b = +",
         round(coef(lm(debt ~ credit_limit, credit_ch6))["credit_limit"], 3)),
       x = "Credit limit (USD)", y = "Debt (USD)") +
  theme_minimal(base_size = 9)

p_c2 <- credit_ch6 %>%
  ggplot(aes(x = income, y = debt)) +
  geom_point(alpha = 0.3, color = "#7d0a52", size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#502479", linewidth = 1.1) +
  labs(title = "Debt ~ Income (bivariate)",
       subtitle = paste0("b = +",
         round(coef(lm(debt ~ income, credit_ch6))["income"], 3),
         "  ← POSITIVE"),
       x = "Income (USD 000)", y = "Debt (USD)") +
  theme_minimal(base_size = 9)

p_c1 + p_c2
```

```{r credit-simpson}
#| fig-height: 3.5
# Disaggregate by credit_limit quartile to reveal Simpson
credit_ch6 %>%
  mutate(limit_q = cut(
    credit_limit,
    breaks = quantile(credit_limit, probs = c(0,.25,.5,.75,1), na.rm = TRUE),
    labels = c("Q1 low", "Q2", "Q3", "Q4 high"),
    include.lowest = TRUE
  )) %>%
  ggplot(aes(x = income, y = debt, color = limit_q)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.0) +
  scale_color_manual(values = c("Q1 low"  = "#D50006",
                                "Q2"      = "#7d0a52",
                                "Q3"      = "#502479",
                                "Q4 high" = "#686868")) +
  labs(
    title    = "Debt ~ Income — disaggregated by credit limit quartile",
    subtitle = "Within each quartile: NEGATIVE slope — sign reversal vs. bivariate",
    x = "Income (USD 000)", y = "Debt (USD)",
    color = "Credit limit"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

```{r credit-model}
debt_model <- lm(debt ~ credit_limit + income, data = credit_ch6)

debt_model %>%
  get_regression_table() %>%
  select(term, estimate) %>%
  gt() %>%
  fmt_number(columns = estimate, decimals = 3) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(
    "Multiple model: income slope reverses sign",
    subtitle = paste0(
      "Bivariate: b_income = +",
      round(coef(lm(debt~income,credit_ch6))["income"], 3),
      "  |  Multiple (controlling for limit): b_income = ",
      round(coef(debt_model)["income"], 3))
  )
```

**What happened:** credit limit is a confounding variable — it correlates
strongly with both income ($r \approx 0.79$) and debt. Within each credit-limit group,
higher income is associated with *lower* debt. The aggregate positive slope
was entirely driven by the confound: high-income people also have high credit limits
which enables more debt.

**Lesson:** In any multiple regression, always ask *"what is being controlled for?"*
The partial slope can differ dramatically from the bivariate slope —
and sometimes reverses sign completely.
:::

---

## The Full Model: Price + Advertising + ShelveLoc

Combining both insights:

$$\widehat{\text{Sales}}_i = b_0 + b_1 \cdot \text{Price}_i
  + b_2 \cdot \text{Advertising}_i
  + b_3 \cdot \mathbf{1}_{\text{Good}}
  + b_4 \cdot \mathbf{1}_{\text{Medium}}$$

```{r full-model}
#| code-fold: false
tb <- mod_full %>%
  get_regression_table() %>%
  select(term, estimate) %>%
  gt()
```
```{r full-model-fmt}
#| code-fold: true
#| code-summary: "Show code: table formatting"
tb %>%
  fmt_number(columns = estimate, decimals = 4) %>%
  tab_style(style = cell_fill(color = "#f5f0fa"),
            locations = cells_body(rows = str_detect(term, "ShelveLoc"))) %>%
  tab_style(style = cell_fill(color = "#fff0f0"),
            locations = cells_body(rows = term == "Price")) %>%
  tab_style(style = cell_fill(color = "#f0fff8"),
            locations = cells_body(rows = term == "Advertising")) %>%
  tab_options(table.font.size = "14px") %>%
  tab_header(
    paste0("Sales ~ Price + Advertising + ShelveLoc  |  adj. R² = ",
           round(summary(mod_full)$adj.r.squared, 3)),
    subtitle = "Purple = ShelveLoc | Red = Price | Green = Advertising"
  )
```

**Live computation — ceteris paribus:**

```{r prediction}
#| code-fold: false
# Predicted Sales: Good shelf, Price = 120, Advertising = 10
predict(mod_full,
        newdata = data.frame(Price = 120, Advertising = 10, ShelveLoc = "Good"))

# Manual — shows each term's contribution
b <- coef(mod_full)
b["(Intercept)"] +
  b["Price"]          * 120 +
  b["Advertising"]    * 10  +
  b["ShelveLocGood"]  * 1
```

:::{.callout-tip}
## `r fontawesome::fa("chalkboard-teacher")` Instructor note — live computation

Type this live. Then vary:
- `ShelveLoc = "Bad"` → *how much does the shelf-location shift cost?*
- `Advertising = 0` → *how much of the prediction came from advertising?*

This makes ceteris paribus tangible: change one input, hold all others fixed,
see what happens.
:::

---

## Model Quality and Selection

### Systematic comparison

```{r model-compare-full}
#| code-fold: false
tb <- bind_rows(
  get_regression_summaries(mod_baseline)  %>% mutate(Model = "① Sales ~ Price",                     k = 2),
  get_regression_summaries(mod_adv)       %>% mutate(Model = "② + Advertising",                     k = 3),
  get_regression_summaries(mod_shelf)     %>% mutate(Model = "③ + ShelveLoc (parallel)",             k = 4),
  get_regression_summaries(mod_full)      %>% mutate(Model = "④ + Advertising + ShelveLoc",          k = 5),
  get_regression_summaries(mod_int_shelf) %>% mutate(Model = "⑤ + ShelveLoc × Price (interaction)", k = 6)
) %>%
  select(Model, k, r_squared, adj_r_squared, rmse) %>%
  gt()
```
```{r model-compare-full-fmt}
#| code-fold: true
#| code-summary: "Show code: table formatting"
tb %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse), decimals = 4) %>%
  tab_style(style = cell_fill(color = "#f0fff0"),
            locations = cells_body(rows = 4)) %>%
  tab_style(style = cell_fill(color = "#fff8e1"),
            locations = cells_body(rows = 5)) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(
    "Model comparison across all specifications",
    subtitle = "Green = selected model ④ | Yellow = interaction (not warranted)"
  ) %>%
  tab_footnote("adj. R² penalises additional parameters — use it (not R²) to compare model sizes")
```

### Adjusted R²

**R² always increases** when a regressor is added — even random noise marginally improves
in-sample fit. **Adjusted R²** corrects for this:

$$\bar{R}^2 = 1 - \frac{n-1}{n-k} \cdot (1 - R^2)$$

where $k$ = number of parameters (including intercept).
If a regressor adds less explanatory power than a random variable would on average,
$\bar{R}^2$ decreases. Use $\bar{R}^2$ — not $R^2$ — when comparing models of different sizes.

### OVB correction: the Price coefficient across specifications

```{r ovb-table}
#| code-fold: false
tb <- tibble(
  Model      = c("① Price only", "② + Advertising",
                 "③ + ShelveLoc", "④ + Both", "⑤ Interaction"),
  `b̂_Price`  = map_dbl(
    list(mod_baseline, mod_adv, mod_shelf, mod_full, mod_int_shelf),
    ~ coef(.x)["Price"]
  ),
  `adj. R²`  = map_dbl(
    list(mod_baseline, mod_adv, mod_shelf, mod_full, mod_int_shelf),
    ~ summary(.x)$adj.r.squared
  )
) %>%
  mutate(`Δ from ①` = `b̂_Price` - coef(mod_baseline)["Price"]) %>%
  gt()
```
```{r ovb-table-fmt}
#| code-fold: true
#| code-summary: "Show code: table formatting"
tb %>%
  fmt_number(columns = c(`b̂_Price`, `adj. R²`, `Δ from ①`), decimals = 4) %>%
  tab_style(
    style = cell_fill(color = "#fff0f0"),
    locations = cells_body(rows = abs(`Δ from ①`) > 0.005)
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(
    "OVB correction: b̂_Price moves when ShelveLoc is controlled",
    subtitle = "Adding Advertising barely changes b̂_Price — ShelveLoc is the confounder"
  )
```

:::{.callout-tip}
## `r fontawesome::fa("chalkboard-teacher")` Instructor note — OVB moment

*"The Price coefficient in model ① is `r round(coef(mod_baseline)["Price"], 3)`.
After controlling for ShelveLoc it becomes `r round(coef(mod_shelf)["Price"], 3)`.
Why more negative?"*

*"Good-shelf stores can charge higher prices AND sell more.
Without ShelveLoc, that positive association leaks into the Price coefficient
and makes it look less negative than it truly is.
That is Omitted Variable Bias — and we just corrected it."*
:::

### Forward specification — best practice

| Strategy | Procedure | Risk |
|----------|-----------|------|
| **Forward** | Start minimal, add theory-grounded variables | Transparent; easy to miss confounders |
| **Backward** | Start full, remove stepwise | Nothing missed; temptation to data-mine |

For this course: **forward, theory-grounded**.
Automated `step()` optimises in-sample fit and inflates false discovery rates —
that is the **p-hacking problem** introduced in PVA2.

---

## Underfitting vs. Overfitting

```{r fit-progression}
#| fig-height: 3.5
fit_df <- bind_rows(
  get_regression_summaries(mod_baseline)  %>% mutate(k = 2, label = "① Price"),
  get_regression_summaries(mod_adv)       %>% mutate(k = 3, label = "② +Adv"),
  get_regression_summaries(mod_shelf)     %>% mutate(k = 4, label = "③ +Shelf"),
  get_regression_summaries(mod_full)      %>% mutate(k = 5, label = "④ Both"),
  get_regression_summaries(mod_int_shelf) %>% mutate(k = 6, label = "⑤ +Inter")
)

p_r2 <- fit_df %>%
  pivot_longer(c(r_squared, adj_r_squared),
               names_to = "metric", values_to = "val") %>%
  ggplot(aes(x = k, y = val, color = metric, group = metric)) +
  geom_line(linewidth = 1.1) + geom_point(size = 3) +
  scale_color_manual(values = c("r_squared"     = "#D50006",
                                "adj_r_squared" = "#502479"),
                     labels = c("R²", "adj. R²")) +
  scale_x_continuous(breaks = c(2,3,4,5,6), labels = fit_df$label) +
  labs(title = "R² vs. adj. R²", x = NULL, y = NULL, color = NULL) +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1),
        legend.position = "bottom")

p_rmse <- fit_df %>%
  ggplot(aes(x = k, y = rmse)) +
  geom_line(color = "#7d0a52", linewidth = 1.1) +
  geom_point(color = "#7d0a52", size = 3) +
  scale_x_continuous(breaks = c(2,3,4,5,6), labels = fit_df$label) +
  labs(title = "In-sample RMSE (000 units)", x = NULL, y = NULL) +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

(p_r2 + p_rmse) +
  plot_annotation(
    title    = "Underfitting → Adequate fit → Overfitting risk",
    subtitle = "R² always rises | adj. R² penalises complexity | RMSE alone insufficient without out-of-sample data"
  )
```

:::{.callout-important}
## The fundamental tension

**Underfitting** (models ①–②): Too simple — misses real patterns.
Symptom: low adj. R², systematic residual patterns.

**Adequate fit** (model ④): Captures the main variation with interpretable,
theory-grounded predictors and justified modelling assumptions.

**Overfitting risk** (model ⑤): Adds complexity without improving adj. R².
In-sample fit looks fine; performance on new data degrades.

**Today we only see in-sample fit** — diagnosing overfitting requires
new data or cross-validation (ER017).

> *"Discovery without overfit. The challenge of data science is to find real
> patterns — not to memorize the noise in your sample."*
> — @taddy_business_2019, p. 8
:::

---

## Residual Diagnostics

```{r resid-3way}
#| fig-height: 4
pts_base <- augment(mod_baseline, data = tbl_carseats)
pts_adv2 <- augment(mod_adv,      data = tbl_carseats)
pts_full <- augment(mod_full,      data = tbl_carseats)

mk <- function(pts, title, r2) {
  pts %>%
    ggplot(aes(.fitted, .resid, color = ShelveLoc)) +
    geom_point(alpha = 0.4, size = 1.3) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "#333") +
    geom_smooth(aes(group = 1), method = "loess", se = FALSE,
                color = "black", linewidth = 0.9) +
    scale_color_manual(values = col_shelf_resid,
                       guide = "none") +
    labs(title    = title,
         subtitle = paste0("adj. R² = ", round(r2, 3)),
         x = "Fitted", y = "Residual") +
    theme_minimal(base_size = 9)
}

(mk(pts_base, "① Price only",         summary(mod_baseline)$adj.r.squared) +
 mk(pts_adv2, "② + Advertising",      summary(mod_adv)$adj.r.squared)      +
 mk(pts_full, "④ + Adv + ShelveLoc",  summary(mod_full)$adj.r.squared)) +
  plot_annotation(
    title    = "Residual diagnostics — descriptive purpose: misspecification detection",
    subtitle = paste0(
      "Signal 1: LOESS not flat → systematic omission  |  ",
      "Signal 2: colour clusters → confound not controlled")
  )
```

:::{.callout-note}
## How to read these plots — two diagnostic signals

**Signal 1 — LOESS trend:**
A flat black line at zero = no systematic pattern remaining.
A curved line = the model misses a systematic relationship.

**Signal 2 — colour clustering:**
Colours randomly mixed = the coloured variable is controlled for.
Colours clustered (purple above, red below) = the variable belongs in the model.

**① Price only:** LOESS bends; purple above zero, red below →
ShelveLoc confounds, systematic omission.

**② + Advertising:** LOESS flatter but not flat; colour separation persists →
ShelveLoc still missing.

**④ + Adv + ShelveLoc:** LOESS approximately flat; colours randomly mixed ✅ →
main systematic variation absorbed. Remaining scatter:
random noise + variables outside the dataset.

**What these plots do NOT check** in the descriptive setting:
normality and homoskedasticity — those are needed for inference (PVA2),
not for the descriptive conditional mean.
:::

:::{.callout-tip}
## `r fontawesome::fa("chalkboard-teacher")` Instructor note — pointing out the signals

Students scan the plots globally and often miss the two specific signals.
Point explicitly:

*"In plot ①, where are the purple dots? Above zero.
Where are the red ones? Below zero. That is not random —
ShelveLoc systematically shifts the residuals."*

*"In plot ④, the colour separation is gone.
The model now accounts for shelf location — the residual is a cleaner signal."*
:::

---

## Multicollinearity

```{r vif-check}
#| code-fold: false
vif(mod_full)
```

```{r corr-matrix}
#| code-fold: false
tb <- tbl_carseats %>%
  select(Price, Advertising, Income, Age, Population) %>%
  cor() %>% round(3) %>%
  as.data.frame() %>% rownames_to_column("Variable") %>%
  gt()
tb
```
```{r corr-matrix-fmt}
#| code-fold: true
#| code-summary: "Show code: table formatting"
tb %>%
  fmt_number(columns = -Variable, decimals = 3) %>%
  tab_options(table.font.size = "12px") %>%
  tab_header("Correlation matrix — numeric regressors")
```

:::{.callout-note}
## Multicollinearity: descriptive vs. inferential impact

| | Descriptive (today) | Stochastic / inference (PVA2) |
|-|---|---|
| **Fitted values $\hat{y}_i$** | Unaffected | Unaffected |
| **Individual coefficients** | Partial effects unreliable when regressors correlated | Unbiased but high variance → wide CIs |
| **Diagnosis** | Correlation matrix, VIF | VIF, condition number |
| **Threshold** | \|r\| > 0.8 between regressors: caution | VIF > 5: problematic |

For model ④: all VIFs are low — no multicollinearity concern.
:::

---

## Performance Scores → Bridge to Lab 3

```{r performance}
#| code-fold: false
store_scored <- mod_full %>%
  get_regression_points() %>%
  mutate(performance_gap = residual) %>%
  arrange(performance_gap)

tb <- store_scored %>%
  head(5) %>%
  select(ID, Sales, Sales_hat, residual) %>%
  gt()
```
```{r performance-fmt}
#| code-fold: true
#| code-summary: "Show code: table formatting"
tb %>%
  fmt_number(columns = c(Sales, Sales_hat, residual), decimals = 2) %>%
  tab_style(style = cell_fill(color = "#fff0f0"), locations = cells_body()) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header("Top 5 Underperformers — strongest negative residuals")
```

```{r perf-viz}
#| code-fold: false
gg <- store_scored %>%
  ggplot(aes(x = Sales_hat, y = performance_gap,
             color = performance_gap < -2)) +
  geom_point(alpha = 0.6, size = 1.8) +
  geom_hline(yintercept = 0, color = "#D50006",
             linewidth = 1.1, linetype = "dashed") +
  geom_hline(yintercept = c(-2, 2), color = "#686868",
             linewidth = 0.5, linetype = "dotted")
```
```{r perf-viz-fmt}
#| code-fold: true
#| code-summary: "Show code: labels & theme"
gg +
  scale_color_manual(values  = c("FALSE" = "#b3b2b2", "TRUE" = "#D50006"),
                     labels  = c("Normal", "Underperformer")) +
  labs(title    = "Performance score: residuals from model ④",
       subtitle = "Price- / Advertising- / ShelveLoc-adjusted underperformance",
       x = "Fitted Sales (000 units)",
       y = "Performance gap (residual)",
       color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

:::{.callout-tip}
## `r fontawesome::fa("chalkboard-teacher")` Instructor note — handover to Lab 3

*"This is the output of the demo: a performance score for every store,
adjusted for price, advertising budget, and shelf location.
A store with a strongly negative residual sells less than comparable stores
under the same conditions — that is the underperformer signal."*

*"In Lab 3 you will rank these stores, look at their characteristics,
and write the Managerial Memo. Open `PVA1_Lab3_PerformanceMining.qmd`."*
:::

---

## What We Learned: Four Models at a Glance

Now that we have worked through each model step by step,
here is the comparison that motivated the sequence:

```{r orientation-table}
#| code-fold: false
tb <- bind_rows(
  get_regression_summaries(mod_adv)       %>% mutate(Model = "Price + Advertising",       k = 3),
  get_regression_summaries(mod_int_adv)   %>% mutate(Model = "Price × Advertising",       k = 4),
  get_regression_summaries(mod_shelf)     %>% mutate(Model = "Price + ShelveLoc",          k = 4),
  get_regression_summaries(mod_int_shelf) %>% mutate(Model = "Price × ShelveLoc",          k = 6)
) %>%
  select(Model, k, r_squared, adj_r_squared, rmse) %>%
  gt()
```
```{r orientation-table-fmt}
#| code-fold: true
#| code-summary: "Show code: table formatting"
tb %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse), decimals = 3) %>%
  tab_style(style = cell_fill(color = "#f0fff0"),
            locations = cells_body(rows = str_detect(Model, "ShelveLoc"))) %>%
  tab_style(style = cell_fill(color = "#fff0f0"),
            locations = cells_body(rows = str_detect(Model, "Advertising"))) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(
    "Four two-variable models — why we started with ShelveLoc",
    subtitle = paste0(
      "Green = ShelveLoc (adj. R² ≈ 0.54)  |  ",
      "Red = Advertising (adj. R² ≈ 0.28)  |  ",
      "Interaction adds nothing in either case")
  )
```

:::{.callout-note}
## The numbers explain the sequence

**ShelveLoc** alone explains nearly twice as much variance as Advertising —
and adding an interaction term changes nothing (adj. R² identical at 0.539).
That is why we started with ShelveLoc and chose the parallel-slopes model:
the data guided the decision, not convention.

**Advertising** still adds real signal on top of ShelveLoc
(adj. R² rises from 0.539 to ~0.54 in the full model) — small but consistent.
:::

---

## Summary

```{r final-summary}
#| echo: false
bind_rows(
  get_regression_summaries(mod_baseline) %>% mutate(Model = "① Sales ~ Price"),
  get_regression_summaries(mod_shelf)    %>% mutate(Model = "② + ShelveLoc (parallel)"),
  get_regression_summaries(mod_full)     %>% mutate(Model = "③ + Advertising + ShelveLoc  ← selected")
) %>%
  select(Model, r_squared, adj_r_squared, rmse) %>%
  gt() %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse), decimals = 4) %>%
  tab_style(style = cell_fill(color = "#f0fff0"), locations = cells_body(rows = 3)) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-important}
## Key concepts established in this demo

1. **Descriptive Regression Notation:** $b_j$ / $e_i$ (descriptive) — distinguished from $\hat\beta_j$ / $\varepsilon_i$ (stochastic, Session 3)

2. **Orientation table first:** empirical adj. R² comparison guides model-building sequence before any individual coefficient is examined

3. **`geom_smooth()` reveals the interaction model** implicitly — always check whether non-parallel slopes are warranted

4. **Occam's Razor operationalised by adj. R²:** Carseats → parallel slopes (adj. R² unchanged); UT Austin → interaction (adj. R² improves)

5. **Geometry:** line (k=1) → plane (k=2 numeric) → parallel planes (+ categorical, no interaction) → tilted planes (+ interaction)

6. **Partial effects** = partial derivatives = ceteris paribus — descriptive, not causal

7. **Confounding and control variables:** ShelveLoc biases the Price coefficient; controlling for it corrects the OVB (visible numerically)

8. **Centering:** makes $b_0$ meaningful as predicted outcome at mean $\mathbf{x}$

9. **Simpson's Paradox:** aggregate slope can differ from or reverse the partial slope — always ask what is being controlled for

10. **adj. R² for model comparison**, eyeball test (Occam), forward specification

11. **Residual diagnostics (descriptive):** two signals — LOESS trend (misspecification) and colour clustering (uncontrolled confound)

12. **Multicollinearity:** affects individual coefficients (not fitted values); VIF > 5 as warning threshold
:::




## References {.unnumbered}
