---
title: "Demo: Multiple Regression"
subtitle: "CRISP-DM Phase 4 — Modeling"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: de
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Inhalt"
    number-sections: false
    code-fold: false
    code-tools: true
    theme: cosmo
    css: ../../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse)
library(moderndive)
library(broom)
library(gt)
library(patchwork)
library(ISLR2)

tbl_carseats <- tibble(Carseats)
```

:::{.callout-note}
## Purpose of this Document

This document serves **two purposes:**

1. **For the instructor:** A fully annotated step-by-step guide for the live demo in RStudio.
   Step by step—with explanations of what is being demonstrated and why.

2. **For students:** Available as a reference document after the session.
   All steps are reproducible and can be followed independently.
:::

---

## Starting Point: What Do We Have So Far?

In Lab 2, we estimated the **baseline model** `Sales ~ Price`:

```{r baseline-recap}
mod_baseline <- tbl_carseats %>%
  lm(Sales ~ Price, data = .)

mod_baseline %>%
  get_regression_summaries() %>%
  select(r_squared, rmse) %>%
  mutate(modell = "Baseline: Sales ~ Price") %>%
  relocate(modell) %>%
  gt() %>%
  fmt_number(columns = c(r_squared, rmse), decimals = 4) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-important}
## The Problem: Omitted Variable Bias

$R^2 \approx$ `r round(summary(mod_baseline)$r.squared, 2)` — price alone explains very little.
And the residual plot showed a systematic pattern based on `ShelveLoc`.

**Diagnosis:** `ShelveLoc` is a **confounder** —
it correlates with both `Price` (stores with good shelf space can charge higher prices)
and with `Sales` (good shelf space directly increases sales).

**Solution:** Include `ShelveLoc` and `Advertising` in the model.
:::

---

## Step 1: Understanding Dummy Encoding

### How R Handles Categorical Variables

```{r dummy-check}
#| code-fold: false
# ShelveLoc is in class character — R converts automatically
class(tbl_carseats$ShelveLoc)
levels(factor(tbl_carseats$ShelveLoc))
```

R automatically encodes `ShelveLoc` into **dummy variables** (one-hot encoding with a reference category):

```{r dummy-viz}
#| echo: false
tibble(
  ShelveLoc       = c("Bad", "Medium", "Good"),
  `(Intercept)`   = c("✅ Referenz", "—", "—"),
  ShelveLocGood   = c(0, 0, 1),
  ShelveLocMedium = c(0, 1, 0)
) %>%
  gt() %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = ShelveLoc == "Bad")
  ) %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = ShelveLoc == "Good")
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("Bad = Referenzkategorie (alphabetisch erste Kategorie)")
```

:::{.callout-note}
## Why a reference category?

With $k$ categories, we only need $k-1$ dummy variables—otherwise, **perfect multicollinearity** occurs
(the dummy variables always sum to 1).

The reference category (**Bad**) is included in the **intercept**.
All other coefficients measure the **difference** from the reference.
:::

### Visualization: Sales by ShelveLoc

```{r shelveloc-viz}
tbl_carseats %>%
  ggplot(aes(x = ShelveLoc, y = Sales, fill = ShelveLoc)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.4) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, size = 3, fill = "white") +
  scale_fill_manual(values = c("Bad"    = "#D50006",
                               "Medium" = "#686868",
                               "Good"   = "#502479")) +
  labs(
    title    = "Sales nach Regalplatz (ShelveLoc)",
    subtitle = "Raute = Mittelwert | Linie = Median",
    x = "Regalplatz", y = "Sales (Tsd. Einheiten)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

---

## Step 2: Estimate multiple regression model

### The model

$$\widehat{\text{Sales}} = \hat\beta_0 + \hat\beta_1 \cdot \text{Price} + \hat\beta_2 \cdot \text{Advertising} + \hat\beta_3 \cdot \mathbf{1}_{\text{Good}} + \hat\beta_4 \cdot \mathbf{1}_{\text{Medium}}$$

```{r multipel-fit}
#| code-fold: false
mod_multipel <- tbl_carseats %>%
  lm(Sales ~ Price + Advertising + ShelveLoc, data = .)

mod_multipel %>%
  get_regression_table() %>%
  gt() %>%
  fmt_number(columns = c(estimate, std_error, statistic,
                         p_value, lower_ci, upper_ci),
             decimals = 4) %>%
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_body(rows = str_detect(term, "ShelveLoc"))
  ) %>%
  tab_style(
    style     = cell_fill(color = "#fff0f0"),
    locations = cells_body(rows = term == "Price")
  ) %>%
  tab_style(
    style     = cell_fill(color = "#f0fff8"),
    locations = cells_body(rows = term == "Advertising")
  ) %>%
  tab_options(table.font.size = "14px") %>%
  tab_footnote("Lila = ShelveLoc-Dummies | Rot = Price | Grün = Advertising")
```

---

## Step 3: Interpret coefficients

### Ceteris-Paribus-Logic

:::{.callout-important}
## What does *ceteris paribus* mean?

Each coefficient in the multiple model measures the effect of **one** variable
while **holding all others constant** (“all else being equal”).

> *"Multiple regression does not tell us what happens if we change Price. 
> It tells us what the data shows about Price AFTER removing the influence 
> of all other variables."*
> — @provost_data_2013, S. 147
:::

```{r interpretation-table}
#| echo: false
coefs <- coef(mod_multipel)

tibble(
  Koeffizient    = names(coefs),
  Schätzwert     = round(coefs, 4),
  Interpretation = c(
    paste0("Ø Sales = ", round(coefs[1], 2),
           " Tsd. bei Price=0, Adv=0, ShelveLoc=Bad"),
    paste0("Pro +1 USD Preis → ",
           round(coefs[2], 4),
           " Tsd. weniger Sales (c.p.)"),
    paste0("Pro +1 Tsd. USD Werbung → +",
           round(coefs[3], 4),
           " Tsd. Sales (c.p.)"),
    paste0("Good vs. Bad Regalplatz: +",
           round(coefs[4], 3),
           " Tsd. Sales (c.p.)"),
    paste0("Medium vs. Bad Regalplatz: +",
           round(coefs[5], 3),
           " Tsd. Sales (c.p.)")
  )
) %>%
  gt() %>%
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_body(rows = str_detect(Koeffizient, "ShelveLoc"))
  ) %>%
  tab_options(table.font.size = "13px")
```

### Live Computation: What is the typical branch selling?

```{r prediction}
#| code-fold: false
# Beispiel: Filiale mit Price=120, Advertising=10, ShelveLoc="Good"
predict(mod_multipel,
        newdata = data.frame(
          Price       = 120,
          Advertising = 10,
          ShelveLoc   = "Good"
        ))
```

```{r prediction-manual}
# Manuell nachrechnen (zeigt ceteris paribus live)
beta <- coef(mod_multipel)
beta["(Intercept)"] +
  beta["Price"] * 120 +
  beta["Advertising"] * 10 +
  beta["ShelveLocGood"] * 1   # Good = 1
```

---

## Step 4:  Compare model quality (fit)

### Baseline vs. Multiple Model

```{r model-comparison}
bind_rows(
  mod_baseline %>%
    get_regression_summaries() %>%
    mutate(modell = "Baseline: Sales ~ Price"),
  mod_multipel %>%
    get_regression_summaries() %>%
    mutate(modell = "Multipel: Sales ~ Price + Advertising + ShelveLoc")
) %>%
  select(modell, r_squared, adj_r_squared, rmse, sigma) %>%
  gt() %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse, sigma),
             decimals = 4) %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = str_detect(modell, "Multipel"))
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("Grün = deutliche Verbesserung durch zusätzliche Variablen")
```

### Change in the price coefficient (OVB adjustment)

```{r ovb-correction}
#| code-fold: false
cat("β_Price ohne ShelveLoc:", round(coef(mod_baseline)["Price"], 4), "\n")
cat("β_Price mit ShelveLoc: ", round(coef(mod_multipel)["Price"], 4), "\n")
cat("Veränderung:           ",
    round(coef(mod_multipel)["Price"] - coef(mod_baseline)["Price"], 4), "\n")
```

:::{.callout-note}
## Why does the price coefficient change?

In the baseline model, $\hat\beta_{\text{Price}}$ was **biased** by the confounder `ShelveLoc`:

Stores with good shelf space can charge higher prices AND have higher sales.
This “hidden” pattern made the price coefficient appear less negative than it actually is.

In the multiple model, this influence is **controlled for**—the price effect is measured **ceteris paribus**.
:::

---

## Step 5: Residual diagnostics

```{r resid-comparison}
pts_base <- augment(mod_baseline, data = tbl_carseats)
pts_mult <- augment(mod_multipel, data = tbl_carseats)

p1 <- pts_base %>%
  ggplot(aes(.fitted, .resid, color = ShelveLoc)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#686868") +
  scale_color_manual(values = c("Bad" = "#D50006",
                                "Medium" = "#b3b2b2",
                                "Good" = "#502479")) +
  labs(title = "Baseline", subtitle = "Muster sichtbar!",
       x = "Ŷ", y = "Residuum") +
  theme_minimal(base_size = 10) + theme(legend.position = "none")

p2 <- pts_mult %>%
  ggplot(aes(.fitted, .resid, color = ShelveLoc)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#686868") +
  scale_color_manual(values = c("Bad" = "#D50006",
                                "Medium" = "#b3b2b2",
                                "Good" = "#502479")) +
  labs(title = "Multiples Modell",
       subtitle = "Muster weitgehend verschwunden ✅",
       x = "Ŷ", y = "Residuum") +
  theme_minimal(base_size = 10) + theme(legend.position = "bottom")

p1 + p2 +
  plot_annotation(
    title    = "Residualplot: Baseline vs. Multiples Modell",
    subtitle = "Nach Aufnahme von ShelveLoc verschwindet das systematische Muster"
  )
```

---

## Step 6: Check for Multicollinearity

```{r multicollinearity}
#| code-fold: false
# Korrelationsmatrix der numerischen Prädiktoren
tbl_carseats %>%
  select(Price, Advertising, Income, Age, Population) %>%
  cor() %>%
  round(3)
```

```{r vif}
# Variance Inflation Factor (VIF) — Faustregel: VIF > 5 problematisch
library(car)
vif(mod_multipel)
```

:::{.callout-note}
## Rule of thumb: VIF-Interpretation

| VIF | Assessment |
|-----|-------------|
| 1–2 | No problem |
| 2–5 | Mild multicollinearity |
| 5–10 | Problematic — coefficients unstable |
| > 10 | Very problematic |

For our model: all VIF values are low → no multicollinearity problem.
:::

---

## Step 7: Compute Performance-Scores 

```{r performance-scores}
#| code-fold: false
filial_scored <- mod_multipel %>%
  get_regression_points() %>%
  mutate(performance_gap = residual) %>%
  arrange(performance_gap)

# Top 5 Underperformer
filial_scored %>%
  head(5) %>%
  select(ID, Sales, Sales_hat, residual) %>%
  gt() %>%
  fmt_number(columns = c(Sales, Sales_hat, residual), decimals = 2) %>%
  tab_style(
    style     = cell_fill(color = "#fff0f0"),
    locations = cells_body()
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header("Top 5 Underperformer (stärkste negative Residuen)")
```

```{r performance-viz}
filial_scored %>%
  ggplot(aes(x = Sales_hat, y = performance_gap)) +
  geom_point(
    aes(color = performance_gap < -2),
    alpha = 0.6, size = 1.8
  ) +
  geom_hline(yintercept = 0, color = "#D50006",
             linewidth = 1.1, linetype = "dashed") +
  geom_hline(yintercept = c(-2, 2), color = "#686868",
             linewidth = 0.6, linetype = "dotted") +
  scale_color_manual(values = c("FALSE" = "#b3b2b2",
                                "TRUE"  = "#D50006"),
                     labels = c("Normal", "Underperformer")) +
  labs(
    title    = "Performance-Score: Residuen des multiplen Modells",
    subtitle = "Roter Bereich = stark negatives Residuum = Underperformer",
    x = "Vorhergesagter Umsatz (Sales_hat)",
    y = "Performance-Gap (Residuum)",
    color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

## Overall Comparison: All Three Models

```{r full-comparison}
mod_adv <- lm(Sales ~ Price + Advertising, data = tbl_carseats)

bind_rows(
  get_regression_summaries(mod_baseline) %>%
    mutate(modell = "① Sales ~ Price"),
  get_regression_summaries(mod_adv) %>%
    mutate(modell = "② Sales ~ Price + Advertising"),
  get_regression_summaries(mod_multipel) %>%
    mutate(modell = "③ Sales ~ Price + Advertising + ShelveLoc")
) %>%
  select(modell, r_squared, adj_r_squared, rmse) %>%
  gt() %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse), decimals = 4) %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = 2)
  ) %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = 3)
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("Grün = bestes Modell | Gelb = mittleres Modell")
```

:::{.callout-important}
## Summary of the Demo

| Model | R² | Key Takeaway |
|--------|-----|----------------|
| Baseline `Sales ~ Price` | ~0.15 | OVB via ShelveLoc — Price alone is not enough |
| + Advertising | ~0.27 | Advertising budget explains additional variance |
| + ShelveLoc | ~0.53 | Shelf location is the strongest predictor |

**Next Step (Lab 3):** Use the residuals from the multiple model as a **performance score** 
to create the list of underperformers and draft the managerial memo.
:::

## References {.unnumbered}
