---
title: "Lab 4: Performance Mining & Managerial Memo"
subtitle: "CRISP-DM Phase 5 & 6 — Evaluation & Deployment"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en-US
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse)
library(moderndive)
library(broom)
library(gt)
library(patchwork)
library(GGally)
library(gapminder)
library(ISLR2)

tbl_carseats <- tibble(Carseats)

# Multiple model from the live demo — available directly
mod_multipel <- tbl_carseats %>%
  lm(Sales ~ Price + Advertising + ShelveLoc, data = .)
```

## Learning Objectives and Overview

:::{.callout-note}
## Learning Objectives of This Lab

- Use the multiple regression model as a **scoring tool**
- Calculate and rank **residuals** as a quantitative measure of underperformance
- Communicate results in a **managerial memo** that is suitable for management (translator role)
- Critically reflect on the model's limitations: R², OVB, **data drought**, make-or-buy
:::

**Scenario:** The live demo showed that the multiple model
`Sales ~ Price + Advertising + ShelveLoc` explains approximately 54% of the variance in sales.
Now we use the **residuals** as a performance score and generate the list of underperformers —
this corresponds to CRISP-DM Phase 5 (Evaluation) and Phase 6 (Deployment).

:::{.callout-tip}
## Schedule
⏱ **15 min.** Required tasks A1–A3 | **15 min.** Group work: Managerial Memo
:::

---

## Background Material: An In-Depth Look at Multiple Regression

This section contains supplementary background material that you can read before or after
the lab. It explores the concepts from the live demo in greater depth using
additional datasets.

::: {.panel-tabset}

### Simpson's Paradox

**The problem:** An aggregated trend may be opposite to the trend within
subgroups — if a confounder influences both.

In the car-seat dataset: stores with good shelf placement (Good) can charge higher prices
**and** have higher sales. This masks the negative price effect in the simple model.

```{r}
#| label: simpson-viz
#| message: false
p_agg <- tbl_carseats %>%
  ggplot(aes(x = Price, y = Sales)) +
  geom_point(alpha = 0.3, color = "#686868", size = 1.5) +
  geom_smooth(method = "lm", color = "#D50006",
              se = FALSE, linewidth = 1.3) +
  annotate("text", x = 70, y = 16,
           label = paste0("Overall: r = ",
                          round(cor(tbl_carseats$Price,
                                    tbl_carseats$Sales), 3)),
           color = "#D50006", size = 3.5, hjust = 0) +
  labs(title = "Aggregated (without groups)",
       x = "Price (USD)", y = "Sales (thousands)") +
  theme_minimal(base_size = 10)

p_grp <- tbl_carseats %>%
  ggplot(aes(x = Price, y = Sales, color = ShelveLoc)) +
  geom_point(alpha = 0.35, size = 1.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.1) +
  scale_color_manual(values = c("Bad"    = "#D50006",
                                "Medium" = "#686868",
                                "Good"   = "#502479")) +
  labs(title = "By shelf location (ShelveLoc)",
       subtitle = "Within each group: same negative trend",
       x = "Price (USD)", y = "Sales (thousands)",
       color = "ShelveLoc") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")

(p_agg + p_grp) +
  plot_annotation(
    title    = "Simpson's Paradox: Aggregate vs. Subgroups",
    subtitle = "Same data — but two very different pictures"
  )
```

**Takeaway:** Always look for confounders and perform subgroup analyses
before interpreting aggregate results. The multiple model solves this problem
by controlling for ShelveLoc.

### Categorical Regression (Gapminder)

**How are dummy variables interpreted?**
The following example using the Gapminder dataset illustrates the logic
using continents — directly applicable to `ShelveLoc`.

```{r gapminder-setup}
gapminder2007 <- gapminder %>%
  filter(year == 2007) %>%
  select(country, continent, lifeExp, gdpPercap)

# Descriptive statistics
gapminder2007 %>%
  group_by(continent) %>%
  summarise(
    n            = n(),
    mean_lifeExp = round(mean(lifeExp), 1),
    sd_lifeExp   = round(sd(lifeExp),   1)
  ) %>%
  arrange(mean_lifeExp) %>%
  gt() %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = continent == "Africa")
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("Yellow = Reference group (alphabetical order)")
```

```{r gapminder-boxplot}
gapminder2007 %>%
  ggplot(aes(x = reorder(continent, lifeExp),
             y = lifeExp, fill = continent)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, size = 3, fill = "white") +
  scale_fill_manual(values = c(
    "Africa"   = "#D50006", "Americas" = "#7d0a52",
    "Asia"     = "#502479", "Europe"   = "#686868",
    "Oceania"  = "#b3b2b2"
  )) +
  labs(title   = "Life Expectancy 2007 by Continent",
       subtitle = "Diamond = Mean | Sorted by average life expectancy",
       x = NULL, y = "Life Expectancy (years)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

```{r gapminder-model}
mod_lifeexp <- gapminder2007 %>%
  lm(lifeExp ~ continent, data = .)

mod_lifeexp %>%
  get_regression_table() %>%
  gt() %>%
  fmt_number(columns = c(estimate, std_error,
                         statistic, p_value,
                         lower_ci, upper_ci),
             decimals = 2) %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = term == "(Intercept)")
  ) %>%
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_body(rows = term != "(Intercept)")
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(
    "Regression: lifeExp ~ continent",
    subtitle = "Intercept = Average Africa | All others = Difference from Africa"
  )
```

**Interpretation of coefficients (directly applicable to Carseats):**

| Gapminder | Carseats |
|-----------|---------| 
| Intercept = Average life expectancy in Africa (reference) | Intercept = Average sales at Bad (reference) |
| `continentAmericas` = Difference between Americas and Africa | `ShelveLocGood` = Difference between Good and Bad |
| Reference comes first alphabetically | Reference comes first alphabetically |

**Verify numerically:**

```{r gapminder-check}
# Intercept = average lifeExp in Africa
gapminder2007 %>%
  filter(continent == "Africa") %>%
  summarise(mean_africa = round(mean(lifeExp), 2))

# Intercept + continentAmericas = average Americas
coef(mod_lifeexp)["(Intercept)"] +
  coef(mod_lifeexp)["continentAmericas"]

gapminder2007 %>%
  filter(continent == "Americas") %>%
  summarise(mean_americas = round(mean(lifeExp), 2))
```

### Multiple Regression: 2 Numerical Regressors

**Concept:** When there are two numerical regressors, OLS does not estimate a straight line,
but rather a **plane** in three-dimensional space.

The following example uses credit card data from ISLR2.

```{r credit-eda}
tbl_credit <- tibble(Credit) %>%
  select(Balance, Income, Limit, Rating, Age)

# Scatterplot matrix
tbl_credit %>%
  select(Balance, Income, Limit) %>%
  ggpairs(
    upper    = list(continuous = wrap("cor", size = 4)),
    lower    = list(continuous = wrap("points",
                                      alpha = 0.3, size = 0.8)),
    diag     = list(continuous = wrap("densityDiag", alpha = 0.6)),
    progress = FALSE
  ) +
  theme_minimal(base_size = 9) +
  labs(title = "Credit card data: Balance, Income, Limit")
```

```{r credit-model}
mod_credit <- tbl_credit %>%
  lm(Balance ~ Income + Limit, data = .)

mod_credit %>%
  get_regression_table() %>%
  gt() %>%
  fmt_number(columns = c(estimate, std_error,
                         statistic, p_value,
                         lower_ci, upper_ci),
             decimals = 4) %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = term != "(Intercept)")
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(
    "Multiple Regression: Balance ~ Income + Limit",
    subtitle = "Both coefficients: ceteris paribus"
  )
```

**Interpretation of the coefficients:**

- **Income** (`r round(coef(mod_credit)["Income"], 4)`):
  $1,000 more income → credit card balance changes by
  `r round(coef(mod_credit)["Income"], 2)` USD, **ceteris paribus** (limit constant).
- **Limit** (`r round(coef(mod_credit)["Limit"], 4)`):
  $1 more credit limit → balance changes by
  `r round(coef(mod_credit)["Limit"], 4)` USD, **ceteris paribus** (income constant).

```{r credit-r2}
mod_credit %>%
  get_regression_summaries() %>%
  select(r_squared, adj_r_squared, rmse) %>%
  gt() %>%
  fmt_number(decimals = 4) %>%
  tab_options(table.font.size = "13px")
```

**Application to Carseats:** The Carseats model combines
*numeric* predictors (Price, Advertising) with a
*categorical* predictor (ShelveLoc) — this is the general case
of multiple regression.

:::

---

## Tasks: Performance Mining

### Step 1 — Review the Model

The multiple model from the live demo:

```{r model-recap}
#| code-fold: false
# Model identical to the demo
mod_multipel <- tbl_carseats %>%
  lm(Sales ~ Price + Advertising + ShelveLoc, data = .)

# Model quality
mod_multipel %>%
  get_regression_summaries() %>%
  select(r_squared, adj_r_squared, rmse, sigma, nobs) %>%
  gt() %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse, sigma),
             decimals = 4) %>%
  tab_options(table.font.size = "14px")
```

```{r model-coefs}
# Coefficients
mod_multipel %>%
  get_regression_table() %>%
  gt() %>%
  fmt_number(columns = c(estimate, std_error,
                         statistic, p_value,
                         lower_ci, upper_ci),
             decimals = 4) %>%
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_body(rows = str_detect(term, "ShelveLoc"))
  ) %>%
  tab_options(table.font.size = "13px")
```

### Step 2 — Compute Performance Scores

```{r scores}
#| code-fold: false
store_scored <- mod_multipel %>%
  get_regression_points() %>%
  mutate(
    performance_gap = residual,
    performance_cat = case_when(
      residual < -2 ~ "Underperformer",
      residual >  2 ~ "Outperformer",
      TRUE          ~ "Normal"
    )
  )

# Overview
store_scored %>%
  count(performance_cat) %>%
  mutate(share = scales::percent(n / sum(n), accuracy = 0.1)) %>%
  arrange(desc(n)) %>%
  gt() %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-note}
## Residual = Measure for Underperformance

$$e_i = \text{Sales}_i - \widehat{\text{Sales}}_i$$

- $e_i \ll 0$: Store sells **significantly less** than the benchmark → **underperformer**
- $e_i \approx 0$: Store is at the market average
- $e_i \gg 0$: Store sells **more** than expected → **outperformer**

**Important:** The benchmark is already *adjusted* for price, advertising, and shelf space
— so an underperformer is truly worse than
stores with comparable starting conditions.
:::

### Step 3 — Top-10 Ranking of Underperformers

```{r top10}
# Add contextual store metadata
store_scored %>%
  arrange(performance_gap) %>%
  head(10) %>%
  left_join(
    tbl_carseats %>%
      mutate(ID = row_number()) %>%
      select(ID, ShelveLoc, Urban, US, Income, Age),
    by = "ID"
  ) %>%
  select(ID, Sales, Sales_hat, performance_gap,
         ShelveLoc.x, Urban, US, Income, Age) %>%
  rename(ShelveLoc = ShelveLoc.x) %>%
  gt() %>%
  fmt_number(columns = c(Sales, Sales_hat, performance_gap),
             decimals = 2) %>%
  fmt_number(columns = c(Income, Age), decimals = 0) %>%
  tab_style(
    style     = cell_fill(color = "#fff0f0"),
    locations = cells_body(rows = performance_gap < -3)
  ) %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = performance_gap >= -3 &
                             performance_gap < -2)
  ) %>%
  tab_options(table.font.size = "12px") %>%
  tab_header(
    "Top-10 Underperformers",
    subtitle = "Dark red: gap < −3 | Yellow: gap between −3 and −2"
  )
```

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A1 — Analyse Underperformers

**1.** Which **ShelveLoc category** dominates the top 10 underperformers?
Test this empirically:

```r
store_scored %>%
  left_join(
    tbl_carseats %>% mutate(ID = row_number()) %>%
      select(ID, ShelveLoc),
    by = "ID"
  ) %>%
  filter(performance_cat == "Underperformer") %>%
  count(ShelveLoc) %>%
  mutate(share = scales::percent(n / sum(n)))
```

**2.** Is there a **store with prime shelf space** among the top 10 underperformers?
If so, why is this particularly interesting for management?

**3.** What do the **top 3 underperformers** have in common?
Compare income, age, and urban/rural location.
:::

### Step 4 — Visualise the Performance Gap

```{r performance-plot}
#| fig-height: 5
# Plot 1: Scatter Performance Gap vs. Predicted
p1 <- store_scored %>%
  ggplot(aes(x = Sales_hat, y = performance_gap,
             color = performance_cat)) +
  geom_point(alpha = 0.6, size = 1.8) +
  geom_hline(yintercept = 0, color = "#333",
             linewidth = 0.8, linetype = "dashed") +
  geom_hline(yintercept = c(-2, 2), color = "#b3b2b2",
             linewidth = 0.5, linetype = "dotted") +
  scale_color_manual(
    values = c("Underperformer" = "#D50006",
               "Normal"         = "#b3b2b2",
               "Outperformer"   = "#502479")
  ) +
  labs(
    title    = "Performance Gap by Predicted Sales",
    subtitle = "Threshold: ±2k units",
    x        = "Predicted Ŝales",
    y        = "Performance Gap (Residual)",
    color    = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")

# Plot 2: Top/Bottom 10 as horizontal bar chart
p2 <- bind_rows(
  store_scored %>% arrange(performance_gap) %>%
    head(10) %>% mutate(group = "Underperformer"),
  store_scored %>% arrange(desc(performance_gap)) %>%
    head(10) %>% mutate(group = "Outperformer")
) %>%
  ggplot(aes(x = reorder(factor(ID), performance_gap),
             y = performance_gap, fill = group)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = 0, color = "#333",
             linewidth = 0.5) +
  scale_fill_manual(values = c("Underperformer" = "#D50006",
                               "Outperformer"   = "#502479")) +
  coord_flip() +
  labs(
    title = "Top-10 Under- and Outperformers",
    x     = "Store ID",
    y     = "Performance Gap (in thousands)",
    fill  = NULL
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "bottom")

p1 + p2
```

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A2 — Interpreting the Visualisation

**1.** Is there a **pattern**: are stores with high predicted sales
more likely to be outperformers, or is the distribution even?

**2.** Calculate performance scores **separately by ShelveLoc**:

```r
store_scored %>%
  left_join(
    tbl_carseats %>% mutate(ID = row_number()) %>%
      select(ID, ShelveLoc),
    by = "ID"
  ) %>%
  group_by(ShelveLoc) %>%
  summarise(
    mean_gap = round(mean(performance_gap), 3),
    sd_gap   = round(sd(performance_gap),   3),
    n_under  = sum(performance_cat == "Underperformer"),
    n_over   = sum(performance_cat == "Outperformer"),
    n_total  = n()
  )
```

**3.** A **store that is outperforming despite having a prime shelf position** might be worth investigating:
Why? What would management do with this information?
:::

### Step 5 — Model Evaluation

```{r}
#| label: model-eval-plots
#| message: false
pts <- augment(mod_multipel, data = tbl_carseats)

# Residual plot
p_res <- pts %>%
  ggplot(aes(.fitted, .resid)) +
  geom_point(alpha = 0.4, color = "#7d0a52", size = 1.5) +
  geom_hline(yintercept = 0, color = "#D50006",
             linewidth = 1, linetype = "dashed") +
  geom_smooth(method = "loess", color = "#502479",
              se = FALSE, linewidth = 0.8) +
  labs(title = "Residual Plot: Multiple Model",
       subtitle = "No clear systematic pattern — but remaining scatter",
       x = "Predicted value", y = "Residual") +
  theme_minimal(base_size = 10)

# QQ plot
p_qq <- pts %>%
  ggplot(aes(sample = .resid)) +
  stat_qq(color = "#7d0a52", alpha = 0.6) +
  stat_qq_line(color = "#D50006", linewidth = 0.8) +
  labs(title = "QQ Plot of Residuals",
       subtitle = "Approximately normal → important for inference in PVA2",
       x = "Theoretical quantiles", y = "Observed quantiles") +
  theme_minimal(base_size = 10)

p_res + p_qq
```

```{r r2-comparison}
# Three-model comparison
mod_basis <- lm(Sales ~ Price, data = tbl_carseats)
mod_adv   <- lm(Sales ~ Price + Advertising, data = tbl_carseats)

bind_rows(
  get_regression_summaries(mod_basis) %>%
    mutate(model = "① Sales ~ Price"),
  get_regression_summaries(mod_adv) %>%
    mutate(model = "② Sales ~ Price + Advertising"),
  get_regression_summaries(mod_multipel) %>%
    mutate(model = "③ Sales ~ Price + Advertising + ShelveLoc")
) %>%
  select(model, r_squared, adj_r_squared, rmse) %>%
  gt() %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse),
             decimals = 4) %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = 2)
  ) %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = 3)
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header("Model Comparison: R² Progression")
```

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A3 — Model Evaluation

**1. Residual plot:** Has the ShelveLoc pattern from Lab 3 disappeared?
What residual scatter do you still see?

**2. R² ≈ 0.54:** For which decisions is this explanatory power sufficient?
Based on this model alone, would you recommend closing a store?

**3. Data Drought:** Which variables are missing from the model?
Classify according to data strategy:

| Category | Examples |
|-----------|----------|
| Present in the dataset, not yet in the model | ? |
| Externally obtainable (commodity data) | ? |
| Can only be collected internally (proprietary data) | ? |
:::

---

## Group Assignment: Managerial Memo

:::{.callout-caution icon=false}
## `r fontawesome::fa("users")` Group assignment (3–4 persons, 15 minutes)

### Task

Write a short **memo to the CEO** (max. 300 words).
You are the data science team — the CEO is not a statistician.

Answer all three questions clearly, concisely, and in a way that is suitable for management.

---

### Question 1 — Model Evaluation: Is R² Sufficient?

> *"Our model explains R² ≈ 54% of the variance in sales.
> Would you recommend that management close stores based on this model?
> Explain your recommendation."*

**What makes for a strong answer?**

<table style="font-size:0.84em; border-collapse:collapse; width:100%;">
<thead>
<tr>
  <th style="padding:5px 10px; background:#f5f5f5; border:1px solid #ddd; width:50%;">Weak answer</th>
  <th style="padding:5px 10px; background:#f5f5f5; border:1px solid #ddd; width:50%;">Strong answer</th>
</tr>
</thead>
<tbody>
<tr>
  <td style="padding:5px 10px; border:1px solid #ddd; vertical-align:top;">
    "R² = 0.54. More than 50% is good enough."
  </td>
  <td style="padding:5px 10px; border:1px solid #ddd; vertical-align:top;">
    "R² = 0.54 means that 46% of the variance in revenue is attributable to unknown factors.
    A decision to close a location based on this information carries a significant risk of error —
    we recommend first supplementing the data in a targeted manner before taking irreversible
    action."
  </td>
</tr>
</tbody>
</table>

---

### Question 2 — OVB: What Variables Are Missing?

> *"What factors are missing from our data that could explain
> why a store is underperforming?
> Give three specific examples and explain why."*

**What makes for a strong answer?**

<table style="font-size:0.84em; border-collapse:collapse; width:100%;">
<thead>
<tr>
  <th style="padding:5px 10px; background:#f5f5f5; border:1px solid #ddd; width:50%;">Weak answer</th>
  <th style="padding:5px 10px; background:#f5f5f5; border:1px solid #ddd; width:50%;">Strong answer</th>
</tr>
</thead>
<tbody>
<tr>
  <td style="padding:5px 10px; border:1px solid #ddd; vertical-align:top;">
    "We need more data about our clients."
  </td>
  <td style="padding:5px 10px; border:1px solid #ddd; vertical-align:top;">
    "Population density within a 5-km radius: stores in sparsely populated regions have
    structurally lower sales — regardless of price, advertising, and shelf space.
    This factor is missing and systematically skews the score for rural stores downward."
  </td>
</tr>
</tbody>
</table>

---

### Question 3 — Data Strategy: Make or Buy?

> *"Should we purchase external geodata and demographic data from a data vendor
> (Commodity Data) or build our own customer survey (Proprietary Data)?
> Argue using the AI Value Creator concept from Thomas et al. (2025)."*

**What makes for a strong answer?**

<table style="font-size:0.84em; border-collapse:collapse; width:100%;">
<thead>
<tr>
  <th style="padding:5px 10px; background:#f5f5f5; border:1px solid #ddd; width:50%;">Weak answer</th>
  <th style="padding:5px 10px; background:#f5f5f5; border:1px solid #ddd; width:50%;">Strong answer</th>
</tr>
</thead>
<tbody>
<tr>
  <td style="padding:5px 10px; border:1px solid #ddd; vertical-align:top;">
    "We should buy data — it is faster and cheaper."
  </td>
  <td style="padding:5px 10px; border:1px solid #ddd; vertical-align:top;">
    "Geodata (Commodity Data) is available to all competitors —
    no sustainable advantage. An own loyalty card with purchase history
    (Proprietary Data) creates a data moat.
    Recommendation: buy commodity data short-term (pilot project), build
    a loyalty programme long-term for a lasting competitive advantage
    (Thomas et al., 2025: AI Value Creator)."
  </td>
</tr>
</tbody>
</table>
:::

---

## Debrief: Make-or-Buy and Expected Value Framework

### When Does Purchasing Data Pay Off?

$$\underbrace{E[\Delta\text{Profit}]}_{\text{Value of better decisions}} > \underbrace{C_{\text{Data}}}_{\text{Acquisition cost}}$$

```{r ev-table}
#| echo: false
tibble(
  Parameter = c(
    "Cost of demographic data (commodity)",
    "Cost of a wrong store closure",
    "P(Wrong decision without data)",
    "P(Wrong decision with data)",
    "E[Δ Profit]",
    "Decision"
  ),
  Value = c(
    "EUR 50,000/year",
    "EUR 500,000",
    "30%",
    "10%",
    "(0.30 − 0.10) × 500,000 = **EUR 100,000**",
    "✅ Purchasing data pays off — 100,000 > 50,000"
  )
) %>%
  gt() %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = Parameter == "Decision")
  ) %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = Parameter == "E[Δ Profit]")
  ) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-note}
## From AI User to AI Value Creator

The pilot project (today) has shown: **Commodity Data** (the public Carseats dataset)
is sufficient for a proof of concept — but not for sustainable competitive advantages.

The path to becoming an **AI Value Creator** (Thomas et al., 2025):

1. **Short-term:** Purchase external geodata (commodity) → improve the model
2. **Medium-term:** Build a loyalty programme → own transaction data
3. **Long-term:** Train models with proprietary data → not replicable by competitors
:::

---

## CRISP-DM: Wrap-Up

```{r crisp-summary}
#| echo: false
tibble::tribble(
  ~Phase,                       ~Result,
  "① Business Understanding",   "Underperformer = store with a strongly negative residual",
  "② Data Understanding",       "EDA: Sales slightly right-skewed, ShelveLoc as confounder",
  "③ Data Preparation",         "ShelveLoc as dummy variable, Advertising included",
  "④ Modeling",                 paste0("Multiple model: R² ≈ ",
                                       round(summary(mod_multipel)$r.squared, 2),
                                       " | three predictors"),
  "⑤ Evaluation",               "Residual plot: OVB resolved, no ShelveLoc pattern remaining",
  "⑥ Deployment",               "Underperformer ranking + Managerial Memo + Make-or-Buy"
) %>%
  gt() %>%
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = Phase == "⑥ Deployment")
  ) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-important}
## Key Messages of This Lab

1. **Residual = Performance Score:** The multiple model provides a price-,
   advertising-, and shelf-location-adjusted benchmark for every store.

2. **R² ≈ 0.54** is a starting point — 46% unexplained variance motivates
   targeted data enrichment, not immediate store closures.

3. **Translator role:** Not the model, but the actionable recommendation is the value —
   translating statistical findings into decision logic.

4. **Make or Buy:** Commodity data for immediate improvement,
   proprietary data for lasting competitive advantage.
:::

:::{.callout-note}
## Outlook: What Comes in PVA2?

Today we worked **descriptively**: all 400 stores = population.

**Open question:** What if the 400 stores were only a **sample**?
How certain are our coefficients? Would we have obtained the same results in a different season?
And how likely is it that the measured advertising effect is purely coincidental?

These are the questions of **statistical inference** — the topic of PVA2:
Bootstrap, confidence intervals, p-hacking, FDR, and the Expected Value Framework
for managers.
:::

## References {.unnumbered}
