<!-- ============================================================
INSTRUCTOR CUSTOMISATION — update before rendering for each cohort
Search for the marker text rather than relying on exact line numbers.

  "buli_raw_{date}.rds"   Dataset filename — update date if new scrape
  xfun::from_root(...)    Adjust path if project structure differs
  set.seed(2024)          Change seed for different cohort results
============================================================ -->
---
title: "Session 2 · Lab 1: Resampling & Only One Test"
subtitle: "From Sampling Distribution to Bootstrap — and Why Every Test Is a Regression"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse)    # data wrangling & plotting
library(moderndive)   # get_regression_table(), get_regression_summaries()
library(infer)        # specify-hypothesize-generate-calculate
library(gt)           # formatted tables

date <- '2023-09-06'
my_in_file <- glue::glue('buli_raw_{date}.rds')
tbl_buli_raw <- read_rds(file = xfun::from_root("data","raw", my_in_file))

tbl_buli <- tbl_buli_raw %>%
              mutate(
                age       = lubridate::year(Sys.Date()) -
                            lubridate::year(player_dob),
                log_value = log(player_market_value_euro),
                pos_bin   = if_else(player_position == "Goalkeeper",
                                    "Goalkeeper", "Outfield"),
                # pos_cat: works with both broad ("Defender","Midfield","Attack")
                # and detailed ("Centre-Back","Left Winger") position labels
                pos_cat   = case_when(
                              str_detect(player_position, "(?i)goal|keeper")              ~ "Goalkeeper",
                              str_detect(player_position, "(?i)back|defen")               ~ "Defender",
                              str_detect(player_position, "(?i)mid")                      ~ "Midfielder",
                              str_detect(player_position, "(?i)attack|forw|wing|strik")   ~ "Forward",
                              TRUE                                                        ~ "Forward"
                            ) %>% factor(levels = c("Goalkeeper","Defender",
                                                    "Midfielder","Forward")) %>%
                            droplevels()
              ) %>%
              rename(name = player_name,
                     mv   = player_market_value_euro) %>%
              filter(!is.na(mv), !is.na(age))

options(scipen = 999)
```

## Scenario & Orientation

In Session 1 we fitted a regression of log market value on age and found a positive coefficient. But that was **one dataset, one fit**. Would we get the same coefficient from a different season, a different sample? This lab makes uncertainty *tangible* by simulating what "repeated sampling" actually looks like — and then shows that bootstrap does the same thing from a single sample.

Along the way, we discover a unifying principle: the t-test, ANOVA, and regression are **not** separate techniques. They are all special cases of one framework.

```{r crisp-overview}
#| echo: false
tibble::tribble(
  ~Phase, ~Name,                    ~Content,                                ~Status,
  "1",    "Business Understanding", "Identify under-/overvalued players",    "Done (Session 1)",
  "2",    "Data Understanding",     "EDA of Bundesliga market values",       "Done (Session 1)",
  "3",    "Data Preparation",       "log-transform, position dummy",         "Done (Session 1)",
  "4",    "Modelling",              "OLS: log_value ~ age",                  "Done (Session 1)",
  "5",    "Evaluation",             "Is the age effect real? (inference)",    "← This lab",
  "6",    "Deployment",             "Transfer decision framework",           "Session 2 lecture"
) %>%
  gt() %>%
  tab_style(cell_fill("#f0fff0"), cells_body(rows = Phase == "5")) %>%
  tab_style(cell_fill("#f5f0fa"), cells_column_labels()) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-note}
## Learning Objectives

- Understand the **sampling distribution** as the foundation of all inference
- Experience the difference between a **sampling distribution** and a **bootstrap distribution** visually
- Compute **bootstrap confidence intervals** and **permutation p-values** with the `infer` package
- Recognise that the t-test, ANOVA, and regression are **all the same test** — the "Only One Test" principle
:::

**Navigation guide**

| Symbol | Box colour | Meaning | When to read |
|--------|-----------|---------|--------------|
| ⚠️ | Red | Mandatory task | Live session |
| 💡 | Green | Tip / best practice | Anytime |
| 📚 | Blue (collapsed) | Background theory | At home |
| ℹ️ | Blue (open) | Short clarification | When needed |

:::{.callout-tip}
## ⏱ Time Budget
**45 minutes** — complete mandatory tasks (A1–A5) first; bonus tasks if time permits.
:::

---

## Icebreaker: One Population, Three Stories

Three random subsamples of 30 players each from the same season produce three different regression lines.

```{r icebreaker-data}
set.seed(23)
buli_s1 <- tbl_buli %>% slice_sample(n = 30)
set.seed(7)
buli_s2 <- tbl_buli %>% slice_sample(n = 30)
set.seed(42)
buli_s3 <- tbl_buli %>% slice_sample(n = 30)
```

```{r icebreaker-plot}
#| echo: false
tbl_buli %>%
  ggplot(aes(x = age, y = log_value)) +
  geom_point(alpha = 0.15, color = "#b3b2b2", size = 0.7) +
  geom_smooth(method = "lm", se = FALSE,
              color = "#333", linewidth = 1.4) +
  geom_smooth(data = buli_s1, method = "lm", se = FALSE,
              color = "#D50006", linewidth = 0.9, linetype = "dashed") +
  geom_smooth(data = buli_s2, method = "lm", se = FALSE,
              color = "#502479", linewidth = 0.9, linetype = "dashed") +
  geom_smooth(data = buli_s3, method = "lm", se = FALSE,
              color = "#7d0a52", linewidth = 0.9, linetype = "dashed") +
  labs(x = "Age", y = "log(Market Value)",
       title = "Population (black) vs. 3 random samples (coloured)") +
  theme_minimal(base_size = 11)
```

Each dashed line tells a slightly different story. Some are steeper, some flatter. If we had only seen one, we might draw a very different conclusion.

:::{.callout-important}
## The Core Problem
A single regression coefficient is **one dart thrown at a board**. To judge the thrower's skill, we need many throws. The distribution of those throws is the **sampling distribution** — and it is the foundation of *all* statistical inference. This lab makes that distribution visible.
:::

---

## Step 1: The Point Estimate

Before measuring uncertainty, we need the thing we are uncertain about: the observed slope.

The `infer` package provides a pipeline that works identically for any statistic:

```
data %>% specify() %>% [hypothesize()] %>% generate() %>% calculate()
```

We start with just `specify()` + `calculate()` to get the observed value:

```{r obs-slope}
#| code-fold: false
obs_slope <- tbl_buli %>%
               specify(log_value ~ age) %>%
               calculate(stat = "slope")
obs_slope
```

This is our point estimate: $\widehat{\beta}_{\text{age}}$ = `r round(pull(obs_slope), 4)`. Because the outcome is log-transformed, the coefficient has a percentage interpretation: each additional year of age is associated with approximately $(e^{\hat\beta} - 1) \times 100$ = `r round((exp(pull(obs_slope)) - 1) * 100, 1)`% higher market value.

But a single number without a measure of precision is just a guess. How much would it change if we had a different sample?

---

## Step 2: Sampling Distribution vs. Bootstrap Distribution

This is the conceptual heart of the lab. We compare two distributions:

1. **Sampling distribution** — what happens when we draw many samples from the *population*
2. **Bootstrap distribution** — what happens when we resample from *one sample*

### 2.1 The Sampling Distribution (The Ideal We Cannot Usually See)

We treat the full Bundesliga season (≈`r nrow(tbl_buli)` players) as our population. This is a simplification — in the lecture we discussed that the season is itself one realisation of a data-generating process. But *for the purpose of understanding the sampling distribution*, we pretend these are all the players that could ever exist.

We now draw 500 random samples of n = 50, compute the slope each time, and plot the distribution:

```{r sampling-dist}
#| code-fold: false
#| cache: true
set.seed(2024)

# --- the "true" population slope ---
pop_slope <- tbl_buli %>%
               lm(log_value ~ age, data = .) %>%
               coef() %>%
               pluck("age")

# --- draw 500 samples of n=50, compute slope each time ---
sampling_dist <- map_dbl(1:500, function(i) {
  tbl_buli %>%
    slice_sample(n = 50) %>%
    lm(log_value ~ age, data = .) %>%
    coef() %>%
    pluck("age")
})
```

```{r sampling-dist-plot}
#| echo: false
tibble(slope = sampling_dist) %>%
  ggplot(aes(x = slope)) +
  geom_histogram(bins = 30, fill = "#502479", color = "white", alpha = 0.7) +
  geom_vline(xintercept = pop_slope, color = "#D50006",
             linewidth = 1.2, linetype = "solid") +
  annotate("text", x = pop_slope + 0.002, y = Inf,
           label = paste0("Population slope = ", round(pop_slope, 4)),
           hjust = 0, vjust = 1.5, color = "#D50006", size = 3.5) +
  labs(title    = "Sampling Distribution: 500 samples of n = 50",
       subtitle = "Each bar = one sample's slope coefficient",
       x        = expression(hat(beta)[age]),
       y        = "Count") +
  theme_minimal(base_size = 11)
```

This is what we **normally never see** — the distribution of estimates across many hypothetical replications. It is centred on the "true" population slope (red line) and its spread is the **standard error**.

### 2.2 The Bootstrap Distribution (What We Can Actually Compute)

In reality we have *one* sample, not 500. Bootstrap mimics the sampling process by drawing 1,000 samples **with replacement** from that single sample:

```{r boot-dist}
#| code-fold: false
#| cache: true
set.seed(2024)

# --- take ONE sample of n=50 ---
tbl_sample <- tbl_buli %>% slice_sample(n = 50)

# --- bootstrap: resample 1000x with replacement from that ONE sample ---
boot_dist <- tbl_sample %>%
               specify(log_value ~ age) %>%
               generate(reps = 1000, type = "bootstrap") %>%
               calculate(stat = "slope")
```

### 2.3 The Comparison — Do They Match?

Now we overlay both distributions. If bootstrap works, the two should look similar:

```{r comparison-plot}
#| echo: false
#| fig-height: 5
tbl_compare <- bind_rows(
  tibble(slope = sampling_dist, type = "Sampling (500 × n=50 from population)"),
  tibble(slope = pull(boot_dist, stat),
         type  = "Bootstrap (1000 × n=50 from one sample)")
)

tbl_compare %>%
  ggplot(aes(x = slope, fill = type)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.55,
                 position = "identity") +
  geom_vline(xintercept = pop_slope, color = "#D50006",
             linewidth = 1, linetype = "dashed") +
  scale_fill_manual(values = c(
    "Sampling (500 × n=50 from population)" = "#502479",
    "Bootstrap (1000 × n=50 from one sample)"   = "#D50006"
  )) +
  labs(title    = "Sampling Distribution vs. Bootstrap Distribution",
       subtitle = "Dashed red line = population slope",
       x        = expression(hat(beta)[age]),
       y        = "Count",
       fill     = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

The two distributions have similar shape and spread. The bootstrap is centred on the *sample* slope (slightly off from the population slope), but the **width** — the standard error — is well approximated.

:::{.callout-important}
## The Key Insight
The **sampling distribution** shows what happens with many real samples — we normally can't see it. The **bootstrap distribution** approximates it from a single sample. This is why bootstrap works: it estimates the *variability* of your statistic without needing repeated experiments.
:::

:::{.callout-note collapse="true"}
## 📚 Background: Why Does Bootstrap Work?
The bootstrap treats the observed sample as a stand-in for the population. The central limit theorem guarantees that — for large enough samples — the bootstrap distribution converges to the true sampling distribution, regardless of the underlying data distribution. The bootstrap gets the *spread* right (standard error), even though its *centre* is the sample statistic rather than the population parameter. This is why bootstrap confidence intervals have correct coverage rates.
:::

For the remainder of this lab, we switch to using the **full population** (all ~`r nrow(tbl_buli)` players) as our data and bootstrap from it directly — the same approach as in Lab 2.

```{r boot-full}
#| code-fold: false
#| cache: true
set.seed(2024)

# --- point estimate from full data ---
obs_slope <- tbl_buli %>%
               specify(log_value ~ age) %>%
               calculate(stat = "slope")

# --- bootstrap from full data ---
boot_dist_full <- tbl_buli %>%
                    specify(log_value ~ age) %>%
                    generate(reps = 1000, type = "bootstrap") %>%
                    calculate(stat = "slope")
```

---

## Step 3: Bootstrap Confidence Interval

The bootstrap distribution from Step 2 gives us two ways to construct a 95% confidence interval:

::: {.panel-tabset}

### Percentile Method

Take the middle 95% of the bootstrap distribution — the simplest approach:

```{r ci-pct}
#| code-fold: false
ci_pct <- boot_dist_full %>%
            get_confidence_interval(type = "percentile", level = 0.95)
ci_pct
```

### Standard Error Method

Centre the interval around the observed estimate, using the bootstrap standard error:

```{r ci-se}
#| code-fold: false
ci_se <- boot_dist_full %>%
           get_confidence_interval(
             type           = "se",
             point_estimate = obs_slope,
             level          = 0.95
           )
ci_se
```

### Visualisation

```{r ci-viz}
#| echo: false
visualize(boot_dist_full) +
  shade_confidence_interval(
    endpoints = ci_pct, fill = NULL,
    linetype = "solid", color = "#D50006"
  ) +
  shade_confidence_interval(
    endpoints = ci_se, fill = NULL,
    linetype = "dashed", color = "#502479"
  ) +
  labs(
    title    = "Bootstrap CI: Percentile (red) vs. Standard Error (purple)",
    subtitle = "Both methods yield similar intervals",
    x        = expression(hat(beta)[age]),
    y        = "Count"
  )
```

:::

Neither interval contains zero → the age effect is **statistically significant** at the 5% level.

:::{.callout-important}
## Correct Interpretation of the CI
Under repeated sampling, 95% of confidence intervals constructed this way would contain the true parameter $\beta_{\text{age}}$.

**Not:** "There is a 95% probability that the true value lies in this interval."
:::

---

## Step 4: Permutation Test

The permutation test approaches inference from the opposite direction: instead of estimating how much the coefficient *varies*, it simulates a **world where there is no relationship at all**. By randomly shuffling the age values, any systematic association between age and market value is destroyed.

```{r perm-test}
#| code-fold: false
#| cache: true
set.seed(2024)

null_dist <- tbl_buli %>%
               specify(log_value ~ age) %>%
               hypothesize(null = "independence") %>%
               generate(reps = 1000, type = "permute") %>%
               calculate(stat = "slope")

p_val <- null_dist %>%
           get_p_value(obs_stat = obs_slope, direction = "two-sided")
p_val
```

```{r perm-viz}
#| echo: false
null_dist %>%
  visualize() +
  shade_p_value(obs_stat = obs_slope, direction = "two-sided") +
  labs(
    title    = "Null Distribution (Permutation Test)",
    subtitle = glue::glue("p-value = {round(p_val$p_value, 4)} | Red area = p-value"),
    x        = expression(hat(beta)~"under H"[0]*": independence"),
    y        = "Count"
  )
```

The null distribution is centred at zero. Our observed slope lies far outside — the p-value is `r round(p_val$p_value, 4)`.

:::{.callout-note collapse="true"}
## 📚 Background: Bootstrap vs. Permutation — When to Use Which?
| | **Bootstrap** | **Permutation** |
|---|---|---|
| `hypothesize()` | No | Yes (`null = "independence"`) |
| `generate()` type | `"bootstrap"` | `"permute"` |
| Question | "How much does my estimate vary?" | "Could this result arise by chance?" |
| Output | Confidence interval | p-value |

Both are resampling methods, but they answer *different questions*. Bootstrap estimates variability; permutation tests a null hypothesis.
:::

---

## Step 5: Only One Test — Regression Subsumes Everything

In many introductory statistics courses, the t-test, ANOVA, and regression are taught as separate procedures. They are not. **They are all special cases of the linear model.**

### 5.1 The t-Test Is a Regression

Do goalkeepers and outfield players have different log market values? The classic way to test this is a two-sample t-test. But we can also frame it as a regression:

$$\log(\text{mv}) = \beta_0 + \beta_1 \cdot \mathbb{I}_{\text{Outfield}} + \varepsilon$$

Here, $\beta_1$ is the *difference in group means* — exactly what the t-test estimates.

```{r ttest-as-regression}
#| code-fold: false
# --- approach 1: t-test ---
t_result <- t.test(log_value ~ pos_bin, data = tbl_buli, var.equal = TRUE)

# --- approach 2: regression ---
lm_pos <- lm(log_value ~ pos_bin, data = tbl_buli)
```

```{r ttest-comparison}
#| echo: false
tibble::tribble(
  ~Approach,       ~Estimate,                    ~`t-statistic`,          ~`p-value`,
  "t-test",        round(diff(t_result$estimate), 4),
                                                  round(t_result$statistic, 4),
                                                                           round(t_result$p.value, 4),
  "Regression",    round(coef(lm_pos)["pos_binOutfield"], 4),
                                                  round(summary(lm_pos)$coefficients["pos_binOutfield","t value"], 4),
                                                                           round(summary(lm_pos)$coefficients["pos_binOutfield","Pr(>|t|)"], 4)
) %>%
  gt() %>%
  fmt_number(decimals = 4) %>%
  tab_style(cell_fill("#f0fff0"), cells_body()) %>%
  tab_options(table.font.size = "14px") %>%
  tab_footnote("The values are identical (up to sign convention for the t-statistic)")
```

The estimates, t-statistics, and p-values are identical. **The t-test is just a regression with a binary predictor.**

:::{.callout-note collapse="true"}
## 📚 Background: Why Are They the Same?
The two-sample t-test with equal variances computes:

$$t = \frac{\bar{x}_1 - \bar{x}_2}{s_p \sqrt{1/n_1 + 1/n_2}}$$

The regression t-test for $\beta_1$ in $Y = \beta_0 + \beta_1 X$ with a binary $X$ computes:

$$t = \frac{\hat\beta_1}{\text{SE}(\hat\beta_1)}$$

These are algebraically identical. The regression simply uses matrix algebra to express the same quantity.
:::

### 5.2 ANOVA Is a Regression with a Factor

What if we have more than two groups? Instead of the binary goalkeeper/outfield split, we use four position categories: Goalkeeper, Defender, Midfielder, Forward.

```{r pos-check}
#| code-fold: false
# --- verify position groups ---
tbl_buli %>% count(pos_cat)
```

```{r anova-as-regression}
#| code-fold: false
# --- regression with a factor (k levels) ---
lm_pos4 <- lm(log_value ~ pos_cat, data = tbl_buli)
```

```{r anova-table}
#| echo: false
get_regression_table(lm_pos4) %>%
  gt() %>%
  fmt_number(columns = c(estimate, std_error, statistic,
                         p_value, lower_ci, upper_ci), decimals = 3) %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = p_value < 0.05)
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("Yellow = significant at α = 0.05. Reference category: Goalkeeper")
```

Each coefficient is a t-test comparing that position to the reference (Goalkeeper). But there is also a **global question**: does position matter *at all*? That is the F-test:

```{r f-test}
#| code-fold: false
# --- the F-test: does ANY position coefficient differ from zero? ---
anova(lm_pos4)
```

```{r f-summary}
#| echo: false
get_regression_summaries(lm_pos4) %>%
  gt() %>%
  fmt_number(decimals = 3) %>%
  tab_style(cell_fill("#fff8e1"),
            cells_body(columns = statistic)) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("The F-statistic tests: do ALL position coefficients jointly equal zero?")
```

The t-test asks: "Is *this one* coefficient different from zero?" The F-test asks: "Is *any* coefficient different from zero?" Both live inside `get_regression_table()` and `get_regression_summaries()`.

:::{.callout-note collapse="true"}
## 📚 Background: The Testing Hierarchy
| Test | Question | Model |
|------|----------|-------|
| **t-test (2 groups)** | Is the mean difference ≠ 0? | `lm(y ~ x_binary)` |
| **ANOVA (k groups)** | Is any group mean different? | `lm(y ~ x_factor)` |
| **Simple regression** | Is the slope ≠ 0? | `lm(y ~ x_numeric)` |
| **Multiple regression** | Are any slopes ≠ 0? | `lm(y ~ x1 + x2 + ...)` |

All of these are the *same* linear model $Y = X\beta + \varepsilon$, just with different design matrices $X$. This is the "Only One Test" principle from [ModernDive](https://moderndive.com): learn one framework and you can handle any hypothesis.

For a comprehensive overview of which test maps to which regression, see the flowchart in [Appendix D of ModernDive](https://moderndive.com/D-regression).
:::

### 5.3 The Resampling Version

All of the above relies on theoretical distributions (t, F). With infer, we can replace *any* of them with a permutation test. The logic is always identical — shuffle the predictor, recompute, compare.

```{r perm-pos4}
#| code-fold: false
#| cache: true
set.seed(2024)

# --- observed F-statistic ---
obs_f <- tbl_buli %>%
           specify(log_value ~ pos_cat) %>%
           calculate(stat = "F")

# --- null distribution via permutation ---
null_f <- tbl_buli %>%
            specify(log_value ~ pos_cat) %>%
            hypothesize(null = "independence") %>%
            generate(reps = 1000, type = "permute") %>%
            calculate(stat = "F")

p_f <- null_f %>% get_p_value(obs_stat = obs_f, direction = "greater")
```

```{r perm-pos4-viz}
#| echo: false
null_f %>%
  visualize() +
  shade_p_value(obs_stat = obs_f, direction = "greater") +
  labs(title    = "Permutation Test for Position Effect (4 groups)",
       subtitle = glue::glue("Observed F = {round(pull(obs_f), 2)} | p = {round(p_f$p_value, 4)}"),
       x        = "F-statistic under H₀",
       y        = "Count")
```

Same conclusion, no distributional assumptions required.

---

## Step 6: Verification — Traditional Regression Output

We close the loop by showing `get_regression_table()` for the simple age model — the same numbers we computed via resampling in Steps 2–4:

```{r trad-table}
#| code-fold: false
reg_buli <- lm(log_value ~ age, data = tbl_buli)
get_regression_table(reg_buli)
```

```{r trad-table-fmt}
#| echo: false
get_regression_table(reg_buli) %>%
  gt() %>%
  fmt_number(columns = c(estimate, std_error, statistic,
                         p_value, lower_ci, upper_ci), decimals = 4) %>%
  tab_style(cell_fill("#fff8e1"),
            cells_body(rows = term == "age")) %>%
  tab_options(table.font.size = "14px") %>%
  tab_footnote("Yellow: age coefficient — compare CI with bootstrap result from Step 3")
```

### Three Methods, One Answer

```{r three-methods}
#| echo: false
trad_age <- get_regression_table(reg_buli) %>% filter(term == "age")

tibble::tribble(
  ~Method,                    ~`CI Lower`,                ~`CI Upper`,  ~`Significant?`,
  "Bootstrap (Percentile)",    round(ci_pct$lower_ci, 4),
                               round(ci_pct$upper_ci, 4),               "Yes",
  "Bootstrap (SE)",            round(ci_se$lower_ci, 4),
                               round(ci_se$upper_ci, 4),                "Yes",
  "Traditional (t-test)",      round(trad_age$lower_ci, 4),
                               round(trad_age$upper_ci, 4),             "Yes"
) %>%
  gt() %>%
  tab_style(cell_text(weight = "bold"), cells_column_labels()) %>%
  tab_style(cell_fill("#f0fff0"), cells_body()) %>%
  tab_options(table.font.size = "14px")
```

All three converge. This is **not** a coincidence — they all estimate the same quantity (the standard error of the slope), just by different routes.

---

## Mandatory Tasks

:::{.callout-important icon=false}
## ⚠️ A1: Sampling vs. Bootstrap Distribution

Look at the overlay plot from Step 2.3.

1. In what respect are the two distributions **similar**?
2. In what respect do they **differ**?
3. Why is the bootstrap distribution centred on the *sample* slope rather than the *population* slope?
:::

:::{.callout-important icon=false}
## ⚠️ A2: Confidence Interval Interpretation

The bootstrap CI from Step 3 is [`r round(ci_pct$lower_ci, 4)`, `r round(ci_pct$upper_ci, 4)`].

1. Does this interval contain zero? What does that imply?
2. Write the correct interpretation of "95% confidence" in one sentence.
3. A colleague says: "There's a 95% chance the true slope is in this interval." Why is that wrong?
:::

:::{.callout-important icon=false}
## ⚠️ A3: Permutation Test

The permutation p-value is `r round(p_val$p_value, 4)`.

1. In your own words: what does the null distribution in Step 4 represent?
2. Why is the null distribution centred at zero?
3. The p-value from `get_regression_table()` is `r round(trad_age$p_value, 4)`. Are they consistent?
:::

:::{.callout-important icon=false}
## ⚠️ A4: t-Test = Regression

Look at the comparison table in Step 5.1.

1. Which column confirms that the t-test and regression give the same result?
2. A colleague insists on using a t-test "because it's the right test for comparing two groups." What would you say?
3. What advantage does the regression framing have over the classical t-test framing?
:::

:::{.callout-important icon=false}
## ⚠️ A5: ANOVA & F-Test

Look at the regression table and F-test in Step 5.2.

1. What is the reference category? How do you know?
2. Pick one of the position coefficients from the regression table. What does its sign and magnitude mean?
3. Why do we need the F-test when we already have t-tests for each coefficient?
:::

---

## Bonus Tasks

:::{.callout-tip icon=false}
## 💡 B1: Sample Size and Bootstrap Precision

Repeat the Step 2 comparison with n = 20 and n = 100. How does sample size affect the bootstrap's ability to approximate the sampling distribution?

```r
set.seed(2024)
tbl_n20  <- tbl_buli %>% slice_sample(n = 20)
tbl_n100 <- tbl_buli %>% slice_sample(n = 100)
# bootstrap each and compare spread
```
:::

:::{.callout-tip icon=false}
## 💡 B2: Permutation Test for the Position Effect

Use `infer` to run a permutation test for the binary variable `pos_bin` (Goalkeeper vs. Outfield) with `stat = "diff in means"`. Compare the p-value with the regression p-value from Step 5.1.

```r
tbl_buli %>%
  specify(log_value ~ pos_bin) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("Outfield", "Goalkeeper"))
```
:::

:::{.callout-tip icon=false}
## 💡 B3: ModernDive Flowchart

Open the [hypothesis testing flowchart in ModernDive Appendix D](https://moderndive.com/D-regression). Pick two tests from the flowchart and show — using the Bundesliga data — that they are equivalent to a linear regression.
:::

---

## Summary

```{r summary-table}
#| echo: false
tibble::tribble(
  ~`Finding`,                       ~`Key Result`,                                ~`Implication for Lab 2`,
  "Sampling ≈ Bootstrap",           "Bootstrap approximates the sampling dist.",   "Justifies resampling for inference",
  "Bootstrap CI excludes zero",     "Age effect is statistically significant",     "→ interpret coefficients in Lab 2",
  "Permutation confirms CI",        paste0("p = ", round(p_val$p_value, 4)),       "Two routes to the same verdict",
  "t-test = regression",            "Identical estimates, t-stats, p-values",      "Regression is the universal framework",
  "ANOVA = regression with factor", "F-test = global significance test",           "→ multiple regression in Session 3"
) %>%
  gt() %>%
  tab_style(cell_fill("#f5f0fa"), cells_column_labels()) %>%
  tab_style(cell_fill("#f0fff0"), cells_body(rows = c(1, 2, 4))) %>%
  tab_style(cell_fill("#fff8e1"), cells_body(rows = c(3, 5))) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-note}
## What Comes Next

In **Lab 2** we move from resampling to the full machinery of **stochastic regression analysis**: the population model vs. the sample model (Stocker notation), all three `moderndive` helper functions, and the `infer::fit()` function for multiple regression. The resampling logic from this lab remains the foundation — Lab 2 simply adds the classical counterparts.

**Dataset used:** `buli_raw_2023-09-06.rds` (Bundesliga players, season 2022/23)
:::
