---
title: "Data Science and Strategy for Business"
subtitle: "ER114"
author: "Prof. Dr. Jörg Schoder"
institute: "FFHS"
date: "`r Sys.Date()`"
bibliography: ../../../lit/my_bib.bib
lang: en-US
reference-section-title: References
output:
  xaringan::moon_reader:
    self_contained: true
    css:
         - default
         - ../../../css/ffhs-theme_js.css
         - xaringan-themer.css
    includes:
      after_body: ../../../css/insert-logo.html
    lib_dir: ../../../libs
    nature:
      slideNumberFormat: "%current%/%total%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: false



---
class: title-slide

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_xaringan(text_color = "#d50006",inverse_text_color = "#FFFFFF",inverse_background_color = "#d50006", title_slide_background_color = "#d50006",header_background_color = "#d50006",header_color = "#FFFFFF",header_h1_font_size = "32px",
  header_h2_font_size = "26px",link_color="#502479",
  header_h3_font_size = "20px",text_slide_number_color = "#d50006",text_slide_number_font_size = "0.5em")
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_progress_bar(color = "#d50006", location = "bottom")
xaringanExtra::use_xaringan_extra(c("tile_view","scribble","panelset","tachyons"))
xaringanExtra::style_panelset_tabs(font_family = "inherit")
xaringanExtra::use_editable(expires = 1)
xaringanExtra::use_freezeframe(trigger = "hover")
```

```{r}
#| label: setup
#| include: false
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(kableExtra)
library(latex2exp)
library(fontawesome)
source(xfun::from_root("lit","helper.R"))
library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           style = "markdown",
           dashed = TRUE)
bib <- ReadBib(xfun::from_root("lit","my_bib.bib"))
Zitiere <- function(bib, key, ...) {
  gsub(" and ", " und ", Citet(bib, key, ...))
}
```

# ER114 - Data Science and Strategy for Business

## Session 3

### Complexity, Generalisation & Causality




<br>
<br>
<br>
<br>
<br>
<br>
<br>
### FS 2026
<br>
### Prof. Dr. Jörg Schoder
.mycontacts[
`r fa('github')` @FFHS-EconomicResearch
`r fa('linkedin')` @jfschoder
]


---
layout: true

<div class="my-footer"></div>

<div style="position: absolute;left:400px;bottom:10px;font-size:9px">`r fa('creative-commons')``r rmarkdown::metadata$author`</div>




---
class: left

.blockquote[Programme]

## Overview Session 3

<div style="display:grid; grid-template-columns:80px 1fr; gap:0.3em 1em;
            font-size:0.82em; margin-top:0.5em; align-items:center;">

<div style="color:#686868;">10'</div>
<div>`r fa('rotate-left')` <strong>Recap & Big Picture</strong> Session 1, 2 & 3 in context</div>

<div style="color:#686868;">25'</div>
<div>`r fa('layer-group')` <strong>From describing to modelling</strong> — inference vs. prediction · induction & deduction · the two dangers</div>

<div style="color:#686868;">25'</div>
<div>`r fa('layer-group')` <strong>Econometric foundations</strong> — A-B-C assumptions (von Auer) · BLUE & Bias-Variance · the two paths</div>

<div style="color:#686868;">25'</div>
<div>`r fa('layer-group')` <strong>Path A: Predictive Modelling</strong> — train-test split · stepwise selection · regularisation · cross-validation</div>

<div style="color:#D50006; font-weight:bold;">40'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab 1</strong> — Reality check: out-of-sample validation</div>

<div style="color:#686868; font-style:italic;">10'</div>
<div style="color:#686868; font-style:italic;">`r fa('mug-hot')` Break</div>

<div style="color:#686868;">25'</div>
<div>`r fa('layer-group')` <strong>Path B: Causal Modelling</strong> — endogeneity · causality · research designs · A/B testing</div>


<div style="color:#D50006; font-weight:bold;">25'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab 2</strong> — Multicollinearity & regularisation (new)</div>

<div style="color:#D50006; font-weight:bold;">25'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab 3</strong> — Design your experiment</div>


<div style="color:#686868;">10'</div>
<div>`r fa('flag-checkered')` <strong>Debrief & Outlook</strong></div>

</div>


???

NOTE for instructor: The original PVA3 slide listed only two labs in the
programme but three in the closing slide. This corrected version has three
labs aligned with the Programme_Session_3 document: (1) OOS validation,
(2) multicollinearity & regularisation, (3) experiment design.


---
class: inverse, center, middle

## Recap & Big Picture

.blockquote[Session 1 → Session 2 → **Session 3**]


---
class: left

.blockquote[Session 1 → Session 2 → **Session 3**]

## Recap & Big Picture

### Session 1: See & Describe
- Descriptive statistics
- Exploratory Data Analysis (EDA) vs. Data Mining: *recognising patterns*
- Descriptive regression: *line as a summary of the data*
- Residual = what the model does **not** explain
- Underfitting, omitted variable bias and overfitting

### Session 2: Quantifying Uncertainty
- Resampling & bootstrap
- *"There is only one test"* (ModernDive)
- Significance vs. relevance
- p-hacking & HARKing


---
class: left

.blockquote[Session 1 → Session 2 → **Session 3**]

## Session 3: Today

**Complexity** — How much complexity is good? What are the dangers?

**Generalisation** — When does a model still work tomorrow?

**Causality** — When are data-driven decisions (un)safe?

> *"There is a ton of data on the graph but little useful information."*
> — Taddy (2019)

📌 Goal: A reflective approach to data-driven decisions.


---
class: inverse, center, middle

## Part 1

.blockquote[From describing to modelling]

.blockquote[Inference vs. prediction]

.blockquote[Epistemological foundation: induction & deduction]

.blockquote[The two dangers]


---
class: left

.blockquote[From describing to modelling]

## Two Worlds of Regression: A Shift in Model Interpretation

.pull-left[
### Descriptive regression
$$y_i = b_0 + b_1 x_i + e_i$$

- Data as population (no randomness, no error term in a stochastic sense)
- **Summary** of observed data
- $\hat{y}_i$: prediction as conditional mean
- Coefficients: marginal effects (ceteris paribus)
- Residual $e_i$: variation not explained by the model


*→ This was our world in Session 1.*
]

.pull-right[
### Stochastic regression model

$$Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i$$

- **Model for a data-generating process (DGP)**
- $\beta_j$: (unknown!) parameters of the *population*
- $\varepsilon_i \sim (0, \sigma^2)$: **stochastic** error term
- Goal: From the *sample*, infer the population
- Requires **assumptions** for the inference to be valid

*→ This is the shift of today's session.*
]


---
class: left

.blockquote[From describing to modelling]

## Why Model? Three Perspectives

```{r purpose-table}
#| echo: false
tibble(
  ` ` = c("🔬 Statistics / Econometrics", "🤖 Data Science / ML", "⛏ Data Mining"),
  `Goal` = c("Inference (sample → population)",
             "Prediction (forecast new cases)",
             "Pattern recognition (exploration)"),
  `Focus` = c("β-coefficients, confidence intervals, tests",
              "ŷ — out-of-sample performance",
              "Hidden structures, clusters, associations"),
  `Typical question` = c("Does advertising work? (causal)",
                       "Who buys tomorrow? (predictive)",
                       "What customer groups exist? (descriptive)"),
  `Key criterion` = c("Unbiasedness, consistency",
                           "Cross-validation, RMSE, AUC",
                           "Interpretability, novelty")
) %>%
  kable(format = "html", escape = FALSE) %>%
  kable_styling(font_size = 15, full_width = TRUE) %>%
  row_spec(0, bold = TRUE, background = "#16213e", color = "white")
```

📌 **Inference and prediction place different demands on models.**


---
class: left

.blockquote[Inference vs. prediction]

## Inference vs. Prediction: What Counts?

.pull-left[
### Inference ("Understanding")

- Goal: $\boldsymbol{\hat{\beta}}$ close to $\boldsymbol{\beta}$
 (unbiased)
- Requires correct **model assumptions**
- Overfitting = assumption violation
- Multicollinearity is **poison** (unstable $\hat{\beta}$)
- Endogeneity is **poison** (biased $\hat{\beta}$)

*Taddy: "Statisticians have concentrated on how close $\hat{\beta}$ is to the true $\beta$."*
]

.pull-right[
### Prediction ("Forecasting")

- Goal: $\boldsymbol{\hat{y}}$ close to $\boldsymbol{y_\text{new}}$ (out-of-sample)
- Requires **good generalisation**
- Overfitting = poor generalisation
- Multicollinearity **often tolerable**
- The model is a **black box** — that is fine!
]



???


.center[
### ⚡ But: Prediction alone is not enough for business decisions!

*"ML will not tell you what will happen if you **change** something."* — Taddy, Lucas critique
]


---
class: left

.blockquote[Epistemological foundation: induction & deduction]

## Induction and Deduction in Data Science

To understand the challenges of data science, we split the process into two
logical steps. **Our example: a price-demand model (linear regression) for
hotel rooms.**

.pull-left[
### 1. Induction (the "learning")
* **What happens:** We use thousands of historical data points to derive a
  general rule.
* **Data-science term:** *Model training / data mining*.
* **The pattern found:** The algorithm learns the rule: *"When room prices
  are high, sales are extremely high (high occupancy)."*
]

.pull-right[
### 2. Deduction (the "application")
* **What happens:** We take the induced rule and apply it to new
  individual cases to make predictions.
* **Data-science term:** *Scoring / prediction*.
* **The application:** For a coming week with high prices, the model
  correctly predicts an occupancy of nearly 100%.
]

???

"Aha-moment": data mining (model training) is nothing other than machine
induction.


---
class: left

.blockquote[The two dangers]

## Failed Generalisation and the Causality Trap

When managers blindly trust algorithmic rules, they fall into one of two traps:

### Danger 1: Failed induction (lack of generalisation)
* The model has not learned a real rule but has memorised random
  fluctuations of our specific sample (*overfitting*).
* **Consequence:** Predictions on new data fail.
* **Solution:** Out-of-sample evaluation (train-test split / cross-validation).

### Danger 2: Blind deduction with interventions (causality trap)
* Our induced rule ("high price = high sales") is based on pure
  *correlation*. There is an invisible confounder: **school holidays**. In
  holidays, demand rises massively, and *that is why* prices are high.
* **Consequence:** Management doubles prices in November. The correlation
  collapses, revenue plummets.
* **Solution:** Causal inference (e.g. A/B tests for true price elasticity).


???

* Generalisation: We must ensure that our induction is not overfitting.
* Causality: Correlation-based rules must not be applied deductively when
  intervening. Matt Taddy stresses: A model based purely on correlation
  collapses when we manipulate the system.


---
class: left

.blockquote[The two dangers]

## From AI User to AI Value Creator

.pull-left[
### The AI User (passive)
* Consumes analytical results and trusts the algorithmic "black box".
* **The risk:** Makes deductive decisions based on models that, due to
  violated foundational assumptions, collapse immediately in reality.
]

.pull-right[
### The AI Value Creator (active)
* Acts as a **"critical validator"** and steers models with strong data
  governance and exclusive data.
* **The advantage:** Critically checks the statistical logic to ensure
  that genuine causality or generalisation is present before budgets are
  released.
]



---
class: inverse, center, middle

## Part 2: Econometric Foundations

.blockquote[A-B-C assumptions (von Auer)]

.blockquote[BLUE & Bias-Variance — common bridge for both paths]

.blockquote[The two paths]


---
class: left

.blockquote[A-B-C assumptions (von Auer)]

## Model Assumptions in Brief

.panelset[
.panel[.panel-name[A-Assumptions]
### A-Assumptions: the model

- **A1 (variable selection)** — No relevant variables missing, no irrelevant ones included
- **A2 (linearity)** — The true relationship is linear in the parameters
- **A3 (parameter constancy)** — Parameters are constant across all $N$ observations
]
.panel[.panel-name[B-Assumptions]
### B-Assumptions: the error term

- **B1 (expectation)** — $E(\varepsilon_i) = 0$
- **B2 (homoscedasticity)** — $\text{Var}(\varepsilon_i) = \sigma^2$
- **B3 (no autocorrelation)** — $\text{Cov}(\varepsilon_i, \varepsilon_j) = 0$ for $i \neq j$
- **B4 (normality)** — $\varepsilon_i \sim N$
]

.panel[.panel-name[C-Assumptions]
### C-Assumptions: the data

- **C1 (exogeneity)** — $\text{Cov}(X_i, \varepsilon_i) = 0$
- **C2 (no perfect multicollinearity)** — No regressor is an exact linear combination of others

### ⚡ Why C1 and C2 are decisive for management:

| Violation | Consequence | Strategy |
|:----------|:----------|:----------|
| **C2** (multicollinearity) | Overfitting (Danger 1) | **Path A:** regularisation |
| **C1** (endogeneity) | Biased causal conclusions (Danger 2) | **Path B:** experiment |

*→ A1 violation (missing variable) leads to C1 violation — the "domino effect"!*
]
.quellePanURL[cf. `r Citet(bib, "vonAuer_2023")`, p. 167ff
]
]


---
class: left

.blockquote[BLUE & Bias-Variance]


## Bias vs. Variance and the BLUE Property of OLS

.panelset[
.panel[.panel-name[Refresher]
### Bias vs. Variance
```{r}
#| echo: false
#| out-width: "80%"
#| fig-align: "center"
knitr::include_graphics(xfun::from_root("img","PVA3","bias_variance_zielscheiben.svg"))
```
]
.panel[.panel-name[BLUE]
### Gauss-Markov Theorem

If A, B, and C assumptions hold, OLS is **BLUE**:

| | Meaning |
|:--|:----------|
| **B**est | Smallest variance among... |
| **L**inear | ...linear... |
| **U**nbiased | ...unbiased... |
| **E**stimators | ...estimators |

📌 "Best" = **only variance**, not total error!

]
.panel[.panel-name[Trade-off]
### Bias-Variance Trade-off

$$\text{MSE} = \text{Bias}^2 + \text{Variance}$$

- BLUE requires: bias = 0
- → lower total error (MSE) is possible!

⚠️ BLUE is theoretically beautiful, but **not always optimal in practice**.

In the regularisation context, we will actively use the trade-off:
*some* bias, **much** less variance.

*→ Taddy's CAPM example shows: a simple model ($\alpha$, $\beta$) extracts
more **usable information** from 7 years of S&P-500 data than a "fancy
plot" with all individual values.*

]
]


---
class: left, alt-slide

.blockquote[ALTERNATIVE — BLUE as a common bridge for both paths]

## BLUE: The Common Bridge for Both Paths

*This alternative slide reframes BLUE explicitly as a bridge concept that
applies to both paths, before path A and B are introduced.*

.pull-left[
### BLUE for Path B (Inference)

When we want to **explain** something, the BLUE property is the relevant
optimality criterion:

- Unbiased estimates → coefficients tell us about the true effect
- Minimum variance among unbiased → tightest confidence intervals
- Holds *if* model is correctly specified

→ **BLUE is the gold standard when we trust the assumptions.**
]

.pull-right[
### BLUE for Path A (Prediction)

When we want to **predict**, BLUE is *not* the right criterion:

- Unbiased is *not* what we need — we need low total error (MSE)
- Accepting some bias for much less variance can lower MSE
- Lasso and Ridge are *not* BLUE — by design

→ **BLUE explains why predictive methods deliberately leave the
unbiasedness regime.**
]

📌 The same theoretical foundation (Gauss-Markov, MSE decomposition)
explains *both* paths — but leads to *different* optimality criteria.



---
class: left

.blockquote[The two paths]

## The Management Checklist: Assumptions as Guardrails

For the *AI Value Creator*, the assumptions are the ultimate checklist:

* **A-Assumptions (specification):** *Did we build the right business model?*
  - **A1** (variable selection): Missing relevant variable → **omitted
    variable bias** → C1 violation (endogeneity) → Danger 2
  - **A3** (parameter constancy): **Lucas critique** — as soon as
    management actively intervenes, a structural break is provoked.

* **B-Assumptions (error term):** *How does the noise behave?*
  - B2/B3 are nearly always violated in big data → **robust standard
    errors** & **bootstrapping** (Session 2) solve this.

* **C-Assumptions (data) — the great fork:**
  - **C2** (multicollinearity) violated → unstable coefficients,
    overfitting → **Path A: regularisation**
  - **C1** (exogeneity) violated → biased causal conclusions →
    **Path B: experiment**

`r fa('circle-right')` *In practice, data-science projects fail at A1
(missing variables) and its consequence C1 (endogeneity). This is where
your domain knowledge counts.*


---
class: left

.blockquote[The two paths]

## Conclusion for the Translator

### When we violate C2 (multicollinearity)

→ We regularise (Path A: predictive).

### When we violate C1 (exogeneity)

→ We experiment (Path B: causal).

???

OLS assumptions (BLUE) are not an academic end in themselves but
guardrails on the way to safe induction (C2: multicollinearity disturbs
pattern recognition) and safe deduction under intervention (A1 → C1:
endogeneity obscures causality).

Depending on which OLS assumption we violate and which business goal we
pursue, our strategy splits:

### Path A: Predictive Modelling (securing induction)
We want **generalisation** for predictions (forecasting customer
behaviour). The main problem: **multicollinearity** and too many
variables lead to overfitting.

### Path B: Causal Modelling (securing the intervention)
We want true **causality** for management decisions (pricing strategy).
The main problem: **lack of exogeneity** (endogeneity / omitted variable
bias due to factors like "seasonality").



---
class: inverse, center, middle

## Part 3: Path A — Predictive Modelling

.blockquote[Multicollinearity]

.blockquote[Train-test split and stepwise variable selection]

.blockquote[Regularisation & cross-validation]


---
class: left

.blockquote[Multicollinearity]

## Multicollinearity: The Twin Problem

.pull-left[
### What happens?

Predictors are **strongly correlated** → OLS cannot **separate** their effects.

### Symptoms

- High $R^2$ but coefficients *not significant* (large standard errors)
- $\hat{\beta}$ unstable (sign changes for small data perturbations in the extreme)

### Diagnosis: Variance Inflation Factor (VIF)

$$VIF_j = \frac{1}{1 - R^2_j}$$

VIF > 5 problematic, VIF > 10 severely problematic.
]

.pull-right[
### Differentiated view and management playbook

| | Inference | Prediction |
|:--|:---------|:-----------|
| Problem? | ⚠️ **Yes!** ($\hat{\beta}$ unusable) | 🟢 Often tolerable |
| Why? | Coefficients unstable | $\hat{y}$ remains stable |
| Solution | Drop variable, build index | Regularisation |

*"Multicollinearity is problematic for interpretation, less so for prediction!"*
]


???

### In R

```r
library(car)
lm(sales ~ advertising + marketing, data = data) %>% vif()
```
### The management rule

* As an *AI Value Creator* you know:

  * For pure prediction, multicollinearity can be **tolerated**:
      * coefficients are unstable, but predictions remain stable;
      * out-of-sample accuracy is the only criterion that matters
        (**generalisation**).
  * If management wants to understand the drivers (**interpretation**
    of individual effects in **confirmatory analyses**), multicollinearity
    is dangerous and must be addressed.



---
class: left

.blockquote[Multicollinearity]

## Underfitting vs. Overfitting: Is There a Sweet Spot?


.panelset[
.panel[.panel-name[Example]
```{r}
#| label: overfitting-demo
#| echo: false
#| message: false
#| fig-height: 5
#| fig-width: 11
set.seed(42)
n <- 30
x <- sort(runif(n, 0, 10))
y <- sin(x) + rnorm(n, 0, 0.3)

df_fit <- data.frame(x = x, y = y)

p1 <- ggplot(df_fit, aes(x, y)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#3498DB", linewidth = 1.2) +
  labs(title = "Underfitting (linear)", subtitle = "High bias, low variance") +
  theme_minimal(base_size = 13)

p2 <- ggplot(df_fit, aes(x, y)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 3), se = FALSE, color = "#27AE60", linewidth = 1.2) +
  labs(title = "Good fit (polynomial degree 3)", subtitle = "Balance of bias and variance") +
  theme_minimal(base_size = 13)

p3 <- ggplot(df_fit, aes(x, y)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 20), se = FALSE, color = "#E74C3C", linewidth = 1.2) +
  labs(title = "Overfitting (polynomial degree 20)", subtitle = "Low bias, high variance") +
  theme_minimal(base_size = 13)

library(patchwork)
p1 + p2 + p3
```

`r fa('circle-right')` Illusion of in-sample fitting
]
.panel[.panel-name[Sweet spot?]
```{r}
#| label: bias-var-simple
#| echo: false
#| out-width: "50%"
#| fig-align: "center"
set.seed(42)
complexity <- seq(1, 10, by = 0.1)
bias2 <- 3 * exp(-0.5 * complexity)
variance <- 0.1 * exp(0.4 * complexity)
total <- bias2 + variance + 0.3

df <- data.frame(
  x = rep(complexity, 3),
  y = c(bias2, variance, total),
  type = rep(c("Bias²", "Variance", "Total error"), each = length(complexity))
)

ggplot(df, aes(x, y, color = type, linetype = type)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = complexity[which.min(total)], linetype = "dashed", alpha = 0.5) +
  annotate("text", x = complexity[which.min(total)] + 0.3, y = 3.5,
           label = "Optimum", angle = 90, size = 3.5) +
  scale_color_manual(values = c("Bias²" = "#E74C3C", "Variance" = "#3498DB", "Total error" = "#2C3E50")) +
  labs(x = "Model complexity", y = "Error", color = "", linetype = "") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```
]
]

???
.pull-left[
### Bias → Underfitting
- Systematic error
- Model **too simple**
- Important relationships missing
- *Example: linear regression for a quadratic relationship*

### Variance → Overfitting
- Random fluctuation
- Model **too complex**
- Learns noise
- *Example: polynomial degree 20 for 10 data points*
]


---
class: left

.blockquote[Multicollinearity]

## Overfitting: Too Much of a Good Thing

.pull-left[
### The basic problem

The model learns **noise** instead of **patterns**.

$$E[(y - \hat{y})^2] = \underbrace{\text{Bias}^2}_{\text{too simple}} + \underbrace{\text{Var}(\hat{y})}_{\text{too complex}} + \sigma_{\varepsilon}^2$$

*Taddy: "The entire field of ML is calibrated against out-of-sample experiments."*

Rule of thumb: at least **10–20 observations per predictor**.
]

.pull-right[
### Counter-strategies
.small[
1. Collect more data
2. Simplify the model
  a. Feature selection
    i. Substantive justification (e.g. through expert knowledge)
    ii. Statistical procedures:
      * Stepwise selection (AIC/BIC)
      * Regularisation (Lasso)
  b. Feature shrinkage: regularisation (Ridge)
  c. Feature extraction: PCA (dimensionality reduction → ER017)
3. Check generalisation
  a. Train/test split
  b. Cross-validation
]

> Key question: *Does the model generalise to **new, unseen** data?*

]

???
📌 → **Lab 1: Train-test split & cross-validation**

When the model learns noise, the balance in the formula shifts:
* $\text{Bias}^2$ falls strongly: the model fits training data perfectly,
  appearing "precise".
* $\text{Var}(\hat{y})$ explodes: a small change in the dataset would
  yield a completely different model.
* Total error $E[(y - \hat{y})^2]$ rises: predictive power for the real
  world is gone.

Analogy: a student does not learn the logical relationships, but
memorises where the lecturer's coffee stains are on each page. In the
exam, they receive a fresh sheet without stains and fail despite "having
studied a lot".


---
class: left

.blockquote[Train-test split and stepwise variable selection]

## The Classical Attempt: Stepwise Selection

In traditional statistics, complexity was controlled by manual or
stepwise addition/removal of variables.

### Backward vs. forward selection
* **Backward selection:** Start with the full model and discard
  insignificant variables. *Problem:* In big-data settings (when the
  number of variables $p$ approaches the number of observations $n$),
  the full model is unstable or not even mathematically computable.

* **Forward stepwise selection:** A "greedy algorithm" starting from
  zero, adding the variable that maximises in-sample $R^2$ at each step.

**The problem of greedy algorithms:** They are myopic. Each step is
considered in isolation (local optimum), without considering the global
optimum across the whole model path.

???

how the problem was solved "in the past": variables were thrown in or
out based on p-values (backward). Make it clear: in big-data scenarios
where there are more variables than data points (p>n), these methods
fail outright mathematically.


---
class: left

.blockquote[Train-test split and stepwise variable selection]

## The Classical Attempt: Train-Test Split

Checking generalisability via a single split into training and test sets.

```{r}
#| label: train-test-split
#| echo: false
#| out-width: "80%"
#| fig-align: "center"
knitr::include_graphics(xfun::from_root("img","PVA3","train_test_split.svg"))
```


---
class: left

.blockquote[Regularisation & cross-validation]

## Cross-Validation: **Systematic, Repeated Splitting and Testing**

```{r}
#| label: cv-schematic
#| echo: false
#| out-width: "70%"
#| fig-align: "center"
knitr::include_graphics(xfun::from_root("img","PVA3","kfold_crossvalidation.svg"))
```
$$\text{CV-Score} = \frac{1}{K} \sum_{k=1}^{K} MSE_k$$



???

We want to simulate the future by strategically dividing the data into
K parts and "hiding one part each time to test predictive power". That
is the essence of cross-validation.

### K-Fold Cross-Validation (the gold standard)
Instead of splitting the data only once into training and test, we use
the data more efficiently:
1. Data is divided into $K$ random equal-sized folds.
2. The model is trained on $K-1$ folds.
3. We measure predictive power (out-of-sample deviance) on the hidden fold.
4. This is repeated $K$ times.

**The advantage:** Every data point is used exactly once as an unseen
test scenario. We get the true out-of-sample error.




---
class: left

.blockquote[Regularisation & cross-validation]

## Regularisation: Controlled Complexity

.pull-left[
### Lasso (L1) vs. Ridge (L2)

**Ridge:** $\min \sum(y_i - \hat{y}_i)^2 + \lambda \sum \beta_j^2$
→ Shrinks all, removes none

**Lasso:** $\min \sum(y_i - \hat{y}_i)^2 + \lambda \sum |\beta_j|$
→ Shrinks **and** sets to 0 → **variable selection**

$\lambda$ is chosen via **cross-validation**.

```r
library(glmnet)
cv_fit <- cv.glmnet(X, y, alpha = 1)
coef(cv_fit, s = "lambda.min")
```
]

.pull-right[
### Lasso coefficient paths

```{r}
#| label: lasso-paths
#| echo: false
#| message: false
#| fig-height: 5
#| fig-width: 6
set.seed(42)
n <- 200; p <- 5
X <- matrix(rnorm(n*p), n, p)
colnames(X) <- c("Price", "Advertising", "Quality", "Age", "Random")
y <- 2*X[,1] - 1.5*X[,2] + 1*X[,3] + 0.3*X[,4] + 0*X[,5] + rnorm(n)

library(glmnet)
fit <- glmnet(X, y, alpha = 1)

beta_df <- as.data.frame(as.matrix(t(coef(fit)[-1,])))
beta_df$lambda <- fit$lambda
beta_long <- pivot_longer(beta_df, -lambda, names_to = "Variable", values_to = "Coefficient")

ggplot(beta_long, aes(log10(lambda), Coefficient, color = Variable)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3) +
  scale_color_brewer(palette = "Set1") +
  labs(x = expression(log[10](lambda)), y = "Coefficient",
       title = "Lasso: coefficient paths") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")
```
]


---
class: left

.blockquote[Regularisation & cross-validation]

## Lasso vs. Ridge: When What?

.pull-left[
### Prefer Ridge when:
- Many **small effects**
- **All** variables matter
- **Multicollinearity** present
- Stable coefficients desired

### Prefer Lasso when:
- **Sparsity** expected (few important variables)
- **Variable selection** desired
- **Interpretability** important
- Many features
]

.pull-right[
### Elastic Net — the best of both worlds

$$\lambda_1 \sum |\beta_j| + \lambda_2 \sum \beta_j^2$$

### In R (`glmnet`)

```r
library(glmnet)

# alpha = 1: Lasso
# alpha = 0: Ridge
# alpha = 0.5: Elastic Net

cv_fit <- cv.glmnet(X, y, alpha = 1)
best_lambda <- cv_fit$lambda.min
```

💡 **Practice:** Often Elastic Net with CV for both α and λ.
]


---
class: left

.blockquote[Regularisation & cross-validation]

## The Synthesis: Cross-Validation + Lasso

Lasso provides, by varying the penalty $\lambda$, a whole **regularisation
path** of candidate models. Cross-validation identifies the sweet spot on
this path that guarantees the best generalisation.

.panelset[
.panel[.panel-name[Analogy]
### The $\lambda$ parameter as a squelch
* $\lambda$ works like the squelch on a radio.
* Turn it too high, you hear nothing (underfitting). Turn it too low,
  you hear only noise (overfitting / Danger 1).

]
.panel[.panel-name[Decision rule]
### Final model selection (after Taddy)

* **CV-min rule:** The model with the lowest out-of-sample error.
  *Focus: maximum predictive accuracy.*

* **1se rule:** The *simplest* model within one standard error of the
  minimum. *Focus: conservative, more robust models.*

**`r fa('circle-right')` Danger 1 averted: we have a generalising model!**
]

.panel[.panel-name[CV variants]

| Variant | When? |
|:---------|:------|
| K-Fold | Standard ($K = 5$ or $10$) |
| Stratified | Unbalanced classes |
| Time-series CV | Time series (no future leakage!) |
| Nested CV | Model selection + evaluation |

*Stage direction: When a data team presents a predictive model without a
clean train-test split, send it back!*
]
]


---
class: left

.blockquote[Lab 1 — Reality check via out-of-sample validation]

## `r fa('laptop-code', fill='#D50006')` Lab 1
<div style="border:3px solid #D50006; border-radius:8px; padding:20px 28px;
            background:#fff5f5; margin-top:1em;">
<div style="font-family:sans-serif; font-size:16px; font-weight:bold;
            color:#D50006; margin-bottom:10px;">
`r fa('flask')` Lab 1: Reality check — train-test split & cross-validation
</div>
<div style="font-size:0.85em; line-height:1.8;">
<strong>Learning objective:</strong> Experience overfitting on your own
model and detect it via out-of-sample validation. Apply train-test split
and K-fold CV.
<br>
<strong>Content:</strong> Fit a model → great $R^2$! → On test data:
💥 experience overfitting · Cross-validation as the systematic remedy.
<br>
`r fa('clock')` <strong>40 minutes</strong> &nbsp;·&nbsp;
`r fa('users')` <strong>Individual or pairs</strong>
<br>
`r fa('circle-right')` <strong>Open lab document:</strong>
<code>Session3_Lab1_OutOfSample_OJ.qmd</code>
</div>
</div>
<br>
.center[.small[Continue after the lab → <strong>Lab 2: Multicollinearity & Regularisation</strong>]]

???

NOTE for instructor: The original lab included an LM → GLM transition
that has been removed in this revised version, in line with the Logit/GLM
deferral to ER017. The lab now focuses purely on the OOS / CV core idea,
with continuous Y throughout.


---
class: left, alt-slide

.blockquote[ALTERNATIVE — Lab 2 (new): Multicollinearity & Regularisation]

## `r fa('laptop-code', fill='#D50006')` Lab 2 (new)
<div style="border:3px solid #D50006; border-radius:8px; padding:20px 28px;
            background:#fff5f5; margin-top:1em;">
<div style="font-family:sans-serif; font-size:16px; font-weight:bold;
            color:#D50006; margin-bottom:10px;">
`r fa('flask')` Lab 2: Multicollinearity & Regularisation
</div>
<div style="font-size:0.85em; line-height:1.8;">
<strong>Learning objective:</strong> Detect multicollinearity via VIF,
apply Lasso / Ridge / Elastic Net, choose $\lambda$ via cross-validation.
Understand the difference between coefficient stability (inference) and
prediction stability.
<br>
<strong>Content:</strong> Correlated features → diagnose with VIF · Lasso
zeroes out features → variable selection · Ridge stabilises · Elastic
Net combines both · CV path → choose $\lambda$ · Interpret coefficient
paths.
<br>
`r fa('clock')` <strong>25 minutes</strong> &nbsp;·&nbsp;
`r fa('users')` <strong>Individual or pairs</strong>
<br>
`r fa('circle-right')` <strong>Open lab document:</strong>
<code>Session3_Lab2_Regularisation_OJ.qmd</code>
</div>
</div>
<br>
.center[.small[Continue after the lab → <strong>Part 4: Path B — Causal Modelling</strong>]]

???

NOTE for instructor: This lab was promised in the closing slide of the
original PVA3 but never had its own slide. Reconstructed here based on
the Programme_Session_3 timing and the Path A content of the lecture.


---
class: inverse, center, middle

## Part 4: Path B — Causal Modelling

.blockquote[Endogeneity]

.blockquote[Correlation vs. causality]

.blockquote[Research designs]

.blockquote[A/B testing]


---
class: left

.blockquote[Endogeneity]

## Endogeneity: The Hidden Poison

.pull-left[
### Definition

$\text{Cov}(X_i, u_i) \neq 0$ — assumption **C1 violated!**

→ OLS is **biased and inconsistent**

### Three causes (A1 → C1 domino effect)

1. **Omitted variable bias** (confounding)
   *Advertising ↔ revenue, season forgotten*

2. **Simultaneity / reverse causality**
   *Police ↔ crime*

3. **Measurement error** in $X$
]

.pull-right[
### The Lucas critique (Danger 2)

A purely predictive model learns the biased historical correlation
("high prices → high occupancy"). When management actively intervenes,
we change the system — the historical correlations collapse (**A3
violation**: parameter constancy breaks).

> Under endogeneity (C1 violation), neither regularisation nor a larger
> sample helps.

*→ This is where ML reaches its limits. Here we need **causal analysis**.*
]


---
class: left, alt-slide

.blockquote[ALTERNATIVE — where do the data come from?]

## Where Do the Data Come From?

*This alternative slide adds the data-provenance perspective that is
implicit in PVA3 but should be explicit (Hernán-anchored).*

The validity of inference depends *not* primarily on the method but on
the **data foundation**:

| Type of data collection | Exogeneity? | Method |
|---|---|---|
| **Survey** (cross-section, longitudinal) | Plausibility argued, never guaranteed; risks: self-selection, non-response | OLS under MLR.4 — Wooldridge world |
| **Experiment** (RCT, A/B test) | Exogeneity *created* by randomisation | Comparison of means, regression adjustment |
| **Natural experiment** (IV, DiD, RDD) | Exogeneity argued via the "as-if-random" allocation | IV / DiD / RDD as identification strategies |

📌 **Hernán's central message:** *"Causal inference is just inference
with the right data."* Method and data are inseparable in the causal
question.

→ Path B unfolds along this axis: Wooldridge-OLS first (under
exogeneity-by-assumption), then design-based identification (when the
assumption fails).


---
class: left

.blockquote[Correlation vs. causality]

## Correlation ≠ Causality

.pull-left[
### Directed Acyclic Graphs (DAGs) and spurious correlations

```
       Z (confounder)
      ↙        ↘
      X  ←---→  Y
    (spurious correlation)
```

**Business:** Marketing ↔ revenue (confounder: economy)

*Taddy: "ML will not tell you what happens if you **change** something."*
]

.pull-right[
### Gold standard: experiment

Randomisation balances **all** confounders — even unknown ones!

### Causal criteria

1. ✅ Correlation present
2. ✅ Temporal order
3. ✅ **No** confounding
4. ✅ Plausible mechanism
5. ✅ Dose-response relationship

*→ Without an experiment: only association!*
]


---
class: left

.blockquote[Correlation vs. causality]

## Practical Relevance: Innovator's Hypothesis (Schrage 2014)

.pull-left[
### The sobering reality (Schrage 2014)
.small[
Microsoft's **Ronny Kohavi** shows real A/B tests — the audience guesses
the winner:

> *No consensus. Ever.*

- Kaushik: *"80% of the time we are wrong."*
- Kohavi: only **⅓** of ideas improve metrics
- Hadiaris: hit rate after 5 years ≈ **33 %**

> *"We are poor at assessing the value of ideas."*
]

### The lesson

.small[
Conviction ≠ evidence. Experts systematically overestimate themselves
(Kahneman, Tetlock).
]
]

.pull-right[
### Schrage's central thesis

.small[

> *"They run experiments not because they can confidently predict outcomes
> — but precisely because they can't."*
]

### Big Data + little experiments

.small[
- Big data: **patterns** and **correlations**
- Experiments: **causal evidence**
- Together: *"virtuous cycles of profitable insight"*
]

### Connection to Session 2

.small[
- p-hacking problem → fix the hypothesis **in advance**!
- *"Data trumps intuition"* — but with humility
]
📌 → **Lab 3: Design your experiment**
]


???

* On Session 2: reminder of the "multiple measures" problem: when we run
  an experiment and then choose which metric counts as success, we
  inflate the false-positive rate.
* The design (Lab 3) must come *before* data collection.

**The mindset:** "Data trumps intuition — but with humility." This is
the ultimate summary of the AI Value Creator role. They use big data to
generate ideas but are humble enough to test those ideas in small, hard
A/B tests on their causal effectiveness before risking the company's
budget.


---
class: left

.blockquote[Correlation vs. causality]

## The Origin of the Problem: The A1 Violation

Why are experts wrong 66 % to 80 % of the time when they actively
intervene in markets (e.g. raise prices, launch features)?

.pull-left[
### Assumption A1: variable selection
* A1 demands that no relevant exogenous variables are missing from the
  model.
* **The reality:** In practice, important drivers are often missing
  because they are hard to measure (e.g. an employee's "selling talent"
  or general "holiday season").
]

.pull-right[
### The domino effect to assumption C1
* When a relevant variable is missing, its influence is shifted into
  the error term.
* If this omitted variable correlates with our action variable (e.g.
  price), a **contemporary correlation** between the action variable
  and the error term arises.
* **The result:** Assumption C1 (exogeneity) is violated. The notorious
  **omitted variable bias** appears.
]


???

In management we often want not only to predict but to actively
intervene (e.g. change prices). For this, we need models that capture
true causality. The most common reason this fails starts with model
specification.

Students often learn in pure statistics only that violated exogeneity
(C1) is bad. Only by linking it to assumption A1 does it become clear
that this is a deeply substantive, economic problem: the data team has
not thought through the business model and overlooked a "confounder"
(such as the holiday season).



---
class: left

.blockquote[Correlation vs. causality]

## The Lucas Critique and the Failure of OLS

When assumption C1 is violated (endogeneity), classical OLS estimation
and its coefficients are biased and worthless for decisions.

### The Lucas critique (the danger of intervention)
* A purely predictive model (Path A) learns the biased historical
  correlation (e.g. "in holidays prices are high and sales are high
  `r fa('circle-right')` higher prices bring more revenue").
* When management actively intervenes and raises the price on a normal
  Tuesday, we change the system.
* The historical correlations collapse because the hidden driver
  (holiday season) is missing. The action fails catastrophically.

`r fa('circle-right')` **How do we heal the C1 violation? We need
research designs that enforce or isolate exogenous (independent)
variation in our data!**



---
class: left

.blockquote[Research designs]

## Refresher: How Do We Evaluate a Research Design?

### Internal vs. external validity

To assess how well a research design heals the C1 violation
(endogeneity) and whether it is suitable for business decisions, we use
two central quality criteria:

.pull-left[
### 1. Internal validity
**The question:** *Have we cleanly isolated causality in our specific study?*
* **Meaning:** Are there really no other explanations for the measured
  effect? Have all confounders been controlled or eliminated?
* **The enemy:** Omitted variable bias (A1 violation) and endogeneity.
  If our action variable still correlates with the noise, internal
  validity is destroyed.
]

.pull-right[
### 2. External validity
**The question:** *Can the result be transferred to the real market?*
* **Meaning:** Does our causal effect also apply to the entire customer
  base in the real business world?
* **The enemy:** Artificial test environments or unrepresentative
  samples (e.g. we test a pricing feature only on student employees but
  want to roll it out to premium customers).
]

**`r fa('circle-right')` The holy grail for the AI Value Creator is a
research design that maximises *both* validities!**



---
class: left

.blockquote[Research designs]

## The Landscape of Research Designs

To prove causality and heal C1, we distinguish three design types in
empirical research:

**1. The laboratory experiment**
* **Concept:** Perfectly controlled environment, systematically isolated
  from the outside world. Confounders are eliminated.
* **Business assessment:** Very high internal validity, but often low
  external validity (lab conditions do not match real markets).

**2. The field experiment (the gold standard in business)**
* **Concept:** Active, random intervention in the real world (e.g. on
  the live customer base).
* **Business assessment:** High internal and external validity.

**3. The natural experiment (observational data)**
* **Concept:** We analyse historical data not generated for our purpose.
  "Nature" or chance has created a quasi-experimental situation.

???

The further we move from the perfectly controlled lab, the more
vulnerable we are to C1 violations.


---
class: left

.blockquote[Research designs]

## The Field Experiment: A/B Testing as RCT

The **Randomised Controlled Trial (RCT)** is the most important tool in
the technology and business world, where it goes by the term **A/B
testing**.

.pull-left[
### How A/B testing heals assumption C1
* Users are assigned by strict randomness (coin flip) to group A
  (status quo) or B (intervention, e.g. new feature).
* This randomness guarantees that the treatment is **completely
  independent** of all hidden confounders.
* **C1 is artificially restored!**
]

.pull-right[
### Evaluation method
* In a perfect A/B test, the simple comparison of means (Average
  Treatment Effect, ATE) often suffices.
* Regression can additionally be used (regression adjustment) to
  control for known confounders and reduce variance.
]

---
class: left



## Natural Experiments & Evaluation Methods

A perfect A/B test is often impossible (ethical, legal, or technical
reasons). The "critical validator" must then check which method the
data team uses on historical observational data (or imperfect
experiments):

### 1. Instrumental Variables (IV) / Two-Stage Least Squares (2SLS)
* **Scenario ("Intent-to-treat"):** We randomly assign users a new
  feature (e.g. a smart thermostat), but not everyone uses it.
  Treatment and confounders (e.g. tech affinity) are entangled.
* **Method:** Use the *random assignment* as an instrumental variable.
  In the first stage (2SLS), it filters out uncorrelated variation; in
  the second stage, the true effect is estimated consistently.

### 2. Difference-in-Differences (DiD)
* **Scenario:** A measure was introduced only in one specific market
  (e.g. country A) but not in another (country B).
* **Method:** Compare the *trend* in the treated group before and after
  the measure with the trend in the control group.

### 3. Regression Discontinuity Design (RDD)
* **Scenario:** A measure takes effect at a hard threshold (e.g. bonus
  payment only above CHF 100 000 in revenue).
* **Method:** Compare only people just below and just above the
  threshold; near the sharp boundary these groups are nearly identical.

*(Note for panel data: to filter out time-invariant unobserved
characteristics of individuals/firms, fixed/random effects are added.)*


???

This is the manager's "weapon" in data meetings. They do not need to
compute the formulas for IV (2SLS) themselves. But when the data team
arrives with an "intent-to-treat" problem, the manager must be able to
mention the keyword "instrumental variables". When a date-bound measure
was introduced in one market, they should ask for a
difference-in-differences design.


---
class: left

.blockquote[A/B testing]

## A/B Testing: The Essentials

.pull-left[
### Design (Kohavi / Schrage)

✅ Formulate hypothesis **in advance**
✅ Set the **primary metric**
✅ Define minimal effect → **sample size**
✅ **Do not** peek early!
✅ Report **effect size**

### Cohen's d

$$d = \frac{\bar{X}_B - \bar{X}_A}{s_{\text{pooled}}}$$

| d | Interpretation |
|:--:|:--:|
| 0.2 | small |
| 0.5 | medium |
| 0.8 | large |
]

.pull-right[
### Outlook: DiD & IV

**Difference-in-Differences:**

```
        Before  After
Treated:  A  →  B     ΔT = B-A
Control:  C  →  D     ΔK = D-C
DiD = ΔT - ΔK
```

**Instrumental variables:**
Find $Z$ that influences $X$ but $Y$ only through $X$.

### Philosophy Eats AI (Schrage 2025)

> *Which questions we ask and which metrics we choose are philosophical
> decisions.*
]


---
class: left

.blockquote[Lab 3 — Design your experiment]

## `r fa('laptop-code', fill='#D50006')` Lab 3
<div style="border:3px solid #D50006; border-radius:8px; padding:20px 28px;
            background:#fff5f5; margin-top:1em;">
<div style="font-family:sans-serif; font-size:16px; font-weight:bold;
            color:#D50006; margin-bottom:10px;">
`r fa('flask')` Lab 3: Design your experiment — Innovator's Hypothesis
</div>
<div style="font-size:0.85em; line-height:1.8;">
<strong>Learning objective:</strong> Recognise endogeneity and design an
appropriate A/B test. Understand why historical correlation collapses
under intervention.
<br>
<strong>Content:</strong> Experience the endogeneity problem (omitted
variable bias) on a concrete example · Formulate hypothesis · Design
A/B test · Compute Cohen's d · Reveal the spurious correlation ·
<em>"Embrace your ignorance"</em>
<br>
`r fa('clock')` <strong>25 minutes</strong> &nbsp;·&nbsp;
`r fa('users')` <strong>Individual or pairs</strong>
<br>
`r fa('circle-right')` <strong>Open lab document:</strong>
<code>Session3_Lab3_Experiment_OJ.qmd</code>
</div>
</div>
<br>
.center[.small[Continue after the lab → <strong>Debrief & Summary</strong>]]

???

* Goal: Students experience the endogeneity problem on a concrete
  example
* Design an A/B test as the solution
* Connection to p-hacking (Session 2): set hypothesis BEFORE the test!


---
class: inverse, center, middle

## Summary & Take-aways



---
class: left

.blockquote[Summary]

## Conclusion: Leading Questions for Management

As a *critical validator*, you no longer let yourself be dazzled by
in-sample $R^2$ or bare p-values. Before budgets are released:
**Do we want to predict or to intervene?**

.pull-left[
### When you predict...
**(Predictive modelling / induction)**
* **The danger:** Overfitting (Danger 1).
* **The diagnosis:** C2 violation (multicollinearity) is standard in
  big data.
* **Your demand to the team:** *"I do not trust any in-sample fit. Show
  me **out-of-sample performance** (cross-validation) and use
  regularisation (Lasso)!"*
]

.pull-right[
### When you intervene...
**(Causal modelling / intervention)**
* **The danger:** Lucas critique (Danger 2).
* **The diagnosis:** A1 violation (missing variables) → C1 violation
  (endogeneity).
* **Your demand to the team:** *"Historical correlation collapses under
  intervention. We absolutely need a causal design (**A/B testing,
  DiD, IV**)!"*
]


---
class: left

.blockquote[Summary]

## Block 3: The red thread

```{r}
#| label: red-thread
#| echo: false
#| warning: false
#| fig-height: 4
#| fig-width: 12
tibble(
  x = 1:6,
  label = c("Descriptive\nregression\n(Session 1)",
            "Stochastic\nmodel\n(shift)",
            "Inference vs.\nprediction\n(purpose)",
            "Assumptions\n(A/B/C)\n& BLUE bridge",
            "Multicollinearity\n& overfitting\n(regularisation)",
            "Endogeneity\n& causality\n(experiments)"),
  color = c("#95a5a6", "#95a5a6", "#3498db", "#2ecc71", "#e74c3c", "#e74c3c")
) %>%
  ggplot(aes(x, y = 0)) +
  geom_segment(aes(xend = x, yend = 0.5), color = "grey70", linewidth = 2) +
  geom_point(aes(y = 0.5, color = I(color)), size = 8) +
  geom_text(aes(y = 0.8, label = label), size = 3.3, lineheight = 0.9) +
  geom_segment(aes(x = 1, xend = 6, y = 0, yend = 0),
               arrow = arrow(length = unit(0.3, "cm")), linewidth = 1, color = "#2c3e50") +
  annotate("text", x = 1, y = -0.15, label = "Sessions 1–2", size = 3, color = "grey50") +
  annotate("text", x = 3.5, y = -0.15, label = "Session 3: Complexity, Generalisation & Causality",
           size = 3.5, fontface = "bold") +
  coord_cartesian(ylim = c(-0.3, 1.2)) +
  theme_void(base_size = 14)
```


---
class: left

## Labs: Inductive Learning

.pull-left[
### Lab 1: Train-test split & CV
*Entry point to prediction*

- Fit a model → great $R^2$!
- On test data: 💥 → experience overfitting
- Cross-validation as the systematic remedy
- Choose $K$ wisely

### Lab 2: Multicollinearity & regularisation

- Correlated features → VIF
- Lasso, Ridge, Elastic Net
- CV for $\lambda$ → what happens to $\hat{\beta}$?
]

.pull-right[
### Lab 3: Design your experiment
*A/B testing (Schrage)*

- Simulate an A/B test
- Hypothesis **before** vs. **after** the test
- Compute Cohen's d
- Reveal spurious correlation
- *"Embrace your ignorance"*
]


---
class: center, middle

> *"An interpretable model translates raw data into information that is
> directly relevant to decision-making."* — Matt Taddy

<br>

### The value creation of the future does not arise from blindly applying algorithms.

### It arises from **translators** who critically validate models and steer them strategically.


---
class: inverse, center, middle

# Thank you!

### Questions? → `r fa('comments')` Discussion & feedback
