---
title: "Data Science and Strategy for Business"
subtitle: "ER114 — Session 3"
author: "Prof. Dr. Jörg Schoder"
institute: "FFHS"
date: "`r Sys.Date()`"
bibliography: ../../../lit/my_bib.bib
lang: en-US
reference-section-title: References
output:
  xaringan::moon_reader:
    self_contained: true
    css:
         - default
         - ../../../css/ffhs-theme_js.css
         - xaringan-themer.css
    includes:
      after_body: ../../../css/insert-logo.html
    lib_dir: ../../../libs
    nature:
      slideNumberFormat: "%current%/%total%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: false
---
class: title-slide

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_xaringan(text_color = "#d50006",inverse_text_color = "#FFFFFF",inverse_background_color = "#d50006", title_slide_background_color = "#d50006",header_background_color = "#d50006",header_color = "#FFFFFF",header_h1_font_size = "32px",
  header_h2_font_size = "26px",link_color="#502479",
  header_h3_font_size = "20px",text_slide_number_color = "#d50006",text_slide_number_font_size = "0.5em")
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_progress_bar(color = "#d50006", location = "bottom")
xaringanExtra::use_xaringan_extra(c("tile_view","scribble","panelset","tachyons"))
xaringanExtra::style_panelset_tabs(font_family = "inherit")
xaringanExtra::use_editable(expires = 1)
xaringanExtra::use_freezeframe(trigger = "hover")
```

```{r}
#| label: setup
#| include: false
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(kableExtra)
library(latex2exp)
library(fontawesome)
library(patchwork)
source(xfun::from_root("lit","helper.R"))
library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           style = "markdown",
           dashed = TRUE)
bib <- ReadBib(xfun::from_root("lit","my_bib.bib"))
Zitiere <- function(bib, key, ...) {
  gsub(" and ", " und ", Citet(bib, key, ...))
}
```

# ER114 — Data Science and Strategy for Business

## Session 3

### Predict or Intervene? Two Paths Beyond Inference

<br>
<br>
<br>
<br>
<br>
### FS 2026
<br>
### Prof. Dr. Jörg Schoder
.mycontacts[
`r fa('github')` @FFHS-EconomicResearch
`r fa('linkedin')` @jfschoder
]


---
layout: true

<div class="my-footer"></div>

<div style="position: absolute;left:400px;bottom:10px;font-size:9px">`r fa('creative-commons')``r rmarkdown::metadata$author`</div>


---
class: left

.blockquote[Programme]

## Today's Programme

<div style="display:grid; grid-template-columns:80px 1fr; gap:0.3em 1em;
            font-size:0.82em; margin-top:0.5em; align-items:center;">

<div style="color:#686868;">10'</div>
<div>`r fa('rotate-left')` <strong>Recap</strong> — where Sessions 1 & 2 left us, two open questions</div>

<div style="color:#686868;">25'</div>
<div>`r fa('layer-group')` <strong>From describing to modelling</strong> — induction & deduction · two dangers · AI Value Creator</div>

<div style="color:#686868;">25'</div>
<div>`r fa('layer-group')` <strong>BLUE: the common bridge</strong> — A-B-C assumptions (Auer) · bias-variance · why both paths share this foundation</div>

<div style="color:#686868;">15'</div>
<div>`r fa('layer-group')` <strong>Path A: Predictive Modelling</strong> — multicollinearity · OOS · Lasso · CV</div>

<div style="color:#D50006; font-weight:bold;">40'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab 1</strong> — out-of-sample on Dominick's panel</div>

<div style="color:#686868; font-style:italic;">10'</div>
<div style="color:#686868; font-style:italic;">`r fa('mug-hot')` Break</div>

<div style="color:#D50006; font-weight:bold;">25'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab 2</strong> — Lasso on demographics</div>

<div style="color:#686868;">25'</div>
<div>`r fa('layer-group')` <strong>Path B: Causal Modelling</strong> — endogeneity · data origin · exogeneity by assumption vs. design · A/B testing</div>

<div style="color:#D50006; font-weight:bold;">25'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab 3</strong> — design A/B test for advertising effect</div>

<div style="color:#686868;">10'</div>
<div>`r fa('flag-checkered')` <strong>Synthesis & Outlook</strong></div>

</div>


---
class: inverse, center, middle

## Recap & Big Picture

.blockquote[Sessions 1 → 2 → **3**]


---
class: left

.blockquote[The story so far]

## Sessions 1 → 2 → **3**

.pull-left[
### Session 1 — Describe
- Dominick's Store 51, three brands
- Tropicana / Minute Maid / Dominick's
- $n = 336$ observations
- EDA, multiple regression, interaction terms
- Findings: brand effect ≈ 0.4, advertising ≈ 0.25 log-units
- *Population mode, no inference*
]

.pull-right[
### Session 2 — Test hypotheses
- Same data, new question: how reliable?
- Sample-splitting — and its limits at $n = 336$
- Replication on Store 50 — robust!
- Inference machine: bootstrap, t-test = ANOVA = regression
- *Stochastic mode, with inference*
]

📌 **Two unanswered questions remain.**


---
class: left

.blockquote[Two open questions]

## Two Questions Session 3 Tackles

.pull-left[
### Question 1
*Does our model also work for stores **we have never seen**?*

This is the **generalisation** question.
It is the question of out-of-sample performance.

→ **Path A: Predictive Modelling.**
]

.pull-right[
### Question 2
*If the advertising effect is 0.25 — what happens **when we actively run advertising tomorrow**?*

This is the **intervention** question.
It is the question of causal inference.

→ **Path B: Causal Modelling.**
]

<br>
<br>
📌 **The two paths are not in opposition. They are answers to different
questions.** And they share a common foundation: the BLUE/bias-variance
framework.


---
class: inverse, center, middle

## Part 1: From Describing to Modelling

.blockquote[Induction & deduction]

.blockquote[The two dangers]

.blockquote[AI User to AI Value Creator]


---
class: left

.blockquote[Epistemological foundation]

## Induction and Deduction in Data Science

To understand the challenges of data science, we split the process into
two logical steps.

.pull-left[
### 1. Induction (the "learning")
* **What happens:** We use thousands of historical data points to derive a
  general rule.
* **Data-science term:** *model training / data mining*.
* **The Dominick's pattern:** *"In the historical data, weeks with
  advertising have ~25% more sales."*
]

.pull-right[
### 2. Deduction (the "application")
* **What happens:** We apply the induced rule to new individual cases.
* **Data-science term:** *scoring / prediction*.
* **The application:** For next week's advertising decision: "we expect
  +25% more sales".
]

📌 **Aha-moment:** Data mining (model training) is nothing other than
machine induction.


---
class: left

.blockquote[The two dangers]

## Failed Generalisation and the Causality Trap

When managers blindly trust algorithmic rules, they fall into one of two traps:

### Danger 1: Failed induction (lack of generalisation)
* The model has not learned a real rule but memorised random fluctuations
  of our specific sample (*overfitting*).
* **Consequence:** Predictions on new data fail.
* **Solution:** Out-of-sample evaluation (train-test split / cross-validation).

### Danger 2: Blind deduction with interventions (causality trap)
* Our induced rule (*"advertising → +25% sales"*) is based on
  *correlation*. There may be a hidden confounder: in our Dominick's
  data, advertising and price-promotions are perfectly entangled
  (`feat = 1 ⟹ deal = 1`).
* **Consequence:** Management runs advertising **without** a price
  promotion → effect collapses.
* **Solution:** Causal inference (e.g. A/B test).


---
class: left

.blockquote[The two dangers]

## From AI User to AI Value Creator

.pull-left[
### The AI User (passive)
* Consumes analytical results and trusts the algorithmic "black box".
* **The risk:** Makes deductive decisions based on models that, due to
  violated foundational assumptions, collapse immediately in reality.
]

.pull-right[
### The AI Value Creator (active)
* Acts as a **"critical validator"** and steers models with strong data
  governance and exclusive data.
* **The advantage:** Critically checks the statistical logic to ensure
  that genuine causality or generalisation is present before budgets are
  released.
]


---
class: inverse, center, middle

## Part 2: BLUE — The Common Bridge

.blockquote[A-B-C assumptions (Auer)]

.blockquote[Why BLUE matters for both paths]

.blockquote[The bias-variance decomposition]


---
class: left

.blockquote[A-B-C assumptions (Auer)]

## Model Assumptions in Brief

.panelset[
.panel[.panel-name[A-Assumptions]
### A-Assumptions: the model

- **A1 (variable selection)** — No relevant variables missing, no irrelevant ones included
- **A2 (linearity)** — The true relationship is linear in the parameters
- **A3 (parameter constancy)** — Parameters are constant across all $N$ observations
]
.panel[.panel-name[B-Assumptions]
### B-Assumptions: the error term

- **B1 (expectation)** — $E(\varepsilon_i) = 0$
- **B2 (homoscedasticity)** — $\text{Var}(\varepsilon_i) = \sigma^2$
- **B3 (no autocorrelation)** — $\text{Cov}(\varepsilon_i, \varepsilon_j) = 0$ for $i \neq j$
- **B4 (normality)** — $\varepsilon_i \sim N$
]

.panel[.panel-name[C-Assumptions]
### C-Assumptions: the data

- **C1 (exogeneity)** — $\text{Cov}(X_i, \varepsilon_i) = 0$
- **C2 (no perfect multicollinearity)** — No regressor is an exact linear combination of others

### ⚡ Why C1 and C2 are decisive for management:

| Violation | Consequence | Strategy |
|:----------|:----------|:----------|
| **C2** (multicollinearity) | Overfitting (Danger 1) | **Path A:** regularisation |
| **C1** (endogeneity) | Biased causal conclusions (Danger 2) | **Path B:** experiment |

*→ A1 violation (missing variable) leads to C1 violation — the "domino effect"!*
]
.quellePanURL[cf. `r Citet(bib, "vonAuer_2023")`, p. 167ff
]
]


---
class: left

.blockquote[BLUE & bias-variance]

## Why BLUE Matters for Both Paths

.pull-left[
### What BLUE says
Under the Gauss-Markov assumptions (A1, A2, A3, B1, B2, B3, C1, C2),
OLS is the **B**est **L**inear **U**nbiased **E**stimator:

- **B**est = minimum variance
- among all **L**inear, **U**nbiased **E**stimators

Note the order: *first* unbiased, *then* minimum variance among the unbiased.
]

.pull-right[
### Why this matters for both paths
**Path B (inference):** unbiasedness is what we want. The estimate must
reflect the true effect. → BLUE is the gold standard.

**Path A (prediction):** unbiasedness is *not* what we need. We want
low total error — and that may require accepting some bias.

→ Same foundation, **different optimality criterion**.
]

📌 The same theoretical foundation explains both paths — but with
opposite priorities.


---
class: left

.blockquote[BLUE & bias-variance]

## Bias-Variance Decomposition: The Key Equation

$$\text{MSE}(\hat{Y}) = \underbrace{\text{Bias}^2(\hat{Y})}_{\text{systematic error}} + \underbrace{\text{Var}(\hat{Y})}_{\text{random error}} + \underbrace{\sigma_{\varepsilon}^2}_{\text{irreducible}}$$

.pull-left[
### Path A reasoning
Total error matters. We *trade* bias for variance:

- accept some bias
- gain much more in variance reduction
- → lower MSE on **new data**

This is exactly what Lasso and Ridge do.
]

.pull-right[
### Path B reasoning
Bias is the enemy. Variance can be quantified (CIs, bootstraps).

- bias = invalid inference
- variance = honest uncertainty
- → BLUE is the gold standard
- → confounders ≠ noise, they are bias
]

📌 **Same theory — opposite direction.**


---
class: left

.blockquote[BLUE & bias-variance]

## The Trade-off, Visualised

```{r}
#| label: bias-var-tradeoff
#| echo: false
#| out-width: "55%"
#| fig-align: "center"
set.seed(42)
complexity <- seq(1, 10, by = 0.1)
bias2 <- 3 * exp(-0.5 * complexity)
variance <- 0.1 * exp(0.4 * complexity)
total <- bias2 + variance + 0.3

df <- data.frame(
  x = rep(complexity, 3),
  y = c(bias2, variance, total),
  type = rep(c("Bias²", "Variance", "Total error (MSE)"), each = length(complexity))
)

ggplot(df, aes(x, y, color = type, linetype = type)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = complexity[which.min(total)], linetype = "dashed", alpha = 0.5) +
  annotate("text", x = complexity[which.min(total)] + 0.3, y = 3.5,
           label = "Sweet spot", angle = 90, size = 3.5) +
  scale_color_manual(values = c("Bias²" = "#E74C3C", "Variance" = "#3498DB",
                                 "Total error (MSE)" = "#2C3E50")) +
  labs(x = "Model complexity", y = "Error", color = "", linetype = "") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

📌 The optimum is **not** at zero bias.


---
class: inverse, center, middle

## Part 3: Path A — Predictive Modelling

.blockquote[Multicollinearity]

.blockquote[Out-of-sample evaluation]

.blockquote[Regularisation: Lasso, Ridge, Elastic Net]


---
class: left

.blockquote[Path A]

## The Predictive Question

> *"How well does my model predict for **new stores** I have never seen?"*

We expand the data world from one store (Sessions 1+2) to **all 83
Dominick's stores**. Plus their demographic data.

| Variable | Type |
|---|---|
| `log_units`, `log_price`, `feat`, `deal` | weekly observation per store-brand |
| `EDUC`, `INCOME`, `HVAL150`, `AGE60`, etc. | store-level demographics (12 variables) |

The big-data world: $n \approx 28{,}000$ observations, 12+ demographic
predictors. **What was a teaching example is now practical scale.**


---
class: left

.blockquote[Path A]

## Multicollinearity in the Demographics

```{r}
#| label: demo-cor
#| echo: false
#| out-width: "60%"
#| fig-align: "center"
demo_cor <- matrix(c(
   1.00, -0.31, -0.09, -0.15, -0.33, -0.63, -0.11,
  -0.31,  1.00, -0.34,  0.66, -0.39,  0.55,  0.89,
  -0.09, -0.34,  1.00, -0.72,  0.24, -0.28, -0.42,
  -0.15,  0.66, -0.72,  1.00, -0.08,  0.40,  0.64,
  -0.33, -0.39,  0.24, -0.08,  1.00, -0.28, -0.47,
  -0.63,  0.55, -0.28,  0.40, -0.28,  1.00,  0.45,
  -0.11,  0.89, -0.42,  0.64, -0.47,  0.45,  1.00
), nrow = 7, byrow = TRUE)
rownames(demo_cor) <- colnames(demo_cor) <-
  c("AGE60","EDUC","ETHNIC","INCOME","HHLARGE","WORKWOM","HVAL150")

df_cor <- as.data.frame(as.table(demo_cor))
names(df_cor) <- c("Var1","Var2","cor")

ggplot(df_cor, aes(Var1, Var2, fill = cor)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", cor)), size = 3) +
  scale_fill_gradient2(low = "#502479", mid = "white", high = "#D50006",
                        midpoint = 0, limits = c(-1,1)) +
  labs(x = "", y = "", fill = "r") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**EDUC ↔ HVAL150** = 0.89 — practically the same dimension.
**INCOME ↔ ETHNIC** = −0.72 — substantial overlap.

→ The "wealth cluster": EDUC + INCOME + HVAL150 capture the same
underlying construct.


---
class: left

.blockquote[Path A]

## Multicollinearity: What It Does

.pull-left[
### For inference (Path B-style)
- $\hat{\beta}$ becomes **unstable**
- Standard errors blow up
- Confidence intervals widen
- Coefficients can switch signs with tiny data changes

→ **Multicollinearity is poison for explanation.**

Diagnose with VIF: $\text{VIF}_j = 1/(1 - R^2_j)$.
Above 5: warning. Above 10: serious problem.
]

.pull-right[
### For prediction (Path A)
- $\hat{\beta}$ unstable, but...
- ...$\hat{y}$ stays stable
- The model still predicts well
- Multicollinearity is **often tolerable**

→ **For prediction, what matters is total error.**
]


---
class: left

.blockquote[Path A]

## Underfitting vs. Overfitting

```{r}
#| label: overfitting-demo
#| echo: false
#| message: false
#| fig-height: 4.2
#| fig-width: 11
set.seed(42)
n <- 30
x <- sort(runif(n, 0, 10))
y <- sin(x) + rnorm(n, 0, 0.3)
df_fit <- data.frame(x = x, y = y)

p1 <- ggplot(df_fit, aes(x, y)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#3498DB", linewidth = 1.2) +
  labs(title = "Underfitting (linear)", subtitle = "High bias, low variance") +
  theme_minimal(base_size = 12)

p2 <- ggplot(df_fit, aes(x, y)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 3), se = FALSE, color = "#27AE60", linewidth = 1.2) +
  labs(title = "Good fit (poly. degree 3)", subtitle = "Balance") +
  theme_minimal(base_size = 12)

p3 <- ggplot(df_fit, aes(x, y)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 20), se = FALSE, color = "#E74C3C", linewidth = 1.2) +
  labs(title = "Overfitting (poly. degree 20)", subtitle = "Low bias, high variance") +
  theme_minimal(base_size = 12)

p1 + p2 + p3
```

`r fa('circle-right')` **The middle one generalises. The right one memorises.**


---
class: left

.blockquote[Path A]

## Train-Test Split → K-Fold CV

```{r}
#| label: cv-figure
#| echo: false
#| out-width: "70%"
#| fig-align: "center"
knitr::include_graphics(xfun::from_root("img","PVA3","kfold_crossvalidation.svg"))
```

$$\text{CV-Score} = \frac{1}{K} \sum_{k=1}^{K} \text{MSE}_k$$

| Variant | When? |
|:---------|:------|
| K-Fold | Standard ($K = 5$ or $10$) |
| Stratified | Unbalanced classes |
| Time-series CV | Time series (no future leakage!) |
| Nested CV | Model selection + evaluation |


---
class: left

.blockquote[Path A]

## Lasso vs. Ridge vs. Elastic Net

.pull-left[
### Ridge (L2)
$$\lambda \sum \beta_j^2$$
- Shrinks all
- Removes none
- Best for **multicollinearity** with many small effects

### Lasso (L1)
$$\lambda \sum |\beta_j|$$
- Shrinks **and** zeroes
- Variable selection
- Best for **sparsity** (few active variables)
]

.pull-right[
### Elastic Net
$$\lambda_1 \sum |\beta_j| + \lambda_2 \sum \beta_j^2$$
- Combines L1 + L2
- Selects *and* keeps groups of correlated features
- Often best in practice

### In R
```r
library(glmnet)
cv_fit <- cv.glmnet(X, y, alpha = 1)  # Lasso
cv_fit <- cv.glmnet(X, y, alpha = 0)  # Ridge
cv_fit <- cv.glmnet(X, y, alpha = 0.5) # Elastic Net
```
]


---
class: left

.blockquote[Path A]

## The CV Path: Choosing $\lambda$

```{r}
#| label: lasso-cv-path
#| echo: false
#| message: false
#| fig-height: 4.2
#| fig-width: 7
#| fig-align: "center"
set.seed(42)
n <- 200; p <- 8
X <- matrix(rnorm(n*p), n, p)
colnames(X) <- c("EDUC","INCOME","HVAL150","ETHNIC","AGE60","WORKWOM","HHLARGE","Random")
y <- 2*X[,1] + 1.5*X[,2] + 0.5*X[,3] - 1*X[,4] + 0.5*X[,5] + rnorm(n)

library(glmnet)
fit <- glmnet(X, y, alpha = 1)
beta_df <- as.data.frame(as.matrix(t(coef(fit)[-1,])))
beta_df$lambda <- fit$lambda
beta_long <- pivot_longer(beta_df, -lambda, names_to = "Variable", values_to = "Coefficient")

ggplot(beta_long, aes(log10(lambda), Coefficient, color = Variable)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3) +
  scale_color_brewer(palette = "Set2") +
  labs(x = expression(log[10](lambda)), y = "Coefficient",
       title = "Lasso path: coefficients shrink as λ grows") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")
```

- Each line = one feature
- As $\lambda$ grows → coefficients shrink → some hit 0
- **CV picks the $\lambda$** that minimises out-of-sample MSE


---
class: left

.blockquote[Lab 1 — Out-of-sample on Dominick's]

## `r fa('laptop-code', fill='#D50006')` Lab 1
<div style="border:3px solid #D50006; border-radius:8px; padding:20px 28px;
            background:#fff5f5; margin-top:1em;">
<div style="font-family:sans-serif; font-size:16px; font-weight:bold;
            color:#D50006; margin-bottom:10px;">
`r fa('flask')` Lab 1: Out-of-sample on Dominick's full panel
</div>
<div style="font-size:0.85em; line-height:1.8;">
<strong>Learning objective:</strong> Experience the gap between in-sample
and out-of-sample performance on the full 83-store Dominick's panel.
Apply train-test split and K-fold cross-validation.
<br>
<strong>Content:</strong>
Fit on Store 51 → great $R^2$! → predict for Store 50: realistic.
Predict for a *random* store: 💥 — out-of-sample reality.
Then K-fold CV across all 83 stores.
<br>
`r fa('clock')` <strong>40 minutes</strong> &nbsp;·&nbsp;
`r fa('users')` <strong>Individual or pairs</strong>
<br>
`r fa('circle-right')` <strong>Open lab document:</strong>
<code>Session3_Lab1_OutOfSample_OJ.qmd</code>
</div>
</div>


---
class: left

.blockquote[Lab 2 — Lasso on demographics]

## `r fa('laptop-code', fill='#D50006')` Lab 2
<div style="border:3px solid #D50006; border-radius:8px; padding:20px 28px;
            background:#fff5f5; margin-top:1em;">
<div style="font-family:sans-serif; font-size:16px; font-weight:bold;
            color:#D50006; margin-bottom:10px;">
`r fa('flask')` Lab 2: Multicollinearity & regularisation
</div>
<div style="font-size:0.85em; line-height:1.8;">
<strong>Learning objective:</strong> Diagnose multicollinearity in the
demographic variables, apply Lasso/Ridge/Elastic Net, choose $\lambda$
by cross-validation.
<br>
<strong>Content:</strong>
VIF on the 12 demographics → identify the wealth cluster.
Lasso path: which variables survive? Ridge for stable coefficients.
Compare on out-of-sample MSE.
<br>
`r fa('clock')` <strong>25 minutes</strong> &nbsp;·&nbsp;
`r fa('users')` <strong>Individual or pairs</strong>
<br>
`r fa('circle-right')` <strong>Open lab document:</strong>
<code>Session3_Lab2_Regularisation_OJ.qmd</code>
</div>
</div>


---
class: inverse, center, middle

## Part 4: Path B — Causal Modelling

.blockquote[Endogeneity]

.blockquote[Where do the data come from?]

.blockquote[Exogeneity by assumption vs. by design]

.blockquote[A/B testing]


---
class: left

.blockquote[Path B]

## The Intervention Question

> *"If we run advertising **tomorrow** for the Dominick's brand — what
> happens to sales?"*

This is *not* a prediction question. It is an **intervention** question.

| Prediction | Intervention |
|---|---|
| World stays as is | World changes through our action |
| Past correlations apply | Past correlations may break |
| Lasso, CV, OOS | Causal inference, RCT |

📌 **Lucas critique** (1976): a model trained on observational data may
collapse the moment we intervene.


---
class: left

.blockquote[Path B]

## Endogeneity: Why OLS Fails

When **C1 (exogeneity)** is violated:

$$\text{Cov}(X_i, \varepsilon_i) \neq 0$$

OLS is **biased and inconsistent**. The coefficient does not reflect
the causal effect.

Three classic sources:
1. **Omitted variable bias** (confounding)
2. **Simultaneity / reverse causality**
3. **Measurement error in $X$**

📌 In our Dominick's data: the perfect implication
$$\text{feat} = 1 \;\;\Longrightarrow\;\; \text{deal} = 1$$
means we **cannot disentangle** the advertising effect from the
price-promotion effect — using observational data alone.


---
class: left

.blockquote[Path B]

## Where Do the Data Come From?

The validity of causal claims depends not primarily on the method,
but on the **type of data collection**.

| Type of data | Exogeneity? | Method |
|---|---|---|
| **Survey** (cross-section, longitudinal) | Plausibility argued, never guaranteed; risks: self-selection, non-response | OLS under MLR.4 — Wooldridge world |
| **Experiment** (RCT, A/B test) | Exogeneity *created* by randomisation | Comparison of means, regression adjustment |
| **Natural experiment** (IV, DiD, RDD) | Exogeneity argued via "as-if-random" allocation | IV / DiD / RDD as identification strategies |

📌 **Hernán's central message:** *"Causal inference is just inference
with the right data."*

The method is the same OLS we know — what differs is the **data
foundation**.


---
class: left

.blockquote[Path B]

## Two Routes to Exogeneity

.pull-left[
### Route 1 — Exogeneity by assumption
We **assume** that all relevant confounders are observed and included.

- The Wooldridge MLR.4 condition
- Rest of the world goes into the error term
- We **trust** that the error is mean-independent of X

→ Risk: A1 violation (missing variable) → C1 violation → biased.

→ Tool: **regression adjustment** with rich covariates.
]

.pull-right[
### Route 2 — Exogeneity by design
We **engineer** exogeneity through the data collection.

- RCT / A/B test: random assignment guarantees exogeneity
- IV: instrument exogenous by argument (lottery, weather)
- DiD: parallel trends replaces exogeneity
- RDD: discontinuity at threshold replaces exogeneity

→ Tool: **identification strategy** built into the data.
]

📌 **The two routes share the same regression machinery — what differs
is the data.**


---
class: left

.blockquote[Path B]

## Internal vs. External Validity

.pull-left[
### Internal validity
*Have we cleanly isolated causality in **our** study?*

- Did we control for all confounders?
- Is the design mathematically tight?
- The enemy: omitted variable bias, selection bias.
]

.pull-right[
### External validity
*Does the result transfer to **the real market**?*

- Is our sample representative?
- Do test conditions match deployment?
- The enemy: artificial environments, narrow sampling.
]

📌 **The Holy Grail: a design that maximises both.**


---
class: left

.blockquote[Path B]

## A/B Testing: The Gold Standard

.pull-left[
### How A/B testing heals C1
1. Users assigned by **strict randomisation** to A or B
2. Randomness guarantees treatment is independent of all hidden confounders
3. **C1 is artificially restored**

### Cohen's d for effect size
$$d = \frac{\bar{X}_B - \bar{X}_A}{s_{\text{pooled}}}$$
| d | Interpretation |
|:--:|:---|
| 0.2 | small |
| 0.5 | medium |
| 0.8 | large |
]

.pull-right[
### Schrage's principle
*"Run experiments not because you can confidently predict outcomes —
but precisely because you can't."*

Microsoft's Ronny Kohavi: only **⅓** of A/B-tested features improve
metrics. Even experts get the answer wrong **80%** of the time.

→ **Conviction ≠ evidence.**

📌 **In Lab 3:** Design an A/B test for the Dominick's advertising
effect.
]


---
class: left

.blockquote[Path B]

## Quasi-Experiments: When RCTs Are Impossible

Sometimes ethics, law, or technology rule out a clean A/B test. The
data team then turns to **quasi-experimental designs**:

| Design | Idea | When? |
|---|---|---|
| **Instrumental Variables (IV)** | A *random* element shifts X but only affects Y through X | Compliance with random assignments; lottery, distance, weather |
| **Difference-in-Differences (DiD)** | Compare *trends* of treated and control before/after a policy change | Geographic rollout, regulatory change |
| **Regression Discontinuity (RDD)** | Compare units just above and just below a hard cutoff | Bonus thresholds, age cutoffs, eligibility scores |

📌 The **Critical Validator** (manager) must recognise *which* quasi-
experimental tool fits the situation — and demand it.


---
class: left

.blockquote[Lab 3 — Design an A/B test]

## `r fa('laptop-code', fill='#D50006')` Lab 3
<div style="border:3px solid #D50006; border-radius:8px; padding:20px 28px;
            background:#fff5f5; margin-top:1em;">
<div style="font-family:sans-serif; font-size:16px; font-weight:bold;
            color:#D50006; margin-bottom:10px;">
`r fa('flask')` Lab 3: Design an A/B test for the advertising effect
</div>
<div style="font-size:0.85em; line-height:1.8;">
<strong>Learning objective:</strong> Experience the limit of observational
data for causal claims. Design a clean A/B test that breaks the feat ↔
deal entanglement.
<br>
<strong>Content:</strong>
Why does observational analysis fail here? Sketch the A/B design.
What is randomised, what is the metric, what sample size? Compute
Cohen's d under realistic effect sizes.
<br>
`r fa('clock')` <strong>25 minutes</strong> &nbsp;·&nbsp;
`r fa('users')` <strong>Individual or pairs</strong>
<br>
`r fa('circle-right')` <strong>Open lab document:</strong>
<code>Session3_Lab3_Experiment_OJ.qmd</code>
</div>
</div>


---
class: inverse, center, middle

## Synthesis & Outlook


---
class: left

.blockquote[Synthesis]

## Two Demands the Translator Should Make

As an *AI Value Creator* you no longer let yourself be dazzled by
in-sample $R^2$ or bare p-values. **Before budgets are released:**

.pull-left[
### When you predict...
*(Path A)*
- The danger: **overfitting** on training data
- The diagnosis: high VIF, complex model
- Your demand: *"Show me out-of-sample performance via cross-validation. Use Lasso."*
]

.pull-right[
### When you intervene...
*(Path B)*
- The danger: **Lucas critique** — past correlations may break
- The diagnosis: missing variables, observational data only
- Your demand: *"Design an A/B test or find a credible quasi-experimental identification."*
]


---
class: left

.blockquote[Synthesis]

## Six Foundations Bridge — Active in Session 3

| # | Classical tradition | Bridge | Data Science |
|---|---|---|---|
| 1 | E[Y\|X] — conditional mean | market benchmark | residual as arbitrage |
| 2 | sampling distribution | uncertainty | bootstrap, infer |
| 3 | expected value & error types | ROI calculus | EV framework |
| 4 | p-value, multiplicity | signal vs. noise | FDR, p-hacking, HARKing |
| 5 | correlation ≠ causality | intervention | A/B testing, causal ML |
| **6** | **BLUE, Gauss-Markov, bias-variance** | **complexity control** | **CV, Lasso, regularisation** |

📌 **Foundation 6 is what we have built today.**


---
class: left

.blockquote[Synthesis]

## Block 3: The Red Thread

```{r}
#| label: red-thread
#| echo: false
#| warning: false
#| fig-height: 4
#| fig-width: 12
tibble(
  x = 1:6,
  label = c("Descriptive\nregression\n(Session 1)",
            "Stochastic\nmodel\n(shift)",
            "Inference vs.\nprediction\n(purpose)",
            "BLUE bridge\n(common\nfoundation)",
            "Path A:\nLasso, CV,\nregularisation",
            "Path B:\nendogeneity,\nA/B testing"),
  color = c("#95a5a6", "#95a5a6", "#3498db", "#2ecc71", "#e74c3c", "#e74c3c")
) %>%
  ggplot(aes(x, y = 0)) +
  geom_segment(aes(xend = x, yend = 0.5), color = "grey70", linewidth = 2) +
  geom_point(aes(y = 0.5, color = I(color)), size = 8) +
  geom_text(aes(y = 0.8, label = label), size = 3.3, lineheight = 0.9) +
  geom_segment(aes(x = 1, xend = 6, y = 0, yend = 0),
               arrow = arrow(length = unit(0.3, "cm")), linewidth = 1, color = "#2c3e50") +
  annotate("text", x = 1, y = -0.15, label = "Sessions 1–2", size = 3, color = "grey50") +
  annotate("text", x = 3.5, y = -0.15, label = "Session 3: Predict or Intervene?",
           size = 3.5, fontface = "bold") +
  coord_cartesian(ylim = c(-0.3, 1.2)) +
  theme_void(base_size = 14)
```


---
class: left

.blockquote[Outlook]

## What Comes Next

.pull-left[
### ER017 — Machine Learning
- Logit / GLM (the binary case)
- Random Forests, gradient boosting
- Causal ML (Causal Forests, double ML)
- Multilevel / hierarchical models
]

.pull-right[
### Time-series methods
- Autocorrelation in panel data
- ARIMA, exponential smoothing
- Forecasting metrics
- Block-bootstrap, time-series CV

### Beyond
- Causal discovery (DAG inference)
- Reinforcement learning
- Bayesian methods
]

→ Many doors open. The ER114 toolkit is the **foundation**.


---
class: left

.blockquote[Closing]

## The Translator's Role

Two sessions taught us to **describe** and **infer**.

This third session taught us **two paths beyond inference**:

- **predict** under stable structure
- **intervene** by changing the data foundation

> *"An interpretable model translates raw data into information that is
> directly relevant to decision-making."* — Matt Taddy

> *"Causal inference is just inference with the right data."* — Miguel Hernán

📌 **The value creation of the future does not arise from blindly
applying algorithms. It arises from translators who critically validate
models and steer them strategically.**


---
class: inverse, center, middle

# Thank you!

### Questions? → `r fa('comments')` Discussion & feedback
