---
title: "Data Science and Strategy for Business"
subtitle: "Complexity, Generalization & Causality: Predict or Intervene?"
author: "Prof. Dr. Jörg Schoder"
institute: "FFHS"
date: "`r Sys.Date()`"
bibliography: ../../lit/my_bib.bib
lang: en-US
reference-section-title: References
output:
  xaringan::moon_reader:
    self_contained: true
    css:
         - default
         - ../../css/ffhs-theme_js.css
         - xaringan-themer.css
    includes:
      after_body: ../../css/insert-logo.html
    lib_dir: ../../libs
    nature:
      slideNumberFormat: "%current%/%total%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: false
---
class: title-slide

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_xaringan(text_color = "#d50006",
               inverse_text_color = "#FFFFFF",
               inverse_background_color = "#d50006",
               title_slide_background_color = "#d50006",
               header_background_color = "#d50006",
               header_color = "#FFFFFF",
               header_h1_font_size = "32px",
               header_h2_font_size = "26px",
               link_color = "#502479",
               header_h3_font_size = "20px",
               text_slide_number_color = "#d50006",
               text_slide_number_font_size = "0.5em")
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_progress_bar(color = "#d50006", location = "bottom")
xaringanExtra::use_xaringan_extra(c("tile_view","scribble","panelset","tachyons"))
xaringanExtra::style_panelset_tabs(font_family = "inherit")
xaringanExtra::use_editable(expires = 1)
xaringanExtra::use_freezeframe(trigger = "hover")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 150)
options(htmltools.dir.version = FALSE)
library(latex2exp)
library(fontawesome)
library(tidyverse)
library(kableExtra)
library(gt)
source(xfun::from_root("lit","helper.R"))
library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           style = "markdown",
           dashed = TRUE)
bib <- ReadBib(xfun::from_root("lit","my_bib.bib"))
Zitiere <- function(bib, key, ...) {
  gsub(" and ", " and ", Citet(bib, key, ...))
}

# colour palette (consistent with labs)
col_primary   <- "#D50006"   # red — Path E1 / OLS / current
col_secondary <- "#502479"   # purple — Path P / Lasso / Lab 2
col_tertiary  <- "#7d0a52"   # dark accent — Path E2 / Causal / Lab 3
col_neutral   <- "#686868"
```


# ER114 — Data Science and Strategy for Business

## Session 3

### Complexity, Generalization & Causality. *Predict or Intervene?*

<br>
<br>
<br>
<br>
<br>
### FS 2026
<br>
### Prof. Dr. Jörg Schoder
.mycontacts[
`r fa('github')` @FFHS-EconomicResearch
`r fa('linkedin')` @jfschoder
]


---
layout: true

<div class="my-footer"></div>

<div style="position: absolute;left:400px;bottom:10px;font-size:9px">`r fa('creative-commons')``r rmarkdown::metadata$author`</div>






---
class: inverse, center, middle

## Recap and Agenda

.blockquote[Recap Session 1 & 2]

.blockquote[Today: Big Picture & Agenda]


---
class: left

.blockquote[Recap]

## Sessions 1 → 2 → **3**

.panelset[
.panel[.panel-name[Session 1]
###  *See &amp; Describe*

`r fa('eye')` Descriptive regression as **conditional mean**:
$$\hat{y} = \mathbb{E}[y\,|\,\mathbf{x}]$$

`r fa('eye')` CRISP-DM as project skeleton

`r fa('eye')` Tidyverse pipeline + EDA

`r fa('eye')` First contact with **OVB** — missing variables bias all coefficients

]
.panel[.panel-name[Session 2]
### *Hypotheses Testing*

`r fa('dice')` Population vs. sample · **stochastic** regression

`r fa('dice')` Bootstrap &amp; permutation — *"only one test"*

`r fa('dice')` Type I/II errors as economic risks (EV-Framework)

`r fa('dice')` p-Hacking, FDR, BH

]
.panel[.panel-name[Session 3]
### *Today*

`r fa('circle-question')` **Inference vs. Prediction** — when is which the right purpose?

`r fa('circle-question')` **Multicollinearity** &amp; the BLUE limits

`r fa('circle-question')` **Bias-variance trade-off** &amp; regularization

`r fa('circle-question')` **Cross-validation** as the out-of-sample yardstick

`r fa('circle-question')` **Causal engineering** — when correlation is not enough

.blockquote[
"The Block&nbsp;3 question is no longer *what do we see?* &mdash; it is *what should we believe and what should we act on?*"
]
]
]

???
* The recap stresses that Sessions 1 and 2 stayed on the **same dataset** with the **same model** — only the lens changed
* Today we change the lens twice more: out-of-sample (Lab 1+2) and counterfactual (Lab 3)


---
class: left

.blockquote[Big picture for today]

## What should we believe — and act on?

<br>

.center[
### `r fa('bridge', fill='#502479')` Lab&nbsp;1: Role of Gauss-Markov-Assumptions
.small[Inference vs. Prediction. Checking Assumptions and assessing consequences.]
]

<br>

.pull-left[
### `r fa('shuffle', fill='#502479')` Will the pattern generalize?

The **prediction** question — out-of-sample.

`r fa('arrow-right', fill='#502479')` Path **P**

`r fa('arrow-right', fill='#502479')` Today's gateway: **Lab&nbsp;2**

]

.pull-right[
### `r fa('diagram-project', fill='#502479')` What changes if I intervene?

The **causal** question — counterfactual.

`r fa('arrow-right', fill='#D50006')` Path **E** (E2 sub-question)

`r fa('arrow-right', fill='#D50006')` Today's gateway: **Lab&nbsp;3**


]

???
* Two simple questions, one whole session
* Lab 1 is the **gateway** — not an afterthought; it is the diagnostic bridge that exposes where each path needs to begin
* Path E1 (classical inference, what is the partial effect?) is what Lab 1 stress-tests; both E2 (causal counterfactual) and P (prediction) inherit lessons from where E1 reaches its limits
* Icon "shuffle" for generalize evokes "across draws / across folds" — more apt than a time-series chart
.small[`r Citet(bib, "breiman_2001")`]
.small[`r Citet(bib, "shmueli_2010")`]


---
class: left

.blockquote[Agenda]

## Session 3 Overview

<div style="display:grid; grid-template-columns:80px 1fr; gap:0.3em 1em;
            font-size:0.82em; margin-top:0.5em; align-items:center;">

<div style="color:#686868;">10'</div>
<div>`r fa('rotate-left')` <strong>Welcome &amp; Recap</strong> — where Sessions 1+2 left us</div>

<div style="color:#686868;">15'</div>
<div>`r fa('lightbulb')` <strong>Prediction vs. Explanation</strong> — the central dilemma of Block&nbsp;3</div>

<div style="color:#686868;">15'</div>
<div>`r fa('chart-line')` <strong>The In-Sample Illusion</strong> — why high R² alone is dangerous</div>

<div style="color:#D50006; font-weight:bold;">45'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab&nbsp;1</strong> — Reality Check via Out-of-Sample Validation</div>

<div style="color:#686868;">15'</div>
<div>`r fa('triangle-exclamation')` <strong>When OLS Fails</strong> — the BLUE bridge, bias–variance, Lasso preview</div>

<div style="color:#686868; font-style:italic;">10'</div>
<div style="color:#686868; font-style:italic;">`r fa('mug-hot')` Break</div>

<div style="color:#D50006; font-weight:bold;">25'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab&nbsp;2</strong> — Multicollinearity &amp; Regularization</div>

<div style="color:#686868;">25'</div>
<div>`r fa('compass')` <strong>Causal-Inference Primer</strong> — beyond correlation: A/B&nbsp;tests, quasi-experiments, IV, DML</div>

<div style="color:#D50006; font-weight:bold;">30'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab&nbsp;3</strong> — Design your Experiment (Innovator's Hypothesis)</div>

<div style="color:#686868;">10'</div>
<div>`r fa('forward')` <strong>Outlook ER017</strong> — Machine Learning &amp; Causal+ML</div>

<div style="color:#686868;">10'</div>
<div>`r fa('flag-checkered')` <strong>Debrief &amp; Evaluation</strong></div>

</div>

???
* Red blocks = labs (Lab 1 is longer because it covers the OLS diagnostic foundation)
* Three connecting threads: DGP-paths E1/E2/P, Spiess credibility argument, Sarah Chen narrative
* Total ≈ 195 min including break



---
class: inverse, center, middle

## Part 1: Foundations

.blockquote[Two Paths from the Data-Generating Process]


---
class: left

.blockquote[Part 1 · The model]

## Why model? The conditional expectation E[y|x]

.pull-left[
For a response $y$ and inputs $\mathbf{x}$:

$$\mathbb{E}[y\,|\,\mathbf{x}] = f(\mathbf{x}'\boldsymbol{\beta})$$

`r fa('arrow-right')` **Marginal** distribution (top): the unconditional shape of $y$ — single number for the mean.

`r fa('arrow-right')` **Conditional** expectation (bottom): a *function* of $\mathbf{x}$ — different mean per slice.

<br>

.blockquote[
.small[
All three paths — E1, E2, P — make claims about parts of this conditional distribution.
]
]


]

.pull-right[
```{r marginal-conditional-stack}
#| echo: false
#| warning: false
#| fig-width: 4.5
#| fig-height: 4.8
#| fig-align: center
#| out-width: "100%"
set.seed(20260522)
n <- 2400
tbl <- tibble(
  bathrooms = sample(0:7, n, replace = TRUE,
                     prob = c(0.04, 0.18, 0.40, 0.22, 0.10, 0.04, 0.015, 0.005)),
  home_value = exp(11.2 + 0.32 * bathrooms + rnorm(n, sd = 0.45))
)

p_marg <- ggplot(tbl, aes(home_value)) +
  geom_histogram(bins = 35, fill = col_neutral, color = "white") +
  scale_x_continuous(labels = scales::label_dollar(scale = 1e-3, suffix = "k"),
                     limits = c(0, 1.6e6)) +
  labs(x = "Home value", y = "Frequency",
       title = "(a) Marginal distribution") +
  theme_minimal(base_size = 9) +
  theme(plot.title = element_text(size = 10, face = "bold"))

p_cond <- ggplot(tbl, aes(factor(bathrooms), home_value)) +
  geom_boxplot(aes(fill = factor(bathrooms)),
               outlier.size = 0.5,
               outlier.colour = col_neutral) +
  scale_y_continuous(labels = scales::label_dollar(scale = 1e-3, suffix = "k")) +
  scale_fill_manual(values = c("#a8e6cf","#80d6c4","#5fc6b8","#43b3a8",
                               "#3196a8","#2576a0","#1c5594","#502479"),
                    guide = "none") +
  labs(x = "Number of bathrooms", y = "Home value",
       title = "(b) Conditional distribution") +
  theme_minimal(base_size = 9) +
  theme(plot.title = element_text(size = 10, face = "bold"))

cowplot::plot_grid(p_marg, p_cond, ncol = 1, rel_heights = c(1, 1.3))
```
]
.quelle[.small[`r Citet(bib, "taddy_modern_2023")` Ch. 1, Fig. 1.1]]

???
* Mirrors Taddy/Hendrix/Harding 2023 Figure 1.1 — both panels (a) marginal histogram and (b) conditional boxplots by bathrooms
* Marginal mean = one number; conditional mean = a function of bathrooms
* The shift from a single number to a function is what regression IS
* All three paths today work on this conditional distribution — but each extracts something different from it


---
class: left

.blockquote[Part 1 · The logic]

## Models and Data: An Epistemiological Perspective

```{r induction-deduction-svg}
#| echo: false
#| out-width: "100%"
#| fig-align: center
knitr::include_graphics(xfun::from_root("img","Session3","induction_deduction.svg"))
```

???
* Two visually distinct halves: Induction (purple) at top, Deduction (red) at bottom
* **Key point:** the role of data is fundamentally different
  - In induction: data is the *source material* from which we generate a hypothesis (data mining, model training)
  - In deduction: data is the *arbiter* that tests an a-priori hypothesis (classical inference) or yields a prediction
* Path P = induction (training) + deduction (predicting new x)
* Path E1 = deduction only (hypothesis pre-specified, then tested)
* No claim of abduction here — it would not add clarity for this audience
* Lab 1 lives in pure deduction; Lab 2 adds explicit induction (CV-tuned model); Lab 3 returns to deduction (test of pre-specified design)


---
class: left

.blockquote[Part 1 · Translator role in Block 3]

## From AI User to AI Value Creator — *applied to method choice*

.panelset[
.panel[.panel-name[Recap from Session 1]
<svg viewBox="0 0 680 200" width="90%" style="display:block;margin:auto">
  <defs>
    <style>
      .rtxt { font-family:sans-serif; font-size:11px; font-weight:bold;
              fill:white; text-anchor:middle; }
      .rarr { stroke:#686868; stroke-width:2; fill:none;
              marker-end:url(#arr); }
      .anno { font-family:sans-serif; font-size:10px; fill:#686868;
              text-anchor:middle; }
    </style>
    <marker id="arr" markerWidth="8" markerHeight="8"
            refX="6" refY="3" orient="auto">
      <path d="M0,0 L0,6 L8,3 z" fill="#686868"/>
    </marker>
  </defs>
  <rect x="10"  y="140" width="130" height="40" rx="6" fill="#b3b2b2"/>
  <text class="rtxt" x="75"  y="165">1. Collect</text>
  <rect x="180" y="140" width="130" height="40" rx="6" fill="#686868"/>
  <text class="rtxt" x="245" y="165">2. Organise</text>
  <rect x="350" y="140" width="130" height="40" rx="6" fill="#7d0a52"/>
  <text class="rtxt" x="415" y="165">3. Analyse</text>
  <rect x="520" y="140" width="140" height="40" rx="6" fill="#502479"/>
  <text class="rtxt" x="590" y="165">4. Infuse</text>
  <line class="rarr" x1="145" y1="160" x2="175" y2="160"/>
  <line class="rarr" x1="315" y1="160" x2="345" y2="160"/>
  <line class="rarr" x1="485" y1="160" x2="515" y2="160"/>
  <line x1="340" y1="20" x2="340" y2="135"
        stroke="#D50006" stroke-width="2" stroke-dasharray="5,3"/>
  <text style="font-family:sans-serif;font-size:10px;fill:#D50006;
               text-anchor:middle;font-weight:bold" x="340" y="15">
    AI User | AI Value Creator
  </text>
  <text class="anno" x="170" y="125">Standard Tools</text>
  <text class="anno" x="170" y="137">Commodity Data</text>
  <text class="anno" x="510" y="125">Proprietary Data</text>
  <text class="anno" x="510" y="137">Sustainable Advantage</text>
</svg>
.center[.small[Session 1 framing — data and tooling determine commodity vs. moat.]]
]
.panel[.panel-name[New in Session 3]
### The same line, applied to *method choice*

.pull-left[
`r fa('user', fill='#686868')` **AI User** *(method-level)*

* Picks the textbook method
* Reports $R^2$, p-value, default outputs
* Confuses Path E and Path P
* Optimizes within a given **map** — without knowing it
]

.pull-right[
`r fa('user-gear', fill='#D50006')` **AI Value Creator** *(method-level)*

* Asks: *prediction or intervention?*
* Picks the **path** before the estimator
* Validates the diagnostic, not just the point estimate
* **Chooses the map** — and defends it
]


.blockquote[
.small[
Today's labs make this concrete. **Lab 1** is the diagnostic discipline. **Lab 2** is the prediction discipline. **Lab 3** is the causal discipline. Knowing which to pick is the Value-Creator skill.
]
]
]
.panel[.panel-name[The data dimension]
### Why method choice matters for the data moat

`r fa('rocket', fill='#D50006')` **Proprietary data** alone is not enough — it must be paired with the **right path**:

* Predicting customer churn from your own data → Path **P** with regularization (Lab&nbsp;2 family)
* Estimating the effect of a *new* pricing policy → Path **E2** with experimental design (Lab&nbsp;3 family)

`r fa('triangle-exclamation', fill='#686868')` Pairing the wrong path with great data wastes the moat:

* Lasso on a one-off intervention question gives confident, biased nonsense
* OLS controls on observational data without a DAG yield estimates the CEO cannot defend

.quellePanURL[`r Citet(bib, "thomasAIValueCreators2025")`]
]
]

???
* Three panels: (1) Session 1 framing as recall, (2) the NEW Session 3 framing applied to method choice, (3) why path + data must align
* Key shift from Session 1: Value Creation in Block 3 is no longer just about *data* — it is about the *method-path-decision* itself
* Lab 3's deliverable T_final is exactly the artefact a Value Creator produces: a defensible methodological recommendation


---
class: left

.blockquote[Part 1 · The deeper layer]

## Philosophy Eats AI — the layer beneath every method

```{r philosophy-svg}
#| echo: false
#| out-width: "98%"
#| fig-align: center
knitr::include_graphics(xfun::from_root("img","Session3","philosophy_eats_ai.svg"))
```

???
* Schrage & Kiron (2025): every AI stack rests on philosophical commitments — usually unarticulated, sometimes catastrophically wrong
* Three pillars are theirs (Ontology, Epistemology, Teleology); Ethics is treated separately as a fourth
* **Ontology examples**: Starbucks defined the "Starbucks experience" as connection (not transactions) before designing Deep Brew AI. Amazon Prime ontologically defined "best customers" — not "most profitable" — as the optimization target
* **Epistemology examples**: prediction (Path P) and counterfactual evidence (Path E2) are two different theories of what counts as knowing. Confusing them is the root cause of most bad analytics
* **Teleology examples**: Google's Gemini fiasco — accuracy and diversity goals were misaligned; the model fell to whichever objective dominated
* "The map is not the territory" (Korzybski via Wittgenstein, quoted by Schrage & Kiron): metrics are maps; loyalty is the territory; RFM/NPS/churn are metrics that *might* represent loyalty — or might define it by fiat
* The Translator's job: surface these choices before picking a method
* Forward link: Lab 3 (what counts as causal evidence?), back to Lab 1+2 (what counts as a "good" prediction?)


---
class: left

.blockquote[Part 1 · The DGP]

## Two Paths from the Data-Generating Process

```{r dgp-paths-slide}
#| echo: false
#| out-width: "80%"
#| fig-align: center
knitr::include_graphics(xfun::from_root("img","Session3","DGP-Paths_gemini.svg"))
```

.center[
.small[
Same data, same DGP — two analytical purposes. Path E asks *what is*; Path P asks *what next*.
]
]

???
* Same SVG students see in Lab 1 — recognition cue intentional
* Path E (Explanation/Inference): two sub-questions — E1 "what is the partial effect?" (next slide) and E2 "what would change under intervention?"
* Path P (Prediction): a single goal — generalise to new data
* The DGP is the unobservable urn; data are realisations; methods differ by what they extract from that realisation
* Breiman 2001 was the first to make this dichotomy programmatic ("Two Cultures")


---
class: left

.blockquote[Part 1 · Path E branches]

## Path E has two sub-questions

```{r path-e-branches}
#| echo: false
tibble::tribble(
  ~`Sub-question`,                                       ~Goal,                              ~Identification,                                    ~Lab,
  "E1 — What is the partial effect?",                    "Estimate β under the model",       "by assumption (A/B/C compass)",                    "Lab 1",
  "E2 — What would change under intervention?",          "Estimate a counterfactual",        "by research design (RCT, DiD, IV, RDD, DML)",      "Lab 3"
) %>%
  gt() %>%
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) %>%
  tab_style(style = cell_fill(color = "#fff5f5"),
            locations = cells_body(rows = 1)) %>%
  tab_style(style = cell_fill(color = "#f5e6f0"),
            locations = cells_body(rows = 2)) %>%
  tab_options(table.font.size = "13px",
              table.width = "100%",
              column_labels.font.size = "12px",
              column_labels.background.color = "#f5f5f5")
```


.pull-left[
### E1 — model given
`r fa('arrow-right', fill='#D50006')` Identification settled **before** the data, by specifying the equation.

`r fa('arrow-right', fill='#D50006')` Same path E logic students saw in Sessions 1 + 2.

`r fa('forward', fill='#D50006')` Today: **Lab 1** stress-tests E1.
]

.pull-right[
### E2 — model earns its claim
`r fa('arrow-right', fill='#7d0a52')` Identification comes from the **data structure** — randomization, parallel trends, instrument validity.

`r fa('arrow-right', fill='#7d0a52')` "Design-based identification" (Angrist &amp; Pischke 2010).

`r fa('forward', fill='#7d0a52')` Today: **Lab 3** operationalizes E2.
]

???
* E1 and E2 both live under Path E — they ask "what is" — but they differ in *what counts* as identification
* E1 is the classical econometrics curriculum; E2 is the credibility-revolution addition (Angrist/Pischke 2010)
* Path P (Prediction) sits beside them — it asks "what next?" and uses out-of-sample generalisation as its yardstick
* The labs are mapped: Lab 1 → E1 (and gateway to E2); Lab 2 → P; Lab 3 → E2


---
class: inverse, center, middle

## Part 2: BLUE Bridge

.blockquote[When Assumptions Bite — and what unites the paths]


---
class: left

.blockquote[Part 2 · The A/B/C compass]

## Model Assumptions — the A/B/C compass

```{r abc-compass}
#| echo: false

tibble::tribble(
  ~Group, ~Assumption,                          ~`What we check`,                                                            ~Tool,                                                                ~`What residuals cannot tell us`,
  "A",    "A1 (variable selection)",            "OVB and overfitting are not directly testable",                            "Substantive reasoning, domain knowledge, sensitivity analysis",      "Whether the model is the 'right' one",
  "A",    "A2 (linearity)",                     "Systematic curvature in residuals vs. fitted",                             "Residuals vs. Fitted plot",                                          "—",
  "A",    "A3 (parameter constancy)",           "Influential observations that shift parameters",                           "Cook's distance, Residuals vs. Leverage",                            "Whether parameters are truly constant in the population",
  "B",    "B1 (E[ε] = 0)",                      "Automatically satisfied with intercept",                                    "(No diagnostic needed)",                                             "—",
  "B",    "B2 (homoscedasticity)",              "Non-constant residual spread",                                              "Scale-Location, Residuals vs. Fitted; Breusch-Pagan",                "—",
  "B",    "B3 (no autocorrelation)",            "Residual correlation across time or observations",                          "ACF plot, Durbin-Watson test",                                       "Non-time-based dependence (e.g. clustering)",
  "B",    "B4 (normality of errors)",           "Tail deviations in Q-Q plot",                                               "Q-Q plot, Shapiro-Wilk test",                                        "— (relevant only for small-sample inference)",
  "C",    "C1 (exogeneity)",                    "Cov(X, ε) = 0 — endogeneity not directly observable",                       "Substantive reasoning, instruments, study design",                   "Whether unobserved confounders correlate with X",
  "C",    "C2 (no perfect multicollinearity)",  "Numerical near-singularity of X'X",                                         "VIF, condition number",                                              "—"
) %>%
  gt(groupname_col = "Group", row_group_as_column = TRUE) %>%
  gt::tab_style(
    style = gt::cell_fill(color = "#FBE6E7"),
    locations = gt::cells_body(rows = Group == "A")
  ) %>%
  gt::tab_style(
    style = gt::cell_fill(color = "#F0F0F0"),
    locations = gt::cells_body(rows = Group == "B")
  ) %>%
  gt::tab_style(
    style = gt::cell_fill(color = "#EFEBF4"),
    locations = gt::cells_body(rows = Group == "C")
  ) %>%
  gt::tab_style(
    style = gt::cell_borders(sides = "bottom", color = "#686868", weight = gt::px(2)),
    locations = gt::cells_body(rows = Assumption %in% c("A3 (parameter constancy)", "B4 (normality of errors)"))
  ) %>%
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_row_groups()
  ) %>%
  gt::tab_style(
    style = gt::cell_fill(color = "#F5F0FA"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::tab_options(table.font.size = "11px",
                  data_row.padding = px(2))
```

.quelle[.small[`r Citet(bib, "vonAuer_2023")` · Lab 1 Step 2]]

???
* Identical table to Lab 1 — students see it again for recognition
* Two empirical groups: **A** (specification) and **C1** (exogeneity) require substantive reasoning, not residual checks
* **A2, A3, B2, B3, B4, C2** can be checked from residuals; **A1, C1** cannot
* This is the diagnostic compass — what violation does which path-specific damage is on the next slide


---
class: left

.blockquote[Part 2 · BLUE]

## What BLUE is — and what violations mean for each path

```{r blue-consequences}
#| echo: false

tibble::tribble(
  ~`Violation type`,         ~Examples,         ~`Effect on β̂`,                            ~`Path E (inference)`,                                                       ~`Path P (prediction)`,
  "Bias violations",         "A1, A2, A3, C1",  "Biased and inconsistent",                  "β̂ estimates the wrong parameter — CIs and tests are meaningless",          "ŷ systematically wrong, but may still predict adequately if relationships stay stable (Lucas caveat)",
  "Efficiency violations",   "B2, B3, C2",      "Unbiased, but inefficient",                "SE incorrect → CIs and tests invalid; robust / HAC SE can repair",          "Point predictions remain unbiased; prediction is comparatively robust",
  "Distribution violations", "B4",              "Still BLUE",                               "Exact small-sample tests invalid; large n → CLT keeps tests valid",         "Unaffected for point prediction"
) %>%
  gt() %>%
  gt::tab_style(
    style = gt::cell_fill(color = "#FBE6E7"),
    locations = gt::cells_body(rows = 1)
  ) %>%
  gt::tab_style(
    style = gt::cell_fill(color = "#F0F0F0"),
    locations = gt::cells_body(rows = 2)
  ) %>%
  gt::tab_style(
    style = gt::cell_fill(color = "#EFEBF4"),
    locations = gt::cells_body(rows = 3)
  ) %>%
  gt::tab_style(
    style = gt::cell_fill(color = "#F5F0FA"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::tab_options(table.font.size = "11px",
                  data_row.padding = px(3))
```

.quelle[.small[Lab 1 Step 2 (BLUE bridge box)]]

???
* Identical structure to Lab 1's BLUE-bridge background box — students see the same logic in slides and lab
* **Bias violations** (A1, A2, A3, C1): break Path E entirely; Path P survives only if the DGP stays stable (Lucas caveat)
* **Efficiency violations** (B2, B3, C2): SE wrong for Path E (fix with robust / HAC SE); Path P unaffected
* **Distribution violations** (B4): no consequences for BLUE; only matters for small-n exact inference
* This is the diagnostic logic that lets the translator decide which path is still defensible under which violation


---
class: left

.blockquote[Part 2 · The bridge]

## BLUE as the bridge between paths

```{r blue-bridge-svg}
#| echo: false
#| out-width: "100%"
#| fig-align: center
knitr::include_graphics(xfun::from_root("img","Session3","blue_bridge.svg"))
```

???
* BLUE is the common starting point both paths share — under assumptions A/B/C, OLS is the unbiased benchmark
* Path P branches *upward*: deliberately accept bias for variance reduction (the "bias-variance trade-off", Hastie/Tibshirani/Friedman 2009). The toolkit is Ridge / Lasso / Elastic Net + cross-validation. Lab 2 operates here.
* Path E branches *downward*: preserve unbiasedness, and add design-based identification (Angrist & Pischke 2010). The toolkit is robust SE, RCT, DiD, IV, DML. Lab 1 & Lab 3 operate here.
* The Spiess (2025) point at the bottom: preserving unbiasedness is also an *institutional* discipline — it prevents the researcher from tuning the analysis toward a favoured result. Pre-commitment makes the estimate credible.
* This is the punchline of Part 2 — every method in today's labs sits on one of these two branches


---
class: left

.blockquote[Narrative anchor → Lab 1]

## `r fa('envelope', fill='#D50006')` Memo from Sarah Chen → Lab 1

.small[
> *Two distinct questions this quarter: **(1)** Does **feature advertising** cause incremental sales? → Path **E** &nbsp;·&nbsp; **(2)** I need an **inventory-planning tool** that forecasts weekly sales. → Path **P**. Answers must survive an audit by my CEO. — Sarah Chen, VP Marketing, Dominick's Finer Foods*
]

<div style="border:3px solid #D50006; border-radius:8px; padding:14px 24px;
            background:#fff5f5; margin-top:0.8em;">
<div style="font-family:sans-serif; font-size:15px; font-weight:bold;
            color:#D50006; margin-bottom:10px;">
`r fa('flask')` Lab&nbsp;1: Reality Check — Stress-testing OLS under A/B/C
</div>
<div style="font-size:0.82em; line-height:1.7;">
<strong>Learning objectives:</strong> Fit a baseline log-log OLS on Dominick's Store&nbsp;114 &middot; Apply A/B/C diagnostics (autoplot, ACF, Durbin-Watson, VIF) &middot; Choose robust&nbsp;SE / HAC when B2/B3 are violated &middot; Trace OVB across 83 stores &middot; Draft Memo&nbsp;T4 for Sarah's Q1
<br>
<strong>Dataset:</strong>
<code>tbl_stores</code> (Dominick's, 83 stores × 3 brands; <code>tbl_s114</code> for Store&nbsp;114)
<br>
`r fa('clock')` <strong>45 minutes</strong> &nbsp;·&nbsp;
`r fa('users')` <strong>Individual work</strong>
<br>
`r fa('circle-right')` <strong>Open lab document:</strong>
<code>Session3_Lab1.qmd</code>
</div>
</div>

.center[.small[Three translator memos as deliverables: **T4** (Lab 1) · **T7** (Lab 2) · **T<sub>final</sub>** (Lab 3).]]

???
* The memo block is now a short teaser — the full memo lives in the lab .qmd, where students read it before Step 0
* Q1 maps to Path E (Lab 1 diagnoses, Lab 3 answers); Q2 maps to Path P (Lab 2 builds the tool)
* Lab-1 finale: classical OLS on historical data cannot answer Q1 even when fit perfectly — motivates Lab 3
* Open the .qmd and walk through Step 0 together; then individual work begins


---
class: inverse, center, middle

## Lab 1

.blockquote[Reality Check — stress-testing OLS]
