<!-- ============================================================
INSTRUCTOR CUSTOMISATION — update before rendering for each cohort
Line references are approximate; search for the marker text.

  ~Line 18    Moodle download link for dominicks_full.rds
  ~Line 19    Moodle download link for dominicks_demo.rds
  ~Line 65    Folder name for current cohort (default: "ER114_FS26")

============================================================ -->
---
title: "Session 3 · Lab 1: BLUE Bridge — Predict or Explain?"
subtitle: "CRISP-DM Phases 5–6 --- Evaluation & Deployment"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en-US
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
bibliography: ../../lit/my_bib.bib
---

```{r setup}
#| include: false
library(tidyverse)        # data wrangling and plotting
library(gt)               # publication-quality tables
library(broom)            # tidy model output
library(patchwork)        # combine ggplots
library(car)              # VIF for multicollinearity diagnostics

# shared visual identity
theme_set(theme_minimal(base_size = 11))

# colour palette
col_primary   <- "#D50006"   # main highlight
col_secondary <- "#502479"   # second highlight
col_neutral   <- "#686868"   # grey reference

# load data (provided as .rds files in data/raw/)
tbl_full <- readRDS(xfun::from_root("data", "raw", "dominicks_full.rds"))
tbl_demo <- readRDS(xfun::from_root("data", "raw", "dominicks_demo.rds"))

# convenient aliases
tbl <- tbl_full
```

## Scenario

You work as a data analyst for **Dominick's Finer Foods**, a Chicago
supermarket chain. The marketing department has asked for your help on
two upcoming decisions. The two questions look similar on the surface
— but they require very different methodological approaches.

This lab is the **bridge** between two paths of modern data analysis:
the BLUE/inference world that you know from Sessions 1 and 2, and the
out-of-sample world of predictive modelling.

```{r crisp-overview}
#| echo: false

tibble::tribble(
  ~Phase, ~Name,                   ~Content,                                                ~Status,
  "1",    "Business Understanding","Marketing decisions: predict and intervene",             "Done",
  "2",    "Data Understanding",    "OJ scanner data: 83 stores × 3 brands × 112 weeks",      "Done (S1)",
  "3",    "Data Preparation",      "Wrangled, ready",                                        "Done (S1)",
  "4",    "Modelling",             "Linear regression with diagnostics (← here)",            "← This lab (1a)",
  "5",    "Evaluation",            "Out-of-sample evaluation, K-fold CV (← here)",           "← This lab (3)",
  "6",    "Deployment",            "Translator memos: what can we deliver?",                 "← This lab (4)"
) |>
  gt() |>
  gt::tab_style(gt::cell_fill("#f0fff0"), gt::cells_body(rows = Phase %in% c("4","5","6"))) |>
  gt::tab_style(gt::cell_fill("#f5f0fa"), gt::cells_column_labels()) |>
  gt::tab_options(table.font.size = "13px")
```

## Navigation

| Symbol | Box | Meaning | When to read |
|--------|-----|---------|--------------|
| ⚠️ | Red | Action required: mandatory tasks for the live session | Live |
| 💡 | Green | Tip / best practice: coding hints, shortcuts | Live |
| 📚 | Blue (collapsed) | Background theory: concepts, derivations | At home |
| ℹ️ | Blue (open) | Short clarification | Live |

---

# Phase 0 — Translator Briefing

*The icebreaker for this lab is the translator's task itself. The
business request opens up the two methodological paths that the lab
will pursue.*

## The Memo from Marketing

:::{.callout-note icon=false collapse="false"}
## 📧 Memo from Sarah Chen, VP Marketing — Dominick's Finer Foods

*"We have found in the previous sessions that advertising (`feat`)
increases our sales by about 25 % — measured in Store 51, replicated
in Store 50. Impressive.*

*Now we need the model for two decisions:*

***1.*** *We want a tool that **predicts the weekly OJ sales for each
of our 83 stores** in Chicago. The tool will be used for inventory
planning.*

***2.*** *We are considering launching a **major advertising campaign**
for the Dominick's house brand in Q4. Can we derive from the historical
data whether this will pay off?*

*How would you approach this?"*

---

— Sarah
:::

## Your task as Translator

:::{.callout-important icon=false}
## ⚠️ T1: Two memos to Sarah

Your task is to write **two short memos back to Sarah at the end of
this lab** — one per question. For each question:

1. Translate Sarah's business question into a methodological operation.
2. State what you can deliver and what you cannot.
3. Reference what you would need additionally if anything is missing.

**A hint on order:** The two questions look similar, but they require
different methods. We work through **Question 2 (campaign effect) first**,
because it builds on what you already know from Sessions 1 and 2
(regression with inference). **Question 1 (prediction tool) comes after**,
because it requires a different optimality criterion that we will
develop along the way.
:::

## What we already know from Sessions 1 and 2

Below is the OJ panel for **all 83 stores** of the top three brands
(Tropicana, Minute Maid, Dominick's). This is a substantial enlargement
of the data world we worked with so far.

```{r data-summary}
#| echo: false

tbl |>
  summarise(
    `# stores`       = n_distinct(store),
    `# brands`       = n_distinct(brand),
    `# weeks`        = n_distinct(week),
    `# observations` = n()
  ) |>
  gt() |>
  gt::tab_options(table.font.size = "13px")
```

The advertising effect we found in Session 1, replicated in Session 2,
is the baseline that Sarah refers to in her memo.

---

# Phase 1a — Question 2 (Explaining the Advertising Effect): Classical Inference

> **🔄 Modularity note for instructor:** This sub-phase (1a) is
> structured as a stand-alone block. In the summer iteration (HS26),
> this entire block is *cut out and moved to Session 2 Lab 2 (Multiple
> Regression with Diagnostics)*. In the current semester (FS26), we
> build up the diagnostics here, because Session 2 has not yet covered
> them.

We start with the question that is closest to what you know from
Sessions 1 and 2: a multiple regression with inference. The translator's
job is to give Sarah a quantitative answer on whether advertising
*causes* sales to rise.

## Naive Model — All 83 Stores

Let us fit the natural starting model: weekly log-sales explained by
advertising, price, and brand, across all 83 stores.

```{r naive-model}
#| code-fold: false
mod_naive <- lm(logmove ~ feat + log(price) + brand, data = tbl)

mod_naive |>
  tidy(conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  gt() |>
  gt::tab_options(table.font.size = "13px")
```

The advertising coefficient is around `r round(coef(mod_naive)["feat"], 2)`.
That looks like Sarah's expectation: roughly a 25 % uplift from
advertising — confirmed across 83 stores.

But before we send Sarah a recommendation, we have to check whether the
model assumptions hold. The whole inferential machinery rests on them.

## A-B-C Assumptions (von Auer) — Diagnostic Checks

The Gauss-Markov framework relies on a set of assumptions
[@vonAuer_2023, ch. 4]. Let us check the ones that can be diagnosed
empirically.

:::{.callout-note collapse="true" #abc-assumptions}
## 📚 Background: The A-B-C Assumptions in Brief

**A-Assumptions (the model):**

- A1 (variable selection): No relevant variable is missing, no
  irrelevant one is included.
- A2 (linearity): The true relationship is linear in the parameters.
- A3 (parameter constancy): Parameters are constant across all
  $N$ observations.

**B-Assumptions (the error term):**

- B1: $E(\varepsilon_i) = 0$.
- B2 (homoscedasticity): $\text{Var}(\varepsilon_i) = \sigma^2$.
- B3 (no autocorrelation): $\text{Cov}(\varepsilon_i, \varepsilon_j) = 0$
  for $i \neq j$.
- B4 (normality): $\varepsilon_i \sim N$.

**C-Assumptions (the data):**

- C1 (exogeneity): $\text{Cov}(X_i, \varepsilon_i) = 0$.
- C2 (no perfect multicollinearity).

In this lab, we focus on the diagnostic checks: VIF for C2, residual
plots for the B-assumptions, and substantive reasoning for A1 and C1.

What the *consequences* of a violation are — that is the topic of
Session 3's BLUE bridge. For now, we just want to know **whether** the
assumptions hold.
:::

### C2: Multicollinearity (VIF)

```{r vif-naive}
#| code-fold: false
vif_naive <- car::vif(mod_naive)
vif_naive
```

VIF values below 5 are unproblematic; values between 5 and 10 warrant
attention; values above 10 indicate severe multicollinearity. In the
naive model, the maximum VIF is `r round(max(vif_naive), 2)` —
unproblematic.

### B-Assumptions: Residual Plots

```{r residual-plots}
#| fig-height: 6
#| fig-width: 9
par(mfrow = c(2, 2))
plot(mod_naive)
par(mfrow = c(1, 1))
```

What do these four plots tell us?

- **Residuals vs Fitted (top-left):** Looks for non-linearity (A2)
  and heteroscedasticity (B2). A horizontal red line with even spread
  is good.
- **Q-Q Residuals (top-right):** Tests normality (B4). Points on the
  diagonal mean Gaussian errors.
- **Scale-Location (bottom-left):** Specifically for heteroscedasticity
  (B2). A flat red line means equal variance.
- **Residuals vs Leverage (bottom-right):** Identifies influential
  outliers.

:::{.callout-tip}
## 💡 Reading residual plots

Don't worry if the diagnostics are not "perfect". With many thousand
observations from real scanner data, perfect Gaussian residuals are
not expected. The question is **whether the violations are severe
enough to invalidate inference** — and the *consequences* of violations
are exactly what Session 3's BLUE bridge will address.

In Session 2, you learned **how to check** — here in Session 3, you
learn **what it means**.
:::

:::{.callout-important icon=false}
## ⚠️ T2: Diagnostic readout

Looking at the four residual plots above:

a. Is there visible heteroscedasticity? Where in the fitted-value
   range is variance largest?
b. Do the Q-Q residuals follow the diagonal — or are there
   substantial deviations at the tails?
c. Are there leverage points that stand out?

Write down a one-sentence verdict per assumption (B2, B4, leverage).
:::


## A1: The Stumbling Block — feat × deal Entanglement

So far, the diagnostics suggest a usable inferential model. The big
remaining question is **A1**: are all relevant variables in the model?

Here we hit a fundamental issue with the Dominick's data. Advertising
(`feat`) and price-promotion (`deal`) are not independent — they are
*structurally entangled*.

```{r feat-deal-table}
#| code-fold: false
tbl |>
  count(feat = round(feat), deal) |>
  pivot_wider(names_from = deal, values_from = n, names_prefix = "deal=") |>
  gt() |>
  gt::tab_header(title = "Joint distribution of feat and deal") |>
  gt::tab_options(table.font.size = "13px")
```

Whenever there is advertising (`feat = 1`), there is *always* a price
promotion (`deal = 1`). The implication is **perfect** in the data.

:::{.callout-warning icon=false}
## ⚠️ The A1 violation

Our naive model includes `feat` but not `deal`. If `feat = 1` always
co-occurs with `deal = 1`, then the estimated `feat` coefficient
absorbs **both** effects: the advertising effect *and* the deal effect.
This is the classic **omitted variable bias** (A1 violation).

Through the domino effect, A1 violation produces a C1 violation
(endogeneity): the omitted `deal` is now in the error term, and it is
correlated with `feat`.
:::

### What happens when we add `deal`?

```{r model-with-deal}
#| code-fold: false
mod_full <- lm(logmove ~ feat + deal + log(price) + brand, data = tbl)

mod_full |>
  tidy(conf.int = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  gt() |>
  gt::tab_options(table.font.size = "13px")
```

Comparing the two models:

```{r coef-comparison}
#| code-fold: true
#| code-summary: "Show code: coefficient comparison table"
bind_rows(
  tidy(mod_naive) |> mutate(model = "Naive (without deal)"),
  tidy(mod_full)  |> mutate(model = "Full (with deal)")
) |>
  filter(term %in% c("feat", "deal")) |>
  select(model, term, estimate) |>
  pivot_wider(names_from = term, values_from = estimate) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  gt() |>
  gt::tab_options(table.font.size = "13px")
```

The `feat` coefficient has changed substantially:
from `r round(coef(mod_naive)["feat"], 3)` (naive)
to `r round(coef(mod_full)["feat"], 3)` (with deal).

:::{.callout-important icon=false}
## ⚠️ T3: Translator memo no. 1 (Question 2 — campaign effect)

Based on what you have just seen, draft the first memo to Sarah:

- Can we tell her, based on the historical data, what an advertising
  campaign *causes* in terms of sales?
- What is the methodological obstacle?
- What would be needed to answer her question properly?

Keep it short: 3-5 sentences in business language. Bonus points if
you mention what kind of study design would resolve the problem
(this is a forward reference to Lab 3).
:::

---

# Phase 2 — The BLUE Bridge Becomes Operative

We have hit a wall on Question 2. The classical inference machinery
delivers a coefficient — but we cannot interpret it causally because
the data foundation does not support it (A1 violated → C1 violated).

But Sarah's **Question 1** is different. She wants a *prediction tool*
for inventory planning. The question is not: *"What would advertising
cause?"* It is: *"What will sales be, given conditions we observe?"*

For prediction, the BLUE criterion is not what we need.

:::{.callout-note collapse="true" #blue-bridge}
## 📚 Background: BLUE for inference vs. prediction

Under the Gauss-Markov assumptions, OLS is the **B**est **L**inear
**U**nbiased **E**stimator. Note the order: *first* unbiased, *then*
minimum variance **among the unbiased**.

**For inference (explanation),** this is the gold standard. The
coefficient should reflect the true effect — unbiasedness is
essential.

**For prediction,** the criterion is different. What we want is
*minimum total error on new data*:

$$\text{MSE}(\hat{Y}) = \text{Bias}^2(\hat{Y}) + \text{Var}(\hat{Y}) + \sigma_{\varepsilon}^2$$

In prediction, accepting *some* bias for *much less variance* can
lower the total MSE. This is the rationale for regularised methods
like Lasso and Ridge — they are *not* BLUE by design.

→ The same theoretical foundation explains both paths, with opposite
optimality criteria. This is the **BLUE bridge**.
:::

This conceptual shift sets up the rest of the lab. We assume the
business-relevant relationships are **stable** for inventory planning
purposes (this is a working assumption, not a guarantee — Lucas
critique applies if the system is later perturbed). Given that
assumption, we now ask: how well does the model **generalise to stores
we have not seen**?

---

# Phase 3 — Question 1 (Prediction): Out-of-Sample Evaluation

## Step 1 — Train on Store 51, predict on Store 50

We start the way Session 2 left us: with a single-store model trained
on Store 51 and evaluated on Store 50.

```{r train-51-predict-50}
#| code-fold: false
mod_51 <- lm(logmove ~ feat + log(price) + brand,
             data = tbl |> filter(store == 51))

tbl_50 <- tbl |> filter(store == 50)
pred_50 <- predict(mod_51, newdata = tbl_50)

rmse_50 <- sqrt(mean((pred_50 - tbl_50$logmove)^2))
rmse_50_rounded <- round(rmse_50, 3)
```

The RMSE on Store 50 is `r rmse_50_rounded`. Given that `logmove`
values range from roughly 3 to 12, this is a useful prediction. Not
surprising — Stores 50 and 51 have similar demographics, as we
chose Store 50 in Session 2 deliberately for replication.

## Step 2 — Predict on stores with different demographics

What happens when we apply the same Store-51-trained model to stores
that are *demographically different*?

```{r predict-all-other-stores}
#| code-fold: true
#| code-summary: "Show code: store-level RMSE"
other_stores <- setdiff(unique(tbl$store), 51)

store_rmses <- tibble(store = other_stores) |>
  rowwise() |>
  mutate(
    rmse = {
      d <- tbl |> filter(store == !!store)
      sqrt(mean((predict(mod_51, newdata = d) - d$logmove)^2))
    }
  ) |>
  ungroup() |>
  arrange(desc(rmse))

# Summary statistics
rmse_summary <- store_rmses |>
  summarise(
    min_rmse    = round(min(rmse),    3),
    median_rmse = round(median(rmse), 3),
    max_rmse    = round(max(rmse),    3),
    n_stores    = n()
  )

rmse_summary |>
  gt() |>
  gt::tab_options(table.font.size = "13px")
```

The RMSE varies from `r rmse_summary$min_rmse` to
`r rmse_summary$max_rmse` across `r rmse_summary$n_stores` stores.
The model trained on Store 51 fits Store 50 well — but for other
stores, the prediction error can be **substantially larger**.

```{r rmse-distribution-plot}
#| code-fold: true
#| code-summary: "Show code: RMSE distribution"
#| fig-height: 4
#| fig-width: 8
ggplot(store_rmses, aes(x = rmse)) +
  geom_histogram(bins = 25, fill = col_primary, alpha = 0.8, colour = "white") +
  geom_vline(xintercept = rmse_50, linetype = "dashed",
             colour = col_secondary, linewidth = 1) +
  annotate("text", x = rmse_50, y = Inf,
           label = paste0("Store 50: RMSE = ", rmse_50_rounded),
           hjust = -0.1, vjust = 2, colour = col_secondary) +
  labs(
    title    = "Out-of-sample RMSE when applying the Store-51 model to other stores",
    subtitle = "The model generalises poorly to stores with different demographics",
    x        = "RMSE on out-of-sample store",
    y        = "Number of stores"
  )
```

:::{.callout-warning icon=false}
## 💥 The aha-moment

We trained on Store 51, replicated on Store 50, and might have been
tempted to conclude *"the model works"*. But out-of-sample evaluation
on **all** other 81 stores shows that the model's predictive
performance varies dramatically. **A single replication is not
enough** for a prediction tool that should work in any of Sarah's
83 stores.
:::

## Step 3 — Train/test split across stores

Instead of training on one store and predicting on others, let us
**split the set of stores** itself: 80 % of stores for training,
20 % for testing.

```{r train-test-split}
#| code-fold: false
set.seed(42)
stores <- unique(tbl$store)
test_stores <- sample(stores, size = floor(0.20 * length(stores)))

tbl_train <- tbl |> filter(!store %in% test_stores)
tbl_test  <- tbl |> filter(store  %in% test_stores)

mod_train <- lm(logmove ~ feat + log(price) + brand, data = tbl_train)
pred_test <- predict(mod_train, newdata = tbl_test)

rmse_test <- sqrt(mean((pred_test - tbl_test$logmove)^2))
rmse_train <- sqrt(mean((mod_train$residuals)^2))
```

| Metric | Value |
|---|---|
| In-sample RMSE (training stores) | `r round(rmse_train, 3)` |
| Out-of-sample RMSE (test stores) | `r round(rmse_test, 3)` |

:::{.callout-tip}
## 💡 Splitting across stores, not across weeks

We split the **set of stores**, not the set of weeks. Splitting at the
week level would mean that the model has seen the same store before
— and predictive performance would be artificially inflated. For a
prediction tool that should work *in new stores*, we need
store-level splitting.

In time-series settings, this concept extends to **temporal CV** to
avoid future-leakage. We will revisit this in ER017.
:::

## Step 4 — K-fold cross-validation

A single 80/20 split depends on which 20 % we happened to pick.
**K-fold CV** averages over all possible splits:

```{r kfold-cv}
#| code-fold: false
set.seed(42)
K <- 5
stores_shuffled <- sample(unique(tbl$store))
folds <- cut(seq_along(stores_shuffled), breaks = K, labels = FALSE)

cv_rmses <- map_dbl(1:K, \(k) {
  test_st  <- stores_shuffled[folds == k]
  train_st <- stores_shuffled[folds != k]

  mod_k <- lm(logmove ~ feat + log(price) + brand,
              data = tbl |> filter(store %in% train_st))
  pred_k <- predict(mod_k,
                    newdata = tbl |> filter(store %in% test_st))
  actual_k <- tbl |> filter(store %in% test_st) |> pull(logmove)
  sqrt(mean((pred_k - actual_k)^2))
})

cv_score <- mean(cv_rmses)
```

The 5-fold CV scores per fold:

```{r cv-scores-table}
#| echo: false
tibble(Fold = 1:K, RMSE = round(cv_rmses, 3)) |>
  gt() |>
  gt::tab_options(table.font.size = "13px")
```

**Mean CV-RMSE: `r round(cv_score, 3)`** — this is our realistic
out-of-sample performance estimate for the simple model with
`feat + log(price) + brand`.

:::{.callout-important icon=false}
## ⚠️ T4: Translator memo no. 2 (Question 1 — prediction tool)

Draft the second memo to Sarah:

- Can we deliver an inventory planning tool that works across all
  83 stores?
- What is the realistic out-of-sample accuracy?
- What variables are *not yet* in the model that you would add to
  improve it? (Hint: think about what makes Store 50 different from
  another store far away in Chicago.)
:::

---

# Phase 4 — Synthesis

```{r synthesis-table}
#| echo: false
tibble::tribble(
  ~`Question`,                            ~`Methodological path`,                  ~`Verdict`,
  "1. Inventory prediction tool",         "Path A: out-of-sample, CV",             "Deliverable with caveats",
  "2. Causal effect of a Q4 campaign",    "Path B: causal identification",         "Not answerable from these data"
) |>
  gt() |>
  gt::tab_style(gt::cell_fill("#f0fff0"), gt::cells_body(rows = 1)) |>
  gt::tab_style(gt::cell_fill("#fff8e1"), gt::cells_body(rows = 2)) |>
  gt::tab_style(gt::cell_fill("#f5f0fa"), gt::cells_column_labels()) |>
  gt::tab_options(table.font.size = "13px")
```

The same dataset gave us two different methodological operations and
two different verdicts. **This is the BLUE bridge in action:**

- Path A (prediction): we accept that we cannot disentangle advertising
  from price promotion, because for *forecasting under stable
  structure*, that does not matter. The relationship is consistent
  in the data, and the model predicts.

- Path B (explanation/intervention): we cannot deliver a causal claim,
  because the data foundation does not support identification. To
  answer Sarah's Q2 properly, we need a different study design — an
  A/B test.

This is the **central lesson of the lab**: the optimality criterion
depends on the question. Same data, two operations, two answers.

:::{.callout-note collapse="true" #lucas-critique}
## 📚 Background: Lucas critique

The Path A prediction relies on a working assumption: the structure
of the relationships does not change. If management *actively
intervenes* (e.g. launches a big campaign), this assumption may break
— a phenomenon known as the **Lucas critique** [@lucas_1976]. The
historical correlations were generated under one regime; the new
regime may behave differently.

This is the deeper reason why Sarah's Q2 is fundamentally different
from Q1: it asks about a *change* to the system, not about predicting
within it. Path A is appropriate when the system stays as it is. Path
B is required when we change the system.
:::

---

# Bonus — For independent study

The next lab (Demo: Regularisation on Demographics) will refine the
prediction tool by adding the 12 demographic variables. As preparation,
you can already explore:

:::{.callout-important icon=false collapse="true"}
## ⚠️ B1 (Bonus): Lasso preview on demographics

Add the 12 demographic variables from `tbl_demo` (joined by `store`)
to the model. Fit a Lasso regression with cross-validation
(`glmnet::cv.glmnet`). Which variables survive at the optimal
$\lambda$? How does the cross-validated RMSE compare to the simple
model we built in Phase 3?

```r
library(glmnet)
tbl_joined <- tbl |>
  left_join(tbl_demo, by = "store")
# build the design matrix, then:
# cv_fit <- cv.glmnet(X, y, alpha = 1)
```

This is exactly the question we will work through together in the
upcoming Demo. Coming with your own attempt makes the Demo more useful.
:::

---

# What comes next

| Lab | Question it answers |
|---|---|
| **Demo: Regularisation on Demographics** | Can we improve the prediction tool by adding demographic variables — and which of them matter? |
| **Lab 3: A/B-Test Design** | How would we set up a study that answers Sarah's Question 2 properly? |

