<!-- ============================================================
INSTRUCTOR CUSTOMISATION — update before rendering for each cohort
Line references are approximate; search for the marker text.

  ~Line 18   Moodle download link for dominicks_full.rds
  ~Line 19   Moodle download link for dominicks_demo.rds

============================================================ -->
---
title: "Session 3 · Lab 1: BLUE Bridge — Explain or Predict?"
subtitle: "CRISP-DM Phases 4–6 --- Modelling, Evaluation, Deployment"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en-US
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
bibliography: ../../lit/my_bib.bib
---

```{r setup}
#| include: false
library(tidyverse)        # data wrangling and plotting
library(gt)               # publication-quality tables
library(broom)            # tidy model output
library(patchwork)        # combine ggplots
library(car)              # VIF for multicollinearity diagnostics

# shared visual identity
theme_set(theme_minimal(base_size = 11))

# colour palette
col_primary   <- "#D50006"   # main highlight (Store 114)
col_secondary <- "#502479"   # contrast (Store 83)
col_neutral   <- "#686868"   # grey reference

# load data — both files come from data/raw/
tbl_full <- read_rds(xfun::from_root("data", "raw", "dominicks_full.rds"))
tbl_demo <- read_rds(xfun::from_root("data", "raw", "dominicks_demo.rds"))

# convenience subsets — filtered on-the-fly from the full panel
tbl_114 <- tbl_full %>% filter(store == 114)
tbl_83  <- tbl_full %>% filter(store == 83)
```

## Scenario

You work as a data analyst for **Dominick's Finer Foods**, a Chicago
supermarket chain. The marketing department has asked for your help on
two upcoming decisions. The questions look similar on the surface — but
they require very different methodological approaches. This lab is the
**bridge** between two paths of modern data analysis: the inferential
world you know from Sessions 1 and 2, and the out-of-sample world of
predictive modelling.

```{r crisp-overview}
#| echo: false

tibble::tribble(
  ~Phase, ~Name,                      ~Content,                                                  ~Status,
  "1",    "Business Understanding",   "Marketing decisions: explain and predict",                "Done",
  "2",    "Data Understanding",       "OJ scanner data: 83 stores × 3 brands × 121 weeks",       "Done (S1)",
  "3",    "Data Preparation",         "Wrangled, ready",                                          "Done (S1)",
  "4",    "Modelling",                "Linear regression with diagnostics (← here)",              "← This lab (1a/b)",
  "5",    "Evaluation",               "Out-of-sample evaluation, replication, K-fold CV (← here)","← This lab (3)",
  "6",    "Deployment",               "Translator memos: what can we deliver?",                   "← This lab (4)"
) %>%
  gt() %>%
  gt::tab_style(gt::cell_fill("#f0fff0"),
                gt::cells_body(rows = Phase %in% c("4","5","6"))) %>%
  gt::tab_style(gt::cell_fill("#f5f0fa"), gt::cells_column_labels()) %>%
  gt::tab_options(table.font.size = "13px")
```

## Navigation

| Symbol | Box | Meaning | When to read |
|--------|-----|---------|--------------|
| ⚠️ | Red | Action required: mandatory tasks for the live session | Live |
| 💡 | Green | Tip / best practice: coding hints, shortcuts | Live |
| 📚 | Blue (collapsed) | Background theory: concepts, derivations | At home |
| ℹ️ | Blue (open) | Short clarification | Live |

---

# Step 0 — Translator Briefing

*The icebreaker for this lab is the translator's task itself. The
business request opens up the two methodological paths that the lab
will pursue.*

## The Memo from Marketing

:::{.callout-note icon=false collapse="false"}
## 📧 Memo from Sarah Chen, VP Marketing — Dominick's Finer Foods

*"Team, two decisions are pending and I need your help.*

***1. Q4 advertising campaign for the Dominick's house brand.***
*Marketing wants to invest in a feature campaign. Before signing off,
the executive board wants to know: **what is the effect of advertising
on sales?** We have years of weekly data. Surely we can answer this
from the historical record?*

***2. Inventory planning tool for all 83 stores.***
*Operations is building a forecasting tool to set weekly OJ stock
across our 83 stores. Can you give them **a prediction model** that
works across the whole chain?*

*How would you approach these?"*

---

— Sarah
:::

## Your task as Translator

:::{.callout-important icon=false}
## ⚠️ T1: Two memos to Sarah

By the end of this lab you will write two short memos to Sarah — one
per question.

For each question:

1. Translate the business question into a methodological operation.
2. State what you can deliver, what you cannot, and why.
3. Reference what additional data or studies would be needed.

**A note on order.** We work through **Question 1 (the advertising
effect)** first, because it builds on what you already know from
Sessions 1 and 2 (regression with inference). **Question 2 (the
prediction tool)** comes second, because it requires a different
optimality criterion — the **BLUE bridge** between the two paths.

Keep a notepad open. We will collect findings step by step.
:::

## What is in the data

Below is the OJ panel for **all 83 stores** of the top three brands
(Tropicana, Minute Maid, Dominick's).

```{r data-summary}
#| echo: false

tbl_full %>%
  summarise(
    `# stores`       = n_distinct(store),
    `# brands`       = n_distinct(brand),
    `# weeks`        = n_distinct(week),
    `# observations` = n()
  ) %>%
  gt() %>%
  gt::tab_options(table.font.size = "13px")
```

We start with a single store — **Store 114** — to keep the inferential
framing manageable. Later in the lab we will scale up to all 83 stores.

```{r store114-summary}
#| echo: false

tbl_114 %>%
  summarise(
    Store              = first(store),
    `# weeks`          = n_distinct(week),
    `# observations`   = n(),
    `weeks min/max`    = paste(min(week), "/", max(week)),
    `missing weeks`    = max(week) - min(week) + 1 - n_distinct(week)
  ) %>%
  gt() %>%
  gt::tab_options(table.font.size = "13px")
```

---

# Step 1 — Question 1 (The Advertising Effect): Classical Inference

We start with the question that is closest to what you know from
Sessions 1 and 2: a multiple regression with inference. The translator's
job is to give Sarah a quantitative answer on whether advertising
*causes* sales to rise.

::: {.callout-note icon=false collapse="true" #s2-bridge}
## 📚 Background: Where this connects to Session 2

In Session 2 you learned to fit models with `infer::fit()`, build
bootstrap confidence intervals for coefficients, and run regression
diagnostics on the A–B–C assumptions (Auer's notation).

In this lab we work with `glm()` / `lm()` and `broom::tidy()` for
simplicity, but you can run the same models through the infer pipe
if you prefer. The diagnostic question is identical: **do the
Gauss-Markov assumptions hold?** What changes in Session 3 is the
*meaning* of a violation — we will return to that in Step 2.
:::

## Step 1a — A naïve model on Store 114

Let us fit the natural starting model: weekly log-sales explained by
advertising, price, and brand. We treat Store 114 as our analytical
sample for now.

```{r mod-naive-114}
#| code-fold: false
mod_naive_114 <- lm(logmove ~ feat + log(price) + brand, data = tbl_114)

mod_naive_114 %>%
  tidy(conf.int = TRUE) %>%
  mutate(across(where(is.numeric), \(x) round(x, 3))) %>%
  gt() %>%
  gt::tab_options(table.font.size = "13px")
```

The advertising coefficient is
**`r round(coef(mod_naive_114)["feat"], 3)`**. Converted to the
multiplicative scale, this means feature advertising is associated
with sales of about
**`r round((exp(coef(mod_naive_114)["feat"]) - 1) * 100, 0)` %** above
baseline. That sounds like an answer for Sarah — but before we recommend
anything, we must check the model's assumptions.

## The A–B–C assumptions (von Auer) — diagnostic checks

The Gauss-Markov framework relies on a set of assumptions
[@vonAuer_2023]. Let us check the ones that can be diagnosed
empirically.

:::{.callout-note collapse="true" #abc-assumptions}
## 📚 Background: The A–B–C assumptions in brief

**A-Assumptions (the model):**

- A1 (variable selection): no relevant variable is missing, no
  irrelevant one is included
- A2 (linearity): the true relationship is linear in the parameters
- A3 (parameter constancy): parameters are constant across all
  $N$ observations

**B-Assumptions (the error term):**

- B1: $E(\varepsilon_i) = 0$
- B2 (homoscedasticity): $\text{Var}(\varepsilon_i) = \sigma^2$
- B3 (no autocorrelation): $\text{Cov}(\varepsilon_i, \varepsilon_j) = 0$
  for $i \neq j$
- B4 (normality): $\varepsilon_i \sim N$

**C-Assumptions (the data):**

- C1 (exogeneity): $\text{Cov}(X_i, \varepsilon_i) = 0$
- C2 (no perfect multicollinearity)

In this lab we focus on the diagnostic checks: VIF for C2, residual
plots for the B-assumptions, and substantive reasoning for A1 and C1.

What the *consequences* of a violation are — that is the topic of
Session 3's BLUE bridge later in this lab. For now we ask: **do** the
assumptions hold?
:::

### C2 — Multicollinearity (VIF)

```{r vif-naive-114}
#| code-fold: false
vif_naive_114 <- car::vif(mod_naive_114)
vif_naive_114
```

The maximum VIF in the naïve model is
**`r round(max(vif_naive_114), 2)`**. VIF values below 5 are
unproblematic, 5–10 deserve attention, above 10 indicate severe
multicollinearity. We are safely below.




### B-Assumptions — residual plots

```{r residual-plots-114}
#| fig-height: 6
#| fig-width: 9
par(mfrow = c(2, 2))
plot(mod_naive_114)
par(mfrow = c(1, 1))
```

Reading these four plots:

- **Residuals vs Fitted (top-left):** looks for non-linearity (A2) and
  heteroscedasticity (B2). A horizontal red line with even spread is good.
- **Q-Q Residuals (top-right):** tests normality (B4). Points on the
  diagonal indicate Gaussian errors.
- **Scale-Location (bottom-left):** specifically for heteroscedasticity
  (B2). A flat red line means equal variance.
- **Residuals vs Leverage (bottom-right):** identifies influential
  outliers.

:::{.callout-tip}
## 💡 Reading residual plots

Don't expect "perfect" diagnostics from real scanner data. The
question is **whether the violations are severe enough to invalidate
inference** — and the *consequences* of violations are exactly what
the BLUE bridge in this lab will address.

In Session 2 you learned **how to check** — here in Session 3 you
will learn **what it means**.
:::

:::{.callout-important icon=false}
## ⚠️ T2: Diagnostic readout for Store 114

Looking at the four residual plots above:

a. Is there visible heteroscedasticity? Where in the fitted-value
   range is the residual spread largest?
b. Do the Q-Q residuals follow the diagonal — or are there
   substantial tail deviations?
c. Are there leverage points that stand out?

Write a one-sentence verdict per assumption (B2, B4, leverage).
:::


## Step 1b — The stumbling block: feat × deal entanglement

The diagnostics in Step 1a look reasonable. The maximum VIF is 2.51, which is safely below the standard danger threshold of 5. The software is happy. 

But as data translators, we should pause and think about **why** the VIF is 2.51. In our naïve model, this VIF primarily indicates a correlation between advertising (`feat`) and the current price level (`log(price)`). 

Why would advertising be systematically correlated with the shelf price? 
Domain knowledge gives us the answer: retailers rarely advertise a product at full price. They advertise *discounts*. This points directly to an **A1 violation (omitted variable)**. We modeled the advertising (`feat`), but we left out the operational trigger: the price promotion (`deal`). 

Let us cross-tabulate `feat` and `deal` to see how deeply these two marketing levers are entangled in Store 114.

```{r feat-deal-table-full}
#| code-fold: false
tbl_114%>%
  count(feat, deal) %>%
  tidyr::complete(feat = 0:1, deal = 0:1, fill = list(n = 0)) %>%
  group_by(feat) %>%
  mutate(
    share_within_feat = n / sum(n),
    share_within_feat = scales::percent(share_within_feat, accuracy = 0.1)
  ) %>%
  ungroup() %>%
  mutate(
    feat = paste0("feat = ", feat),
    deal = paste0("deal = ", deal)
  ) %>%
  gt() %>%
  gt::tab_header(title = "feat × deal for Store 114") %>%
  gt::cols_label(
    feat = "Advertising",
    deal = "Price promotion",
    n = "Count",
    share_within_feat = "Share within feat"
  ) %>%
  gt::tab_options(table.font.size = "13px")
```

:::{.callout-tip icon=false}
## 💡 Why doesn't a model with BOTH variables crash? (The asymmetric split)

Looking at the table, `feat = 1` perfectly implies `deal = 1`. If we add `deal` to our regression, shouldn't the model crash due to perfect multicollinearity (a C2 violation)? 

Look at the `feat = 0` rows. When there is *no* advertising, the store still runs price promotions roughly 38% of the time (101 out of 268 weeks). This asymmetric variation is the model's lifeline:
1. It learns the pure effect of a price drop (`deal`) from the weeks without advertising.
2. It estimates the `feat` coefficient as the *additional* boost that advertising provides *on top* of a deal. 

But because `feat = 1` never exists without a deal, the model has to strictly assume that effects are simply additive. It has never actually seen advertising work on its own.
:::

```{r feat-deal-table-114}
#| code-fold: false
tbl_114 %>%
  count(feat, deal) %>%
  pivot_wider(names_from = deal,
              values_from = n,
              names_prefix = "deal=",
              values_fill = 0) %>%
  gt() %>%
  gt::tab_header(title = "Joint distribution of feat and deal in Store 114") %>%
  gt::tab_options(table.font.size = "13px")
```

In every single week where Store 114 ran feature advertising
(`feat = 1`), there was also a price promotion (`deal = 1`). Feature
advertising never appears alone in the data.

:::{.callout-warning icon=false}
## ⚠️ The A1 violation

Our naïve model includes `feat` but not `deal`. If `feat = 1` always
co-occurs with `deal = 1`, then the estimated `feat` coefficient
absorbs **both** effects: the advertising effect *and* the
deal effect.

This is the classic **omitted variable bias** (A1 violation). Through
the domino effect, A1 violation produces a C1 violation
(endogeneity): the omitted `deal` is now part of the error term, and
it is correlated with `feat`.
:::

### What happens when we add `deal`?

```{r mod-full-114}
#| code-fold: false
mod_full_114 <- lm(logmove ~ feat + deal + log(price) + brand, data = tbl_114)

mod_full_114 %>%
  tidy(conf.int = TRUE) %>%
  mutate(across(where(is.numeric), \(x) round(x, 3))) %>%
  gt() %>%
  gt::tab_options(table.font.size = "13px")
```

Comparing the two models on Store 114:

```{r coef-comparison-114}
#| code-fold: true
#| code-summary: "Show code: coefficient comparison table"
bind_rows(
  tidy(mod_naive_114) %>% mutate(model = "Naïve (without deal)"),
  tidy(mod_full_114)  %>% mutate(model = "Full (with deal)")
) %>%
  filter(term %in% c("feat", "deal")) %>%
  select(model, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  mutate(across(where(is.numeric), \(x) round(x, 3))) %>%
  gt() %>%
  gt::tab_options(table.font.size = "13px")
```

The `feat` coefficient drops from
**`r round(coef(mod_naive_114)["feat"], 3)`** (naïve) to
**`r round(coef(mod_full_114)["feat"], 3)`** (with `deal`). The
difference of
**`r round(coef(mod_naive_114)["feat"] - coef(mod_full_114)["feat"], 3)`**
log units is the empirical size of the omitted-variable bias.

:::{.callout-important icon=false}
## ⚠️ T3: Interpret the OVB shift

a. By how much does the `feat` coefficient change when `deal` is added
   to the model? (Read from the comparison table.)
b. What does this say about the naïve estimate from Step 1a?
c. Could you, based on these data alone, tell Sarah what an
   advertising campaign **without** a deal would do? Why or why not?

Keep your notes — they go into the first translator memo at the
end of the lab.
:::

## Step 1c — Does this generalise? Replication and the full panel

We have an OVB story for Store 114. Is this a peculiarity of Store 114,
or is it a structural feature of the data? Two paths to find out:

1. **Replicate** in a second store
2. **Pool** all 83 stores

We start with the replication, because it brings out something that
pooling would hide.

### Replication in Store 83

We fit the same two models — naïve and full — on **Store 83**, a
demographically different store.

```{r mod-83}
#| code-fold: false
mod_naive_83 <- lm(logmove ~ feat + log(price) + brand, data = tbl_83)
mod_full_83  <- lm(logmove ~ feat + deal + log(price) + brand, data = tbl_83)
```

```{r replication-comparison}
#| code-fold: true
#| code-summary: "Show code: side-by-side comparison"
bind_rows(
  tibble(store = "Store 114", model = "naïve",     feat = coef(mod_naive_114)["feat"]),
  tibble(store = "Store 114", model = "with deal", feat = coef(mod_full_114)["feat"]),
  tibble(store = "Store 83",  model = "naïve",     feat = coef(mod_naive_83)["feat"]),
  tibble(store = "Store 83",  model = "with deal", feat = coef(mod_full_83)["feat"])
) %>%
  pivot_wider(names_from = model, values_from = feat) %>%
  mutate(
    ovb_diff = naïve - `with deal`,
    across(where(is.numeric), \(x) round(x, 3))
  ) %>%
  gt() %>%
  gt::tab_options(table.font.size = "13px")
```

In **Store 114** the OVB difference is
**`r round(coef(mod_naive_114)["feat"] - coef(mod_full_114)["feat"], 3)`** —
positive, as expected from the theoretical argument.

In **Store 83** the OVB difference is
**`r round(coef(mod_naive_83)["feat"] - coef(mod_full_83)["feat"], 3)`** —
the sign **flips**. Adding `deal` does not reduce the `feat`
coefficient here; it slightly increases it.

:::{.callout-warning icon=false}
## 💥 Aha-moment: replication as a stress test

Two stores from the same chain, same brands, same time window — and
they tell different stories about how `feat` and `deal` interact
**locally**. Within each store, classical inference gives a confident
answer. Across stores, the answers do not match.

This is what **replication** is for. It is not a confirmation ritual.
It is a stress test of external validity.

If we had only seen Store 114, we might have written Sarah a
confident memo. Store 83 forces us to qualify it.
:::

### The full panel: what does pooling tell us?

When we fit the same models on **all 83 stores**, we get an average
view that smooths over local heterogeneity.

```{r mod-full-panel}
#| code-fold: false
mod_naive_full <- lm(logmove ~ feat + log(price) + brand, data = tbl_full)
mod_full_full  <- lm(logmove ~ feat + deal + log(price) + brand, data = tbl_full)
```

```{r panel-overview}
#| code-fold: true
#| code-summary: "Show code: full-panel comparison"
tibble(
  scope    = c("Store 114 (n = 351)", "Store 83 (n = 333)", "Full panel (n ≈ 28k)"),
  feat_naive = c(coef(mod_naive_114)["feat"], coef(mod_naive_83)["feat"], coef(mod_naive_full)["feat"]),
  feat_full  = c(coef(mod_full_114)["feat"],  coef(mod_full_83)["feat"],  coef(mod_full_full)["feat"])
) %>%
  mutate(
    ovb_diff = feat_naive - feat_full,
    across(where(is.numeric), \(x) round(x, 3))
  ) %>%
  gt() %>%
  gt::tab_options(table.font.size = "13px")
```

In the full panel the OVB difference is
**`r round(coef(mod_naive_full)["feat"] - coef(mod_full_full)["feat"], 3)`**.
The structural confounding is positive on average — but the local
picture is heterogeneous.

```{r feat-deal-table-full}
#| code-fold: true
#| code-summary: "Show code: feat × deal in the full panel"
tbl_full %>%
  count(feat, deal) %>%
  pivot_wider(names_from = deal,
              values_from = n,
              names_prefix = "deal=",
              values_fill = 0) %>%
  gt() %>%
  gt::tab_header(title = "feat × deal across all 83 stores") %>%
  gt::tab_options(table.font.size = "13px")
```

Globally, the structural rule is exact: **whenever `feat = 1`, `deal = 1`**.
There is no week, in any store, where a brand was featured without
also being on price promotion.

:::{.callout-important icon=false}
## ⚠️ T4: Translator memo no. 1 (Question 1 — the advertising effect)

Draft Sarah's first memo. Cover:

- Can the historical data tell us what an advertising campaign
  *causes* in terms of sales?
- What is the methodological obstacle?
- What would be needed to answer her question properly?

**3–5 sentences, business language.** Bonus points if you mention what
kind of **study design** would resolve the problem (forward reference
to Lab 3).
:::

---

# Step 2 — The BLUE bridge becomes operative

We have hit a wall on Question 1. Classical inference delivers
coefficients with tight confidence intervals — but we cannot interpret
them causally, because the data foundation does not support it
(A1 violated → C1 violated). And the replication in Store 83 shows
that even within the chain, the local story changes.

Sarah's **Question 2** is different. She wants a *prediction tool* for
inventory planning. The question is not *"what would advertising
cause?"*. It is *"what will sales be, given conditions we observe?"*.

For prediction, BLUE is not the right criterion.

:::{.callout-note collapse="true" #blue-bridge}
## 📚 Background: BLUE for inference vs. prediction

Under the Gauss-Markov assumptions, OLS is the **B**est **L**inear
**U**nbiased **E**stimator. Note the order: *first* unbiased, *then*
minimum variance **among the unbiased**.

**For inference (explanation),** this is the gold standard. The
coefficient should reflect the true effect — unbiasedness is essential.

**For prediction,** the criterion is different. What we want is
*minimum total error on new data*:

$$\text{MSE}(\hat{Y}) = \text{Bias}^2(\hat{Y}) + \text{Var}(\hat{Y}) + \sigma_{\varepsilon}^2$$

In prediction, accepting *some* bias for *much less variance* can
lower the total MSE. This is the rationale for regularised methods
like Lasso and Ridge — they are *not* BLUE by design.

→ The same theoretical foundation explains both paths, with opposite
optimality criteria. This is the **BLUE bridge**.
:::

This conceptual shift sets up the rest of the lab. We assume the
business-relevant relationships are **stable** for inventory planning
purposes (this is a working assumption, not a guarantee — the Lucas
critique applies if the system is later perturbed). Given that
assumption, we now ask: **how well does the model generalise to stores
we have not seen?**

---

# Step 3 — Question 2 (Prediction): Out-of-Sample Evaluation

## Step 3a — Train on Store 114, predict on Store 83

We use the same data we already have: train on our analytical sample
(Store 114), evaluate the prediction error on the replication store
(Store 83).

```{r predict-114-to-83}
#| code-fold: false
pred_83  <- predict(mod_naive_114, newdata = tbl_83)
rmse_83  <- sqrt(mean((pred_83 - tbl_83$logmove)^2))
rmse_114 <- sqrt(mean(residuals(mod_naive_114)^2))
```

| Metric                                   | RMSE on log-sales |
|------------------------------------------|------------------:|
| In-sample (Store 114, naïve model)       | `r round(rmse_114, 3)` |
| Out-of-sample (Store 83 from Store 114)  | `r round(rmse_83, 3)`  |

Given that `logmove` ranges roughly from 4 to 14, this is a non-trivial
error. The out-of-sample value is meaningfully larger than the
in-sample value — a first sign that performance on **unseen stores**
is systematically worse than what the training data suggested.

## Step 3b — Train/test split across stores

A single out-of-sample target store gives one number. To get a
realistic estimate of generalisation error, we split the **set of
stores** itself: 80 % of stores for training, 20 % for testing.

```{r train-test-split}
#| code-fold: false
set.seed(42)
stores_all  <- unique(tbl_full$store)
test_stores <- sample(stores_all, size = floor(0.20 * length(stores_all)))

tbl_train <- tbl_full %>% filter(!store %in% test_stores)
tbl_test  <- tbl_full %>% filter(store  %in% test_stores)

mod_train  <- lm(logmove ~ feat + log(price) + brand, data = tbl_train)
pred_test  <- predict(mod_train, newdata = tbl_test)

rmse_train <- sqrt(mean(residuals(mod_train)^2))
rmse_test  <- sqrt(mean((pred_test - tbl_test$logmove)^2))
```

| Metric                          | RMSE on log-sales |
|---------------------------------|------------------:|
| In-sample (training stores)     | `r round(rmse_train, 3)` |
| Out-of-sample (test stores)     | `r round(rmse_test, 3)`  |

:::{.callout-tip}
## 💡 Splitting across stores, not across weeks

We split the **set of stores**, not the set of weeks. Splitting at the
week level would mean the model has already seen the same store —
predictive performance would be artificially inflated. For a
prediction tool that should work *in new stores*, we need store-level
splitting.

In time-series settings the same principle extends to **temporal CV**
to avoid future-leakage. You will revisit this in ER017.
:::

## Step 3c — K-fold cross-validation

A single 80/20 split depends on which 20 % we happened to draw.
**K-fold CV** averages over all possible splits and gives a more
stable estimate.

```{r kfold-cv}
#| code-fold: false
set.seed(42)
K <- 5
stores_shuffled <- sample(unique(tbl_full$store))
folds <- cut(seq_along(stores_shuffled), breaks = K, labels = FALSE)

cv_rmses <- map_dbl(1:K, \(k) {
  test_st  <- stores_shuffled[folds == k]
  train_st <- stores_shuffled[folds != k]

  mod_k <- lm(logmove ~ feat + log(price) + brand,
              data = tbl_full %>% filter(store %in% train_st))
  pred_k   <- predict(mod_k,
                      newdata = tbl_full %>% filter(store %in% test_st))
  actual_k <- tbl_full %>% filter(store %in% test_st) %>% pull(logmove)
  sqrt(mean((pred_k - actual_k)^2))
})

cv_score <- mean(cv_rmses)
```

The five fold-wise RMSEs are
**`r paste(round(cv_rmses, 3), collapse = ", ")`**. The mean of these
gives our cross-validated estimate of out-of-sample error:

| Quantity            | Value |
|---------------------|------:|
| Mean 5-fold CV-RMSE | `r round(cv_score, 3)` |

This is the realistic generalisation-error estimate for the simple
model `feat + log(price) + brand` on a new store.

:::{.callout-important icon=false}
## ⚠️ T5: Translator memo no. 2 (Question 2 — the prediction tool)

Draft Sarah's second memo. Cover:

- Can we deliver an inventory-planning tool that works across all
  83 stores?
- What is the realistic out-of-sample accuracy
  (CV-RMSE ≈ `r round(cv_score, 2)`)?
- What variables are *not yet* in the model that you would add to
  improve it? (Hint: think about what makes Store 83 different from
  Store 114.)
:::

---

# Step 4 — Synthesis

```{r synthesis-table}
#| echo: false
tibble::tribble(
  ~`Question`,                              ~`Methodological path`,            ~`Verdict`,
  "1. Effect of a Q4 advertising campaign", "Path B: causal identification",    "Not answerable from these data alone",
  "2. Inventory prediction tool",           "Path A: out-of-sample, CV",        "Deliverable, with caveats"
) %>%
  gt() %>%
  gt::tab_style(gt::cell_fill("#fff8e1"), gt::cells_body(rows = 1)) %>%
  gt::tab_style(gt::cell_fill("#f0fff0"), gt::cells_body(rows = 2)) %>%
  gt::tab_style(gt::cell_fill("#f5f0fa"), gt::cells_column_labels()) %>%
  gt::tab_options(table.font.size = "13px")
```

The same dataset gave us two methodological operations and two
different verdicts. **This is the BLUE bridge in action:**

- Path A (prediction): we accept that we cannot disentangle advertising
  from price promotion, because for *forecasting under stable
  structure* that does not matter. The relationship is consistent in
  the data, and the model predicts.
- Path B (explanation/intervention): we cannot deliver a causal claim,
  because the data foundation does not support identification. To
  answer Sarah's Q1 properly, we need a different study design — an
  A/B test (forward reference to Lab 3).

This is the **central lesson of the lab**: the optimality criterion
depends on the question. Same data, two operations, two answers.

:::{.callout-note collapse="true" #lucas-critique}
## 📚 Background: Lucas critique

The Path A prediction relies on a working assumption: the structure of
the relationships does not change. If management *actively intervenes*
(e.g. launches a big campaign), this assumption may break — a
phenomenon known as the **Lucas critique** [@lucas_1976]. The
historical correlations were generated under one regime; the new
regime may behave differently.

This is the deeper reason why Sarah's Q1 is fundamentally different
from Q2: it asks about a *change* to the system, not about predicting
within it. Path A is appropriate when the system stays as it is.
Path B is required when we change the system.
:::

:::{.callout-note collapse="true" #pop-sample}
## 📚 Background: Big data is not population

Even when we use all 83 stores, we have not "reached population".
These data come from **one** supermarket chain, in **one** city,
during **one** time period (1989–1994), with **one** selected set of
brands. Inferring from this dataset to "supermarket markets in
general" is an extra-statistical generalisation.

Big data does not dissolve the population/sample distinction — it
just makes the sample large. The Lucas critique and external-validity
arguments still apply.
:::

---

# Bonus — For independent study

Lab 2 will refine the prediction tool by adding the 12 demographic
variables from `tbl_demo`. As preparation, you can explore:

:::{.callout-important icon=false collapse="true"}
## ⚠️ B1 (Bonus): Lasso preview on demographics

Join the demographics to the full panel and fit a Lasso model with
cross-validation:

```r
library(glmnet)
tbl_joined <- tbl_full %>%
  left_join(tbl_demo, by = "store")
# build the design matrix, then:
# cv_fit <- cv.glmnet(X, y, alpha = 1)
```

Which demographic variables survive at the optimal $\lambda$? How does
the cross-validated RMSE compare to the simple model from Step 3?

This is exactly the question we will work through together in Lab 2.
Coming with your own attempt makes the lab more useful.
:::

---

# What comes next

| Lab | Question it answers |
|-----|---------------------|
| **Lab 2 — Regularisation on demographics** | Can we improve the prediction tool by adding demographic variables — and which of them matter? |
| **Lab 3 — A/B-test design**                | How would we set up a study that answers Sarah's Question 1 properly? |



### The full panel: exact rule or strong tendency?

When we move from one store to all 83 stores, we should again verify the
empirical relationship instead of assuming it.

```{r feat-deal-table-full}
#| code-fold: false
tbl_full %>%
  count(feat, deal) %>%
  tidyr::complete(feat = 0:1, deal = 0:1, fill = list(n = 0)) %>%
  group_by(feat) %>%
  mutate(
    share_within_feat = n / sum(n),
    share_within_feat = scales::percent(share_within_feat, accuracy = 0.1)
  ) %>%
  ungroup() %>%
  mutate(
    feat = paste0("feat = ", feat),
    deal = paste0("deal = ", deal)
  ) %>%
  gt() %>%
  gt::tab_header(title = "feat × deal across all 83 stores") %>%
  gt::cols_label(
    feat = "Advertising",
    deal = "Price promotion",
    n = "Count",
    share_within_feat = "Share within feat"
  ) %>%
  gt::tab_options(table.font.size = "13px")
```

```{r feat-deal-summary-full}
#| code-fold: false
tbl_full %>%
  summarise(
    `# obs`                  = n(),
    `# feat = 1`             = sum(feat == 1, na.rm = TRUE),
    `# deal = 1`             = sum(deal == 1, na.rm = TRUE),
    `# feat = 1 & deal = 1`  = sum(feat == 1 & deal == 1, na.rm = TRUE),
    `# feat = 1 & deal = 0`  = sum(feat == 1 & deal == 0, na.rm = TRUE),
    `P(deal = 1 | feat = 1)` = ifelse(sum(feat == 1, na.rm = TRUE) > 0,
                                      sum(feat == 1 & deal == 1, na.rm = TRUE) /
                                      sum(feat == 1, na.rm = TRUE),
                                      NA_real_)
  ) %>%
  mutate(`P(deal = 1 | feat = 1)` = scales::percent(`P(deal = 1 | feat = 1)`,
                                                    accuracy = 0.1)) %>%
  gt() %>%
  gt::tab_header(title = "How close is the full panel to feat => deal?") %>%
  gt::tab_options(table.font.size = "13px")
```

```{r feat-without-deal-by-store}
#| code-fold: true
#| code-summary: "Show code: stores with feature-without-deal cases"
tbl_oj %>%
  group_by(store) %>%
  summarise(
    n_feat = sum(feat == 1, na.rm = TRUE),
    n_feat_without_deal = sum(feat == 1 & deal == 0, na.rm = TRUE),
    p_deal_given_feat = ifelse(n_feat > 0,
                               sum(feat == 1 & deal == 1, na.rm = TRUE) / n_feat,
                               NA_real_)
  ) %>%
  ungroup() %>%
  filter(n_feat > 0) %>%
  arrange(desc(n_feat_without_deal), p_deal_given_feat) %>%
  mutate(
    p_deal_given_feat = scales::percent(p_deal_given_feat, accuracy = 0.1)
  ) %>%
  gt() %>%
  gt::tab_header(title = "Store-level dependence between feature and deal") %>%
  gt::cols_label(
    n_feat = "# feat = 1",
    n_feat_without_deal = "# feat = 1 & deal = 0",
    p_deal_given_feat = "P(deal = 1 | feat = 1)"
  ) %>%
  gt::tab_options(table.font.size = "13px")
```

The full panel therefore tells us **how strong** the dependence between
`feat` and `deal` is on average, while the store-level table shows that
this dependence may still vary locally. This matters for both causal
interpretation and out-of-sample transportability.
