---
title: "Data Science and Strategy for Business"
subtitle: "Complexity, Generalization & Causality: Predict or Intervene?"
author: "Prof. Dr. Jörg Schoder"
institute: "FFHS"
date: "`r Sys.Date()`"
bibliography: ../../lit/my_bib.bib
lang: en-US
reference-section-title: References
output:
  xaringan::moon_reader:
    self_contained: true
    css:
         - default
         - ../../css/ffhs-theme_js.css
         - xaringan-themer.css
    includes:
      after_body: ../../css/insert-logo.html
    lib_dir: ../../libs
    nature:
      slideNumberFormat: "%current%/%total%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: false
---
class: title-slide

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_xaringan(text_color = "#d50006",
               inverse_text_color = "#FFFFFF",
               inverse_background_color = "#d50006",
               title_slide_background_color = "#d50006",
               header_background_color = "#d50006",
               header_color = "#FFFFFF",
               header_h1_font_size = "32px",
               header_h2_font_size = "26px",
               link_color = "#502479",
               header_h3_font_size = "20px",
               text_slide_number_color = "#d50006",
               text_slide_number_font_size = "0.5em")
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_progress_bar(color = "#d50006", location = "bottom")
xaringanExtra::use_xaringan_extra(c("tile_view","scribble","panelset","tachyons"))
xaringanExtra::style_panelset_tabs(font_family = "inherit")
xaringanExtra::use_editable(expires = 1)
xaringanExtra::use_freezeframe(trigger = "hover")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 150)
options(htmltools.dir.version = FALSE)
library(latex2exp)
library(fontawesome)
library(tidyverse)
library(kableExtra)
library(gt)
source(xfun::from_root("lit","helper.R"))
library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           style = "markdown",
           dashed = TRUE)
bib <- ReadBib(xfun::from_root("lit","my_bib.bib"))
Zitiere <- function(bib, key, ...) {
  gsub(" and ", " and ", Citet(bib, key, ...))
}

# colour palette (consistent with labs)
col_primary   <- "#D50006"   # red — Path E1 / OLS / current
col_secondary <- "#502479"   # purple — Path P / Lasso / Lab 2
col_tertiary  <- "#7d0a52"   # dark accent — Path E2 / Causal / Lab 3
col_neutral   <- "#686868"
```


# ER114 — Data Science and Strategy for Business

## Session 3

### Complexity, Generalization & Causality. *Predict or Intervene?*

<br>
<br>
<br>
<br>
<br>
### FS 2026
<br>
### Prof. Dr. Jörg Schoder
.mycontacts[
`r fa('github')` @FFHS-EconomicResearch
`r fa('linkedin')` @jfschoder
]


---
layout: true

<div class="my-footer"></div>

<div style="position: absolute;left:400px;bottom:10px;font-size:9px">`r fa('creative-commons')``r rmarkdown::metadata$author`</div>




---
class: left

.blockquote[Agenda]

## Session 3 Overview

<div style="display:grid; grid-template-columns:80px 1fr; gap:0.3em 1em;
            font-size:0.82em; margin-top:0.5em; align-items:center;">

<div style="color:#686868;">10'</div>
<div>`r fa('rotate-left')` <strong>Welcome &amp; Recap</strong> — where Sessions 1+2 left us</div>

<div style="color:#686868;">15'</div>
<div>`r fa('lightbulb')` <strong>Prediction vs. Explanation</strong> — the central dilemma of Block&nbsp;3</div>

<div style="color:#686868;">15'</div>
<div>`r fa('chart-line')` <strong>The In-Sample Illusion</strong> — why high R² alone is dangerous</div>

<div style="color:#D50006; font-weight:bold;">45'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab&nbsp;1</strong> — Reality Check via Out-of-Sample Validation</div>

<div style="color:#686868;">15'</div>
<div>`r fa('triangle-exclamation')` <strong>When OLS Fails</strong> — the BLUE bridge, bias–variance, Lasso preview</div>

<div style="color:#686868; font-style:italic;">10'</div>
<div style="color:#686868; font-style:italic;">`r fa('mug-hot')` Break</div>

<div style="color:#D50006; font-weight:bold;">25'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab&nbsp;2</strong> — Multicollinearity &amp; Regularization</div>

<div style="color:#686868;">25'</div>
<div>`r fa('compass')` <strong>Causal-Inference Primer</strong> — beyond correlation: A/B&nbsp;tests, quasi-experiments, IV, DML</div>

<div style="color:#D50006; font-weight:bold;">30'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab&nbsp;3</strong> — Design your Experiment (Innovator's Hypothesis)</div>

<div style="color:#686868;">10'</div>
<div>`r fa('forward')` <strong>Outlook ER017</strong> — Machine Learning &amp; Causal+ML</div>

<div style="color:#686868;">10'</div>
<div>`r fa('flag-checkered')` <strong>Debrief &amp; Evaluation</strong></div>

</div>

???
* Red blocks = labs (Lab 1 is longer because it covers the OLS diagnostic foundation)
* Three connecting threads: DGP-paths E1/E2/P, Spiess credibility argument, Sarah Chen narrative
* Total ≈ 195 min including break


---
class: inverse, center, middle

## Recap

### Where did Sessions 1 + 2 leave us?


---
class: left

.blockquote[Recap]

## Sessions 1 → 2 → **3**

.panelset[
.panel[.panel-name[Session 1]
###  *See &amp; Describe*

`r fa('eye')` Descriptive regression as **conditional mean**:
$$\hat{y} = \mathbb{E}[y\,|\,\mathbf{x}]$$

`r fa('eye')` CRISP-DM as project skeleton

`r fa('eye')` Tidyverse pipeline + EDA

`r fa('eye')` First contact with **OVB** — missing variables bias all coefficients

]
.panel[.panel-name[Session 2]
### *Hypotheses Testing*

`r fa('dice')` Population vs. sample · **stochastic** regression

`r fa('dice')` Bootstrap &amp; permutation — *"only one test"*

`r fa('dice')` Type I/II errors as economic risks (EV-Framework)

`r fa('dice')` p-Hacking, FDR, BH

]
.panel[.panel-name[Session 3]
### *Today*

`r fa('circle-question')` **Inference vs. Prediction** — when is which the right purpose?

`r fa('circle-question')` **Multicollinearity** &amp; the BLUE limits

`r fa('circle-question')` **Bias-variance trade-off** &amp; regularization

`r fa('circle-question')` **Cross-validation** as the out-of-sample yardstick

`r fa('circle-question')` **Causal engineering** — when correlation is not enough

.blockquote[
"The Block&nbsp;3 question is no longer *what do we see?* &mdash; it is *what should we believe and what should we act on?*"
]
]
]

???
* The recap stresses that Sessions 1 and 2 stayed on the **same dataset** with the **same model** — only the lens changed
* Today we change the lens twice more: out-of-sample (Lab 1+2) and counterfactual (Lab 3)


---
class: left

.blockquote[Big picture for today]

## What should we believe — and act on?

<br>

.center[
### `r fa('bridge', fill='#502479')` Lab&nbsp;1: Role of Gauss-Markov-Assumptions
.small[Inference vs. Prediction. Checking Assumptions and assessing consequences.]
]

<br>

.pull-left[
### `r fa('shuffle', fill='#502479')` Will the pattern generalize?

The **prediction** question — out-of-sample.

`r fa('arrow-right', fill='#502479')` Path **P**

`r fa('arrow-right', fill='#502479')` Today's gateway: **Lab&nbsp;2**

]

.pull-right[
### `r fa('diagram-project', fill='#502479')` What changes if I intervene?

The **causal** question — counterfactual.

`r fa('arrow-right', fill='#D50006')` Path **E** (E2 sub-question)

`r fa('arrow-right', fill='#D50006')` Today's gateway: **Lab&nbsp;3**


]

???
* Two simple questions, one whole session
* Lab 1 is the **gateway** — not an afterthought; it is the diagnostic bridge that exposes where each path needs to begin
* Path E1 (classical inference, what is the partial effect?) is what Lab 1 stress-tests; both E2 (causal counterfactual) and P (prediction) inherit lessons from where E1 reaches its limits
* Icon "shuffle" for generalize evokes "across draws / across folds" — more apt than a time-series chart
.small[`r Citet(bib, "breiman_2001")`]
.small[`r Citet(bib, "shmueli_2010")`]

---
class: inverse, center, middle

## Part 1: Foundations

.blockquote[Two Paths from the Data-Generating Process]


---
class: left

.blockquote[Part 1 · The model]

## Why model? The conditional expectation E[y|x]

.pull-left[
For a response $y$ and inputs $\mathbf{x}$:

$$\mathbb{E}[y\,|\,\mathbf{x}] = f(\mathbf{x}'\boldsymbol{\beta})$$

`r fa('arrow-right')` **Marginal** distribution (top): the unconditional shape of $y$ — single number for the mean.

`r fa('arrow-right')` **Conditional** expectation (bottom): a *function* of $\mathbf{x}$ — different mean per slice.


.blockquote[
.small[
All three paths — E1, E2, P — make claims about parts of this conditional distribution.
]
]

.tr[.small[`r Citet(bib, "taddy_modern_2023")` Ch. 1, Fig. 1.1]]
]

.pull-right[
```{r marginal-conditional-stack}
#| echo: false
#| warning: false
#| fig-width: 4.5
#| fig-height: 4.8
#| fig-align: 'center'
#| out-width: "100%"
set.seed(20260522)
n <- 2400
tbl <- tibble(
  bathrooms = sample(0:7, n, replace = TRUE,
                     prob = c(0.04, 0.18, 0.40, 0.22, 0.10, 0.04, 0.015, 0.005)),
  home_value = exp(11.2 + 0.32 * bathrooms + rnorm(n, sd = 0.45))
)

p_marg <- ggplot(tbl, aes(home_value)) +
  geom_histogram(bins = 35, fill = col_neutral, color = "white") +
  scale_x_continuous(labels = scales::label_dollar(scale = 1e-3, suffix = "k"),
                     limits = c(0, 1.6e6)) +
  labs(x = "Home value", y = "Frequency",
       title = "(a) Marginal distribution") +
  theme_minimal(base_size = 9) +
  theme(plot.title = element_text(size = 10, face = "bold"))

p_cond <- ggplot(tbl, aes(factor(bathrooms), home_value)) +
  geom_boxplot(aes(fill = factor(bathrooms)),
               outlier.size = 0.5,
               outlier.colour = col_neutral) +
  scale_y_continuous(labels = scales::label_dollar(scale = 1e-3, suffix = "k")) +
  scale_fill_manual(values = c("#a8e6cf","#80d6c4","#5fc6b8","#43b3a8",
                               "#3196a8","#2576a0","#1c5594","#502479"),
                    guide = "none") +
  labs(x = "Number of bathrooms", y = "Home value",
       title = "(b) Conditional distribution") +
  theme_minimal(base_size = 9) +
  theme(plot.title = element_text(size = 10, face = "bold"))

cowplot::plot_grid(p_marg, p_cond, ncol = 1, rel_heights = c(1, 1.3))
```
]

???
* Mirrors Taddy/Hendrix/Harding 2023 Figure 1.1 — both panels (a) marginal histogram and (b) conditional boxplots by bathrooms
* Marginal mean = one number; conditional mean = a function of bathrooms
* The shift from a single number to a function is what regression IS
* All three paths today work on this conditional distribution — but each extracts something different from it


---
class: left

.blockquote[Part 1 · The logic]

## Induction and Deduction — two roles for data

```{r induction-deduction-svg, echo=FALSE, out.width="100%", fig.align='center'}
knitr::include_graphics(xfun::from_root("img","Session3","induction_deduction.svg"))
```

???
* Two visually distinct halves: Induction (purple) at top, Deduction (red) at bottom
* **Key point:** the role of data is fundamentally different
  - In induction: data is the *source material* from which we generate a hypothesis (data mining, model training)
  - In deduction: data is the *arbiter* that tests an a-priori hypothesis (classical inference) or yields a prediction
* Path P = induction (training) + deduction (predicting new x)
* Path E1 = deduction only (hypothesis pre-specified, then tested)
* No claim of abduction here — it would not add clarity for this audience
* Lab 1 lives in pure deduction; Lab 2 adds explicit induction (CV-tuned model); Lab 3 returns to deduction (test of pre-specified design)


---
class: left

.blockquote[Part 1 · Translator role in Block 3]

## From AI User to AI Value Creator — *applied to method choice*

.panelset[
.panel[.panel-name[Recap from Session 1]
<svg viewBox="0 0 680 200" width="90%" style="display:block;margin:auto">
  <defs>
    <style>
      .rtxt { font-family:sans-serif; font-size:11px; font-weight:bold;
              fill:white; text-anchor:middle; }
      .rarr { stroke:#686868; stroke-width:2; fill:none;
              marker-end:url(#arr); }
      .anno { font-family:sans-serif; font-size:10px; fill:#686868;
              text-anchor:middle; }
    </style>
    <marker id="arr" markerWidth="8" markerHeight="8"
            refX="6" refY="3" orient="auto">
      <path d="M0,0 L0,6 L8,3 z" fill="#686868"/>
    </marker>
  </defs>
  <rect x="10"  y="140" width="130" height="40" rx="6" fill="#b3b2b2"/>
  <text class="rtxt" x="75"  y="165">1. Collect</text>
  <rect x="180" y="140" width="130" height="40" rx="6" fill="#686868"/>
  <text class="rtxt" x="245" y="165">2. Organise</text>
  <rect x="350" y="140" width="130" height="40" rx="6" fill="#7d0a52"/>
  <text class="rtxt" x="415" y="165">3. Analyse</text>
  <rect x="520" y="140" width="140" height="40" rx="6" fill="#502479"/>
  <text class="rtxt" x="590" y="165">4. Infuse</text>
  <line class="rarr" x1="145" y1="160" x2="175" y2="160"/>
  <line class="rarr" x1="315" y1="160" x2="345" y2="160"/>
  <line class="rarr" x1="485" y1="160" x2="515" y2="160"/>
  <line x1="340" y1="20" x2="340" y2="135"
        stroke="#D50006" stroke-width="2" stroke-dasharray="5,3"/>
  <text style="font-family:sans-serif;font-size:10px;fill:#D50006;
               text-anchor:middle;font-weight:bold" x="340" y="15">
    AI User | AI Value Creator
  </text>
  <text class="anno" x="170" y="125">Standard Tools</text>
  <text class="anno" x="170" y="137">Commodity Data</text>
  <text class="anno" x="510" y="125">Proprietary Data</text>
  <text class="anno" x="510" y="137">Sustainable Advantage</text>
</svg>
.center[.small[Session 1 framing — data and tooling determine commodity vs. moat.]]
]
.panel[.panel-name[New in Session 3]
### The same line, applied to *method choice*

.pull-left[
`r fa('user', fill='#686868')` **AI User** *(method-level)*

* Picks the textbook method
* Reports $R^2$, p-value, default outputs
* Confuses Path E and Path P
* Optimizes within a given **map** — without knowing it
]

.pull-right[
`r fa('user-gear', fill='#D50006')` **AI Value Creator** *(method-level)*

* Asks: *prediction or intervention?*
* Picks the **path** before the estimator
* Validates the diagnostic, not just the point estimate
* **Chooses the map** — and defends it
]

<br>

.blockquote[
.small[
Today's labs make this concrete. **Lab 1** is the diagnostic discipline. **Lab 2** is the prediction discipline. **Lab 3** is the causal discipline. Knowing which to pick is the Value-Creator skill.
]
]
]
.panel[.panel-name[The data dimension]
### Why method choice matters for the data moat

`r fa('rocket', fill='#D50006')` **Proprietary data** alone is not enough — it must be paired with the **right path**:

* Predicting customer churn from your own data → Path **P** with regularization (Lab&nbsp;2 family)
* Estimating the effect of a *new* pricing policy → Path **E2** with experimental design (Lab&nbsp;3 family)

`r fa('triangle-exclamation', fill='#686868')` Pairing the wrong path with great data wastes the moat:

* Lasso on a one-off intervention question gives confident, biased nonsense
* OLS controls on observational data without a DAG yield estimates the CEO cannot defend

.quellePanURL[`r Citet(bib, "thomasAIValueCreators2025")`]
]
]

???
* Three panels: (1) Session 1 framing as recall, (2) the NEW Session 3 framing applied to method choice, (3) why path + data must align
* Key shift from Session 1: Value Creation in Block 3 is no longer just about *data* — it is about the *method-path-decision* itself
* Lab 3's deliverable T_final is exactly the artefact a Value Creator produces: a defensible methodological recommendation


---
class: left

.blockquote[Part 1 · The deeper layer]

## Philosophy Eats AI — the layer beneath every method

```{r philosophy-svg, echo=FALSE, out.width="98%", fig.align='center'}
knitr::include_graphics(xfun::from_root("img","Session3","philosophy_eats_ai.svg"))
```

???
* Schrage & Kiron (2025): every AI stack rests on philosophical commitments — usually unarticulated, sometimes catastrophically wrong
* Three pillars are theirs (Ontology, Epistemology, Teleology); Ethics is treated separately as a fourth
* **Ontology examples**: Starbucks defined the "Starbucks experience" as connection (not transactions) before designing Deep Brew AI. Amazon Prime ontologically defined "best customers" — not "most profitable" — as the optimization target
* **Epistemology examples**: prediction (Path P) and counterfactual evidence (Path E2) are two different theories of what counts as knowing. Confusing them is the root cause of most bad analytics
* **Teleology examples**: Google's Gemini fiasco — accuracy and diversity goals were misaligned; the model fell to whichever objective dominated
* "The map is not the territory" (Korzybski via Wittgenstein, quoted by Schrage & Kiron): metrics are maps; loyalty is the territory; RFM/NPS/churn are metrics that *might* represent loyalty — or might define it by fiat
* The Translator's job: surface these choices before picking a method
* Forward link: Lab 3 (what counts as causal evidence?), back to Lab 1+2 (what counts as a "good" prediction?)


---
class: left

.blockquote[Part 1 · The DGP]

## Two Paths from the Data-Generating Process

```{r dgp-paths-slide, echo=FALSE, fig.align='center', out.width="80%"}
knitr::include_graphics(xfun::from_root("img","Session3","DGP-Paths_gemini.svg"))
```

.center[
.small[
Same data, same DGP — two analytical purposes. Path E asks *what is*; Path P asks *what next*.
]
]

???
* Same SVG students see in Lab 1 — recognition cue intentional
* Path E (Explanation/Inference): two sub-questions — E1 "what is the partial effect?" (next slide) and E2 "what would change under intervention?"
* Path P (Prediction): a single goal — generalise to new data
* The DGP is the unobservable urn; data are realisations; methods differ by what they extract from that realisation
* Breiman 2001 was the first to make this dichotomy programmatic ("Two Cultures")


---
class: left

.blockquote[Part 1 · Path E branches]

## Path E has two sub-questions

```{r path-e-branches, echo=FALSE}
tibble::tribble(
  ~`Sub-question`,                                       ~Goal,                              ~Identification,                                    ~Lab,
  "E1 — What is the partial effect?",                    "Estimate β under the model",       "by assumption (A/B/C compass)",                    "Lab 1",
  "E2 — What would change under intervention?",          "Estimate a counterfactual",        "by research design (RCT, DiD, IV, RDD, DML)",      "Lab 3"
) %>%
  gt() %>%
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) %>%
  tab_style(style = cell_fill(color = "#fff5f5"),
            locations = cells_body(rows = 1)) %>%
  tab_style(style = cell_fill(color = "#f5e6f0"),
            locations = cells_body(rows = 2)) %>%
  tab_options(table.font.size = "13px",
              table.width = "100%",
              column_labels.font.size = "12px",
              column_labels.background.color = "#f5f5f5")
```

<br>

.pull-left[
### E1 — model given
`r fa('arrow-right', fill='#D50006')` Identification settled **before** the data, by specifying the equation.

`r fa('arrow-right', fill='#D50006')` Same path E logic students saw in Sessions 1 + 2.

`r fa('forward', fill='#D50006')` Today: **Lab 1** stress-tests E1.
]

.pull-right[
### E2 — model earns its claim
`r fa('arrow-right', fill='#7d0a52')` Identification comes from the **data structure** — randomization, parallel trends, instrument validity.

`r fa('arrow-right', fill='#7d0a52')` "Design-based identification" (Angrist &amp; Pischke 2010).

`r fa('forward', fill='#7d0a52')` Today: **Lab 3** operationalizes E2.
]

???
* E1 and E2 both live under Path E — they ask "what is" — but they differ in *what counts* as identification
* E1 is the classical econometrics curriculum; E2 is the credibility-revolution addition (Angrist/Pischke 2010)
* Path P (Prediction) sits beside them — it asks "what next?" and uses out-of-sample generalisation as its yardstick
* The labs are mapped: Lab 1 → E1 (and gateway to E2); Lab 2 → P; Lab 3 → E2


---
class: inverse, center, middle

## Part 2: BLUE Bridge

.blockquote[When Assumptions Bite — and what unites the paths]


---
class: left

.blockquote[Part 2 · The A/B/C compass]

## Model Assumptions — the A/B/C compass

```{r abc-compass, echo=FALSE}
tibble::tribble(
  ~Group, ~Code,                            ~Assumption,                                         ~`Violation hits Path E`,                ~`Violation hits Path P`,
  "A",    "A1 — variable selection",        "No relevant variable omitted, none irrelevant",     "catastrophic (OVB → wrong β)",          "ok if DGP unchanged",
  "A",    "A2 — linearity",                 "Linear in parameters",                              "biased estimates",                       "biased predictions",
  "A",    "A3 — parameter constancy",       "Parameters constant across observations",           "shifts after intervention (Lucas)",     "fails on new regime",
  "B",    "B1 — E[ε] = 0",                  "Mean-zero error",                                    "automatically met",                      "automatically met",
  "B",    "B2 — homoscedasticity",          "Var(ε) constant",                                   "SE wrong (→ robust SE)",                "unaffected",
  "B",    "B3 — no autocorrelation",        "Cov(εᵢ,εⱼ)=0",                                       "SE wrong (→ HAC)",                       "unaffected",
  "B",    "B4 — normality",                 "ε ~ N (only small samples)",                         "exact inference only",                   "irrelevant",
  "C",    "C1 — exogeneity",                "Cov(X,ε) = 0",                                       "catastrophic — wrong β",                 "ok if DGP unchanged",
  "C",    "C2 — no multicollinearity",      "Predictors not perfectly collinear",                "unstable β, inflated SE",                "predictions ok"
) %>%
  gt() %>%
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) %>%
  tab_style(style = cell_fill(color = "#fff5f5"),
            locations = cells_body(columns = `Violation hits Path E`)) %>%
  tab_style(style = cell_fill(color = "#f5f0fa"),
            locations = cells_body(columns = `Violation hits Path P`)) %>%
  tab_options(table.font.size = "11px",
              table.width = "100%",
              column_labels.font.size = "11px",
              data_row.padding = px(2),
              column_labels.background.color = "#f5f5f5")
```

.tr[.small[`r Citet(bib, "vonAuer_2023")` · Lab 1 Step 2]]

???
* Same compass students saw in Lab 1 — but now read **column-wise by path**: which violation costs Path E vs. Path P
* The big-picture observation: Path E is more vulnerable than Path P to most violations — but Path P is **catastrophically** vulnerable to A3 (parameter shift), which is the *only* violation Path E can sometimes survive
* Lab 2 attacks C2 (multicollinearity); Lab 3 attacks C1 (endogeneity); Lab 1 demonstrates how all of these show up in diagnostics
* Skip slide in time pressure: the take-aways "Prediction needs stable relationships, Inference needs credible explanations" are on the next slide


---
class: left

.blockquote[Part 2 · BLUE]

## BLUE — *what it guarantees, and the take-away for each path*

.pull-left[
### Gauss-Markov says
Under A1–A3, B1–B3, C1–C2 (no B4 needed for large $n$):

OLS is **B**est **L**inear **U**nbiased **E**stimator — the **unbiased linear** estimator with the **smallest variance**.

`r fa('check', fill='#502479')` Identification settled by assumption

`r fa('check', fill='#502479')` Coefficients = partial effects

`r fa('check', fill='#502479')` SE valid for inference

### What it does **not** say

`r fa('triangle-exclamation', fill='#D50006')` "Best" only within **linear + unbiased**

`r fa('triangle-exclamation', fill='#D50006')` Other estimators may beat OLS on MSE

`r fa('triangle-exclamation', fill='#D50006')` BLUE is a property of the *estimator*, not a performance guarantee
]

.pull-right[
### Take-away by path

<div style="background:#f5f0fa; padding:0.6em 0.8em; border-left:5px solid #502479; margin-bottom:0.5em;">
<strong>Path P (Prediction)</strong>: <em>tolerates</em> C2 (multicollinearity); <em>breaks</em> under A3 (Lucas). Predictions need <strong>stable relationships</strong>.
</div>

<div style="background:#fff5f5; padding:0.6em 0.8em; border-left:5px solid #D50006; margin-bottom:0.5em;">
<strong>Path E (Inference / Causal)</strong>: <em>breaks</em> under C1 (endogeneity) and C2; needs <strong>credible explanations</strong>.
</div>

<div style="background:#f5f5f5; padding:0.6em 0.8em; border-left:5px solid #686868;">
<strong>Both paths</strong> can repair B2 / B3 violations with robust / HAC standard errors — Lab&nbsp;1 shows how.
</div>

<br>

.small[Different goals → different vulnerabilities → different repair strategies. The next slide synthesises this as the BLUE bridge.]
]

???
* This slide consolidates what was previously two separate slides — the guarantees and the path-specific consequences
* Skip in time pressure: the BLUE-bridge SVG on the next slide IS the visual punchline
* Tonspur: BLUE is unconditionally correct under assumptions — but what you *value* differs by path. That's why Path P will deliberately abandon unbiasedness, and Path E will hold onto it as a credibility commitment


---
class: left

.blockquote[Part 2 · The bridge]

## BLUE as the bridge between paths

```{r blue-bridge-svg, echo=FALSE, out.width="100%", fig.align='center'}
knitr::include_graphics(xfun::from_root("img","Session3","blue_bridge.svg"))
```

???
* BLUE is the common starting point both paths share — under assumptions A/B/C, OLS is the unbiased benchmark
* Path P branches *upward*: deliberately accept bias for variance reduction (the "bias-variance trade-off", Hastie/Tibshirani/Friedman 2009). The toolkit is Ridge / Lasso / Elastic Net + cross-validation. Lab 2 operates here.
* Path E branches *downward*: preserve unbiasedness, and add design-based identification (Angrist & Pischke 2010). The toolkit is robust SE, RCT, DiD, IV, DML. Lab 1 & Lab 3 operate here.
* The Spiess (2025) point at the bottom: preserving unbiasedness is also an *institutional* discipline — it prevents the researcher from tuning the analysis toward a favoured result. Pre-commitment makes the estimate credible.
* This is the punchline of Part 2 — every method in today's labs sits on one of these two branches


---
class: left

.blockquote[Narrative anchor → Lab 1]

## `r fa('envelope', fill='#D50006')` Memo from Sarah Chen — and into Lab 1

.pull-left[
.small[
**To:** Translator Team &nbsp;·&nbsp; **From:** Sarah Chen, VP Marketing — Dominick's Finer Foods
]

> *Two distinct questions this quarter:*
>
> *1. Does **feature advertising** cause incremental sales?* &nbsp;→ Path **E** (Lab 1 diagnoses, Lab 3 answers)
>
> *2. I need an **inventory-planning tool** that forecasts weekly sales.* &nbsp;→ Path **P** (Lab 2)
>
> *Answers must survive an audit by my CEO.*

<br>

.small[Three translator memos as deliverables: **T4** (Lab 1) · **T7** (Lab 2) · **T<sub>final</sub>** (Lab 3).]
]

.pull-right[
## `r fa('flask', fill='#D50006')` Lab 1

<div style="border:3px solid #D50006; border-radius:8px; padding:14px 18px;
            background:#fff5f5; margin-top:0.4em;">
<div style="font-family:sans-serif; font-size:13px; font-weight:bold;
            color:#D50006; margin-bottom:8px;">
`r fa('flask')` Lab&nbsp;1: Reality Check — Stress-testing OLS under A/B/C
</div>
<div style="font-size:0.78em; line-height:1.55;">
<strong>Learning objectives:</strong> Fit a baseline log-log OLS on Dominick's Store&nbsp;114 · Apply A/B/C diagnostics (autoplot, ACF, Durbin-Watson, VIF) · Choose robust&nbsp;SE / HAC when B2/B3 are violated · Trace OVB across 83 stores · Draft Memo&nbsp;T4 for Sarah's Q1
<br>
<strong>Dataset:</strong>
<code>tbl_stores</code> (Dominick's, 83 stores × 3 brands)
<br>
`r fa('clock')` <strong>45 minutes</strong> &nbsp;·&nbsp;
`r fa('users')` <strong>Individual work</strong>
<br>
`r fa('circle-right')` <strong>Open lab document:</strong>
<code>Session3_Lab1.qmd</code>
</div>
</div>
]

???
* Memo + Lab-box on one slide — the memo's two questions are the through-line of the entire afternoon
* Q1 is the causal question (Path E) — Lab 1 stress-tests E1, Lab 3 operationalises E2
* Q2 is the predictive question (Path P) — Lab 2 builds the tool
* The Lab 1 finale motivates Lab 3: classical OLS on historical data **cannot** answer Q1 even when fit perfectly — we need a designed experiment
* Open the .qmd and walk through Step 0 together; then individual work begins


---
class: inverse, center, middle

## Lab 1

.blockquote[Reality Check — stress-testing OLS]
