---
title: "Data Science and Strategy for Business"
subtitle: "ER114 — Session 3"
author: "Prof. Dr. Jörg Schoder"
institute: "FFHS"
date: "`r Sys.Date()`"
bibliography: ../../../lit/my_bib.bib
lang: en-US
reference-section-title: References
output:
  xaringan::moon_reader:
    self_contained: true
    css:
         - default
         - ../../../css/ffhs-theme_js.css
         - xaringan-themer.css
    includes:
      after_body: ../../../css/insert-logo.html
    lib_dir: ../../../libs
    nature:
      slideNumberFormat: "%current%/%total%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: false
---
class: title-slide

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_xaringan(text_color = "#d50006",inverse_text_color = "#FFFFFF",inverse_background_color = "#d50006", title_slide_background_color = "#d50006",header_background_color = "#d50006",header_color = "#FFFFFF",header_h1_font_size = "32px",
  header_h2_font_size = "26px",link_color="#502479",
  header_h3_font_size = "20px",text_slide_number_color = "#d50006",text_slide_number_font_size = "0.5em")
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_progress_bar(color = "#d50006", location = "bottom")
xaringanExtra::use_xaringan_extra(c("tile_view","scribble","panelset","tachyons"))
xaringanExtra::style_panelset_tabs(font_family = "inherit")
xaringanExtra::use_editable(expires = 1)
xaringanExtra::use_freezeframe(trigger = "hover")
```

```{r}
#| label: setup
#| include: false
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(kableExtra)
library(latex2exp)
library(fontawesome)
library(patchwork)
source(xfun::from_root("lit","helper.R"))
library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           style = "markdown",
           dashed = TRUE)
bib <- ReadBib(xfun::from_root("lit","my_bib.bib"))
Zitiere <- function(bib, key, ...) {
  gsub(" and ", " und ", Citet(bib, key, ...))
}
```

# ER114 — Data Science and Strategy for Business

## Session 3

### Predict or Intervene? Two Paths Beyond Inference

<br>
<br>
<br>
<br>
<br>
### FS 2026
<br>
### Prof. Dr. Jörg Schoder
.mycontacts[
`r fa('github')` @FFHS-EconomicResearch
`r fa('linkedin')` @jfschoder
]


---
layout: true

<div class="my-footer"></div>

<div style="position: absolute;left:400px;bottom:10px;font-size:9px">`r fa('creative-commons')``r rmarkdown::metadata$author`</div>


---
class: left

.blockquote[Programme]

## Today's Programme

<div style="display:grid; grid-template-columns:80px 1fr; gap:0.3em 1em;
            font-size:0.85em; margin-top:0.5em; align-items:center;">

<div style="color:#686868;">10'</div>
<div>`r fa('rotate-left')` <strong>Recap</strong> — where Sessions 1 and 2 left us</div>

<div style="color:#686868;">15'</div>
<div>`r fa('layer-group')` <strong>Two open questions</strong> — generalising and intervening</div>

<div style="color:#686868;">20'</div>
<div>`r fa('layer-group')` <strong>BLUE: the common bridge</strong> — Auer didactics, bias-variance trade-off, why both paths share the same foundation</div>

<div style="color:#686868;">15'</div>
<div>`r fa('layer-group')` <strong>Path A: Predictive Modelling</strong> — out-of-sample, multicollinearity, regularisation</div>

<div style="color:#D50006; font-weight:bold;">40'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab 1</strong> — out-of-sample on Dominick's, full-panel</div>

<div style="color:#686868; font-style:italic;">10'</div>
<div style="color:#686868; font-style:italic;">`r fa('mug-hot')` Break</div>

<div style="color:#D50006; font-weight:bold;">25'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab 2</strong> — Lasso for variable selection on demographics</div>

<div style="color:#686868;">20'</div>
<div>`r fa('layer-group')` <strong>Path B: Causal Modelling</strong> — Where the data come from. Exogeneity by assumption vs. by design</div>

<div style="color:#D50006; font-weight:bold;">25'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab 3</strong> — design an A/B test for the advertising effect</div>

<div style="color:#686868;">10'</div>
<div>`r fa('flag-checkered')` <strong>Synthesis & Outlook</strong></div>

</div>


---
class: inverse, center, middle

## Recap: Where we left off


---
class: left

.blockquote[The story so far]

## Sessions 1 → 2 → **3**

.pull-left[
### Session 1 — Describe
- Dominick's Store 51, three brands
- Tropicana / Minute Maid / Dominick's
- $n = 336$ observations
- Multiple regression, interaction terms
- Findings: brand effect ≈ 0.4 log-units; advertising effect ≈ 0.25 log-units
- *Population mode, no inference*
]

.pull-right[
### Session 2 — Test hypotheses
- Same data, new question: how reliable is what we found?
- Sample-splitting — and its limits at $n = 336$
- Replication on Store 50 — same advertising effect?
- Inference machine: bootstrap, t-test = ANOVA = regression
- *Stochastic mode, with inference*
]

📌 **Two unanswered questions remain.**


---
class: left

.blockquote[Two open questions]

## Two Questions Session 3 Tackles

.pull-left[
### Question 1
*Does our model also work for stores **we have never seen**?*

This is the **generalisation** question. It is the question of out-of-sample performance.

→ Path A: Predictive Modelling.
]

.pull-right[
### Question 2
*If the advertising effect is 0.25 — what happens **when we actively run advertising tomorrow**?*

This is the **intervention** question. It is the question of causal inference.

→ Path B: Causal Modelling.
]

<br>
<br>
📌 **The two paths are not in opposition. They are answers to different
questions.** And they share a common foundation: the BLUE/bias-variance
framework.



---
class: inverse, center, middle

## The Common Bridge: BLUE & Bias-Variance


---
class: left

.blockquote[Common bridge]

## Why BLUE Matters for Both Paths

.pull-left[
### What BLUE says
Under the Gauss-Markov assumptions (A1, A2, A3, B1, B2, B3, C1, C2),
OLS is the **B**est **L**inear **U**nbiased **E**stimator:

- **B**est = minimum variance
- among all **L**inear, **U**nbiased **E**stimators

Note the order: *first* unbiased, *then* minimum variance among the unbiased.
]

.pull-right[
### Why this matters for both paths
**For Path B (inference):** unbiasedness is what we need.
The estimate must reflect the true effect.

**For Path A (prediction):** unbiasedness is *not* what we need.
We want low total error — and that may require accepting some bias.

→ Same foundation, different optimality criterion.
]

.quellePanURL[cf. `r Citet(bib, "vonAuer_2023")`, p. 167ff]


---
class: left

.blockquote[Common bridge]

## Bias-Variance Decomposition: The Key Equation

$$\text{MSE}(\hat{Y}) = \underbrace{\text{Bias}^2(\hat{Y})}_{\text{systematic error}} + \underbrace{\text{Var}(\hat{Y})}_{\text{random error}} + \underbrace{\sigma_{\varepsilon}^2}_{\text{irreducible}}$$

.pull-left[
### Path A reasoning
Total error matters. So we *trade* bias for variance:

- accept some bias
- gain much more in variance reduction
- → lower MSE on **new data**

This is exactly what Lasso and Ridge do.
]

.pull-right[
### Path B reasoning
Bias is the enemy. Variance can be quantified (CIs, bootstraps).

- bias = invalid inference
- variance = honest uncertainty
- → BLUE is the gold standard
- → confounders ≠ noise, they are bias
]

📌 **Same theory — opposite direction.**


---
class: left

.blockquote[Common bridge]

## The Bias-Variance Trade-off, Visualised

```{r}
#| label: bias-var-tradeoff
#| echo: false
#| out-width: "55%"
#| fig-align: "center"
set.seed(42)
complexity <- seq(1, 10, by = 0.1)
bias2 <- 3 * exp(-0.5 * complexity)
variance <- 0.1 * exp(0.4 * complexity)
total <- bias2 + variance + 0.3

df <- data.frame(
  x = rep(complexity, 3),
  y = c(bias2, variance, total),
  type = rep(c("Bias²", "Variance", "Total error (MSE)"), each = length(complexity))
)

ggplot(df, aes(x, y, color = type, linetype = type)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = complexity[which.min(total)], linetype = "dashed", alpha = 0.5) +
  annotate("text", x = complexity[which.min(total)] + 0.3, y = 3.5,
           label = "Sweet spot", angle = 90, size = 3.5) +
  scale_color_manual(values = c("Bias²" = "#E74C3C", "Variance" = "#3498DB",
                                 "Total error (MSE)" = "#2C3E50")) +
  labs(x = "Model complexity", y = "Error", color = "", linetype = "") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

📌 The optimum is **not** at zero bias. It is where bias² + variance is minimal.



---
class: inverse, center, middle

## Path A: Predictive Modelling


---
class: left

.blockquote[Path A]

## The Predictive Question

> *"How well does my model predict for **new stores** I have never seen?"*

We now expand the data world from one store (Session 1+2) to **all 83
Dominick's stores** in the Chicago area. Plus their demographic data.

| Variable | Type |
|---|---|
| `log_units`, `log_price`, `feat`, `deal` | weekly observation per store-brand |
| `EDUC`, `INCOME`, `HVAL150`, `AGE60`, etc. | store-level demographics (12 variables) |

The big-data world is here: $n \approx 28{,}000$ observations, 12+
demographic predictors. **What was a teaching example is now a
practical scale.**


---
class: left

.blockquote[Path A]

## In-Sample $R^2$ Lies

.pull-left[
### The pattern
```{r}
#| label: insample-lie
#| echo: false
#| message: false
#| fig-height: 4
#| fig-width: 5
set.seed(42)
n <- 30
x <- sort(runif(n, 0, 10))
y <- sin(x) + rnorm(n, 0, 0.3)
df_fit <- data.frame(x = x, y = y)
ggplot(df_fit, aes(x, y)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 20), se = FALSE,
              color = "#D50006", linewidth = 1.4) +
  labs(title = "Polynomial degree 20 — perfect in-sample fit",
       subtitle = "$R^2 \\approx 0.99$ — but worthless for new data") +
  theme_minimal(base_size = 13)
```
]

.pull-right[
### The trap
*"My model has $R^2 = 0.94$!"* sounds impressive — until we test on
**new data**.

A model that perfectly memorises the training set:
- has **near-zero** bias on these data
- has **enormous** variance for any other dataset
- is **useless** for prediction

→ The cure: out-of-sample evaluation.
]


---
class: left

.blockquote[Path A]

## Train-Test Split: The Naive Solution

```{r}
#| label: train-test-figure
#| echo: false
#| out-width: "75%"
#| fig-align: "center"
knitr::include_graphics(xfun::from_root("img","PVA3","train_test_split.svg"))
```

- Single split (e.g. 80/20)
- Fit on 80%, evaluate on 20%
- **Problem:** the result depends on which 80% we picked


---
class: left

.blockquote[Path A]

## K-Fold Cross-Validation: The Gold Standard

```{r}
#| label: cv-figure
#| echo: false
#| out-width: "70%"
#| fig-align: "center"
knitr::include_graphics(xfun::from_root("img","PVA3","kfold_crossvalidation.svg"))
```

$$\text{CV-Score} = \frac{1}{K} \sum_{k=1}^{K} \text{MSE}_k$$

- $K = 5$ or $K = 10$ (standard)
- Each observation acts as test exactly once
- Robust against unlucky splits


---
class: left

.blockquote[Path A]

## Multicollinearity in Demographics

The demographics in Dominick's data are not independent. They cluster:

```{r}
#| label: demo-cor-matrix
#| echo: false
#| out-width: "60%"
#| fig-align: "center"
demo_cor <- matrix(c(
   1.00, -0.31, -0.09, -0.15, -0.33, -0.63, -0.11,
  -0.31,  1.00, -0.34,  0.66, -0.39,  0.55,  0.89,
  -0.09, -0.34,  1.00, -0.72,  0.24, -0.28, -0.42,
  -0.15,  0.66, -0.72,  1.00, -0.08,  0.40,  0.64,
  -0.33, -0.39,  0.24, -0.08,  1.00, -0.28, -0.47,
  -0.63,  0.55, -0.28,  0.40, -0.28,  1.00,  0.45,
  -0.11,  0.89, -0.42,  0.64, -0.47,  0.45,  1.00
), nrow = 7, byrow = TRUE)
rownames(demo_cor) <- colnames(demo_cor) <-
  c("AGE60","EDUC","ETHNIC","INCOME","HHLARGE","WORKWOM","HVAL150")

# Heatmap with values
df_cor <- as.data.frame(as.table(demo_cor))
names(df_cor) <- c("Var1","Var2","cor")

ggplot(df_cor, aes(Var1, Var2, fill = cor)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", cor)), size = 3) +
  scale_fill_gradient2(low = "#502479", mid = "white", high = "#D50006",
                        midpoint = 0, limits = c(-1,1)) +
  labs(x = "", y = "", fill = "r") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**EDUC ↔ HVAL150** = 0.89 — practically the same dimension.
**INCOME ↔ ETHNIC** = −0.72 — substantial overlap.



---
class: left

.blockquote[Path A]

## What Multicollinearity Does

.pull-left[
### For inference (Path B)
- $\hat{\beta}$ becomes **unstable**
- Standard errors blow up
- Confidence intervals widen
- Coefficients can switch signs with tiny data changes

→ **Multicollinearity is poison for explanation.**
]

.pull-right[
### For prediction (Path A)
- $\hat{\beta}$ unstable, but...
- ...$\hat{y}$ stays stable
- The model still predicts well
- Multicollinearity is **often tolerable** — or even useful

→ **For prediction, what matters is total error.**
]

📌 *Diagnose with VIF: $\text{VIF}_j = 1/(1 - R^2_j)$. Above 5: warning.
Above 10: serious problem.*


---
class: left

.blockquote[Path A]

## Lasso: Sparsity Through L1 Penalty

$$\min_{\beta} \sum_{i=1}^{n} (y_i - x_i^\top \beta)^2 + \lambda \sum_{j=1}^{p} |\beta_j|$$

.pull-left[
### What Lasso does
- Shrinks coefficients toward 0
- **Sets some exactly to 0** → variable selection
- Solves multicollinearity *and* high-dimensional p > n
- $\lambda$ chosen by cross-validation
]

.pull-right[
### What this means for our case
- Lasso on demographics will **automatically drop one of EDUC/HVAL150**
- The "wealth" cluster gets compressed into one variable
- Final model: 5–8 active demographics instead of 12
]

```r
library(glmnet)
cv_fit <- cv.glmnet(X, y, alpha = 1)
coef(cv_fit, s = "lambda.min")
```


---
class: left

.blockquote[Path A]

## Ridge vs. Lasso vs. Elastic Net

.pull-left[
### Ridge (L2)
$$\lambda \sum \beta_j^2$$
- Shrinks all
- Removes none
- Good for **multicollinearity** with many small effects

### Lasso (L1)
$$\lambda \sum |\beta_j|$$
- Shrinks **and** zeroes
- Variable selection
- Good for **sparsity** (few active variables)
]

.pull-right[
### Elastic Net
$$\lambda_1 \sum |\beta_j| + \lambda_2 \sum \beta_j^2$$
- Combines L1 + L2
- Selects *and* keeps groups of correlated features
- Often best in practice with `glmnet`

💡 In R: `cv.glmnet(X, y, alpha = 0.5)` for Elastic Net.
]

📌 The choice depends on the goal: variable selection (Lasso), stability
(Ridge), or balanced (Elastic Net).


---
class: left

.blockquote[Path A]

## The CV Path: Choosing $\lambda$

```{r}
#| label: lasso-cv-path
#| echo: false
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-align: "center"
set.seed(42)
n <- 200; p <- 8
X <- matrix(rnorm(n*p), n, p)
colnames(X) <- c("EDUC","INCOME","HVAL150","ETHNIC","AGE60","WORKWOM","HHLARGE","Random")
y <- 2*X[,1] + 1.5*X[,2] + 0.5*X[,3] - 1*X[,4] + 0.5*X[,5] + rnorm(n)

library(glmnet)
fit <- glmnet(X, y, alpha = 1)
beta_df <- as.data.frame(as.matrix(t(coef(fit)[-1,])))
beta_df$lambda <- fit$lambda
beta_long <- pivot_longer(beta_df, -lambda, names_to = "Variable", values_to = "Coefficient")

ggplot(beta_long, aes(log10(lambda), Coefficient, color = Variable)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3) +
  scale_color_brewer(palette = "Set2") +
  labs(x = expression(log[10](lambda)), y = "Coefficient",
       title = "Lasso path: how coefficients shrink as λ grows") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

- Each line is one feature
- As $\lambda$ grows, coefficients shrink and (some) hit 0
- **Cross-validation picks the $\lambda$** that minimises out-of-sample MSE


---
class: left

.blockquote[Lab 1 — Out-of-sample on Dominick's]

## `r fa('laptop-code', fill='#D50006')` Lab 1
<div style="border:3px solid #D50006; border-radius:8px; padding:20px 28px;
            background:#fff5f5; margin-top:1em;">
<div style="font-family:sans-serif; font-size:16px; font-weight:bold;
            color:#D50006; margin-bottom:10px;">
`r fa('flask')` Lab 1: Out-of-sample on Dominick's full panel
</div>
<div style="font-size:0.85em; line-height:1.8;">
<strong>Learning objective:</strong> Experience the gap between in-sample
and out-of-sample performance on the full 83-store Dominick's panel.
Apply train-test split and K-fold cross-validation.
<br>
<strong>Content:</strong>
fit on Store 51 → great $R^2$! → predict for Store 50: realistic.
Predict for a *random* store: 💥 — out-of-sample reality.
Then K-fold CV across all 83 stores.
<br>
`r fa('clock')` <strong>40 minutes</strong> &nbsp;·&nbsp;
`r fa('users')` <strong>Individual or pairs</strong>
<br>
`r fa('circle-right')` <strong>Open lab document:</strong>
<code>Session3_Lab1_OutOfSample_OJ.qmd</code>
</div>
</div>


---
class: left

.blockquote[Lab 2 — Lasso on demographics]

## `r fa('laptop-code', fill='#D50006')` Lab 2
<div style="border:3px solid #D50006; border-radius:8px; padding:20px 28px;
            background:#fff5f5; margin-top:1em;">
<div style="font-family:sans-serif; font-size:16px; font-weight:bold;
            color:#D50006; margin-bottom:10px;">
`r fa('flask')` Lab 2: Multicollinearity & regularisation
</div>
<div style="font-size:0.85em; line-height:1.8;">
<strong>Learning objective:</strong> Diagnose multicollinearity in
demographic variables, apply Lasso/Ridge/Elastic Net, choose $\lambda$
by cross-validation.
<br>
<strong>Content:</strong>
VIF on the 12 demographics → identify the wealth cluster.
Lasso path: which variables survive? Ridge for stable coefficients.
Compare the three approaches on out-of-sample MSE.
<br>
`r fa('clock')` <strong>25 minutes</strong> &nbsp;·&nbsp;
`r fa('users')` <strong>Individual or pairs</strong>
<br>
`r fa('circle-right')` <strong>Open lab document:</strong>
<code>Session3_Lab2_Regularisation_OJ.qmd</code>
</div>
</div>


---
class: inverse, center, middle

## Path B: Causal Modelling


---
class: left

.blockquote[Path B]

## The Intervention Question

> *"If we run advertising **tomorrow** for the Dominick's brand — what
> happens to sales?"*

This is *not* a prediction question. It is an **intervention** question.

The crucial difference:

| Prediction | Intervention |
|---|---|
| World stays as is | World changes through our action |
| Past correlations apply | Past correlations may break |
| Lasso, CV, OOS | Causal inference, RCT |

📌 **Lucas critique** (1976): a model trained on observational data may
collapse the moment we intervene.


---
class: left

.blockquote[Path B]

## Endogeneity: Why OLS Fails

When **C1 (exogeneity)** is violated:

$$\text{Cov}(X_i, \varepsilon_i) \neq 0$$

OLS is **biased and inconsistent**. The coefficient does not reflect
the causal effect.

Three classic sources:
1. **Omitted variable bias** (confounding): some Z affects both X and Y, but is missing
2. **Simultaneity / reverse causality**: X causes Y, but Y also causes X
3. **Measurement error in X**

In our Dominick's data: the perfect implication

$$\text{feat} = 1 \quad\Longrightarrow\quad \text{deal} = 1$$

means we **cannot disentangle** the advertising effect from the price-promotion effect — using observational data alone.


---
class: left

.blockquote[Path B]

## Where Do the Data Come From?

The validity of inference depends not primarily on the method, but on
the **type of data**.

| Type of data | Exogeneity? | Method |
|---|---|---|
| **Survey** (cross-section, longitudinal) | Plausibility argued, never guaranteed; risks: self-selection, non-response | OLS under MLR.4 — Wooldridge world |
| **Experiment** (RCT, A/B test) | Exogeneity *created* by randomisation | Comparison of means, regression adjustment |
| **Natural experiment** (IV, DiD, RDD) | Exogeneity argued via "as-if-random" allocation | IV / DiD / RDD as identification strategies |

📌 **Hernán's central message:** *"Causal inference is just inference with the right data."*

The method is the same OLS we know — what differs is the **data foundation**.


---
class: left

.blockquote[Path B]

## Two Routes to Exogeneity

.pull-left[
### Route 1 — Exogeneity by assumption
We **assume** that all relevant confounders are observed and included.

- The Wooldridge MLR.4 condition
- Rest of the world goes into the error term
- We **trust** that the error is mean-independent of X

→ Risk: A1 violation (missing variable) → C1 violation → biased.

→ Tool: **regression adjustment** with rich covariates.
]

.pull-right[
### Route 2 — Exogeneity by design
We **engineer** exogeneity through the data collection.

- RCT / A/B test: random assignment guarantees exogeneity
- IV: the instrument is exogenous by argument (lottery, weather, distance)
- DiD: parallel trends assumption replaces exogeneity
- RDD: discontinuity at threshold replaces exogeneity

→ Tool: **identification strategy** built into the data.
]

📌 **The two routes share the same regression machinery — the difference
is in the data.**


---
class: left

.blockquote[Path B]

## Internal vs. External Validity

To assess any causal design:

.pull-left[
### Internal validity
*Have we cleanly isolated causality in **our** study?*

- Did we control for all confounders?
- Is the design mathematically tight?
- The enemy: omitted variable bias, selection bias.
]

.pull-right[
### External validity
*Does the result transfer to **the real market**?*

- Is our sample representative?
- Do the test conditions match deployment conditions?
- The enemy: artificial test environments, narrow sampling frames.
]

📌 **The Holy Grail: a design that maximises both.**


---
class: left

.blockquote[Path B]

## A/B Testing: The Gold Standard

In the technology and business world, a Randomised Controlled Trial is
called an **A/B test**.

.pull-left[
### How it heals C1
1. Users assigned by **strict randomisation** to A (status quo) or B (treatment)
2. Randomness guarantees the treatment is independent of all hidden confounders
3. **C1 is artificially restored** — by design

### Cohen's d for effect size
$$d = \frac{\bar{X}_B - \bar{X}_A}{s_{\text{pooled}}}$$
| d | Interpretation |
|:--:|:---|
| 0.2 | small |
| 0.5 | medium |
| 0.8 | large |
]

.pull-right[
### Schrage's principle
*"Run experiments not because you can confidently predict outcomes —
but precisely because you can't."*

Microsoft's Ronny Kohavi: only **⅓** of A/B-tested features improve
metrics. Even experts get the answer wrong **80%** of the time.

→ **Conviction ≠ evidence.**

📌 **In Lab 3:** Design an A/B test for the Dominick's advertising
effect. Sketch the randomisation, define the metric, compute Cohen's d
under realistic effect sizes.
]


---
class: left

.blockquote[Path B]

## Quasi-Experiments: When RCTs Are Impossible

Sometimes ethics, law, or technology rule out a clean A/B test. The
data team then turns to **quasi-experimental designs**:

| Design | Idea | When? |
|---|---|---|
| **Instrumental Variables (IV)** | A *random* element shifts X but only affects Y through X | Compliance with random assignments; lottery, distance, weather |
| **Difference-in-Differences (DiD)** | Compare *trends* of treated and control before/after a policy change | Geographic rollout, regulatory change |
| **Regression Discontinuity (RDD)** | Compare units just above and just below a hard cutoff | Bonus thresholds, age cutoffs, eligibility scores |

📌 The **Critical Validator** (manager) must recognise *which* quasi-
experimental tool fits the situation — and demand it.


---
class: left

.blockquote[Lab 3 — Design an A/B test]

## `r fa('laptop-code', fill='#D50006')` Lab 3
<div style="border:3px solid #D50006; border-radius:8px; padding:20px 28px;
            background:#fff5f5; margin-top:1em;">
<div style="font-family:sans-serif; font-size:16px; font-weight:bold;
            color:#D50006; margin-bottom:10px;">
`r fa('flask')` Lab 3: Design an A/B test for the advertising effect
</div>
<div style="font-size:0.85em; line-height:1.8;">
<strong>Learning objective:</strong> Experience the limit of observational
data for causal claims. Design a clean A/B test that breaks the feat ↔
deal entanglement.
<br>
<strong>Content:</strong>
Why does observational analysis fail here? Sketch the A/B design.
What is randomised, what is the metric, what sample size? Compute
Cohen's d under realistic effect sizes.
<br>
`r fa('clock')` <strong>25 minutes</strong> &nbsp;·&nbsp;
`r fa('users')` <strong>Individual or pairs</strong>
<br>
`r fa('circle-right')` <strong>Open lab document:</strong>
<code>Session3_Lab3_Experiment_OJ.qmd</code>
</div>
</div>


---
class: inverse, center, middle

## Synthesis & Outlook


---
class: left

.blockquote[Synthesis]

## Two Demands the Translator Should Make

As an *AI Value Creator* you no longer let yourself be dazzled by
in-sample $R^2$ or bare p-values. **Before budgets are released:**

.pull-left[
### When you predict...
*(Path A)*
- The danger: **overfitting** on training data
- The diagnosis: high VIF, complex model
- Your demand: *"Show me out-of-sample performance via cross-validation. Use Lasso."*
]

.pull-right[
### When you intervene...
*(Path B)*
- The danger: **Lucas critique** — past correlations may break
- The diagnosis: missing variables, observational data only
- Your demand: *"Design an A/B test or find a credible quasi-experimental identification."*
]


---
class: left

.blockquote[Synthesis]

## Six Foundations Bridge — Active in Session 3

| # | Classical tradition | Bridge | Data Science |
|---|---|---|---|
| 1 | E[Y\|X] — conditional mean | market benchmark | residual as arbitrage |
| 2 | sampling distribution | uncertainty | bootstrap, infer |
| 3 | expected value & error types | ROI calculus | EV framework |
| 4 | p-value, multiplicity | signal vs. noise | FDR, p-hacking, HARKing |
| 5 | correlation ≠ causality | intervention | A/B testing, causal ML |
| **6** | **BLUE, Gauss-Markov, bias-variance** | **complexity control** | **CV, Lasso, regularisation** |

📌 **Foundation 6 is what we have built today.**


---
class: left

.blockquote[Outlook]

## Outlook: What Comes Next

.pull-left[
### ER017 — Machine Learning
- Logit / GLM (the binary case)
- Random Forests, gradient boosting
- Causal ML (Causal Forests, double ML)
- Multilevel / hierarchical models
]

.pull-right[
### Time-series methods
- Autocorrelation in panel data
- ARIMA, exponential smoothing
- Forecasting metrics
- Block-bootstrap, time-series CV

### Beyond
- Causal discovery (DAG inference)
- Reinforcement learning
- Bayesian methods
]

→ Many doors open. The ER114 toolkit is the **foundation**.


---
class: left

.blockquote[Closing]

## The Translator's Role

Two sessions taught us to **describe** and **infer**.

This third session taught us **two paths beyond inference**:

- **predict** under stable structure
- **intervene** by changing the data foundation

> *"An interpretable model translates raw data into information that is
> directly relevant to decision-making."* — Matt Taddy

> *"Causal inference is just inference with the right data."* — Miguel Hernán

📌 **The value creation of the future does not arise from blindly
applying algorithms. It arises from translators who critically validate
models and steer them strategically.**


---
class: inverse, center, middle

# Thank you!

### Questions? → `r fa('comments')` Discussion & feedback
