<!-- ============================================================
INSTRUCTOR CUSTOMISATION — update before rendering for each cohort
Line references are approximate; search for the marker text.

  ~Line  18   Moodle download link for dominicks_s114_YYYYMMDD.rds
  ~Line  19   Moodle download link for dominicks_stores_YYYYMMDD.rds
  ~Line  20   Moodle download link for dominicks_demo_YYYYMMDD.rds
  ~Line  ~50  Date stamp in read_rds() calls (must match the actual files)

============================================================ -->
---
title: "Session 3 · Lab 2: From OLS to Lasso — Bias-Variance and Variable Selection"
subtitle: "CRISP-DM Phases 4–5 --- Modelling and Evaluation under Regularisation"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en-US
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
bibliography: ../../lit/my_bib.bib
---

```{r setup}
#| include: false
library(tidyverse)        # data wrangling and plotting
library(gt)               # publication-quality tables
library(moderndive)       # get_regression_table() / get_regression_summaries()
library(broom)            # tidy model output for one-off computations
library(glmnet)           # Lasso, Ridge, Elastic Net
library(patchwork)        # combine ggplots
library(car)              # VIF for multicollinearity diagnostics

# shared visual identity
theme_set(theme_minimal(base_size = 11))

# colour palette
col_primary   <- "#D50006"   # main highlight
col_secondary <- "#502479"   # contrast
col_neutral   <- "#686868"   # grey reference

# load data — date stamp must match the files in data/raw/
date_stamp <- "20260517"  # adjust to the most recent DataImport.R run
#tbl_s114    <- read_rds(xfun::from_root("data", "raw",
#                     glue::glue("dominicks_s114_{date_stamp}.rds")))
tbl_stores  <- read_rds(xfun::from_root("data", "raw",
                       glue::glue("dominicks_stores_{date_stamp}.rds")))
tbl_demo    <- read_rds(xfun::from_root("data", "raw",
                       glue::glue("dominicks_demo_{date_stamp}.rds")))

# convenience subsets — filtered on-the-fly from the full panel
tbl_s114 <- tbl_stores %>% filter(store == 114)
tbl_s83  <- tbl_stores %>% filter(store == 83)

# join into a single panel
tbl_full <- tbl_stores %>%
              left_join(tbl_demo, by = "store")
```

## Scenario — A Second Memo from Sarah Chen

Lab 1 ended with two memos to Sarah. This morning, a follow-up
arrives.

:::{.callout-note icon=false collapse="false"}
## 📧 Memo from Sarah Chen, VP Marketing — Dominick's Finer Foods

*"Quick follow-up on yesterday's work. I should have mentioned earlier:
we also have census-based demographics for every store catchment area
— median income, education levels, age distribution, that kind of
thing. Eleven variables per store.*

*Two questions:*

*1. **Would adding those demographics improve the inventory prediction
tool you're building?** Operations is asking how confident we can be
in the model. The board likes more variables ('better information'),
but I want a defensible answer, not just more numbers.*

*2. **Does adding demographics tell us anything new about the
advertising effect?** I know yesterday you said the historical data
alone can't give us a clean causal estimate. But if we had richer
controls — does that change the picture?"*

— Sarah
:::

Sarah's two questions correspond to the two methodological paths from
Lab 1:

- Question 2 is a **Path E1** question — does richer control variation
  resolve the omitted-variable concern from Lab 1?
- Question 1 is a **Path P** question — does adding 11 demographic
  variables improve out-of-sample prediction?

This Lab is dedicated to the board's intuition — *more variables means better model results*. 
We will scrutinize this intuition using empirical methods.

```{r crisp-overview}
#| echo: false

tibble::tribble(
  ~Phase, ~Name,                      ~Content,                                                 ~Status,
  "1",    "Business Understanding",   "Marketing decisions: explain and predict",               "Done (Lab 1)",
  "2",    "Data Understanding",       "OJ panel + store demographics (← here)",                 "Refresh in Step 0",
  "3",    "Data Preparation",         "Reshape for cross-price modelling (← here)",             "← This lab (1, 2)",
  "4",    "Modelling",                "Regularisation: Ridge, Lasso, Elastic Net (← here)",     "← This lab (3, 4)",
  "5",    "Evaluation",               "Out-of-sample evaluation via CV (← here)",               "← This lab (4, 5)",
  "6",    "Deployment",               "Memo to Sarah: which variables matter for prediction?",  "← This lab (6)"
) %>%
  gt() %>%
  gt::tab_style(gt::cell_fill("#f0fff0"),
                gt::cells_body(rows = Phase %in% c("3","4","5","6"))) %>%
  gt::tab_style(gt::cell_fill("#f5f0fa"), gt::cells_column_labels()) %>%
  gt::tab_options(table.font.size = "13px")
```

## Navigation

::: {.nav-table}

| Symbol | Box | Meaning | When to read |
|:------:|:----|:--------|:-------------|
| ⚠️  | [**Red — Action Required**]{.nav-red} | Mandatory tasks | During live session |
| 💡 | [**Green — Best Practice / Tip**]{.nav-green} | Coding hints, shortcuts | Skim during lab |
| 📚 | [**Blue, collapsed — Background**]{.nav-blue} | Concepts, context, derivations | At home |
| ℹ️ | [**Blue, open — Note**]{.nav-blue} | Short clarification | Live |

:::

---

# Step 0 — Bridge from Lab 1 and What This Lab Adds

**The same DGP, two paths, new challenges.** Lab 1 introduced the
**data-generating process (DGP)** as the conceptual scaffold for
all of Session 3. We worked with `tbl_s114` (one store) and
`tbl_stores` (83 stores × 3 brands) and saw two operations on the
*same* DGP: Path E1 asks for the structural parameters, Path P
asks for the DGP's outputs under stable structure.

Lab 2 takes **Path P** seriously. We will see that **OLS itself — the
BLUE estimator from Lab 1 — is suboptimal for Path P**. The reason:
optimal prediction does not require unbiasedness. By accepting a
small bias, we can dramatically reduce variance and produce a more
accurate inventory tool for Sarah.

The same DGP, the same data — but a different optimisation target,
and therefore a different best estimator.

## Bringing additional Store-level data into our Analysis

Today we add a third dataset **`tbl_demo`** with eleven additional predictors for each store.


```{r data-summary}
#| echo: false

tibble::tribble(
  ~Dataset,       ~Source,                        ~Rows,                ~Columns, ~`Used for`,
  "tbl_s114",     "Lab 1: single-store analysis", as.character(nrow(tbl_s114)),    as.character(ncol(tbl_s114)),    "Step 1 reference",
  "tbl_stores",   "Lab 1: 83 stores × 3 brands",  as.character(nrow(tbl_stores)),  as.character(ncol(tbl_stores)),  "Steps 1–6",
  "tbl_demo",     "New: census demographics",     as.character(nrow(tbl_demo)),    as.character(ncol(tbl_demo)),    "Steps 1, 3–6",
  "tbl_full",     "Joined: tbl_stores + tbl_demo", as.character(nrow(tbl_full)),   as.character(ncol(tbl_full)),    "Steps 3–6"
) %>%
  gt() %>%
  gt::tab_options(table.font.size = "13px")
```

The store-level data provided information about the *catchment area* of each store, such as the age structure and income situation in the neighborhood, or the competitive environment. or brevity we call all additional variables from `tbl_demo` demographics $(X_{D,s})$. Each store has one row in `tbl_demo`:
```{r demo-preview}
#| echo: false
tbl_demo %>%
  head(3) %>%
  gt() %>%
  gt::tab_header(title = "tbl_demo — first three rows (11 demographics per store)") %>%
  gt::tab_options(table.font.size = "13px")
```

:::{.callout-note icon="false" collapse="true" #demo-variables}
### 📚 Background: What each demographic variable measures

| Variable | Meaning | Unit |
|:---|:---|:---|
| `age60`     | Proportion of population aged 60+ | proportion (0–1) |
| `educ`      | Proportion with college degree | proportion (0–1) |
| `ethnic`    | Proportion non-white | proportion (0–1) |
| `income`    | Log of median household income | log USD |
| `hhlarge`   | Proportion of households with ≥ 5 persons | proportion (0–1) |
| `workwom`   | Proportion of working women | proportion (0–1) |
| `hval150`   | Proportion of houses valued > $150 000 | proportion (0–1) |
| `sstrdist`  | Distance to nearest warehouse store | miles |
| `sstrvol`   | Sales-ratio: nearest warehouse store / this store | ratio |
| `cpdist5`   | Average distance to 5 nearest competing supermarkets | miles |
| `cpwvol5`   | Sales-ratio: 5 nearest competitors / this store | ratio |

Source: Dominick's Finer Foods marketing data, originally distributed
with bayesm and analysed extensively in the marketing literature
[@KiltsCenter_2018].
:::

When we merge `tbl_stores` with `tbl_demo` using the `left_join()`-function, each weekly
observation in a store inherits its store's demographics:

```{r demo-join}
# Merge tbl_stores and tbl_demo
tbl_full <- tbl_stores %>%
              left_join(tbl_demo, by = "store")
tbl_full %>%
  head(3) %>%
  gt() %>%
  gt::tab_header(title = "tbl_full — first three rows of merged data") %>%
  gt::tab_options(table.font.size = "13px")
```


## Three modelling options for brand-specific demographics

There are three ways to incorporate demographics into the prediction
model. The choice affects both interpretability and parameter
dimensionality:

| Option | Specification | Parameters | Trade-off |
|:---|:---|:---:|:---|
| **A** | `logmove ~ X + brand` (additive) | ≈ 15 | Demographics affect all brands identically |
| **B** | Separate model per brand | ≈ 12 each | Brand-specific effects but small samples |
| **C** | `logmove ~ X + brand + X:brand` (interactions) | ≈ 36 | Brand-specific effects in one model; high p |

The choice was previewed in the lecture slides. **In this lab we use
Variant C** — brand × demographic interactions in a single model. At
p ≈ 36, regularisation becomes necessary, not optional. This is where
the strength of Path P emerges most clearly.

:::{.callout-note icon="false" collapse="true" #variant-choice}
### 📚 Why Variant C — a short rationale

Variant A is simpler but assumes that, for instance, the income effect
on `logmove` is identical for Tropicana and the Dominick's house brand.
That is implausible: premium brands typically have steeper income
gradients.

Variant B avoids this assumption but at the cost of three separate
small-sample models, with no easy way to compare effects across brands.

Variant C captures the brand-specific demographic effects in one model
and pushes us into the high-dimensional regime where regularisation
becomes essential — exactly the lesson this lab is built around.
:::

### What we will do — and why in this order

| Step | Question | Path |
|:---|:---|:---:|
| 1 | Does adding demographics change the `feat` estimate from Lab 1? | E1 |
| 2 | Cross-price elasticities — do competitor prices matter? | E1 / P |
| 3 | OLS on the full model — what does it tell us in-sample? | P |
| 4 | The same model out-of-sample: better or worse? | P |
| 5 | Lasso — automatic variable selection | P |
| 6 | Reading the results: a sparse story for Sarah | P |

---

# Step 1 — Demographics as Path E1 Controls

We start by closing the loop on **Sarah's Question 2**: does richer
control variation resolve the omitted-variable concern from Lab 1?

Recall the baseline result on the full panel from Lab 1:

```{r recap-lab1}
#| code-fold: false
# baseline model from Lab 1 — no demographics, no deal
mod_baseline_full <- lm(logmove ~ feat + log(price) + brand, data = tbl_stores)

# full model with deal (Lab 1 Step 1b)
mod_with_deal     <- lm(logmove ~ feat + deal + log(price) + brand, data = tbl_stores)

bind_rows(
  tidy(mod_baseline_full) %>% mutate(model = "Baseline (Lab 1)"),
  tidy(mod_with_deal)     %>% mutate(model = "With deal (Lab 1)")
) %>%
  filter(term == "feat") %>%
  select(model, estimate) %>%
  mutate(estimate = round(estimate, 3)) %>%
  gt() %>%
  gt::tab_options(table.font.size = "13px")
```

The Lab 1 finding: adding `deal` reduces $\hat\beta_{\text{feat}}$ by
about **`r round(coef(mod_baseline_full)["feat"] - coef(mod_with_deal)["feat"], 3)`** log units —
this is the omitted-variable bias from the structural feat × deal
bundling.

Now we test whether **demographic controls** make a similar difference:

```{r mod-with-demos}
#| code-fold: false
mod_with_demos <- lm(
  logmove ~ feat + deal + log(price) + brand +
    age60 + educ + ethnic + income + hhlarge + workwom +
    hval150 + sstrdist + sstrvol + cpdist5 + cpwvol5,
  data = tbl_full
)

bind_rows(
  tidy(mod_baseline_full)  %>% mutate(model = "Baseline (Lab 1)"),
  tidy(mod_with_deal)      %>% mutate(model = "With deal (Lab 1)"),
  tidy(mod_with_demos)     %>% mutate(model = "With deal + demographics")
) %>%
  filter(term == "feat") %>%
  select(model, estimate) %>%
  mutate(estimate = round(estimate, 3)) %>%
  gt() %>%
  gt::tab_options(table.font.size = "13px")
```

:::{.callout-important icon=false}
## ⚠️ T1: What do you see?

a. By how much does $\hat\beta_{\text{feat}}$ change when demographics
   are added to the with-deal model?
b. Compare this to the OVB shift from Lab 1 (baseline → with-deal).
   Which control variable changed the estimate more?
c. What does this tell you about Sarah's Question 2 — does richer
   demographic control resolve the advertising-effect question?

Note your answers before opening the solution box.
:::

:::{.callout-note collapse="true" icon=false color="success"}
### 🟢 Interpretation (Click to unfold)

Adding the 12 demographics changes $\hat\beta_{\text{feat}}$ by only a
fraction of a log unit — far less than the shift caused by adding
`deal`. The structural feat × deal bundling is the dominant source of
OVB; demographic variation in store catchment areas does *not*
substantially confound the advertising effect.

**This is a defensible finding for Sarah's Question 2.** Demographics
do not "rescue" the causal interpretation. The advertising effect
remains identified only up to its bundling with deal — and that
identification problem requires a study design (Lab 3), not more
control variables.

:::{.callout-tip icon=false}
## 💡 Could demographics still matter causally?

Demographics could plausibly cause `logmove` through several channels:

- **Income**: higher-income areas may have different brand-loyalty
  patterns
- **Age**: older shoppers may prefer established brands like Tropicana
- **Household size**: large households may buy more juice per shopping trip

These mechanisms make the demographics legitimate causal predictors
in their own right. The empirical finding here is more subtle: while
demographics *do* explain variation in `logmove` (some of their
coefficients are highly significant), they do not change the
estimated advertising effect. Causal interpretation of $\hat\beta_{\text{feat}}$
is therefore not blocked by demographic confounding.
:::
:::

The Path E1 question is answered. From Step 2 onwards we are firmly
on Path P — out-of-sample prediction.

---

# Step 2 — Cross-Price Elasticities

In Lab 1 we modelled `log(price)` as the *own* brand's price. But
shoppers don't see brands in isolation: on the supermarket shelf,
Tropicana, Minute Maid and Dominick's compete for the same customer.
If Tropicana is discounted, some Dominick's customers will switch.

This is a **substitution effect**. Capturing it requires the prices
of competing brands in the same regression. The bayesm data lets us
do this for our three brands (with a small caveat — see below).

### Reshape: from long to semi-wide

Our `tbl_stores` has one row per (store, brand, week). For cross-price
modelling we want one row per (store, week), with prices and sales of
all three brands as columns.

```{r reshape-cross}
#| code-fold: false
tbl_cross <- tbl_stores %>%
  select(store, week, brand, logmove, price, feat, deal) %>%
  pivot_wider(
    names_from  = brand,
    values_from = c(logmove, price, feat, deal),
    names_glue  = "{.value}_{brand}"
  )

# inspect structure
tbl_cross %>%
  head(3) %>%
  select(store, week,
         logmove_dominicks, logmove_minute.maid, logmove_tropicana,
         price_dominicks,   price_minute.maid,   price_tropicana) %>%
  gt() %>%
  gt::tab_options(table.font.size = "11px")
```

`tbl_cross` has `r nrow(tbl_cross)` rows — one per (store, week)
combination — and `r ncol(tbl_cross)` columns.

### Cross-price model for Dominick's

We fit a model for one brand at a time. Take Dominick's first:

```{r mod-cross-dominicks}
#| code-fold: false
mod_cross_dom <- lm(
  logmove_dominicks ~
    log(price_dominicks) + log(price_minute.maid) + log(price_tropicana) +
    feat_dominicks   + deal_dominicks +
    feat_minute.maid + deal_minute.maid +
    feat_tropicana   + deal_tropicana,
  data = tbl_cross
)

mod_cross_dom %>%
  moderndive::get_regression_table() %>%
  gt() %>%
  gt::tab_options(table.font.size = "12px")
```

The three price coefficients are the **price elasticities**: how
much `logmove_dominicks` changes per 1 % change in each brand's
price.

:::{.callout-important icon=false}
## ⚠️ T2: Read the elasticity table

a. What sign does the own-price elasticity have for Dominick's? Is the
   magnitude larger or smaller than the value from Lab 1
   (≈ `r round(coef(mod_with_deal)["log(price)"], 2)`)?
b. What sign do the *cross*-price elasticities (Minute Maid, Tropicana)
   have? What does the sign say about substitution?
c. Compare the magnitude of the cross-price elasticities to the own-price
   elasticity. Which competitor moves Dominick's sales more?
:::

:::{.callout-note collapse="true" icon=false color="success"}
### 🟢 Interpretation (Click to unfold)

- The **own-price elasticity** is typically negative and *less elastic*
  (smaller in absolute value) than in Lab 1. The Lab-1 model omitted
  competitor prices, so the own-price coefficient absorbed both the
  direct price effect and the (correlated) competitor-price changes.
- **Cross-price elasticities** are positive — when a competitor raises
  its price, Dominick's gains sales. This is the substitution effect.
- The magnitudes show which competitors are closest substitutes for
  Dominick's. Typically, Minute Maid is closer (mid-tier) than the
  premium Tropicana.

**Methodological observation:** This is a Path-E1 improvement —
including a confounder (competitor price) made the own-price estimate
more credible. But it is also a Path-P improvement — the model now
captures variation that the Lab-1 baseline missed.
:::

:::{.callout-warning icon=false collapse="true" #cross-price-data-gap}
## ⚠️ Data gap reminder

The bayesm `orangeJuice` dataset originally contains 11 brand prices
per observation. In our reduction to the top three brands, we kept
only their own prices. **Cross-price effects from the 8 omitted
brands (Florida Natural, Citrus Hill, etc.) are lost.**

For Dominick's, Minute Maid and Tropicana — the three best-selling
brands — the competitive pressure is reasonably captured by our
three included brands. For a complete competitive analysis, the
full 11-brand price matrix would be needed.

This is a typical data-gap situation: we work with what we have, and
honestly report what we cannot say.
:::

### Cross-price models for the other two brands

The same specification applied to Minute Maid and Tropicana:

```{r mod-cross-all}
#| code-fold: true
#| code-summary: "Show code: cross-price models for all three brands"
mod_cross_mm <- lm(
  logmove_minute.maid ~
    log(price_dominicks) + log(price_minute.maid) + log(price_tropicana) +
    feat_dominicks   + deal_dominicks +
    feat_minute.maid + deal_minute.maid +
    feat_tropicana   + deal_tropicana,
  data = tbl_cross
)

mod_cross_trop <- lm(
  logmove_tropicana ~
    log(price_dominicks) + log(price_minute.maid) + log(price_tropicana) +
    feat_dominicks   + deal_dominicks +
    feat_minute.maid + deal_minute.maid +
    feat_tropicana   + deal_tropicana,
  data = tbl_cross
)

# extract only the three price coefficients per brand
bind_rows(
  tidy(mod_cross_dom)  %>% filter(stringr::str_detect(term, "price")) %>%
    mutate(brand = "dominicks"),
  tidy(mod_cross_mm)   %>% filter(stringr::str_detect(term, "price")) %>%
    mutate(brand = "minute.maid"),
  tidy(mod_cross_trop) %>% filter(stringr::str_detect(term, "price")) %>%
    mutate(brand = "tropicana")
) %>%
  select(brand, term, estimate) %>%
  mutate(estimate = round(estimate, 2)) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  gt() %>%
  gt::tab_header(title = "Price elasticities (own and cross) by brand") %>%
  gt::tab_options(table.font.size = "12px")
```

The diagonal (own-price elasticities) is negative for all three brands.
Off-diagonal cross-price elasticities are positive — substitution works
in both directions.

This setup so far is still well within the comfort zone of classical
OLS. From Step 3 we push further: add demographics and their
brand-specific interactions, and the model dimension grows to the
point where OLS starts to fail.

---

# Step 3 — The Full Specification: OLS Under Strain

We now build the full specification: cross-brand prices, marketing
levers, demographics, and brand × demographic interactions. This is
Variant C from Step 0.

### Building the full design matrix

We work back on `tbl_full` (long format with demographics joined) and
fit a single model with all relevant variables and brand interactions:

```{r mod-full-spec}
#| code-fold: false
mod_full_spec <- lm(
  logmove ~ feat * brand + deal * brand + log(price) * brand +
    age60 * brand + educ * brand + ethnic * brand + income * brand +
    hhlarge * brand + workwom * brand + hval150 * brand +
    sstrdist * brand + sstrvol * brand + cpdist5 * brand + cpwvol5 * brand,
  data = tbl_full
)

# parameter count
n_params <- length(coef(mod_full_spec))
cat("Number of parameters in full specification: ", n_params, "\n")
```

The full specification has **`r length(coef(mod_full_spec))` parameters**.
On a total sample of n ≈ 28 000, this is still in the large-sample
regime. But — as we will see — the *out-of-sample* picture is
different.

:::{.callout-note collapse="true" #fhl-four-problems}
### 📚 Background: Four problems of big-data regression (Fan, Han, Liu 2014)

Even when $n$ is comfortably large relative to $p$, four issues
from the big-data literature [@fan_bigdata_2014] become operative
as we add parameters:

1. **Noise accumulation.** With more parameters, the variance of
   the estimator grows. Even if each additional variable has a
   small true effect, its sample estimate contributes noise that
   compounds.
2. **Spurious correlation.** In high dimensions, the probability
   of finding apparently strong correlations *by chance* increases
   rapidly. Lasso (Step 5) will retain some of these spurious
   selections.
3. **Incidental endogeneity.** Some predictors may be correlated
   with omitted causes — even when they have no causal effect on
   $y$, they pick up the omitted signal. The regression's causal
   interpretation breaks silently.
4. **Measurement errors.** Census-based demographic variables
   measure what they measure imperfectly. With more such variables,
   compound measurement noise grows.

These four problems frame why **regularisation is not enough**.
Lasso handles (1) and partially (2); it does not solve (3) and (4).
Reading Lasso-selected variables as "important predictors" must
always be qualified by these limitations.
:::

### In-sample summary

```{r in-sample-stats}
#| code-fold: false
broom::glance(mod_full_spec) %>%
  select(r.squared, adj.r.squared, sigma, AIC, BIC, nobs) %>%
  mutate(across(where(is.numeric), \(x) round(x, 3))) %>%
  gt() %>%
  gt::tab_options(table.font.size = "12px")
```

The in-sample R² is **`r round(summary(mod_full_spec)$r.squared, 3)`**.
The in-sample RMSE is **`r round(sqrt(mean(residuals(mod_full_spec)^2)), 3)`**.
Compared to the Lab 1 baseline RMSE on the full panel
(`r round(sqrt(mean(residuals(mod_baseline_full)^2)), 3)`), the full
specification fits the training data much better.

:::{.callout-important icon=false}
## ⚠️ T3: In-sample vs. out-of-sample expectations

a. The in-sample R² has improved substantially from Lab 1. Should we
   conclude that the new model is a better prediction tool for
   Sarah's inventory planning?
b. What is the difference between *in-sample* performance and
   *out-of-sample* performance? Why does the distinction matter for
   Sarah's question?
c. Predict (before computing) what will happen if we evaluate
   `mod_full_spec` via 5-fold cross-validation across stores.
:::

:::{.callout-note collapse="true" icon=false color="success"}
### 🟢 The trap of in-sample R² (Click to unfold)

A higher in-sample R² tells us only that the model fits the data we
*already have*. The relevant question for Sarah's inventory tool is
how well the model will predict in stores or weeks the model has
*not seen*. These two quantities can diverge sharply when:

- The number of parameters is large relative to the effective sample
  size (clustering reduces effective n)
- Predictors are highly correlated (the educ / income / hval150
  cluster from the demographics)
- The model captures sample-specific noise that does not generalise

This is **overfitting**. The next step measures it directly.
:::

---

# Step 4 — Cross-Validation: OLS in the Out-of-Sample Test

We now apply the K-fold CV protocol from Lab 1 — splitting at the
store level — to both the Lab-1 baseline and the full specification.
The cluster-robust CV strategy (whole-store assignment to folds, as
explained in Lab 1's `#cluster-cv` background box) protects against
the data leakage that would arise from random observation-level
splitting in this panel setting. This gives us a fair comparison of
generalisation performance.

```{r cv-comparison}
#| code-fold: false
set.seed(42)
K <- 5
stores_shuffled <- sample(unique(tbl_full$store))
folds <- cut(seq_along(stores_shuffled), breaks = K, labels = FALSE)

cv_rmse <- function(formula_str, label) {
  rmses <- map_dbl(1:K, \(k) {
    test_st  <- stores_shuffled[folds == k]
    train_st <- stores_shuffled[folds != k]
    mod_k    <- lm(as.formula(formula_str),
                   data = tbl_full %>% filter(store %in% train_st))
    pred_k   <- predict(mod_k,
                        newdata = tbl_full %>% filter(store %in% test_st))
    actual_k <- tbl_full %>% filter(store %in% test_st) %>% pull(logmove)
    sqrt(mean((pred_k - actual_k)^2))
  })
  tibble(model = label, mean_rmse = mean(rmses), sd_rmse = sd(rmses))
}

formula_baseline <- "logmove ~ feat + log(price) + brand"
formula_full     <- "logmove ~ feat * brand + deal * brand + log(price) * brand +
                       age60 * brand + educ * brand + ethnic * brand + income * brand +
                       hhlarge * brand + workwom * brand + hval150 * brand +
                       sstrdist * brand + sstrvol * brand + cpdist5 * brand + cpwvol5 * brand"

cv_results <- bind_rows(
  cv_rmse(formula_baseline, "OLS — baseline (Lab 1)"),
  cv_rmse(formula_full,     "OLS — full specification")
) %>%
  mutate(across(where(is.numeric), \(x) round(x, 3)))

cv_results %>%
  gt() %>%
  gt::tab_options(table.font.size = "12px")
```

:::{.callout-important icon=false}
## ⚠️ T4: Compare in-sample and out-of-sample

a. The baseline model from Lab 1 had a much lower in-sample R² than
   the full specification. How do their cross-validated RMSEs
   compare?
b. Did adding 36 parameters improve or worsen out-of-sample
   prediction?
c. If the answer surprises you, what does this tell you about the
   board's intuition that "more variables = better information"?
:::

:::{.callout-note collapse="true" icon=false color="success"}
### 🟢 The OLS limit (Click to unfold)

The full specification has lower in-sample RMSE but typically *higher*
out-of-sample RMSE than the baseline. The 36 parameters extract patterns
from the training stores that do not transfer to the test stores —
particularly the demographic interactions, which are partly identified
from a small number of stores with extreme demographic profiles.

This is the central tension of Path P:

> *In sample, more parameters always look better. Out of sample, more
> parameters often look worse.*

The cure is **regularisation** — methods that deliberately shrink
parameter estimates toward zero, trading some bias for a substantial
reduction in variance. The next step introduces the Lasso, which does
exactly this and additionally performs **automatic variable selection**.
:::

---

# Step 5 — Lasso: Bias Accepted for Lower MSE

:::{.callout-note collapse="true" #bias-variance}
### 📚 Background: The bias-variance decomposition of prediction error

For a fixed query point $x_0$ and an estimator $\hat f(x_0)$
trained on a random sample, the expected squared prediction error
decomposes as:

$$\mathbb{E}\left[(y_0 - \hat f(x_0))^2\right] = \underbrace{\left(\mathbb{E}[\hat f(x_0)] - f(x_0)\right)^2}_{\text{Bias}^2} + \underbrace{\mathrm{Var}(\hat f(x_0))}_{\text{Variance}} + \underbrace{\sigma^2}_{\text{Irreducible error}}$$

- **Bias** measures the systematic error: on average, how far off
  is the estimator?
- **Variance** measures the sensitivity to the training sample: if
  we resampled the data, how much would $\hat f(x_0)$ change?
- **Irreducible error** is the noise in the DGP itself — no
  estimator can eliminate it.

**The Gauss-Markov constraint from Lab 1.** OLS minimises variance
*among all linear unbiased estimators*. It does not minimise
variance *overall*. By accepting a small bias, we move outside the
BLUE restriction and can find estimators with lower total MSE.

**Two readings of the bias-variance trade-off.**

- **Statistical reading.** Accepting a small bias to reduce
  variance can lower the expected MSE. This is the technical
  justification of Ridge and Lasso, developed in this step.
- **Strategic reading [@spiess_optimalEstimation_2025].** A
  researcher with private preferences (publication pressure,
  client expectations, ideological priors) could exploit the
  bias-variance trade-off by tuning regularisation parameters
  until the favoured estimate emerges. From an outside reader's
  perspective, a biased estimate is indistinguishable from a
  tuned-toward-preference estimate. This is why the classical
  econometric insistence on *unbiasedness* — anchored in Lab 1's
  "Note of Caution" — has been defended as more than statistical
  pedantry: it functions as an institutional restriction against
  specification-hacking.

The implication for our lab: when we use Lasso for **prediction**
(Path P), the bias is a deliberate efficiency choice and the
strategic concern does not apply (we never claimed unbiasedness).
When we want to use Lasso outputs for **causal claims** (Path E),
the bias becomes a credibility liability — unless the bias is
fixed *ex ante* through sample-splitting (Lab 3, Double Machine
Learning [@chernozhukov_DoubleML_2018]).
:::

The **Lasso** [@tibshirani_1996] adds an $L_1$ penalty to the
ordinary least-squares loss:

$$\hat{\beta}_{\text{Lasso}} = \arg\min_\beta \left\{ \sum_{i=1}^n (y_i - x_i^\top \beta)^2 + \lambda \sum_{j=1}^p |\beta_j| \right\}$$

For $\lambda = 0$, Lasso equals OLS. For $\lambda \to \infty$, all
coefficients are forced to zero. Cross-validation picks the $\lambda$
that minimises out-of-sample MSE.

:::{.callout-note collapse="true" #lasso-vs-ridge}
### 📚 Background: Lasso vs. Ridge — and why Lasso selects variables

Ridge regression uses an $L_2$ penalty ($\sum \beta_j^2$); Lasso uses
$L_1$. The geometric consequence is that Lasso's penalty corners on
the coordinate axes push small coefficients exactly to zero, while
Ridge shrinks them smoothly toward zero without setting them to
zero.

| Method | Penalty | Shrinks toward 0 | Sets to 0 | Use case |
|:---|:---|:---:|:---:|:---|
| OLS | none | no | no | Low-dim, BLUE conditions |
| Ridge | $L_2$ | yes | no | High correlation, all variables matter |
| Lasso | $L_1$ | yes | **yes** | Variable selection — sparse signal |
| Elastic Net | $L_1 + L_2$ | yes | yes | Mixture: groups + selection |

For our setting — many highly correlated demographics, plus the
suspicion that only a few matter for prediction — Lasso is the
natural choice. It will tell us, automatically, which variables
survive at the $\lambda$ that minimises prediction error.
:::

### Fit Lasso with cross-validation

The `glmnet` package needs a design matrix (no intercept column) and
a response vector:

```{r lasso-prep}
#| code-fold: false
# build the design matrix for the full specification
X_full <- model.matrix(as.formula(formula_full), data = tbl_full)[, -1]
y_full <- tbl_full$logmove

cat("Design matrix dimensions: ", dim(X_full), "\n")
```

```{r lasso-cv}
#| code-fold: false
set.seed(42)
cv_lasso <- cv.glmnet(X_full, y_full, alpha = 1, nfolds = 5)

# the lambda that minimises CV-MSE, and the more parsimonious "1se" choice
lambda_min <- cv_lasso$lambda.min
lambda_1se <- cv_lasso$lambda.1se

cat("lambda.min: ", round(lambda_min, 5), "\n")
cat("lambda.1se: ", round(lambda_1se, 5), "\n")
```

### The cross-validation curve

```{r lasso-cv-plot}
#| fig-height: 5
#| fig-width: 8
plot(cv_lasso)
title("Lasso cross-validation: MSE vs. log(lambda)", line = 2.5)
```

The dashed vertical lines mark **lambda.min** (minimum CV-MSE) and
**lambda.1se** (most regularised model within one standard error).
The top axis shows how many non-zero coefficients survive at each
$\lambda$.

:::{.callout-important icon=false}
## ⚠️ T5: Compare Lasso to OLS on out-of-sample MSE

a. Read off the CV-MSE at lambda.min and at lambda.1se from the plot
   (or from `cv_lasso$cvm` at the corresponding indices).
b. How does the CV-MSE at lambda.min compare to the OLS cross-validation
   results from Step 4?
c. How many variables does Lasso retain at lambda.1se? How many at
   lambda.min?
:::

```{r lasso-comparison-numeric}
#| code-fold: true
#| code-summary: "Show code: numerical Lasso vs. OLS comparison"
idx_min <- which(cv_lasso$lambda == lambda_min)
idx_1se <- which(cv_lasso$lambda == lambda_1se)

tibble(
  Method                = c("OLS — baseline",     "OLS — full",
                            "Lasso (lambda.min)", "Lasso (lambda.1se)"),
  `CV-MSE`              = c(cv_results$mean_rmse[1]^2,
                            cv_results$mean_rmse[2]^2,
                            cv_lasso$cvm[idx_min],
                            cv_lasso$cvm[idx_1se]),
  `Non-zero params`     = c(length(coef(mod_baseline_full)) - 1,
                            length(coef(mod_full_spec)) - 1,
                            cv_lasso$nzero[idx_min],
                            cv_lasso$nzero[idx_1se])
) %>%
  mutate(`CV-MSE` = round(`CV-MSE`, 4)) %>%
  gt() %>%
  gt::tab_options(table.font.size = "12px")
```

:::{.callout-note collapse="true" icon=false color="success"}
### 🟢 The Lasso payoff (Click to unfold)

Lasso typically achieves a CV-MSE close to or below the baseline OLS,
while using far fewer parameters than the full OLS specification.
This is the bias-variance trade-off in action: by accepting a small
bias (shrinkage), Lasso reduces variance enough to lower the total
prediction error.

The key contrast for Sarah's board:

- "More variables = better information" is **wrong out-of-sample**.
  More variables increase variance.
- "Fewer well-chosen variables = better prediction" is the empirical
  reality, when the chosen variables capture the systematic patterns.

Lasso does the choosing automatically, via the $\lambda$ that
cross-validates best.
:::

---

# Step 6 — Which Variables Survive?

The Lasso coefficients at `lambda.1se` tell us which variables carry
prediction power. Variables with zero coefficients have been removed
from the model.

```{r lasso-coefficients}
#| code-fold: false
coef_lasso <- coef(cv_lasso, s = "lambda.1se")
coef_df <- as_tibble(as.matrix(coef_lasso), rownames = "variable") %>%
                rename(estimate = 2) %>%                # ← position-based, version-proof
                filter(variable != "(Intercept)") %>%
                mutate(retained = estimate != 0)
# count retained vs. removed


# count retained vs. removed
coef_df %>%
  count(retained) %>%
  gt() %>%
  gt::tab_header(title = "Variable selection at lambda.1se") %>%
  gt::tab_options(table.font.size = "12px")

```

The retained variables, sorted by magnitude:

```{r lasso-retained}
#| code-fold: true
#| code-summary: "Show code: retained variables sorted"
coef_df %>%
  filter(retained) %>%
  mutate(abs_estimate = abs(estimate)) %>%
  arrange(desc(abs_estimate)) %>%
  select(variable, estimate) %>%
  mutate(estimate = round(estimate, 4)) %>%
  gt() %>%
  gt::tab_options(table.font.size = "11px")
```

:::{.callout-important icon=false}
## ⚠️ T6: Read the Lasso selection

a. Which type of variable dominates the retained list — marketing
   levers (feat, deal), prices, demographics, or interactions?
b. Are demographic main effects retained, or mostly the brand ×
   demographic interactions?
c. Compare to your expectation from Step 0: did Variant C deliver
   what we expected — namely, brand-specific demographic effects?
:::

:::{.callout-note collapse="true" icon=false color="success"}
### 🟢 What the selection tells us (Click to unfold)

Typical findings on the OJ panel:

- **Price and brand effects** survive cleanly — these are the most
  reliable predictors
- **`feat` and `deal`** survive (or their interactions with brand) —
  marketing matters
- **Income, education, household-size and competition variables**
  survive selectively, often as brand × demographic interactions
  rather than main effects — confirming that the *gradient* of the
  demographic effect differs by brand
- **Many redundant demographic interactions are removed** — the
  educ / hval150 / income cluster collapses into one or two
  representative variables

The lesson: with Lasso, **the data picks the relevant variables**.
We do not need to decide in advance which demographics to include —
the algorithm tells us, via cross-validation.
:::

:::{.callout-warning icon=false}
### ⚠️ Important: Lasso coefficients are NOT causal estimates

The Lasso has done what we asked: it found the sparse set of
variables that best predicts on held-out data. But:

- Lasso **deliberately introduces bias** (this was the whole point)
- The coefficients are **shrunk toward zero** — they understate the
  true partial effects
- Variable selection is **not the same as causal identification** —
  Lasso retains predictively useful variables, not causally relevant
  variables
- Two variables that are highly correlated may both be useful;
  Lasso arbitrarily picks one (depending on the random seed). The
  *unselected* variable is not "unimportant".

For Sarah's Question 1 (the causal effect of advertising), Lasso is
**not** the right tool. The Lasso output here serves Path P
(Question 2, the inventory tool) only. Lab 3 will return to the
causal question with appropriate methodology.

**A Lucas-style caveat (recall Lab 1, the BLUE-bridge section).**
The Lasso-selected model predicts well *under the same DGP as in
the training data*. If Sarah's stores undergo a regime change —
new competitors, new pricing strategies, new demographic shifts —
the selected variables may no longer be the right ones. This is
the Lucas critique applied to Path P. Prediction is honest about
its scope: it is interpolation within a stable structure, not
extrapolation across regimes.

**The path forward: sample-splitting.** If we want Lasso (or any
ML method) to help with *causal* questions, we need to fix the
specification choice before estimation. The technique:

1. Split the data into two parts (typically 50/50)
2. Use part 1 to *select* the variables (Lasso, etc.)
3. Use part 2 to *estimate* the final coefficients with the
   selected specification

Part 2 has not seen the specification-search. Inference on part 2
is honest — classical SE and CI become interpretable again. This
is the principle behind **Double Machine Learning**
[@chernozhukov_DoubleML_2018], which Lab 3 develops formally for
causal estimation. The institutional argument from Lab 1's "Note
of Caution" [@spiess_optimalEstimation_2025] is operationalised
here: sample-splitting is what *enforces* the unbiasedness
restriction in practice.
:::

---

## Translator memo for Sarah

:::{.callout-important icon=false}
## ⚠️ T7: Translator memo (Lab 2)

Draft Sarah's follow-up memo. Cover:

- Question 1: Did demographics improve the prediction tool? *By how
  much, and what variables were essential?*
- Question 2: Did demographics change the picture on the advertising
  effect? *No — the structural feat × deal problem remains.*
- A short statement on the board's "more variables = better
  information" intuition: when does it hold, when does it fail?

**3–5 sentences per question, business language.**
:::

---

# Bonus — For independent study

## ⚠️ B1: Ridge vs. Lasso vs. Elastic Net

The three regularisers differ in their penalty geometry. This bonus
task compares them on our data.

:::{.callout-important icon=false}
## ⚠️ B1: Compare three regularisation approaches

```{r b1-three-methods}
#| eval: false
set.seed(42)

# Ridge: alpha = 0
cv_ridge <- cv.glmnet(X_full, y_full, alpha = 0, nfolds = 5)

# Lasso: alpha = 1 (already fit as cv_lasso)

# Elastic Net: alpha = 0.5
cv_elnet <- cv.glmnet(X_full, y_full, alpha = 0.5, nfolds = 5)

# Compare CV-MSE at lambda.min
tibble::tibble(
  method    = c("Ridge",                  "Lasso",                cv_elnet = "Elastic Net (α = 0.5)"),
  cv_mse    = c(min(cv_ridge$cvm),        min(cv_lasso$cvm),      min(cv_elnet$cvm)),
  n_nonzero = c(sum(coef(cv_ridge, s = "lambda.min") != 0) - 1,
                sum(coef(cv_lasso, s = "lambda.min") != 0) - 1,
                sum(coef(cv_elnet, s = "lambda.min") != 0) - 1)
) %>%
  dplyr::mutate(across(where(is.numeric), \(x) round(x, 4)))
```

a. Which method gives the lowest CV-MSE? By how much?
b. Which method retains the most variables? Which the fewest?
c. What does the trade-off look like — does the method with lowest
   MSE also retain the most variables, or is there a clear winner
   on both axes?
:::

## ⚠️ B2: A pure-cross-price Lasso (without demographics)

This bonus isolates the role of cross-price information by fitting
Lasso on the `tbl_cross` data with only prices, feat and deal — no
demographics.

:::{.callout-important icon=false}
## ⚠️ B2: Cross-price Lasso

```{r b2-cross-lasso}
#| eval: false
library(glmnet)

# build a separate model for each of the three brands
for (target_brand in c("dominicks", "minute.maid", "tropicana")) {

  y_brand <- tbl_cross[[paste0("logmove_", target_brand)]]

  X_brand <- tbl_cross %>%
    select(starts_with("price_"), starts_with("feat_"), starts_with("deal_")) %>%
    mutate(across(starts_with("price_"), log)) %>%
    as.matrix()

  cv_b <- cv.glmnet(X_brand, y_brand, alpha = 1, nfolds = 5)

  cat("Brand: ", target_brand, "\n")
  print(coef(cv_b, s = "lambda.1se"))
  cat("\n")
}
```

a. Which cross-prices survive selection for each brand?
b. Do all three brands retain their own price?
c. What does the pattern of retained cross-prices tell you about
   substitution within the OJ category?
:::

## ⚠️ B3: Lasso path visualisation

Plot the full coefficient path as a function of $\lambda$ to see
which variables exit the model first.

:::{.callout-important icon=false}
## ⚠️ B3: Trace coefficient paths

```{r b3-coef-path}
#| eval: false
fit_lasso <- glmnet(X_full, y_full, alpha = 1)
plot(fit_lasso, xvar = "lambda", label = TRUE)
abline(v = log(cv_lasso$lambda.1se), lty = 2, col = "red")
abline(v = log(cv_lasso$lambda.min), lty = 2, col = "blue")
```

a. Which variables are the last to be shrunk to zero as $\lambda$
   increases?
b. Are these the same variables that ended up in the lambda.1se
   selection from Step 6?
c. What does this picture tell you about the *order* in which variables
   contribute to prediction?
:::

## ⚠️ B4: Big-data regime — would we have done this differently?

In Step 0 we noted that p ≈ 36 takes us close to the big-data regime
in spirit but not in scale. This bonus asks what would change if p
were truly large (say, p > n).

:::{.callout-important icon=false}
## ⚠️ B4: A reflection without code

Imagine Sarah comes back next year with the full 11-brand price matrix
*and* loyalty-card-level customer demographics — say, 200 variables.

a. Why would OLS fail completely in this setting? (Hint: think about
   the $X^\top X$ matrix.)
b. Why does Lasso continue to work even when p > n?
c. Which assumption from §0.1a of the lab development schema becomes
   relevant here — and what does it imply for the inference / prediction
   distinction?

This is a written reflection — no R code expected.
:::

## ⚠️ B5: Multilevel as the E1 alternative

Lasso is a Path-P tool. For Path E1 — inferential reading — clustered
or multilevel models address related concerns more directly.

:::{.callout-important icon=false}
## ⚠️ B5: Fit a multilevel model with random intercepts by store

```{r b5-multilevel}
#| eval: false
library(lme4)

mod_lme <- lmer(
  logmove ~ feat + deal + log(price) + brand + income + educ +
    (1 | store),
  data = tbl_full
)

summary(mod_lme)
```

a. How does the `feat` coefficient compare to the OLS estimate from
   Step 1?
b. What does the random-intercept variance tell us about between-store
   heterogeneity?
c. Why is this **not** a substitute for the regularisation pathway —
   what does it add, what doesn't it address?
:::

---

# Summary — what we found

```{r summary-table}
#| echo: false

tibble::tribble(
  ~`Finding type`,                    ~`Key result`,                                                 ~`Implication for Lab 3`,
  "E1 — advertising effect",          "Demographics do not change the feat estimate from Lab 1",     "The A1 problem is structural and needs a study design (Lab 3)",
  "E1 — own-price elasticity",        "Own-price elasticity moderates when cross-prices are added",  "Competitor information matters for both inference and prediction",
  "P — full OLS specification",       "Higher in-sample R², but worse cross-validated MSE",          "Classical OLS limit reached at p ≈ 36",
  "P — Lasso selection",              "Sparse model performs better out-of-sample than full OLS",    "Variable selection is feasible and informative",
  "P — board intuition",              "'More variables = better information' fails out-of-sample",   "The board needs a clearer mental model of prediction"
) %>%
  gt() %>%
  gt::tab_style(gt::cell_fill("#f5f0fa"), gt::cells_column_labels()) %>%
  gt::tab_style(gt::cell_fill("#f0fff0"), gt::cells_body(rows = c(2, 4))) %>%
  gt::tab_style(gt::cell_fill("#fff8e1"), gt::cells_body(rows = c(1, 3, 5))) %>%
  gt::tab_options(table.font.size = "13px")
```

---

# What comes next

| Lab | Question it answers |
|-----|---------------------|
| **Lab 3 — A/B-test design + DiD simulation** | If Sarah really wants the causal effect of advertising, how must she set up the study? And what would she do with a natural experiment (Path E2)? |

**What Lab 3 will add to this story.** Lab 1 showed that OVB
contaminates the OLS estimate of $\beta_{\text{feat}}$. Lab 2 showed
that no amount of Path-P sophistication (Ridge, Lasso, demographic
controls) can fix this — they all live in the same DGP. The fix
must come from a *different DGP*: a study design that manipulates
the treatment instead of observing it.

- **A/B-test design** as the gold standard
- **Difference-in-Differences and Event-Study Designs** as quasi-
  experimental paths when randomisation is infeasible
- **Instrumental Variables** sneak-peek
- **Double Machine Learning** [@chernozhukov_DoubleML_2018] as the
  operative answer to the Spiess-style institutional argument from
  Lab 1's "Note of Caution": ML for nuisance modeling, with
  sample-splitting protecting the causal estimate

The mini-DAG from Lab 1 (`deal` as common cause of `feat` and
`logmove`) is the starting point. Lab 3 develops the DAG framework
systematically.
