<!-- ============================================================
INSTRUCTOR CUSTOMISATION — update before rendering for each cohort
Line references are approximate; search for the marker text.

  Line ~46   Moodle link for tbl_stores download (search: "[MOODLE]")
  Line ~75   Memo date and quarter reference (search: "Q4 2025")

============================================================ -->

---
title: "Session 3 · Lab 3: From Observation to Manipulation — Causal Inference"
subtitle: "CRISP-DM Phase 6 — Deployment, with Path E2 Methodology"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en-US
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
bibliography: ../../lit/my_bib.bib
---

```{r setup}
#| include: false
library(tidyverse)
library(broom)
library(glue)
library(moderndive)
library(gt)
library(xfun)
library(ggdag)         # DAG construction and analysis
library(dagitty)       # backdoor paths and identification
library(plm)           # panel data and DiD
library(AER)           # IV regression (ivreg)
library(fixest)        # high-dimensional fixed effects, modern DiD
library(fontawesome)   # Fontawesome Icons

# shared visual identity
theme_set(theme_minimal(base_size = 11))

# colour palette
col_primary   <- "#D50006"
col_secondary <- "#502479"
col_neutral   <- "#686868"
col_accent    <- "#7d0a52"

# load data — date stamp must match the files in data/raw/
date_stamp <- "20260517"  # adjust to the most recent DataImport.R run
tbl_stores  <- read_rds(xfun::from_root("data", "raw",
                       glue::glue("dominicks_stores_{date_stamp}.rds")))
tbl_demo    <- read_rds(xfun::from_root("data", "raw",
                       glue::glue("dominicks_demo_{date_stamp}.rds")))
# Merge data sets
tbl_full   <- tbl_stores %>% left_join(tbl_demo, by = "store")
```

## Scenario — A Third Memo from Sarah Chen

:::{.callout-tip icon=false}
## 📧 Memo from Sarah Chen, VP Marketing — Dominick's Finer Foods

Date: 2025-11-22
To: Bridge Cohort

Team,

Lab 1 and Lab 2 made things clearer than I had hoped — and harder.
The historical data, even with all the demographic controls Lab 2
added, cannot tell me the *causal* effect of feature advertising
on sales. You called this an "OVB problem" and an "identification
problem."

But then our data steward at the Kilts Center mentioned
something I had not known: **between 1989 and 1994, Dominick's
and Chicago Booth ran randomised pricing and shelf-management
experiments in 25+ product categories across the whole 100-store
chain.** Parts of the variation in our data are not the result of
manager decisions — they are the result of *deliberate
randomisation* by a research team.

That changes things. So I have two questions for this lab:

> ***Q1.*** *What does it actually buy us to know that some of
> our data is experimental? What would we still need to know to
> use it?*
>
> ***Q2.*** *Given what you find out, how would you design — for
> Q4 2026 — an experiment to measure the causal effect of
> feature advertising? Walk me through it. The board needs a
> go/no-go decision on the Q4 budget review.*

I want both: an honest reading of what we have, and a forward
plan for what we still need.

— Sarah
:::

## The bigger picture: when ML meets causal inference

Lab 2 showed that Machine-Learning-style methods (Lasso, Ridge,
Elastic Net) can dramatically improve prediction. They cannot,
however, deliver causal answers — for two reasons:

1. **They optimise the wrong objective.** ML minimises out-of-sample
   MSE for $\hat y$. Causal inference targets $\hat\beta$, the
   structural parameter of the DGP.
2. **They are excellent p-hacking machines.** Their flexibility,
   uncontrolled, can find a model that fits any preferred result.
   This is exactly the institutional concern raised in Lab 1's
   "Note of Caution" [@spiess_optimalEstimation_2025].

The literature on the ML / econometrics interface
[@athey_imbens_2019; @mullainathan_MLApplied_2017;
@spiess_optimalEstimation_2025] has developed three responses:

- **Path E2 — Study design (this lab's main path).** Manipulate the
  DGP through randomisation. The bias is removed *structurally*,
  not via statistical tricks.
- **Path E2+ — Quasi-experiments.** When manipulation is infeasible,
  exploit natural variation (DiD, Event Studies, IV).
- **Path E1 + ML — Double Machine Learning.** Use ML for the
  nuisance components (high-dimensional controls) while protecting
  the causal estimate via sample-splitting. We touch on this at
  the end of the lab.

All three responses share one feature: a clear separation between
*where the algorithm operates* (prediction, nuisance modeling) and
*where the causal claim is made* (the study design). The human in
the loop defines this separation — the algorithm does not.

## Navigation

| Symbol | Box colour | Meaning | When to read |
|:------:|:-----------|:--------|:-------------|
| ⚠️ | Red (Action Required) | Mandatory tasks, live session | Now |
| 💡 | Green (Tip / Best Practice) | Coding tips, shortcuts | When relevant |
| 📚 | Blue collapsed (Background Theory) | Concepts, theory | Read at home |
| ℹ️ | Blue open (Note) | Short clarifications | In flow |

---

# Step 0 — Reflection on Experimental Data

In Lab 1 we discovered a confounder. In Lab 2 we accepted that
prediction is the only honest objective when the DGP cannot be
manipulated. In Lab 3 we change the question itself: can we
*manipulate* the DGP — or find places where someone already did?

Sarah's memo opens with a surprising fact:

> *"Parts of the variation in our data are not the result of
> manager decisions — they are the result of deliberate
> randomisation by a research team."*

This claim is well-documented. The Kilts Center for Marketing
documentation states explicitly:

> *"From 1989 to 1994, the University of Chicago Booth School of
> Business and Dominick's Finer Foods entered into a partnership
> for store-level research into shelf management and pricing.
> Together, they conducted randomized experiments in more than 25
> different categories throughout all stores in this 100-store
> chain."* [@KiltsCenter_2018]

That changes the methodological status of our dataset. The OJ
data is not pure observational — it is a **hybrid**, with
experimental islands inside an observational sea.

:::{.callout-important icon=false}
### ⚠️ T1 — Reflection (5 minutes, in pairs)

Discuss with your neighbour:

a. **What does it buy us** to know that *some* of the variation
   in our data is experimental? Be concrete. Name at least two
   methodological advantages.
b. **What would we still need to know** before we could exploit
   this fact for Sarah's Q1? List the open questions.
c. Could we, in principle, *identify* the experimental
   sub-sample from the data we have today, 30 years later?
:::

:::{.callout-note collapse="true" icon=false}
## 🟢 Discussion (click to unfold)

a. Knowing some weeks/stores/categories are experimental means:
   (i) for those rows, the treatment is **assigned independently
   of unobserved confounders** — the OVB structure we discovered
   in Lab 1 (`deal → feat`) does not apply; (ii) we have
   somewhere in the data a **true Path E2 sub-sample** that can
   serve as a benchmark for any quasi-experimental method we
   build for the rest of the data.

b. The open questions are non-trivial. We would need to know:
   (i) **which weeks** carried experimental treatments;
   (ii) **which stores** were assigned and to what arm;
   (iii) the **randomisation protocol** (was assignment
   store-level, week-level, category-level?);
   (iv) **what the treatment actually was** (price discount?
   shelf re-positioning? both?);
   (v) **how compliance was monitored** — did the store managers
   actually implement the assigned treatment?

c. Without the original Chicago Booth protocols, *not really*.
   The dataset as published does not flag experimental
   observations. This is a sobering lesson in **data
   provenance**: experimental status, like any other
   methodological metadata, has to be preserved alongside the
   numbers, or it is effectively lost. This is one reason
   modern open-science platforms require pre-registered
   protocols.
:::

This reflection sets up the rest of the lab. We will continue to
treat the OJ data as observational for our analyses, because
the experimental sub-sample is no longer identifiable. But we
now understand that:

* **Observational** is a *default state* of data analysis — not a
  natural property of data
* The **distance** between an observational analysis and a
  causal claim depends entirely on the design that produced the
  variation
* For Sarah's Q4 2026 decision, we will need a *new* experimental
  design — one she controls and documents

**Today's roadmap:**

```{r crisp-overview}
#| echo: false
tibble::tribble(
  ~Step, ~Method,                                        ~`Question it answers`,
  "0",   "Reflection on experimental data",              "What does Path E2 status of historical data buy us?",
  "1",   "DAGs as a causal language",                    "What is the structural reason the OLS estimate is biased?",
  "2",   "A/B-test design (Path E2)",                    "What study design removes the bias by construction?",
  "3",   "DiD, Event Studies, IV (Path E2+)",            "When manipulation fails, what quasi-experiments help?",
  "4",   "FE/RE methods + Double ML",                    "What identification strategies use the panel structure?",
  "5",   "Synthesis: memo + 2026 design",                "What do we recommend Sarah do next?"
) %>%
  gt() %>%
  gt::tab_style(gt::cell_fill("#f0fff0"), gt::cells_body(rows = Step == "0")) %>%
  gt::tab_style(gt::cell_fill("#f5f0fa"), gt::cells_column_labels()) %>%
  gt::tab_options(table.font.size = "13px")
```

`r fa('clock')` Time estimate: ~5 min reflection (T1) · ~10 min
DAGs (Step 1) · ~12 min A/B simulation (Step 2) · ~12 min DiD
(Step 3) · ~12 min IV + FE (Step 4) · ~10 min synthesis +
2026 design (Step 5). Total: ~60 minutes.

---

# Step 1 — DAGs as a Tool for Causal Thinking

Recall the structure we discovered in Lab 1: `deal` is a common
cause of both `feat` and `logmove`. Drawn as a graph:

```{r confounder-dag}
#| echo: false
#| fig-align: center
#| out-width: 65%
knitr::include_graphics(xfun::from_root("img", "Session3", "OVB-DAG.svg"))
```

This is a **directed acyclic graph (DAG)**. We now formalise the
language of DAGs and use them to plan causal analyses.

## Three building blocks

DAGs are built from three structural patterns. Each has a precise
implication for what *conditioning* (i.e. controlling) does to the
bias.

### Confounder (common cause)

```
    Z
   / \
  v   v
  X → Y
```

`Z` causes both `X` and `Y`. The unconditional correlation between
`X` and `Y` mixes the causal effect $X \to Y$ with the confounding
path $X \leftarrow Z \to Y$. Conditioning on `Z` **blocks** the
confounding path and identifies the causal effect.

**This is the `deal` situation from Lab 1.**

### Mediator (intermediate variable)

```
  X → M → Y
```

`M` is on the causal pathway from `X` to `Y`. Conditioning on `M`
**blocks** the indirect effect of `X` on `Y`, leaving only the
direct effect. If the question is "what is the total effect of
`X` on `Y`?", we **must not** condition on `M`.

### Collider (common effect)

```
  X      Y
   \    /
    v  v
     C
```

`C` is caused by both `X` and `Y`. `X` and `Y` are unconditionally
independent. But if we condition on `C` (e.g., include it as a
control in our regression), we **open** an artificial path between
`X` and `Y` — this is **collider bias** or selection bias.

:::{.callout-warning icon=false}
### ⚠️ Counter-intuitive: adding controls can HURT

A common Bachelor-level reflex is: "if in doubt, include more
controls." DAGs show why this is wrong. Conditioning on a
collider introduces bias where there was none. Conditioning on
a mediator removes part of the effect you want to measure.

The correct rule is the **backdoor criterion** [@pearl_2009]:
condition on the minimal set of variables that blocks all
backdoor paths from treatment to outcome, **without** opening
new collider paths.
:::

## Self-selection and the identification problem

Confounding is one face of a deeper problem. The same idea
appears under several names in the applied literature; they
describe overlapping mechanisms by which observational data fail
to identify causal effects.

```{r identification-table}
#| echo: false
tibble::tribble(
  ~Mechanism,             ~`What happens in the DGP`,                                                ~`OJ-data example`,
  "Confounding",          "Common cause $Z$ drives both $X$ and $Y$.",                                "Holiday weeks drive both `deal` and `logmove`.",
  "Self-selection",       "Units choose treatment based on unobserved expectations of $Y$.",          "Manager runs `feat` only when she expects strong response — store-week selects itself in.",
  "Simultaneity / reverse causality", "$X$ and $Y$ are jointly determined in the same period.",       "Price is set in response to anticipated demand.",
  "Measurement error in $X$",         "$X$ observed with noise; OLS estimate attenuated.",           "Self-reported feature implementation, week-aggregation mismatch.",
  "Selection on the outcome",         "Whether $(X,Y)$ is observed depends on $Y$ itself.",          "Stores that closed dropped out — survivor sample only."
) %>%
  gt() %>%
  gt::tab_style(gt::cell_fill("#f5f0fa"), gt::cells_column_labels()) %>%
  gt::tab_options(table.font.size = "11.5px")
```

All five share one symptom: $\mathrm{Cov}(x, \varepsilon) \neq 0$,
and OLS becomes biased. They differ in **diagnosis**. Confounding
is a DAG problem — add the right control. Self-selection is a
**sampling** problem — the data we observe are *not random draws
from the DGP*. They are the outcome of unit-level decisions made
on the basis of unobserved information.

This distinction matters for our two questions. Sarah's Q1 asks
*what to do with the observational data we have*; Q2 asks *how to
generate new data that avoids these problems by construction*.

:::{.callout-note collapse="true" #identification-detail}
### 📚 Background: Identification — the more general concept

A parameter is **identified** if it can in principle be recovered
from the population distribution of observables [@huntington-klein_effect_2025,
ch. 5]. Identification is *not* a sample-size question — even
with infinite data, an unidentified parameter cannot be
recovered.

The five mechanisms above are **identification failures**: even
the population OLS slope does not equal the structural effect we
want. The cure is therefore not "more data" but a **redesign of
the variation** — either by intervention (Step 2: A/B-test) or
by exploiting natural variation that satisfies an identifying
assumption (Step 3: DiD, IV).

For a careful treatment of the identification idea — including
the distinction between point identification, set identification
and partial identification — see [@huntington-klein_effect_2025,
ch. 5] and [@cunningham_2021, ch. 1].
:::

:::{.callout-important icon=false}
### ⚠️ T2: Read the DAG of Sarah's situation

The Lab-1 mini-DAG (above) shows `deal → feat → logmove` and
`deal → logmove`. Suppose Sarah also wonders whether **shelf
position** matters. Imagine the following extended structure:

- `deal` is set by the regional manager based on perceived demand
- `feat` is set by the local store manager, often coordinating
  with `deal`
- `shelf_position` is also set by the local manager based on `feat`
- `logmove` depends on `feat`, `deal`, and `shelf_position`

a. Draw this extended DAG (paper or whiteboard).
b. Is `shelf_position` a confounder, mediator, or collider for the
   `feat → logmove` relationship?
c. If we control for `shelf_position` in the regression, what
   happens to the estimate of $\beta_{\text{feat}}$?
:::

:::{.callout-note collapse="true" icon=false color="success"}
### 🟢 Discussion (Click to unfold)

`shelf_position` is on the path `feat → shelf_position → logmove`
— it is a **mediator** of part of the `feat` effect.

If we condition on `shelf_position`, we **remove** the part of the
advertising effect that operates through better shelf placement.
The remaining coefficient on `feat` captures only the *direct*
effect (e.g., the flyer-recall effect), not the *total* effect
that Sarah cares about for her budget decision.

**Lesson:** the right control set depends on the causal question
you are asking. There is no universally "correct" set of controls.
:::

## Using DAGs in code: ggdag

```{r ggdag-example}
#| code-fold: true
#| code-summary: "Show code: ggdag example"
dag <- dagify(
  logmove ~ feat + deal + shelf_position,
  feat ~ deal,
  shelf_position ~ feat,
  coords = list(
    x = c(deal = 1, feat = 2, shelf_position = 3, logmove = 4),
    y = c(deal = 2, feat = 1, shelf_position = 2, logmove = 1)
  )
)

ggdag(dag, layout = "auto") + theme_dag()
```

The `dagitty` package can identify minimal sufficient adjustment
sets automatically:

```{r dagitty-example}
#| code-fold: true
#| code-summary: "Show code: dagitty adjustment set"
dag_obj <- dagitty::dagitty(
  'dag {
     deal -> feat
     deal -> logmove
     feat -> logmove
     feat -> shelf_position
     shelf_position -> logmove
   }'
)

dagitty::adjustmentSets(dag_obj, exposure = "feat", outcome = "logmove",
                        effect = "total")
```

For Sarah's **total effect** question, the minimal sufficient
adjustment set is `{deal}` — confirming what Lab 1 found
empirically.

---

# Step 2 — A/B-test Design: Path E2 in Action

Once we have the DAG, the cleanest solution to OVB is to **break
the confounding arrows by randomisation**. If `feat` is assigned at
random (not by the store manager), the arrow `deal → feat` is
severed.

The DAG of an A/B test:

```
  random_assignment → feat
  deal              → logmove
  feat              → logmove
```

`random_assignment` replaces all other determinants of `feat`. Now
`feat` is unconditionally uncorrelated with `deal`, with
`shelf_position`, with all unmeasured confounders. The OLS estimate
of $\beta_{\text{feat}}$ on randomised data is **unbiased by
construction**.

## A small simulation

Let us see this empirically. We generate synthetic data with a
known true effect, then compare the naive OLS estimate (on
observational data) to the OLS estimate on randomised data.

```{r ab-simulation}
#| code-fold: false
set.seed(42)
n <- 2000

# True parameters
beta_true_feat <- 0.5
beta_true_deal <- 0.4

# Observational DGP: deal causes feat (confounder)
deal_obs <- rbinom(n, 1, 0.3)
feat_obs <- rbinom(n, 1, 0.2 + 0.6 * deal_obs)  # deal increases feat probability
y_obs    <- 5 + beta_true_feat * feat_obs + beta_true_deal * deal_obs +
            rnorm(n, 0, 0.5)

tbl_obs <- tibble(feat = feat_obs, deal = deal_obs, y = y_obs)

# Randomised DGP: feat is independent of deal
deal_rct <- rbinom(n, 1, 0.3)
feat_rct <- rbinom(n, 1, 0.5)  # randomisation: 50/50, independent of deal
y_rct    <- 5 + beta_true_feat * feat_rct + beta_true_deal * deal_rct +
            rnorm(n, 0, 0.5)

tbl_rct <- tibble(feat = feat_rct, deal = deal_rct, y = y_rct)

# Naive OLS on observational data — without deal control
mod_obs_naive <- lm(y ~ feat, data = tbl_obs)

# Naive OLS on RCT data — without deal control
mod_rct_naive <- lm(y ~ feat, data = tbl_rct)

bind_rows(
  tidy(mod_obs_naive) %>% mutate(setting = "Observational"),
  tidy(mod_rct_naive) %>% mutate(setting = "Randomised (A/B)")
) %>%
  filter(term == "feat") %>%
  select(setting, estimate, std.error, p.value) %>%
  mutate(true_value = beta_true_feat,
         bias = round(estimate - beta_true_feat, 3),
         across(where(is.numeric), \(x) round(x, 3))) %>%
  gt() %>%
  gt::tab_options(table.font.size = "12px")
```

:::{.callout-important icon=false}
### ⚠️ T3: Read the simulation output

a. What is the bias of the naive OLS estimate on observational
   data?
b. What is the bias on the randomised data?
c. Why does randomisation work, even though we did *not* control
   for `deal` in either regression?
:::

:::{.callout-note collapse="true" icon=false color="success"}
### 🟢 Discussion (Click to unfold)

a. On observational data, the naive OLS estimate is biased
   upward (overstates the true `feat` effect) because it absorbs
   the `deal` effect that operates through the correlation
   `deal → feat`.
b. On randomised data, the estimate is unbiased — within sampling
   noise it equals the true value 0.5.
c. Randomisation **breaks the arrow** `deal → feat` by construction.
   `feat` is now uncorrelated with `deal` (and with every other
   confounder, observed or unobserved). The OLS estimate of
   $\beta_{\text{feat}}$ is identified from the experimental
   variation alone.

**This is the gold standard for Path E2.** It does not require us
to *measure* every confounder — randomisation handles all of them
at once, including unobservable ones.
:::

## Practical considerations

A/B tests in retail are operationally non-trivial:

- **Unit of randomisation.** Store-week, store-level, or brand
  × store-week? Each choice has implications for power and for
  what causal estimand we identify.
- **Spillover.** If neighbouring stores share customers, treating
  store A while controlling store B contaminates the control.
- **Compliance.** A randomly assigned `feat = 1` may not be
  implemented uniformly by the local manager.
- **Cost.** Sarah needs board approval to *not* advertise in some
  stores during Q4.

For Sarah's purposes, a **cluster-randomised store-level A/B
test** in non-peak weeks is realistic and would identify the
total effect of `feat` cleanly.

:::{.callout-note collapse="true" #ab-vignette-audit}
### 📚 Background: A/B-test, vignette, audit — same logic, different cost

In Part 4 of today's session we placed three closely related
designs on a single validity spectrum (Vignette — Audit — A/B-test).
For Sarah's question, the candidate is clearly the A/B-test
(real users, real behaviour, real outcomes). But the alternatives
are worth keeping in mind:

| Design                        | What Sarah would do for her question                                                                | Cost                                              |
|:------------------------------|:-------------------------------------------------------------------------------------------------|:--------------------------------------------------|
| **Vignette experiment**       | Survey store managers with hypothetical advertising scenarios; measure stated intent.            | Low. Replicable in a thesis project.              |
| **Audit / correspondence**    | Not directly applicable here — audits use real but *fictitious* applicants/applications.          | Not applicable to internal pricing decisions.     |
| **A/B-test (Sarah's choice)** | Randomise `feat = 1/0` across actual store-weeks; measure actual sales.                          | Board approval needed; opportunity cost real.     |

The A/B-test has by far the highest external validity for this
specific business decision. But — as the IAB example from the
slides shows — if budget approval fails, a **vignette experiment
with store managers** can still produce a defensible signal at
near-zero cost. The two designs answer slightly different
questions (intention vs. behaviour) and the gap between them is
itself informative.
:::

---

# Step 3 — Difference-in-Differences, Event Studies, and IV

When randomisation is impossible (politically, operationally, or
ethically), **quasi-experimental designs** exploit natural
variation. The classical example: a policy change rolls out at
different times across regions, giving us "treated" and "control"
groups whose pre-treatment trends should be parallel.

## The DiD logic

Suppose the regional office expanded the advertising programme
in Chicago stores starting week 80, while Suburb stores received
no change. The DiD estimator compares:

$$\hat\delta_{\text{DiD}} = \big(\bar Y_{\text{Chicago}, \text{post}} - \bar Y_{\text{Chicago}, \text{pre}}\big) - \big(\bar Y_{\text{Suburb}, \text{post}} - \bar Y_{\text{Suburb}, \text{pre}}\big)$$

The key identifying assumption is **parallel trends**: in the
absence of treatment, the Chicago and Suburb sales would have
evolved similarly.

## A synthetic DiD example

```{r did-simulation}
#| code-fold: false
set.seed(1)

# Treatment: a subset of stores receives an intensified ad campaign starting week 80
treated_stores <- tbl_stores %>% pull(store) %>% unique() %>% sample(20)

tbl_did <- tbl_stores %>%
  mutate(
    treated = as.numeric(store %in% treated_stores),
    post    = as.numeric(week >= 80),
    treat_post = treated * post,
    # synthetic treatment effect: log-sales rise by 0.15 in treated × post
    logmove_simulated = logmove + 0.15 * treat_post
  )

# DiD regression with two-way fixed effects
mod_did <- fixest::feols(logmove_simulated ~ treat_post | store + week,
                         data = tbl_did)

# extract the DiD coefficient
tidy(mod_did) %>%
  filter(term == "treat_post") %>%
  mutate(across(where(is.numeric), \(x) round(x, 3))) %>%
  gt() %>%
  gt::tab_options(table.font.size = "12px")
```

The estimated DiD coefficient should be close to 0.15 — the true
effect we injected.

## A recent critique: TWFE with staggered adoption

The classical two-way-fixed-effects (TWFE) regression we just
ran has been shown to misbehave when treatment timing varies
across units (staggered rollout). [@borusyak_etal_2024]
demonstrate that TWFE implicitly compares newly-treated units to
already-treated units, producing **negative weights** and
**aggregation biases**.

**The remedy: imputation-based DiD.** Model the counterfactual
outcome for control observations explicitly, then compute
treatment effects only for treated observations:

1. Fit a model on never-treated and not-yet-treated observations
2. Predict counterfactual outcomes for treated observations
3. Treatment effect = actual − counterfactual

This **separates** the modeling of the control trend from the
estimation of the causal effect — a recurring pattern in modern
causal inference, and again an instance of the Spiess-style
institutional argument [@spiess_optimalEstimation_2025]: estimation
and selection / modeling are kept in separate "rooms" to protect
the causal claim.

```{r did-imputation-sketch}
#| eval: false
# Sketch (not run in this lab; uses didimputation package)
library(didimputation)

did_imp <- did_imputation(
  data = tbl_did,
  yname = "logmove_simulated",
  gname = "first_treated_week",  # week at which each unit was first treated
  tname = "week",
  idname = "store"
)
summary(did_imp)
```

The `didimputation` package implements [@borusyak_etal_2024]
directly. Full treatment in →ER017.

## Event-study designs as the dynamic extension

DiD gives a **single average** treatment effect. Event-study
designs are the dynamic extension: they trace how the treatment
effect evolves *over time* relative to the treatment date. Each
"lead" (period before treatment) and "lag" (period after)
receives its own coefficient.

| Approach | What it estimates | When to use |
|:---|:---|:---|
| **DiD** | Average effect post vs. pre | Single sharp policy change |
| **Event study** | Dynamic profile of effect | When timing of effect matters |
| **Synthetic control** | Effect for one treated unit | Single unit, many controls |
| **Imputation DiD** | Robust to staggered adoption | Multiple treatment timings |

The Event-Study design pre-treatment coefficients also provide a
**direct test of parallel trends**: if they are all close to zero,
the parallel-trends assumption is empirically supported.

:::{.callout-note collapse="true" #event-study-detail}
### 📚 Background: How an event study differs from DiD

In a classical DiD with treatment at time $t^*$, we estimate one
coefficient: the average effect post-$t^*$ relative to pre-$t^*$.

In an event study, we estimate one coefficient *per event time*.
If the event is at week 80 and we observe 10 weeks before and
10 weeks after, we estimate 21 coefficients (or 20 with one
omitted as reference):

$$Y_{it} = \alpha_i + \lambda_t + \sum_{\tau = -10}^{10} \beta_\tau \mathbb{1}\{t - t^*_i = \tau\} + \varepsilon_{it}$$

The plot of $\{\hat\beta_\tau\}$ against $\tau$ shows:

- The pre-treatment coefficients should hover around zero
  (parallel trends test)
- The post-treatment coefficients show how the effect builds up,
  peaks, and possibly decays

Event studies are now standard practice for evaluating policy
interventions where timing matters, and they are the natural
host for the imputation-style estimator from
[@borusyak_etal_2024].
:::

---

# Step 4 — Panel Methods, IV, and Double Machine Learning

So far we have looked at experimental designs (Step 2) and at
quasi-experiments that exploit a time-of-treatment (Step 3). One
last family of methods uses the **panel structure** of the data
itself — the fact that we observe the *same* stores repeatedly
over *many* weeks. The same panel structure underpins our
within-store analysis from Lab 1, the DiD specification from
Step 3, and the modern Double-ML pipeline.

## Fixed effects, random effects, and the within-transformation

Lab 1's analysis of `tbl_s114` was implicitly a **within-store**
analysis: by working with one store, we held all store-specific
heterogeneity constant. Fixed-effects panel models generalise
this to all 83 stores while still removing store-level
confounders [@huntington-klein_effect_2025, ch. 16].

Let $i$ index stores and $t$ index weeks. The two-way fixed
effects model is

$$y_{it} = \beta\, x_{it} + \alpha_i + \lambda_t + \varepsilon_{it}$$

where $\alpha_i$ absorbs **all time-invariant** store
characteristics (manager identity, neighbourhood demographics,
shelf-space layout) and $\lambda_t$ absorbs all common time
shocks (national promotions, holidays, weather waves).

```{r fe-demo}
#| code-fold: false
# Two-way fixed effects: store + week
mod_fe <- fixest::feols(
  logmove ~ feat + deal + log(price) | store + week,
  data = tbl_stores
)

bind_rows(
  tidy(lm(logmove ~ feat + deal + log(price), data = tbl_stores)) %>%
    mutate(model = "Pooled OLS"),
  tidy(mod_fe) %>% mutate(model = "Two-way FE (store + week)")
) %>%
  filter(term %in% c("feat", "deal", "log(price)")) %>%
  select(model, term, estimate, std.error) %>%
  mutate(across(c(estimate, std.error), \(x) round(x, 3))) %>%
  gt() %>%
  gt::tab_style(gt::cell_fill("#f5f0fa"), gt::cells_column_labels()) %>%
  gt::tab_options(table.font.size = "12px")
```

Comparing pooled OLS to two-way FE reveals two things at once.
First, $\hat\beta_{\text{feat}}$ typically *changes* when store
fixed effects are added — the pooled estimate was absorbing
between-store differences in baseline performance. Second, the
estimates **gain interpretation**: the FE coefficient is now a
*within-store* effect, identified from variation in `feat`
*within* the same store over time.

:::{.callout-note collapse="true" #fe-vs-re}
### 📚 Background: Fixed effects vs. random effects (FE vs. RE)

Two strategies handle unobserved store heterogeneity $\alpha_i$.
The choice is not aesthetic — it rests on whether $\alpha_i$ is
correlated with the regressors.

**Fixed effects.** $\alpha_i$ is treated as a parameter to
estimate (or, equivalently, demeaned away by the within-
transformation). Identification requires **only** that the
*time-varying* error $\varepsilon_{it}$ is uncorrelated with
$x_{it}$. Time-invariant store characteristics are not
identified — they are absorbed into the $\alpha_i$.

**Random effects.** $\alpha_i$ is modelled as a random draw from
a distribution, *uncorrelated with $x_{it}$ by assumption*. This
gives more efficient estimates (recovers between-store variation
too) — but if the assumption fails (typically: manager skill
$\alpha_i$ correlates with `feat`), the estimates are biased.

**The Hausman test** [@huntington-klein_effect_2025, ch. 16]
compares the two: if RE and FE estimates differ substantially,
RE is rejected. Most modern applied work defaults to FE for
robustness.

In our setting, FE is the right choice: store-level managerial
skill, neighbourhood quality and shelf design almost certainly
correlate with how often `feat` is run.
:::

The within-transformation behind FE is an application of the
**Frisch-Waugh-Lovell theorem** [@huntington-klein_effect_2025,
ch. 13]: regressing $y$ on $x$ after partialling out store
dummies is identical to regressing $y$ on $x$ after demeaning
both within each store. This is the same residualisation logic
that returns — in much greater generality — in the Double-ML
sample-splitting step at the end of this section.

## Instrumental variables and 2SLS — a working demonstration

Sometimes the causal question is about a continuous variable
(price, for example) and we have no natural experiment. We need
an **instrument** $Z$ that:

1. Is correlated with the treatment $D$ (**relevance**)
2. Affects the outcome $Y$ only through $D$ (**exclusion
   restriction**)

For Sarah's setting, **competitor pricing volatility**
(`cpwvol5`) is one candidate: stores facing more volatile
competition adjust their own prices more often, but `cpwvol5`
plausibly affects sales only via the price channel, not
directly. (This argument is debatable — see the warning below.)

```{r iv-2sls-demo}
#| code-fold: false
# Pooled OLS for log(price) — likely biased by demand confounding
mod_ols_price <- lm(logmove ~ log(price) + feat + brand, data = tbl_full)

# 2SLS: log(price) instrumented by cpwvol5
# Syntax in AER::ivreg: outcome ~ regressors | instruments
mod_iv <- AER::ivreg(
  logmove ~ log(price) + feat + brand |
            cpwvol5    + feat + brand,
  data = tbl_full
)

bind_rows(
  tidy(mod_ols_price) %>% mutate(model = "Pooled OLS"),
  tidy(mod_iv)        %>% mutate(model = "2SLS (cpwvol5 as instrument)")
) %>%
  filter(term == "log(price)") %>%
  select(model, estimate, std.error, p.value) %>%
  mutate(across(c(estimate, std.error, p.value), \(x) round(x, 4))) %>%
  gt() %>%
  gt::tab_style(gt::cell_fill("#f5f0fa"), gt::cells_column_labels()) %>%
  gt::tab_options(table.font.size = "12px")
```

The OLS coefficient on `log(price)` reflects whatever in our
data correlates `price` with `logmove`. The 2SLS coefficient
uses *only* the part of price variation that is driven by
`cpwvol5` — and asks: how does *that exogenous* variation
relate to sales?

:::{.callout-warning icon=false}
### ⚠️ The exclusion restriction is untestable

The IV estimate above only has a causal interpretation if
`cpwvol5` affects `logmove` *only* through `log(price)`. If
competitor volatility also drives in-store customer mood, or
correlates with regional events that move sales directly, the
exclusion is violated and the IV estimate is no better than the
OLS one — possibly worse.

This is the central honesty of IV estimation: the assumption
that justifies it cannot be tested with the data the
assumption rests on. The exclusion restriction must be argued
substantively. For Sarah's board, this means: do not present an
IV estimate without a one-paragraph defence of why the
instrument is plausibly exogenous.

For full operational treatment with weak-instrument diagnostics,
overidentification tests and modern IV variants → ER017.
:::

## Double Machine Learning — when controls are high-dimensional

Sometimes the causal question is not about a missing instrument
but about *too many potential controls*. With 12 demographic
variables, 3 brand interactions, and time-varying competitors,
the number of candidate confounders is large. Including all of
them naively in OLS can introduce overfitting and incidental
endogeneity (recall Fan/Han/Liu from Lab 2).

**Double Machine Learning** [@chernozhukov_DoubleML_2018]
generalises the Frisch-Waugh-Lovell logic from FE to *flexible*
nuisance models:

1. **Use ML to model the nuisance parts** of the data:
   - $\hat E[Y \mid X]$: how outcome relates to controls
   - $\hat E[D \mid X]$: how treatment relates to controls
2. **Residualise.** Compute $\tilde Y = Y - \hat E[Y \mid X]$
   and $\tilde D = D - \hat E[D \mid X]$.
3. **Final causal estimate.** Regress $\tilde Y$ on $\tilde D$.
   The coefficient is the causal effect.

**The crucial twist: cross-fitting.** Steps 1 and 2 use *one
half* of the data; step 3 uses the *other half*. This
sample-splitting [@spiess_optimalEstimation_2025] prevents the
ML's flexibility from leaking into the causal estimate.

This is the operative answer to the institutional argument from
Lab 1's "Note of Caution": ML for nuisance modeling, with
sample-splitting protecting the causal estimate.

```{r dml-sketch}
#| eval: false
# Conceptual code — full implementation in ER017
library(DoubleML)
library(mlr3)

data <- DoubleMLData$new(tbl_full,
  y_col  = "logmove",
  d_cols = "feat",
  x_cols = c("log_price", "age60", "income", "educ", "ethnic",
             "hhlarge", "workwom", "hval150", "sstrdist",
             "sstrvol", "cpdist5", "cpwvol5"))

ml_g <- lrn("regr.ranger")  # Random Forest for E[Y|X]
ml_m <- lrn("regr.ranger")  # Random Forest for E[D|X]

dml_model <- DoubleMLPLR$new(data, ml_g, ml_m, n_folds = 5)
dml_model$fit()
dml_model$summary()
```

The result: a causal estimate of $\beta_{\text{feat}}$ that uses
ML's flexibility for the controls but preserves unbiasedness
for the treatment effect. Full operational treatment with the
`DoubleML` package → ER017.

---

# Step 5 — Synthesis: Memo for Sarah and a Design for 2026

The three labs together answer Sarah's two questions in a
structured way:

```{r summary-table}
#| echo: false
tibble::tribble(
  ~Question,           ~Path,  ~`Method (this module)`,                            ~`What was achieved`,
  "Q1 (causal effect)", "E1",   "Lab 1 OLS + diagnostics",                          "Bias diagnosed (`deal` confounder)",
  "Q1 (causal effect)", "E1+P", "Lab 2 Lasso with controls",                        "Bias confirmed structural, not solvable by controls",
  "Q1 (causal effect)", "E2",   "Lab 3 A/B test design",                            "Clean identification possible — design required",
  "Q1 (causal effect)", "E2+",  "Lab 3 DiD / Event Study / IV",                     "Quasi-experimental fallbacks if A/B test infeasible",
  "Q1 (causal effect)", "E1+ML","Lab 3 Double ML",                                  "High-dimensional controls + unbiased estimate",
  "Q2 (prediction)",    "P",    "Lab 2 cv-Lasso with demographics",                 "Inventory tool ready, OOS RMSE quantified",
  "Q2 (design 2026)",   "E2",   "Step 5 — new design for Q4 2026",                  "Forward-looking experimental protocol"
) %>%
  gt() %>%
  gt::tab_style(gt::cell_fill("#f5f0fa"), gt::cells_column_labels()) %>%
  gt::tab_options(table.font.size = "13px")
```

:::{.callout-important icon=false}
### ⚠️ T_final (a): Memo on the historical analysis

Draft Sarah's memo on what the *existing* historical data can —
and cannot — tell her about Q1 (the causal effect of feature
advertising):

a. Which design would you recommend if she had to **measure**
   the effect today? A/B test, DiD, IV or DML? Argue based on
   feasibility, cost, validity, and what causal claim each
   supports.
b. If the board insists on having a recommendation *now*
   (before any new experiment), what is the *best estimate* you
   can provide from observational data — and how do you
   communicate its limitations honestly?
c. How does the Lab-1 "Note of Caution" argument
   [@spiess_optimalEstimation_2025] apply to what you are about
   to write? What does the institutional restriction of
   unbiasedness mean for *you* as the analyst?

**5–8 sentences per question. Business language.**
:::

## A new experiment for Q4 2026

The historical Dominick's data was collected in the 1989–1994
period — a retail world without scanner panels, without loyalty
cards, without geofenced mobile coupons, without programmatic
shelf cameras, without micro-targeted digital advertising. The
methodological *logic* of an A/B-test has not changed in 30
years, but the **technology of experimentation** has.

This is the practical setting for Schrage's *Innovator's
Hypothesis* (introduced in Part 4 of today's session): high-
performing firms run cheap, small, fast experiments — and they
run *many* of them — precisely because they cannot reliably
forecast which interventions will pay off. The cost of running
an experiment in 2026 is a small fraction of what it cost in
1994. The value of the design discipline is correspondingly
higher.

:::{.callout-important icon=false}
### ⚠️ T_final (b): Design Sarah's 2026 experiment

Draft a one-page experimental protocol for Q4 2026 that would
let Sarah answer her Q1 cleanly. Cover the following points,
deliberately exploiting capabilities that did not exist in
1994:

a. **Unit of randomisation.** Store-week, store-day, customer-
   level (via loyalty card)? Which choice best matches the
   causal estimand Sarah cares about?
b. **Treatment definition.** What exactly is randomised — the
   *flyer*, the *display*, the *digital ad*, the *price
   discount*? How do you keep the treatment clean?
c. **Sample size and power.** Given the variability of
   `logmove` in the historical data (use Bonus B2 for
   guidance), how long does the experiment need to run?
d. **Compliance and spillover.** What technology would you use
   in 2026 to monitor compliance? How do you handle customers
   who shop across stores?
e. **What you would pre-register.** Following the Spiess
   institutional argument: which hypotheses, which estimands,
   which subgroups would you commit to *before* seeing data?
f. **One thing the 1994 Chicago Booth team probably could not
   do.** What single methodological move from the 2020s would
   change the design most?

**Half a page. Aimed at Sarah's CFO and CTO — they should each
recognise one of their concerns being addressed.**
:::

:::{.callout-tip icon=false}
### 💡 The point of T_final (b)

This is not an exercise in technological optimism. It is a
disciplined application of Schrage's hypothesis to a concrete
business decision. *Knowing what existed in 1994* is the
sober anchor; *knowing what exists in 2026* is the lever. The
gap between them is precisely the budget conversation Sarah
needs to have with her board.

A good memo here will not list every new technology — it will
identify the *one* design decision that turns a high-cost,
slow experiment into a low-cost, fast one.
:::

## A note on the human in the loop

Across Lab 1, Lab 2, and Lab 3, you have seen three roles for the
analyst:

- **Lab 1 (Path E1):** the analyst diagnoses model misspecification
- **Lab 2 (Path P):** the analyst tunes regularisation to optimise
  prediction
- **Lab 3 (Path E2):** the analyst designs the study to enable
  causal identification

In all three roles, the algorithm operates in a sub-space defined
by the analyst. The algorithm does not choose the question, does
not validate the assumptions, does not bear responsibility for
the recommendation. **The human is in the loop.** Sarah Chen's
memos do not come from `lm()` or `cv.glmnet()` or `DoubleML$fit()`
— they come from your judgment, informed by the algorithms'
output.

This is the central professional skill of the modern data scientist
in business: not the technical mastery of any one tool, but the
ability to map a business question onto the right methodological
path, validate the underlying assumptions, and communicate the
honest answer — including its limitations.

---

# Bonus — For independent study

## ⚠️ B1: Collider-bias simulation

Construct a small DAG where `X → C` and `Y → C` (C is a collider).
Generate data accordingly. Show that:

a. The unconditional regression of `Y` on `X` gives an unbiased
   estimate (which is zero, by construction).
b. Conditioning on `C` (including it as a control) introduces a
   spurious correlation between `X` and `Y`.

This is the empirical version of the warning in Step 1.

## ⚠️ B2: Power analysis for a store-level A/B test

For Sarah's planned A/B test:

a. Given the variability of `logmove` in `tbl_stores`, how many
   store-weeks per arm are needed to detect an effect of size
   $\delta = 0.10$ at 80% power and $\alpha = 0.05$?
b. How does the required sample change if Sarah wants to detect
   a smaller effect ($\delta = 0.05$)?
c. What does this imply for the **duration** of the A/B test
   given the 83 available stores?

Use `pwr::pwr.t.test()` or `pwr::pwr.f2.test()`.

## ⚠️ B3: Event study on the synthetic DiD data

Re-fit the DiD example from Step 3 as an event study:

a. Replace the single `treat_post` indicator with a set of
   indicators for $\tau \in \{-10, -9, \ldots, +10\}$.
b. Plot the coefficient profile.
c. Are the pre-treatment coefficients consistent with parallel
   trends?

## ⚠️ B4: Hausman test for IV validity

Use the IV setup from Step 4. Run a Hausman test comparing OLS
and IV estimates:

a. What does the Hausman test tell you about whether `log(price)`
   is endogenous?
b. What does it *not* tell you (hint: about the validity of the
   instrument itself)?

## ⚠️ B5: Sensitivity analysis for DiD parallel trends

The parallel-trends assumption is **untestable** for the
post-treatment period (by construction we never see it). But
[@rambachan_roth_2023] developed bounds-based sensitivity analyses:
"how much would parallel trends have to be violated to overturn
my conclusion?"

a. Conceptually: why is this question more useful than a
   yes/no test of the assumption?
b. Sketch (in prose) how you would communicate the result to
   Sarah's board.

## ⚠️ B6: Reconstructing the Chicago Booth × Dominick's experimental design

This is the historical research extension of Step 0. The Kilts
Center documents that Dominick's and Booth ran RCTs in 25+
categories between 1989 and 1994 [@KiltsCenter_2018], but the
operational protocols are not in the public data dump.

a. Search the marketing-science literature (Marketing Science,
   JMR, JM) for the original Chicago Booth × Dominick's
   experimental papers. Hoch, Drèze, and Purk are useful
   starting authors.
b. Identify, for one published experiment of your choice: the
   randomisation unit (store/week/SKU), the duration, the
   measured outcomes, and any compliance issues the authors
   discuss.
c. Compare that 1990s protocol to the 2026 design you
   sketched in T_final (b). Which differences are
   methodological, which are technological, and which would
   matter most for Sarah's specific question?

**This bonus connects the historical curiosity of Step 0 with
the forward-looking design of T_final (b). It is the deepest
exercise in the lab — plan an afternoon for it.**

---

# What comes next (→ER117)

| Module | What it adds |
|:---|:---|
| **Time Series & Causal Inference** | Full operational treatment of DiD, Event Studies, IV, Synthetic Control, DML |
| **Panel Data Methods** | Fixed effects, random effects, GMM, dynamic panels |

The DGP framework from Session 3 carries through both modules:
the question is always *what is the DGP*, *which path approaches
it*, and *what does our chosen method actually identify*.
