---
title: "Data Science and Strategy for Business"
subtitle: "Complexity, Generalization & Causality: Predict or Intervene?"
author: "Prof. Dr. Jörg Schoder"
institute: "FFHS"
date: "`r Sys.Date()`"
bibliography: ../../lit/my_bib.bib
lang: en-US
reference-section-title: References
output:
  xaringan::moon_reader:
    self_contained: true
    css:
         - default
         - ../../css/ffhs-theme_js.css
         - xaringan-themer.css
    includes:
      after_body: ../../css/insert-logo.html
    lib_dir: ../../libs
    nature:
      slideNumberFormat: "%current%/%total%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: false

    

      

---

class: title-slide

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_xaringan(text_color = "#d50006",
               inverse_text_color = "#FFFFFF",
               inverse_background_color = "#d50006",
               title_slide_background_color = "#d50006",
               header_background_color = "#d50006",
               header_color = "#FFFFFF",
               header_h1_font_size = "32px",
               header_h2_font_size = "26px",
               link_color = "#502479",
               header_h3_font_size = "20px",
               text_slide_number_color = "#d50006",
               text_slide_number_font_size = "0.5em")
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_progress_bar(color = "#d50006", location = "bottom")
xaringanExtra::use_xaringan_extra(c("tile_view","scribble","panelset","tachyons"))
xaringanExtra::style_panelset_tabs(font_family = "inherit")
xaringanExtra::use_editable(expires = 1)
xaringanExtra::use_freezeframe(trigger = "hover")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 150)
options(htmltools.dir.version = FALSE)
library(latex2exp)
library(fontawesome)
library(tidyverse)
library(kableExtra)
library(gt)
source(xfun::from_root("lit","helper.R"))
library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           style = "markdown",
           dashed = TRUE)
bib <- ReadBib(xfun::from_root("lit","my_bib.bib"))
Zitiere <- function(bib, key, ...) {
  gsub(" and ", " and ", Citet(bib, key, ...))
}

# colour palette (consistent with labs)
col_primary   <- "#D50006"   # red — Path E1 / OLS / current
col_secondary <- "#502479"   # purple — Path P / Lasso / Lab 2
col_tertiary  <- "#7d0a52"   # dark accent — Path E2 / Causal / Lab 3
col_neutral   <- "#686868"
```


# ER114 — Data Science and Strategy for Business

## Session 3

### Complexity, Generalization & Causality

### *Predict or Intervene? Two Paths Beyond Inference*

<br>
<br>
<br>
<br>
<br>
### FS 2026
<br>
### Prof. Dr. Jörg Schoder
.mycontacts[
`r fa('github')` @FFHS-EconomicResearch
`r fa('linkedin')` @jfschoder
]


---
layout: true

<div class="my-footer"></div>

<div style="position: absolute;left:400px;bottom:10px;font-size:9px">`r fa('creative-commons')``r rmarkdown::metadata$author`</div>




---
class: left

.blockquote[Agenda]

## Session 3 Overview

<div style="display:grid; grid-template-columns:80px 1fr; gap:0.3em 1em;
            font-size:0.82em; margin-top:0.5em; align-items:center;">

<div style="color:#686868;">10'</div>
<div>`r fa('rotate-left')` <strong>Welcome &amp; Recap</strong> — where Sessions 1+2 left us</div>

<div style="color:#686868;">15'</div>
<div>`r fa('lightbulb')` <strong>Prediction vs. Explanation</strong> — the central dilemma of Block&nbsp;3</div>

<div style="color:#686868;">15'</div>
<div>`r fa('chart-line')` <strong>The In-Sample Illusion</strong> — why high R² alone is dangerous</div>

<div style="color:#D50006; font-weight:bold;">45'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab&nbsp;1</strong> — Reality Check via Out-of-Sample Validation</div>

<div style="color:#686868;">15'</div>
<div>`r fa('triangle-exclamation')` <strong>When OLS Fails</strong> — the BLUE bridge, bias–variance, Lasso preview</div>

<div style="color:#686868; font-style:italic;">10'</div>
<div style="color:#686868; font-style:italic;">`r fa('mug-hot')` Break</div>

<div style="color:#D50006; font-weight:bold;">25'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab&nbsp;2</strong> — Multicollinearity &amp; Regularization</div>

<div style="color:#686868;">25'</div>
<div>`r fa('compass')` <strong>Causal-Inference Primer</strong> — beyond correlation: A/B&nbsp;tests, quasi-experiments, IV, DML</div>

<div style="color:#D50006; font-weight:bold;">30'</div>
<div style="color:#D50006; font-weight:bold;">`r fa('flask')` <strong>Lab&nbsp;3</strong> — Design your Experiment (Innovator's Hypothesis)</div>

<div style="color:#686868;">10'</div>
<div>`r fa('forward')` <strong>Outlook ER017</strong> — Machine Learning &amp; Causal+ML</div>

<div style="color:#686868;">10'</div>
<div>`r fa('flag-checkered')` <strong>Debrief &amp; Evaluation</strong></div>

</div>

???
* Red blocks = labs (Lab 1 is longer because it covers the OLS diagnostic foundation)
* Three connecting threads: DGP-paths E1/E2/P, Spiess credibility argument, Sarah Chen narrative
* Total ≈ 195 min including break


---
class: inverse, center, middle

## Recap

### Where did Sessions 1 + 2 leave us?


---
class: left

.blockquote[Recap]

## Sessions 1 → 2 → **3**

.pull-left[
### Session 1 — *Sees &amp; Describes*

`r fa('eye')` Descriptive regression as **conditional mean**:
$$\hat{y} = \mathbb{E}[y\,|\,\mathbf{x}]$$

`r fa('eye')` CRISP-DM as project skeleton

`r fa('eye')` Tidyverse pipeline + EDA

`r fa('eye')` First contact with **OVB** — missing variables bias all coefficients

<br>

### Session 2 — *Tests Hypotheses*

`r fa('dice')` Population vs. sample · **stochastic** regression

`r fa('dice')` Bootstrap &amp; permutation — *"only one test"*

`r fa('dice')` Type I/II errors as economic risks (EV-Framework)

`r fa('dice')` p-Hacking, FDR, BH

]

.pull-right[
### Session 3 — *Today*

`r fa('circle-question')` **Inference vs. Prediction** — when is which the right purpose?

`r fa('circle-question')` **Multicollinearity** &amp; the BLUE limits

`r fa('circle-question')` **Bias-variance trade-off** &amp; regularization

`r fa('circle-question')` **Cross-validation** as the out-of-sample yardstick

`r fa('circle-question')` **Causal engineering** — when correlation is not enough

<br>

.blockquote[
"The Block&nbsp;3 question is no longer *what do we see?* &mdash; it is *what should we believe and what should we act on?*"
]
]

???
* The recap stresses that Sessions 1 and 2 stayed on the **same dataset** with the **same model** — only the lens changed
* Today we change the lens twice more: out-of-sample (Lab 1+2) and counterfactual (Lab 3)


---
class: left

.blockquote[Recap]

## Two Questions Session 3 Tackles

.pull-left[
### Question 1
`r fa('chart-line', fill='#502479')` *"Will the pattern I see in the past generalize to new data?"*

This is the **prediction question**. The data don't have to be experimental — they just have to be representative of what we will see tomorrow.

**Tools:** train/test split, cross-validation, regularization.

**Lab focus:** Lab&nbsp;2.

**Path label:** Path&nbsp;**P** (`r Citet(bib, "breiman_2001")`).
]

.pull-right[
### Question 2
`r fa('flask', fill='#7d0a52')` *"If I intervene — change a price, run a campaign — what changes?"*

This is the **causal question**. Pattern in past data is **not enough**; we need a justification for the counterfactual.

**Tools:** RCT, DiD, IV, RDD, DML.

**Lab focus:** Lab&nbsp;3.

**Path label:** Path&nbsp;**E2** (`r Citet(bib, "shmueli_2010")`).
]

<br>

.center[
.blockquote[
Lab&nbsp;1 ends Session&nbsp;2's classical inference path (Path **E1**) by stress-testing it under the A/B/C assumptions — and shows where it breaks.
]
]

???
* Notice we are still inside the same business problem (Dominick's Finer Foods, Sarah Chen)
* The two questions correspond to two different paths through the same data
* Spelling out the path labels here so we can use them throughout the deck and in the labs without confusion


---
class: left

.blockquote[Narrative anchor]

## `r fa('envelope', fill='#D50006')` Memo from Sarah Chen — third instalment

.pull-left[
**To:** Translator Team
**From:** Sarah Chen, VP Marketing &mdash; Dominick's Finer Foods
**Re:** Two strategic decisions, one quarter, no margin for error

<br>

> *We have **two distinct questions** to settle this quarter:*
>
> 1. *Does **feature advertising** really cause incremental sales — or are we just confirming what we already do when we promote?*
>
> 2. *I need an **inventory-planning tool** that forecasts weekly sales for any store-brand combination.*
>
> *I need answers that **survive an audit** by my CEO. No statistical theatre.*
]

.pull-right[
### How today maps to Sarah's two questions

<div style="display:grid; grid-template-columns:140px 1fr; gap:0.4em 0.8em;
            font-size:0.85em; margin-top:0.5em;">

<div style="color:#D50006; font-weight:bold;">Lab&nbsp;1 → Memo&nbsp;T4</div>
<div>Diagnose whether classical OLS can answer Q1 at all. Spoiler: not directly — an A/B test is needed.</div>

<div style="color:#502479; font-weight:bold;">Lab&nbsp;2 → Memo&nbsp;T7</div>
<div>Build the inventory-planning tool. Multicollinearity in demographics; Lasso for out-of-sample MSE.</div>

<div style="color:#7d0a52; font-weight:bold;">Lab&nbsp;3 → Memo T<sub>final</sub></div>
<div>Design the A/B test that answers Q1 credibly. Bridge to DiD, IV, DML for cases when RCT is impossible.</div>

</div>

<br>

.blockquote[
The three memos turn methodological choices into managerial artefacts — *that* is the Translator role.
]
]

???
* The Sarah Chen memos appear in all three lab .qmd files (T4 = Lab 1, T7 = Lab 2, T_final = Lab 3)
* Today's slides reference these memos at three points to anchor the labs
* Audience: VP Marketing — not a statistician. Memos must be defensible to a CEO.


---
class: inverse, center, middle

## Part 1

### Prediction vs. Explanation — *the central dilemma*


---
class: left

.blockquote[Part 1 · Three Paths]

## Three Paths from the DGP

.pull-left[
Every dataset is **one realization** of an unobservable **data-generating process** (DGP).

Three fundamentally different questions can be asked of the DGP:

**Path E1 — Classical Inference**
> "What is the stable conditional partial effect, *given* the model?"
> *Identification: by assumption.*

**Path E2 — Causal Inference**
> "What would change if I intervened?"
> *Identification: by research design.*

**Path P — Prediction**
> "How well do I generalize to new data?"
> *Identification: out-of-sample performance.*
]

.pull-right[
```{r dgp-paths-slide, echo=FALSE, fig.align='center', out.width="100%"}
knitr::include_graphics(xfun::from_root("img","Session3","DGP-Paths_gemini.svg"))
```

<br>

.small[
Same data — three legitimate analytical purposes.
The path choice **drives**: estimator, assumption set, evaluation criterion.
]
]

.center[
.blockquote[
.small[
`r Citet(bib, "shmueli_2010")` · `r Citet(bib, "breiman_2001")` · `r Citet(bib, "mullainathan_MLApplied_2017")`
]
]
]

???
* This is THE conceptual scaffold of Session 3 — and of Lab 1 (where the SVG is reused)
* Throughout today we will sometimes use "Path A / Path B" (Predict / Intervene) as shorthand —
  but in the labs and in published statistics it's E1/E2/P. Both vocabularies map 1:1.
* The SVG itself is the same file students see in Lab 1 — repetition reinforces recognition


---
class: left

.blockquote[Part 1 · Purpose drives method]

## Inference vs. Prediction — what counts?

<div style="display:grid; grid-template-columns:1fr 1fr; gap:0.55em; margin-top:0.2em;">
  <div class="vs2-hdr" style="background:#502479;">`r fa('magnifying-glass', fill='white')` &nbsp;Inference (Paths E1, E2)</div>
  <div class="vs2-hdr" style="background:#D50006;">`r fa('wand-magic-sparkles', fill='white')` &nbsp;Prediction (Path P)</div>
</div>

<div class="vs2-row">
  <div class="vs2-dim">`r fa('bullseye', fill='#777')` &nbsp;Object of interest</div>
  <div class="vs2-cell" style="background:#f5f0fa; border:1px solid #d0b8e8;">
    <strong>Parameter</strong> $\beta_j$ — what does the coefficient *mean*?<br>
    <span style="color:#666;">Stable structural relation, partial effect.</span>
  </div>
  <div class="vs2-cell" style="background:#fff0f0; border:1px solid #f0b8b8;">
    <strong>Prediction</strong> $\hat{y}$ — how close to the truth on **new** data?<br>
    <span style="color:#666;">Coefficient values are means to an end.</span>
  </div>
</div>

<div class="vs2-row">
  <div class="vs2-dim">`r fa('shield-halved', fill='#777')` &nbsp;Validation criterion</div>
  <div class="vs2-cell" style="background:#f5f0fa; border:1px solid #d0b8e8;">
    <strong>Unbiasedness · CI · p-value</strong><br>
    <span style="color:#666;">Did we capture the truth in expectation?</span>
  </div>
  <div class="vs2-cell" style="background:#fff0f0; border:1px solid #f0b8b8;">
    <strong>Out-of-sample MSE / R²</strong><br>
    <span style="color:#666;">Does it generalize to data it has not seen?</span>
  </div>
</div>

<div class="vs2-row">
  <div class="vs2-dim">`r fa('triangle-exclamation', fill='#777')` &nbsp;Main risk</div>
  <div class="vs2-cell" style="background:#f5f0fa; border:1px solid #d0b8e8;">
    <strong>Omitted-variable bias</strong>, model misspecification
  </div>
  <div class="vs2-cell" style="background:#fff0f0; border:1px solid #f0b8b8;">
    <strong>Overfitting</strong>, regularization bias
  </div>
</div>

<div class="vs2-row">
  <div class="vs2-dim">`r fa('user-tie', fill='#777')` &nbsp;Managerial use</div>
  <div class="vs2-cell" style="background:#f5f0fa; border:1px solid #d0b8e8;">
    Decide *whether* to act (price up?, campaign yes/no?)
  </div>
  <div class="vs2-cell" style="background:#fff0f0; border:1px solid #f0b8b8;">
    Decide *how much* to stock, route, target — at scale, automatically
  </div>
</div>

???
* Both columns sit on top of the same regression machinery — what differs is the **purpose** and therefore the **diagnostic**
* The two are not enemies: many real projects need both, in sequence
* Highlight Path E2 missing here on purpose — it adds **counterfactual identification** as a third evaluation axis (next slides)


---
class: left

.blockquote[Part 1 · Logic of Data Science]

## Induction and Deduction in Data Science

.pull-left[
### 1. Induction — *the learning step*

`r fa('arrow-up-from-ground-water', fill='#502479')` From **specific observations** → toward a **general model**

`r fa('arrow-up-from-ground-water', fill='#502479')` Estimating coefficients, fitting trees, training neural nets

`r fa('arrow-up-from-ground-water', fill='#502479')` Risk: the pattern was a **sample idiosyncrasy** — it does not generalize

> *Inductive failure = the model fits the training data but is useless out-of-sample.*

**Path P** lives or dies on the success of induction — measured by **out-of-sample performance**.
]

.pull-right[
### 2. Deduction — *the application step*

`r fa('arrow-down', fill='#D50006')` From **general model** → toward a **specific new case**

`r fa('arrow-down', fill='#D50006')` Predicting $\hat{y}$ for an unseen $\mathbf{x}$, choosing an action

`r fa('arrow-down', fill='#D50006')` Risk: the model is correct **descriptively** but breaks when the world is **changed** by the action

> *Deductive failure = applying a predictive model to an intervention scenario.*

**Path E2** is the systematic answer: identify causal effects *before* deducing actions.
]

<br>

.center[
.blockquote[
.small[
Two failure modes — two different remedies. Lab&nbsp;2 fights inductive failure (regularization &amp; CV). Lab&nbsp;3 fights deductive failure (research design).
]
]
]

???
* This slide is borrowed from PVA3 (German original) — translated and re-anchored to the E1/E2/P scheme
* Many students conflate the two failure modes; teaching them as **distinct** is the lesson
* Box & Hunter, Field 2012, Schmueli 2010 all make this distinction explicitly


---
class: left

.blockquote[Part 1 · Two failure modes]

## Failed Generalization and the Causality Trap

.pull-left[
### Danger 1 — Failed induction
*The model overfits.*

A polynomial of degree 15 fits the training data perfectly. Tested on next month's data, it fails catastrophically.

**Symptoms:**
- High in-sample $R^2$, low out-of-sample $R^2$
- Coefficients are unstable across resamples
- VIF explodes when predictors correlate

**Remedy:** train/test split → cross-validation → regularization (Lab&nbsp;2).
]

.pull-right[
### Danger 2 — Blind deduction
*The model is used for an intervention it was not built for.*

We see: *"stores with higher feature ads have higher sales"*.
We conclude: *"more feature ads cause more sales"*.

But feature ads at Dominick's are **always bundled with a price promotion** — the correlation reflects an operational rule, not a causal mechanism. → see Lab&nbsp;1 OVB diagram

**Symptoms:**
- A1 violation (omitted variable)
- C1 violation (exogeneity)
- Lucas (1976): *the policy itself changes the relationship*

**Remedy:** identification by design — RCT, DiD, IV (Lab&nbsp;3).
]

<br>

.center[
.small[Both failures are invisible at the regression-table level. They only surface under the right diagnostic — that's why Block&nbsp;3 spends time on diagnostics.]
]

???
* The "Lucas trap" gets a one-line treatment here because it returns in Part 4 under Endogeneity
* The OVB story from Lab 1 (feat ⊥/⊥ deal) is the empirical anchor — students recognize it
* "Both failures are invisible at the regression-table level" — that's why we teach diagnostics, not regressions


---
class: left

.blockquote[Part 1 · Translator role]

## From AI User to AI Value Creator

<div style="display:grid; grid-template-columns:1fr 1fr; gap:0.55em; margin-top:0.4em;">

<div style="background:#f5f5f5; padding:0.8em; border-left:5px solid #686868;">
<h3 style="margin-top:0;"><span style="color:#686868;">`r fa('user')` AI User</span> &mdash; *passive*</h3>

`r fa('circle-dot', fill='#686868')` Takes algorithms off the shelf

`r fa('circle-dot', fill='#686868')` Trusts default outputs

`r fa('circle-dot', fill='#686868')` Reports $R^2$, p-value, confusion matrix

`r fa('circle-dot', fill='#686868')` Cannot answer: *"how do I know this is the right tool for this question?"*

<br>

.small[Commodity skill set — replaceable.]
</div>

<div style="background:#fff0f0; padding:0.8em; border-left:5px solid #D50006;">
<h3 style="margin-top:0;"><span style="color:#D50006;">`r fa('user-gear')` AI Value Creator</span> &mdash; *active*</h3>

`r fa('circle-check', fill='#D50006')` Asks: *what is the question — prediction or intervention?*

`r fa('circle-check', fill='#D50006')` Chooses the **path** (E1, E2, P) — and the associated estimator

`r fa('circle-check', fill='#D50006')` Validates diagnostics, not just point estimates

`r fa('circle-check', fill='#D50006')` Translates the result into a decision Sarah can defend

<br>

.small[Translator skill set — the strategic edge.]
</div>

</div>

<br>

.center[
.blockquote[
*The translator role is the through-line of ER114. In Session 1 we drew the line between AI User and AI Value Creator; in Block 3 we operationalize it: every method choice in today's labs is a Value-Creator decision.*
.tr[`r Citet(bib, "thomasAIValueCreators2025")`]
]
]

???
* This slide closes a loop opened in Session 1 — same vocabulary, now applied to Block 3 methodology
* The four Value-Creator checkpoints map to Memo T4/T7/T_final structure in the labs
* Audience cue: many students arrive expecting "the right tool" — instead today's lesson is "the right path"


---
class: left

.blockquote[Part 1 · Foundations]

## Model Assumptions — the A/B/C compass

.pull-left[
### A — *Specification*
- **A1** No relevant variable omitted, no irrelevant variable included
- **A2** Linear in parameters
- **A3** Parameters constant across observations

<br>

### B — *Error term*
- **B1** $\mathbb{E}[\varepsilon] = 0$
- **B2** Homoscedasticity — $\text{Var}(\varepsilon) = \sigma^2$
- **B3** No autocorrelation
- **B4** Normality *(needed only for finite-sample inference)*
]

.pull-right[
### C — *Variables*
- **C1** Exogeneity: $\text{Cov}(X, \varepsilon) = 0$
- **C2** No perfect multicollinearity

<br>
<br>

.blockquote[
**Block&nbsp;3 zooms in on C1 and C2:**

`r fa('arrow-right', fill='#502479')` **C2 (multicollinearity)** → bias-variance, Lasso, regularization → *Lab&nbsp;2*

`r fa('arrow-right', fill='#7d0a52')` **C1 (exogeneity)** → endogeneity, OVB, causal design → *Lab&nbsp;3*
]

<br>

`r fa('book-open', fill='#502479')` Full A/B/C diagnostics: Lab&nbsp;1 Step&nbsp;2 and following.
]

<br>

.center[
.small[`r Citet(bib, "vonAuer_2023")` for the European econometrics tradition · `r Citet(bib, "fan_bigdata_2014")` for the big-data extension.]
]

???
* Two-thirds of this slide is review from Lab 1 — but worth showing here because the rest of the deck is structured around C1 (causal) vs C2 (prediction)
* "needed only for finite-sample inference" — B4 is decoupled here; Block 3 sample sizes (~28k obs) live in the large-n regime
* von Auer 2023 is the German-language econometrics textbook that the labs reference


---
class: left

.blockquote[Part 1 · Bridge to Lab 1]

## Sneak Peek: what Lab&nbsp;1 actually does

.pull-left[
Lab&nbsp;1 is **Path E1 + the gateway to E2**:

`r fa('magnifying-glass', fill='#502479')` **Question 1** &mdash; the advertising effect

- Fit baseline OLS on Store 114 (n ≈ 351)
- A/B/C diagnostics: autoplot, ACF, Durbin-Watson, VIF, scatter-vs-fit
- Robust SE (HC3 / Newey-West) where needed
- Discover OVB: `feat` and `deal` are **perfectly collinear** across all 83 stores
- Translator Memo T4: *"the historical data cannot answer this — we need an A/B test"*

`r fa('forward', fill='#7d0a52')` **Result:** Path E1 confirms its own limits → motivates Lab&nbsp;3.
]

.pull-right[
### Lab 1 timeline (45 min)

<div style="display:grid; grid-template-columns:50px 1fr; gap:0.3em 0.7em; font-size:0.83em;">

<div style="color:#D50006;">Step 0</div>
<div>Memo, navigation, three paths from the DGP</div>

<div style="color:#D50006;">Step 1a</div>
<div>Baseline model, log-log elasticity reading</div>

<div style="color:#D50006;">Step 1b</div>
<div>A/B/C diagnostics — autoplot, ACF, DW, VIF</div>

<div style="color:#D50006;">Step 1c</div>
<div>OVB analysis: feat × deal across 83 stores</div>

<div style="color:#D50006;">T4</div>
<div>Translator memo no. 1</div>

</div>

<br>

.blockquote[
.small[
The Lab 1 finale is the empirical motivation for Lab 3: OLS on historical data **cannot** answer "what would happen under intervention" — even when fit perfectly.
]
]
]

???
* This slide is the orientation map for students about to enter Lab 1
* The two columns reflect the .qmd structure exactly — students should see this map again as the lab's TOC
* Memo T4 is the bridge artefact between Lab 1 and Lab 3


---
class: inverse, center, middle

## Part 2

### The In-Sample Illusion


---
class: left

.blockquote[Part 2 · The trap]

## High In-Sample R² Lies — a quick demonstration

.pull-left[
We simulate $n=30$ observations from the true DGP
$$y_i = 2 + 1.3 \cdot x_i + \varepsilon_i, \quad \varepsilon_i \sim \mathcal{N}(0, 1)$$

and fit polynomial regressions of degree 1, 5, and 15 on the **same training data**.

We then evaluate each model on a **fresh test set** drawn from the same DGP.

What do you expect to see?

`r fa('arrow-right', fill='#686868')` In-sample $R^2$ ↗ with degree (always)

`r fa('arrow-right', fill='#686868')` Out-of-sample $R^2$ — peaks somewhere, then collapses

The collapse is the **overfitting penalty**.
]

.pull-right[
```{r in-sample-demo, echo=FALSE, fig.width=5.5, fig.height=5, fig.align='center', out.width="100%"}
set.seed(20260522)
n_train <- 30
n_test  <- 100

# true DGP
gen <- function(n) {
  x <- runif(n, -2, 2)
  y <- 2 + 1.3 * x + rnorm(n, sd = 1)
  tibble(x = x, y = y)
}

tbl_train <- gen(n_train)
tbl_test  <- gen(n_test)

x_grid <- tibble(x = seq(-2, 2, length.out = 200))

fit_poly <- function(deg) {
  mod <- lm(y ~ poly(x, degree = deg, raw = TRUE), data = tbl_train)
  x_grid %>%
    mutate(degree = deg,
           yhat   = predict(mod, newdata = x_grid))
}

preds <- map_dfr(c(1, 5, 15), fit_poly) %>%
  mutate(degree = factor(degree,
                         labels = c("degree 1 (good)",
                                    "degree 5 (sweet spot?)",
                                    "degree 15 (overfit)")))

ggplot() +
  geom_point(data = tbl_train, aes(x, y),
             color = "#686868", alpha = 0.75, size = 1.8) +
  geom_line(data = preds, aes(x, yhat, color = degree),
            linewidth = 1.05) +
  scale_color_manual(values = c(col_secondary, col_primary, col_tertiary)) +
  facet_wrap(~ degree, nrow = 3) +
  labs(x = NULL, y = NULL,
       title = "Same data, three model complexities",
       subtitle = "Grey points = training data") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"),
        plot.subtitle = element_text(color = "#686868", size = 9))
```
]

???
* The left column primes the expectation; the right column shows the visual.
* This is the classic Schmueli/Taddy/James-et-al overfitting demo — keep it light, no formulas on this slide.
* The degree-15 curve will *visibly* swing wildly near the edges; that's the teaching moment.


---
class: left

.blockquote[Part 2 · The numbers]

## In-Sample vs. Out-of-Sample R² — the gap

```{r in-sample-r2, echo=FALSE, fig.width=8, fig.height=4, fig.align='center', out.width="80%"}
# reuse training data + same DGP for fresh test sets
set.seed(20260522)
n_train <- 30
n_test  <- 1000

gen <- function(n) {
  x <- runif(n, -2, 2)
  y <- 2 + 1.3 * x + rnorm(n, sd = 1)
  tibble(x = x, y = y)
}

tbl_train <- gen(n_train)
tbl_test  <- gen(n_test)

r2_fun <- function(y, yhat) {
  ss_res <- sum((y - yhat)^2)
  ss_tot <- sum((y - mean(y))^2)
  1 - ss_res / ss_tot
}

r2_tbl <- map_dfr(1:15, function(deg) {
  mod   <- lm(y ~ poly(x, degree = deg, raw = TRUE), data = tbl_train)
  yhat_in  <- predict(mod, newdata = tbl_train)
  yhat_out <- predict(mod, newdata = tbl_test)
  tibble(degree   = deg,
         r2_in    = r2_fun(tbl_train$y, yhat_in),
         r2_out   = r2_fun(tbl_test$y,  yhat_out))
}) %>%
  pivot_longer(starts_with("r2"), names_to = "set", values_to = "r2") %>%
  mutate(set = recode(set,
                      r2_in  = "In-sample (training)",
                      r2_out = "Out-of-sample (test)"))

ggplot(r2_tbl, aes(degree, r2, color = set)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.2) +
  scale_color_manual(values = c("In-sample (training)"     = col_primary,
                                "Out-of-sample (test)"     = col_secondary)) +
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11, 13, 15)) +
  scale_y_continuous(limits = c(NA, 1.02)) +
  labs(x = "Polynomial degree (model complexity)",
       y = expression(R^2),
       color = NULL,
       title = "In-sample R² keeps climbing — out-of-sample R² collapses",
       subtitle = "Same DGP as left slide; n_train = 30, n_test = 1,000") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",
        plot.subtitle = element_text(color = "#686868", size = 10))
```

<br>

.pull-left[
### The trap

`r fa('triangle-exclamation', fill='#D50006')` In-sample $R^2$ can be pushed arbitrarily close to 1 by adding complexity.

`r fa('triangle-exclamation', fill='#D50006')` But the model has memorized noise, not signal.

`r fa('triangle-exclamation', fill='#D50006')` The right metric is **out-of-sample**.
]

.pull-right[
### The remedies (preview)

`r fa('shield-halved', fill='#502479')` **Train/test split** — naive baseline

`r fa('shield-halved', fill='#502479')` **K-fold CV** — gold standard for choosing complexity

`r fa('shield-halved', fill='#502479')` **Lasso** — embed complexity control in the loss function
]

???
* Sweet spot here is degree 1-2 — the true DGP IS linear. Higher degrees only fit noise.
* If you have time, run the demo live with set.seed(...) variation — the in-sample curve is monotone, the OOS curve always crashes
* This visual is the conceptual entry point to Lab 1 (train/test) and Lab 2 (CV + Lasso)


---
class: left

.blockquote[Part 2 · Bridge to Lab 1]

## What to take into Lab 1

.pull-left[
### Going in
You will:

`r fa('arrow-right', fill='#D50006')` Fit a baseline OLS on Store 114 (Lab&nbsp;1 Step&nbsp;1a)

`r fa('arrow-right', fill='#D50006')` Read the regression table — coefficient, $R^2$, F-statistic

`r fa('arrow-right', fill='#D50006')` **Resist** the urge to declare victory based on $R^2$ alone (cf. Part&nbsp;2 demo)

`r fa('arrow-right', fill='#D50006')` Move on to **A/B/C diagnostics** before any managerial conclusion
]

.pull-right[
### What to look for in the diagnostics

`r fa('check', fill='#502479')` **A1 violation** &mdash; do `feat` and `deal` move together?

`r fa('check', fill='#502479')` **B2 violation** &mdash; is the residual variance constant across price levels?

`r fa('check', fill='#502479')` **B3 violation** &mdash; are weekly residuals autocorrelated within store?

`r fa('check', fill='#502479')` **C1 violation** &mdash; does feature advertising correlate with something we *cannot* observe?

The diagnostics tell you whether the regression table can be trusted at all.
]

<br>

.center[
.blockquote[
*Lab 1 is the diagnostic discipline; Lab 2 is the prediction discipline; Lab 3 is the causal discipline. All three are needed for a translator's memo.*
]
]

???
* This slide is the "elevator pitch" before students dive into Lab 1
* The four checks correspond to the four diagnostic sections of the Lab 1 .qmd
* Translator framing closes the part — we move from theory to lab


---
class: inverse, center, middle

## Lab 1 — Reality Check via Out-of-Sample Validation

### `r fa('flask', fill='white')` 45 min · Path E1, Steps 1a–1c · Memo T4

