---
title: "Lab 3: Baseline Model --- Simple Linear Regression"
subtitle: "CRISP-DM Phases 3 & 4 --- Data Preparation & Modelling"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en-US
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
bibliography: ../../lit/my_bib.bib
---

```{r setup}
#| include: false
library(tidyverse)    # data wrangling & ggplot2
library(moderndive)   # get_regression_table(), get_regression_points()
library(broom)        # augment() for residual diagnostics
library(gt)           # formatted tables
library(patchwork)    # combine ggplots
library(ISLR2)        # Carseats dataset

tbl_carseats <- tibble(Carseats) %>%
  mutate(ShelveLoc = factor(ShelveLoc,        # correct level order from Lab 2
                            levels = c("Bad", "Medium", "Good")))
```

## Learning Objectives and Overview

:::{.callout-note}
## Learning Objectives

- Understand simple linear regression as a **descriptive tool**: conditional means,
  not causal claims
- Read and interpret the OLS regression equation and its coefficients correctly
- Compute and interpret the **coefficient of determination R²** ---
  and understand why a low R² is information, not failure
- Use **residuals** as a measure of store-level underperformance
- Diagnose **Omitted Variable Bias (OVB)** visually in the residual plot
:::

**Scenario:** In Lab 2 you explored the Carseats data and established that
`Price` is negatively correlated with `Sales` ($r \approx -0.44$).
Now you build the simplest possible regression model --- the **baseline** ---
and critically assess what it reveals and what it hides.

## 🧭 How to Navigate This Lab

| Symbol | Box colour | Meaning | When to read |
|--------|-----------|---------|-------------|
| ⚠️ | <span style="color:#e02424">**Red — Action Required**</span> | Run code, interpret output, answer questions | **Do this in the live session** |
| 💡 | <span style="color:#2f9e44">**Green — Best Practice / Tip**</span> | Coding tips, R functions, shortcuts | Skim in the lab |
| 📚 | <span style="color:#1c7ed6">**Blue, collapsed — Background Theory**</span> | Concepts, theory, derivations | Read at home |
| ℹ️ | <span style="color:#1c7ed6">**Blue, open — Note**</span> | Short clarifications | Read when reached |

⏱ **40 minutes** for mandatory tasks (A1--A4). Bonus tasks if time permits.

---

## Icebreaker: The Same Line --- Different Worlds

In Lab 2 the Datasaurus Dozen showed that identical summary statistics can hide
completely different data structures. The same principle applies to regression:
identical OLS lines can describe fundamentally different relationships.

```{r anscombe-stats}
#| code-fold: true
#| code-summary: "Show code: Anscombe summary statistics"

# Anscombe's Quartet: four datasets with (almost) identical regression statistics
anscombe_tidy <- anscombe %>%
  mutate(observation = seq_len(n())) %>%
  tidyr::pivot_longer(-observation,
                      names_to = c(".value", "set"),
                      names_pattern = "(.)(.)")  %>%
  mutate(set = c("I", "II", "III", "IV")[as.integer(set)])

# Summary: same correlation, same slope, same R² across all four
anscombe_tidy %>%
  group_by(set) %>%
  summarise(
    n        = n(),
    mean_x   = round(mean(x), 2),
    mean_y   = round(mean(y), 2),
    cor_xy   = round(cor(x, y), 3),       # Pearson correlation
    slope    = round(coef(lm(y ~ x))[2], 3),  # OLS slope
    r_sq     = round(cor(x, y)^2, 3)                               # R² = r² for simple regression
  ) %>%
  gt() %>%
  tab_header(
    title    = "Anscombe's Quartet --- summary statistics",
    subtitle = "All four datasets have nearly identical regression statistics"
  ) %>%
  tab_options(table.font.size = "13px")
```

Now look at what the OLS line actually fits:

```{r anscombe-plots}
#| code-fold: true
#| code-summary: "Show code: Anscombe scatter plots"
#| fig-height: 5
#| fig-width: 9

anscombe_tidy %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(color = "#7d0a52", size = 2.5, alpha = 0.8) +   # raw observations
  geom_smooth(method = "lm", se = FALSE,                      # OLS line
              color = "#D50006", linewidth = 1.2) +
  facet_wrap(~set, ncol = 4) +                                # one panel per dataset
  # --- labels ---
  labs(
    title    = "Anscombe's Quartet --- same line, different worlds",
    subtitle = "Panel I: linear ✓ | II: curved | III: outlier in y | IV: leverage point",
    x = "x", y = "y"
  ) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))
```

:::{.callout-important}
## The Central Lesson for Regression

Panel I shows a genuinely linear relationship --- the OLS line is appropriate.
Panels II--IV show cases where the same line misleads: curvature, a single outlier,
and a single leverage point each produce the same fitted line but completely different
underlying structures.

**Consequence:** Fitting a regression model is not enough. The residuals must always
be inspected visually. This is what Step 3 of this lab teaches.

The OLS line in Panel I is also the **benchmark** concept we use today:
the predicted value $\hat{y}_i$ is the *conditional mean* --- what a store *on average*
achieves at a given price. The residual $e_i = y_i - \hat{y}_i$ measures the deviation
from this benchmark.
:::

---

## Background: Where We Are in CRISP-DM

In Lab 2 (Phase 2: Data Understanding) we established the key EDA findings:
`Price` is negatively correlated with `Sales`, `ShelveLoc` creates large group
differences, and the log transformation makes `Sales` distribution worse, not better.
Now we enter Phases 3 & 4: we prepare the data and fit the first model.

```{r crisp-overview}
#| echo: false

tibble::tribble(
  ~Phase, ~Name,                   ~Content,                                           ~Status,
  "1",    "Business Understanding","Goal: identify underperforming stores",            "Done (lecture)",
  "2",    "Data Understanding",    "EDA: distributions, correlations, OVB signals",   "Done (Lab 2)",
  "3",    "Data Preparation",      "Factor recoding, zero handling, variable selection","← This lab",
  "4",    "Modelling",             "Baseline model: Sales ~ Price",                   "← This lab",
  "5",    "Evaluation",            "R², residuals, OVB diagnosis",                    "← This lab",
  "6",    "Deployment",            "Performance ranking, managerial memo",             "Lab 4"
) %>%
  gt() %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = Phase %in% c("3", "4", "5"))
  ) %>%
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_column_labels()
  ) %>%
  tab_options(table.font.size = "13px")
```

---

## Background: Simple Linear Regression as a Descriptive Tool

### What Regression Is --- and What It Is Not

:::{.callout-note collapse="true"}
## 📚 Background: The Etymology and History of Regression

The term *regression* derives from the Latin *regredi* (to go back, to return).
It was introduced into statistics by **Francis Galton (1822--1911)**, who observed
that the children of unusually tall parents tend to be taller than average ---
but not as tall as their parents. Extreme characteristics *regress* towards the mean
across generations.

The modern statistical meaning is broader: regression is any method that describes
how the conditional mean of one variable changes with another. The "returning"
interpretation is now purely etymological.

In German-language statistics, the term *Regressionsanalyse* (@stoetzer_regressionsanalyse_2020)
covers both the descriptive and inferential variants. This lab covers only the
**descriptive** case --- the 400 stores are our complete population, not a sample.
:::

Simple linear regression is often introduced alongside inferential machinery:
standard errors, p-values, confidence intervals. Before we reach that machinery,
it is worth stepping back and asking: *what is regression fundamentally?*

**At its core, simple linear regression is a descriptive technique.**
Given $n$ observations $(x_i, y_i)$, OLS finds the straight line that best summarises
the relationship between $x$ and $y$ in this dataset. In our case: the line that
best describes how average sales vary with price *across the 400 stores we have*.

The 400 Carseats stores are our **population** --- we are not generalising from a
sample to a larger world. We are describing what is already in front of us.

:::{.callout-note collapse="true"}
## 📚 Background: Exact vs. Statistical Relationships

In some situations, the relationship between two variables is exact and deterministic.
At a petrol station, the bill is exactly proportional to the number of litres filled:
$y = p \cdot x$, with no scatter around the line.

In most business situations, the relationship is statistical: the same price does
not always produce the same sales, because other factors --- shelf location,
advertising, local competition, demographics --- also play a role. The OLS line
captures the *systematic* part of this relationship; the residuals capture the rest.

$$y_i = \underbrace{\hat\beta_0 + \hat\beta_1 x_i}_{\text{systematic}} + \underbrace{e_i}_{\text{unsystematic}}$$

The statistical relationship is not a defect --- it is the reality of business data.
The residuals are the starting point for understanding *why* some stores deviate
from the systematic pattern (@stoetzer_regressionsanalyse_2020 [p. 47]).
:::

### Regression as Conditional Means

The OLS line is a machine for computing **conditional means**:

$$\hat{y}_i = \hat\beta_0 + \hat\beta_1 \cdot x_i \approx E[Y \mid X = x_i]$$

For each value of $x$ (price), the line gives the average $y$ (sales) across all
stores with that price. The coefficients describe the *systematic* part:

- $\hat\beta_1$ **(slope):** how much does the conditional mean of Sales change when
  Price increases by one unit?
- $\hat\beta_0$ **(intercept):** predicted Sales when Price = 0 --- rarely interpretable
  in practice, but it anchors the line.

:::{.callout-note collapse="true"}
## 📚 Background: Why OLS? Three Ways to Draw a Line

There are many ways to draw a line through a scatter plot.
The Rmd document @stoetzer_regressionsanalyse_2020 illustrates three:

1. **OLS (vertical distances):** minimise $\sum e_i^2$ where $e_i = y_i - \hat{y}_i$.
   This is the standard regression of Y on X.
2. **Reverse regression (horizontal distances):** minimise horizontal distances ---
   this gives the regression of X on Y, which is a different line.
3. **Orthogonal regression (perpendicular distances):** minimise the shortest distance
   from each point to the line --- used in Principal Component Analysis.

OLS (vertical distances) is appropriate when $x$ is measured without error and
we want to predict $y$. The choice of distance measure is a modelling decision,
not a technical one.

**Key property of OLS residuals** (follows from the first-order conditions):

$$\sum_{i=1}^n e_i = 0 \quad \Rightarrow \quad \bar{e} = 0$$

The residuals always sum to zero for any OLS regression with an intercept.
This means: the OLS line always passes through the point $(\bar{x}, \bar{y})$.
:::

### The OLS Formula and Its Link to Correlation

OLS selects $\hat\beta_0$ and $\hat\beta_1$ to minimise the sum of squared residuals:

$$\min_{\hat\beta_0,\, \hat\beta_1} \sum_{i=1}^{n} e_i^2 = \sum_{i=1}^{n}(y_i - \hat\beta_0 - \hat\beta_1 x_i)^2$$

The analytical solutions (from setting first-order conditions to zero) are:

$$\hat\beta_1 = \frac{\text{Cov}(x, y)}{\text{Var}(x)} = r_{xy} \cdot \frac{s_y}{s_x}
\qquad
\hat\beta_0 = \bar{y} - \hat\beta_1 \cdot \bar{x}$$

The slope $\hat\beta_1$ is directly proportional to the correlation $r_{xy}$ ---
scaled by the ratio of standard deviations to express it in the original units.
This means the sign of the slope always equals the sign of the correlation.

### The Coefficient of Determination R²

The total variation in $y$ decomposes into explained and unexplained parts:

$$\underbrace{\sum_i(y_i - \bar{y})^2}_{\text{TSS}} =
\underbrace{\sum_i(\hat{y}_i - \bar{y})^2}_{\text{ESS}} +
\underbrace{\sum_i e_i^2}_{\text{SSR}}$$

$$R^2 = \frac{\text{ESS}}{\text{TSS}} = 1 - \frac{\text{SSR}}{\text{TSS}} \in [0, 1]$$

:::{.callout-note}
## Interpreting R² Correctly

$R^2 = 0.15$ means: **15% of the variance in Sales** is explained by Price alone.
The remaining 85% is attributable to other factors not yet in the model.

A low R² is **not a failure** --- it is diagnostic information.
It tells us precisely that important variables are missing.
In our case, it motivates adding `ShelveLoc` and `Advertising` in the Demo.

For simple regression only: $R^2 = r_{xy}^2$ --- the square of the Pearson correlation.
:::

### What the Regression Table Shows

When we fit an OLS model on our complete population of 400 stores, the output
from `lm()` includes more columns than we need today:

| Column | Meaning | Today |
|--------|---------|-------|
| `estimate` | The coefficient $\hat\beta$ | ✅ Focus here |
| `std_error` | Standard error (sampling concept) | → PVA2 |
| `statistic` | t-value (sampling concept) | → PVA2 |
| `p_value` | p-value (sampling concept) | → PVA2 |
| `lower_ci` / `upper_ci` | Confidence interval (sampling concept) | → PVA2 |

:::{.callout-important}
## Parameters, Not Estimates

Since we treat the 400 stores as our **population**, the column labelled `estimate`
is somewhat misleading: $\hat\beta$ is not an *estimate* (which implies sampling
uncertainty) but a **parameter** --- the exact slope for our 400 stores.

Standard errors, t-values, p-values, and confidence intervals are sampling concepts.
They quantify how results would vary across hypothetical repeated samples --- which
is not relevant when we already have the complete population.

**For this lab:** we focus exclusively on the `estimate` column.
The remaining columns become meaningful in PVA2.
:::

:::{.callout-note collapse="true"}
## 📚 Background: `moderndive` --- A Tidy-Compatible Wrapper for `lm()`

The base R function `lm()` produces an object of class `lm`, which does not
behave like a tibble and can cause issues in pipes. The `moderndive` package
(@ismay_statistical_2023) provides tidy wrappers:

| Function | What it returns |
|----------|----------------|
| `get_regression_table()` | Coefficients as a tibble (like `tidy()` from `broom`) |
| `get_regression_points()` | Fitted values and residuals for each observation |
| `get_regression_summaries()` | R², RMSE, and other model-level statistics |

These functions work seamlessly in pipes and produce tibbles that can be passed
directly to `gt()` for formatting.

Note: `moderndive` applies a degrees-of-freedom correction internally, consistent
with the descriptive (population) interpretation of regression used in this lab.
The coefficients are identical to base R `lm()` --- only the output format differs.
:::

---

## Step 1: Fit the Baseline Model

Before running `lm()`, write down what we are estimating:

$$\widehat{\text{Sales}}_i = \hat\beta_0 + \hat\beta_1 \cdot \text{Price}_i$$

We expect $\hat\beta_1 < 0$: stores with higher prices should, on average, sell
fewer units. The EDA in Lab 2 confirmed this with $r \approx -0.44$.

```{r baseline-fit}
#| code-fold: false

# Fit the simple linear regression model
mod_baseline <- tbl_carseats %>%
  lm(Sales ~ Price, data = .)          # formula: Sales as function of Price

# Regression table: tidy output via moderndive
mod_baseline %>%
  get_regression_table() %>%
  gt() %>%
  fmt_number(
    columns = c(estimate, std_error, statistic, p_value, lower_ci, upper_ci),
    decimals = 4
  ) %>%
  # --- highlight slope row ---
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = term == "Price")
  ) %>%
  tab_options(table.font.size = "14px") %>%
  tab_footnote("Yellow = slope for Price | Columns 2–6 are sampling concepts → PVA2")
```

```{r baseline-quality}
#| code-fold: false

# Model quality: R², RMSE
mod_baseline %>%
  get_regression_summaries() %>%
  select(r_squared, adj_r_squared, rmse, nobs) %>%
  gt() %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse), decimals = 4) %>%
  tab_options(table.font.size = "14px") %>%
  tab_header("Baseline model quality: Sales ~ Price")
```

:::{.callout-important icon=false}
## ⚠️ A1: Interpret the Coefficients

Write out the regression equation with the actual numbers and interpret each part:

$$\widehat{\text{Sales}} = \hat\beta_0 + \hat\beta_1 \cdot \text{Price}$$

1. **Intercept $\hat\beta_0$:** What does this value mean literally?
   Is it sensibly interpretable for a car-seat store (Price = 0)?

2. **Slope $\hat\beta_1$:** If a store raises its price by 10 USD,
   what is the predicted change in Sales? (Express in units.)

3. **R²:** What percentage of the variance in Sales does Price explain?
   Is this a "good" or "bad" model? Or is that the wrong question?

4. **RMSE:** The Root Mean Squared Error is measured in the same units as Sales.
   What does it tell us about the average prediction accuracy?
:::

---

## Step 2: Visualise the Regression Line

```{r scatter-fit-core}
#| code-fold: false
#| fig-height: 4.5

tbl_carseats %>%
  ggplot(aes(x = Price, y = Sales)) +
  geom_point(alpha = 0.35, color = "#7d0a52", size = 1.8) +    # raw observations
  geom_smooth(method = "lm", color = "#D50006",                 # OLS line = conditional mean
              se = FALSE, linewidth = 1.2)                      # se=FALSE: population, not sample
```

```{r scatter-fit-layout}
#| code-fold: true
#| code-summary: "Show code: labels, annotation & theme"
#| fig-height: 4.5

# Extract coefficients via tidy pipe (no $ operator)
b0 <- mod_baseline %>%
  get_regression_table() %>%
  filter(term == "intercept") %>%
  pull(estimate) %>%
  round(2)

b1 <- mod_baseline %>%
  get_regression_table() %>%
  filter(term == "Price") %>%
  pull(estimate) %>%
  round(2)

r2 <- mod_baseline %>%
  get_regression_summaries() %>%
  pull(r_squared) %>%
  round(3)

tbl_carseats %>%
  ggplot(aes(x = Price, y = Sales)) +
  geom_point(alpha = 0.35, color = "#7d0a52", size = 1.8) +
  geom_smooth(method = "lm", color = "#D50006",
              se = FALSE, linewidth = 1.2) +
  annotate("text",                                              # equation annotation
    x = 70, y = 16,
    label = paste0("Sales\u0302 = ", b0, " + (", b1, ") \u00d7 Price\nR\u00b2 = ", r2),
    color = "#D50006", size = 3.5, hjust = 0, fontface = "bold"
  ) +
  # --- labels ---
  labs(
    title    = "Baseline Model: Sales ~ Price",
    subtitle = "OLS line = conditional mean E[Sales | Price]  |  se = FALSE: population analysis",
    x = "Price (USD)", y = "Sales (thousands of units)"
  ) +
  theme_minimal(base_size = 11)
```

:::{.callout-tip}
## 💡 Why `se = FALSE`?

`geom_smooth()` draws a grey confidence band by default (`se = TRUE`).
This band represents inferential uncertainty --- how much the slope might vary
across hypothetical repeated samples. Since we treat the 400 stores as the
complete population, there is no sampling uncertainty. Setting `se = FALSE`
removes a visually prominent but conceptually misleading element.
:::

:::{.callout-important icon=false}
## ⚠️ A2: Read the Regression Line

1. Pick one point in the scatter plot that lies clearly **above** the OLS line.
   What does its position mean in business terms?

2. Pick one point that lies clearly **below** the line.
   What does its position mean?

3. The OLS line always passes through the point $(\bar{x}, \bar{y})$.
   Verify this: what are the mean values of Price and Sales in the dataset?

```r
tbl_carseats %>%
  summarise(
    mean_price = mean(Price),
    mean_sales = mean(Sales)
  )
```
:::

---

## Step 3: Residual Analysis

### What Are Residuals?

$$e_i = y_i - \hat{y}_i = \text{Sales}_i - \widehat{\text{Sales}}_i$$

The residual measures how far a store's actual sales deviate from the conditional
mean at its price level. It is the **deviation from the market benchmark**:

- $e_i < 0$: store sells **less** than the price benchmark predicts → underperformer
- $e_i > 0$: store sells **more** than expected → outperformer
- $e_i \approx 0$: store is at the market average

This deviation score is the foundation for performance mining in →Lab 4.

### Residual Table

```{r residuals}
#| code-fold: false

# Compute fitted values and residuals for all 400 stores
reg_points <- mod_baseline %>%
  get_regression_points()        # returns: ID, Sales, Price, Sales_hat, residual

# Show first 10 rows
reg_points %>%
  head(10) %>%
  gt() %>%
  fmt_number(columns = c(Sales, Price, Sales_hat, residual), decimals = 3) %>%
  # --- colour-code extreme residuals ---
  tab_style(
    style     = cell_fill(color = "#fff0f0"),
    locations = cells_body(rows = residual < -2)   # underperformers
  ) %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = residual > 2)    # outperformers
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("Red = residual < -2 (underperformer) | Green = residual > 2 (outperformer)")
```

### Residual Plot

```{r residual-plot-core}
#| code-fold: false
#| fig-height: 4

# Augment: add fitted values and residuals to original data
pts <- mod_baseline %>%
  augment(data = tbl_carseats)   # adds .fitted, .resid, etc.

# Residuals vs. fitted values
pts %>%
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4, color = "#7d0a52", size = 1.5) +    # one point per store
  geom_hline(yintercept = 0, color = "#D50006",                # zero reference line
             linewidth = 1.1, linetype = "dashed") +
  geom_smooth(method = "loess", color = "#502479",             # LOESS: detect patterns
              se = FALSE, linewidth = 0.8)
```

```{r residual-plot-layout}
#| code-fold: true
#| code-summary: "Show code: labels & theme"
#| fig-height: 4

pts %>%
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4, color = "#7d0a52", size = 1.5) +
  geom_hline(yintercept = 0, color = "#D50006",
             linewidth = 1.1, linetype = "dashed") +
  geom_smooth(method = "loess", color = "#502479",
              se = FALSE, linewidth = 0.8) +
  # --- labels ---
  labs(
    title    = "Residuals vs. Fitted Values",
    subtitle = "Ideal: random scatter around 0 | LOESS (purple) detects systematic patterns",
    x = "Predicted Sales (Ŝales)", y = "Residual"
  ) +
  theme_minimal(base_size = 11)
```

:::{.callout-tip}
## 💡 What to Look For in a Residual Plot

An ideal residual plot shows **random scatter** around the zero line ---
no systematic upward or downward trends, no U-shape, no funnel shape.

The LOESS curve (purple) helps detect patterns the eye might miss.
If it deviates substantially from zero, it signals a problem with the model.

| Pattern | Signal |
|---------|--------|
| Random scatter | Model captures the relationship well |
| U-shape (LOESS curves up then down) | Non-linearity: need a curved term |
| Funnel shape (wider at right) | Heteroskedasticity: variance grows with fitted values |
| Horizontal bands by colour | Missing categorical variable (→ Step 4) |
:::

:::{.callout-important icon=false}
## ⚠️ A3: Interpret the Residual Plot

1. Does the residual plot show random scatter, or do you detect a systematic pattern?
   Describe what you see.

2. What would an *ideal* residual plot look like for a well-specified model?

3. The LOESS curve (purple) is approximately flat around zero.
   Does this suggest a linearity problem?
:::

---

## Step 4: OVB Diagnosis

The residual plot in Step 3 looks approximately random when we ignore group membership.
But we know from Lab 2 that `ShelveLoc` is strongly associated with `Sales`.
What happens when we colour the residuals by shelf location?

```{r ovb-diagnosis-core}
#| code-fold: false
#| fig-height: 4

pts %>%
  ggplot(aes(x = .fitted, y = .resid, color = ShelveLoc)) +
  geom_point(alpha = 0.5, size = 1.5) +                        # colour = shelf location
  geom_hline(yintercept = 0, color = "#686868",
             linewidth = 1, linetype = "dashed") +
  scale_color_manual(values = c("Bad"    = "#D50006",
                                "Medium" = "#b3b2b2",
                                "Good"   = "#502479"))
```

```{r ovb-diagnosis-layout}
#| code-fold: true
#| code-summary: "Show code: labels & theme"
#| fig-height: 4

pts %>%
  ggplot(aes(x = .fitted, y = .resid, color = ShelveLoc)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_hline(yintercept = 0, color = "#686868",
             linewidth = 1, linetype = "dashed") +
  scale_color_manual(values = c("Bad"    = "#D50006",
                                "Medium" = "#b3b2b2",
                                "Good"   = "#502479")) +
  # --- labels ---
  labs(
    title    = "Residuals coloured by shelf location",
    subtitle = "Systematic colour banding = OVB fingerprint: ShelveLoc is missing from the model",
    x = "Predicted Sales (Ŝales)", y = "Residual",
    color    = "ShelveLoc"
  ) +
  theme_minimal(base_size = 11)
```

The systematic colour banding is the **OVB fingerprint**: Good (purple) clusters
above zero, Bad (red) clusters below. A variable that the model should include
is creating a pattern in what should be random noise.

:::{.callout-note collapse="true"}
## 📚 Background: Omitted Variable Bias (OVB)

OVB occurs when a variable that is correlated with *both* the regressor ($x$) and
the outcome ($y$) is left out of the model. The slope coefficient $\hat\beta_1$
then absorbs part of the omitted variable's effect, making it biased.

In our case, the OVB formula for the Price coefficient is:

$$\hat\beta_{\text{Price, biased}} = \beta_{\text{Price}} +
\underbrace{\beta_{\text{ShelveLoc}} \cdot \frac{\text{Cov(Price, ShelveLoc)}}{\text{Var(Price)}}}_{\text{bias term}}$$

From Lab 2 we know:
- Good shelf locations are associated with **higher** prices (Cov > 0)
- Good shelf locations are associated with **higher** sales ($\beta_{\text{ShelveLoc}} > 0$)

Therefore: the bias term is positive, meaning $\hat\beta_{\text{Price}}$ is
*less negative* than the true price effect. The baseline model underestimates
how strongly price reduces sales.
:::

:::{.callout-important icon=false}
## ⚠️ A4: Diagnose the OVB

1. Describe the colour pattern in the residual plot.
   Where do Bad (red), Medium (grey), and Good (purple) stores tend to cluster?

2. Verify the mechanism from Lab 2: do stores with better shelf locations
   also charge higher prices on average?

```r
tbl_carseats %>%
  group_by(ShelveLoc) %>%
  summarise(
    mean_price = round(mean(Price), 2),
    mean_sales = round(mean(Sales), 2),
    n          = n()
  )
```

3. Now check empirically how much the Price coefficient changes when ShelveLoc
   is added to the model:

```r
# Baseline: without ShelveLoc
mod_without <- tbl_carseats %>% lm(Sales ~ Price, data = .)

# Extended: with ShelveLoc
mod_with <- tbl_carseats %>% lm(Sales ~ Price + ShelveLoc, data = .)

# Compare price coefficients
mod_without %>%
  get_regression_table() %>%
  filter(term == "Price") %>%
  pull(estimate)

mod_with %>%
  get_regression_table() %>%
  filter(term == "Price") %>%
  pull(estimate)
```

   In which direction does the Price coefficient change?
   Is this consistent with the OVB formula in the background box?

4. **R² check:** compute the R² of `mod_with`. How much does it improve over
   the baseline? Does this match what the residual plot suggested?
:::

---

## Summary

```{r summary-table}
#| echo: false

# Pre-compute values for summary table
r2_baseline <- mod_baseline %>%
  get_regression_summaries() %>%
  pull(r_squared) %>%
  round(3)

b1_baseline <- mod_baseline %>%
  get_regression_table() %>%
  filter(term == "Price") %>%
  pull(estimate) %>%
  round(3)

tibble::tribble(
  ~Criterion,         ~`Baseline (Sales ~ Price)`,          ~`Expected with ShelveLoc`,
  "R²",               as.character(r2_baseline),             "~0.50 (substantially better)",
  "β_Price",          paste0(b1_baseline, " per USD"),       "Changes when OVB corrected",
  "Residual plot",    "Random scatter overall...",           "Pattern disappears",
  "...but",           "Systematic colour banding by ShelveLoc", "No residual banding",
  "Conclusion",       "OVB: ShelveLoc missing from model",  "Multiple regression → Demo"
) %>%
  gt() %>%
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style     = cell_fill(color = "#fff0f0"),
    locations = cells_body(rows = Criterion == "Conclusion")
  ) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-important}
## Key Messages of This Lab

1. **Regression = conditional mean** --- not a causal claim, but a benchmark.
   The OLS line gives the average Sales for each Price level.

2. **Population framing:** the 400 stores are our complete dataset.
   $\hat\beta$ is a **parameter**, not an estimate. Standard errors, p-values,
   and confidence intervals are sampling concepts --- they become relevant in PVA2.

3. **Low R² = diagnostic information, not failure.**
   R² = `r r2_baseline` tells us that Price alone explains only
   `r round(r2_baseline * 100, 1)`% of sales variance.
   The missing 85% is a signal: important variables are absent.

4. **The residual plot diagnoses OVB.** Systematic colour banding by ShelveLoc
   reveals that the Price coefficient is biased by a missing variable.

5. **OVB direction:** Good shelves command higher prices *and* generate higher sales.
   This positive correlation pulls $\hat\beta_{\text{Price}}$ towards zero ---
   the baseline underestimates how strongly price reduces sales.
:::

:::{.callout-note}
## Transition to the Demo

In the **live demo**, the instructor adds `ShelveLoc` and `Advertising` to the model.
You will see:

- R² rise from ~`r r2_baseline` to ~0.50
- The Price coefficient become more negative (OVB correction)
- The residual plot lose its systematic colour banding

This is the **multiple regression model** --- and it leads directly to →Lab 4:
Performance Mining, where residuals become store-level performance scores.
:::

---

## Bonus Tasks

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B1: Compute R² Manually

Verify R² from first principles using the variance decomposition TSS = ESS + SSR:

```r
reg_points %>%
  summarise(
    TSS = sum((Sales - mean(Sales))^2),   # total variation in Sales
    SSR = sum(residual^2),                 # unexplained variation
    ESS = TSS - SSR,                       # explained variation
    R2  = ESS / TSS                        # share explained
  )
```

Does this match `get_regression_summaries()`?
:::

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B2: R² = r² in Simple Regression

Show empirically that for simple (not multiple) regression:
$$R^2 = r_{x,y}^2$$

```r
r_xy <- tbl_carseats %>%
  summarise(r = cor(Price, Sales)) %>%
  pull(r)

r_xy^2

mod_baseline %>%
  get_regression_summaries() %>%
  pull(r_squared)
```

Why does this equality hold **only** for simple regression?
(Hint: consider how $\hat{y}_i$ relates to $x_i$ in the simple case.)
:::

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B3: Visualise the Variance Decomposition

For the first 20 stores, draw the three components of variance:

```r
first20 <- reg_points %>%
  slice(1:20) %>%
  mutate(
    ybar   = mean(tbl_carseats$Sales),    # overall mean
    x_rank = row_number()
  )

ggplot(first20, aes(x = x_rank)) +
  geom_segment(aes(xend = x_rank, y = ybar,      yend = Sales),
               color = "#b3b2b2", linewidth = 1) +      # TSS (grey)
  geom_segment(aes(xend = x_rank, y = Sales_hat, yend = Sales),
               color = "#D50006", linewidth = 1.2) +     # SSR (red)
  geom_segment(aes(xend = x_rank, y = ybar,      yend = Sales_hat),
               color = "#502479", linewidth = 1.2) +     # ESS (purple)
  geom_point(aes(y = Sales),     color = "#7d0a52", size = 2) +
  geom_point(aes(y = Sales_hat), color = "#D50006",  shape = 17, size = 2) +
  geom_hline(yintercept = mean(tbl_carseats$Sales),
             linetype = "dashed", color = "#686868") +
  labs(title    = "Variance decomposition: TSS = ESS + SSR",
       subtitle = "Grey = TSS | Purple = ESS (explained) | Red = SSR (residual)",
       x = "Observation (rank)", y = "Sales") +
  theme_minimal()
```
:::

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B4: First Performance Ranking

Using the baseline residuals, identify the 10 most extreme underperformers:

```r
reg_points %>%
  arrange(residual) %>%
  head(10) %>%
  select(ID, Sales, Price, Sales_hat, residual) %>%
  left_join(
    tbl_carseats %>% mutate(ID = row_number()) %>%
      select(ID, ShelveLoc, Advertising, Urban, US),
    by = "ID"
  ) %>%
  gt() %>%
  fmt_number(columns = c(Sales, Sales_hat, residual), decimals = 2) %>%
  tab_style(
    style     = cell_fill(color = "#fff0f0"),
    locations = cells_body(rows = residual < -3)
  )
```

What do the top underperformers have in common?
Look at their ShelveLoc category --- is this consistent with the OVB diagnosis?
:::

## References {.unnumbered}
