---
title: "Lab 3: Baseline Model — Simple Regression"
subtitle: "CRISP-DM Phase 3 & 4 — Data Preparation & Modeling"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en-US
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse)
library(moderndive)
library(broom)
library(gt)
library(patchwork)
library(ISLR2)

tbl_carseats <- tibble(Carseats)
```

## Learning Objectives and Overview

:::{.callout-note}
## Learning Objectives

- Understand simple linear regression as a **descriptive** tool: conditional means, not causal claims
- Read the regression equation as a **model formula** and interpret its coefficients correctly
- Compute and interpret the **coefficient of determination R²**
- Use **residuals** as a measure of underperformance and read the residual plot
- Diagnose **Omitted Variable Bias (OVB)** in the residual plot
:::

**Scenario:** In Lab 2 you explored the Carseats data and established that
`Price` is negatively correlated with `Sales`. Now you build the simplest possible
regression model — the **baseline model** — and critically assess what it can and cannot do.

:::{.callout-tip}
## Schedule
⏱ **15 minutes** — Mandatory tasks (A1–A3), bonus tasks if time permits.
:::

---

## Background: Simple Linear Regression as a Descriptive Tool

### What Is Simple Linear Regression — and What Is It Not?

Simple linear regression is often introduced alongside inferential machinery: standard errors, p-values, confidence intervals. Before we reach that machinery, it is worth stepping back and clarifying what regression *fundamentally is*.

**At its core, simple linear regression is a descriptive technique.**

Given a dataset of $n$ observations $(x_i, y_i)$, OLS finds the straight line that best summarises the relationship between $x$ and $y$ in that dataset. In our case: the line that best describes how average sales vary with price *across the 400 stores we have*.

This is important: the 400 Carseats stores are our **population** (or at least the complete dataset of interest). We are not trying to generalise from a small sample to a larger world — we are describing what is already in front of us. This is the same descriptive spirit as EDA in Lab 2.

### Regression as Conditional Means

The OLS line can be understood as a machine for computing **conditional means**:

$$\hat{y}_i = \hat\beta_0 + \hat\beta_1 \cdot x_i \approx E[Y \mid X = x_i]$$

For each value of $x$ (price), the regression line gives the average $y$ (sales) across all stores with that price. The coefficients describe the *systematic* part of this relationship; residuals capture what is left over.

This perspective — regression as conditional means — makes coefficients intuitive:

- $\hat\beta_1$ is the **slope**: how much does the conditional mean of Sales change when Price increases by one unit?
- $\hat\beta_0$ is the **intercept**: what is the conditional mean of Sales when Price = 0? (Often not directly interpretable.)

### A Note on the Standard Regression Table

When we estimate a regression on a **population** (all 400 stores), the standard output from `lm()` includes more columns than we strictly need right now:

| Column | Meaning | Relevant? |
|--------|---------|-----------|
| `estimate` | The coefficient $\hat\beta$ | ✅ Yes — this is what we need |
| `std_error` | Standard error — how variable would $\hat\beta$ be across repeated samples? | ⚠️ Not meaningful for a population |
| `statistic` | t-value = estimate / std_error | ⚠️ Requires sampling context |
| `p_value` | Probability under $H_0: \beta = 0$ in a sampling framework | ⚠️ Not meaningful for a population |
| `lower_ci` / `upper_ci` | 95% confidence interval | ⚠️ Requires sampling context |

:::{.callout-important}
## Descriptive Context: What the Column Headers Actually Mean

The column labelled `estimate` is somewhat misleading here: in a population analysis, $\hat\beta$ is not an **estimate** (which implies sampling uncertainty) but a **parameter** — the exact value of the slope for our 400 stores. Similarly, standard errors, p-values, and confidence intervals are concepts from sampling theory: they quantify how much results might vary across hypothetical repeated samples. When we treat the 400 stores as the population, there is no such variation — we have *all* the data.

**Practical consequence for this lab:**\
We focus exclusively on the `estimate` column. The other columns (standard error, t-value, p-value, confidence interval) are shown by the software, but they only become conceptually meaningful in PVA2, when we treat our data as a sample from a larger process and ask inferential questions.
:::

### The Model Formula

Before turning to R, let us write down the model we are going to fit and make explicit what its parts mean.

The **simple linear regression model** for our baseline case is:

$$\widehat{\text{Sales}}_i = \hat\beta_0 + \hat\beta_1 \cdot \text{Price}_i$$

**Interpretation of the coefficients:**

- $\hat\beta_1$ (slope): the **partial derivative** of $\widehat{\text{Sales}}$ with respect to $\text{Price}$:
$$\hat\beta_1 = \frac{\partial \widehat{\text{Sales}}}{\partial \text{Price}}$$
In plain language: for each additional dollar in price, the conditional mean of Sales changes by $\hat\beta_1$ thousand units — all else being equal (which here means: within the dataset as it is).

- $\hat\beta_0$ (intercept): the predicted value of $\widehat{\text{Sales}}$ when $\text{Price} = 0$. This is rarely interpretable in practice (no store charges zero), but it anchors the line.

**How are the coefficients chosen?**

OLS selects $\hat\beta_0$ and $\hat\beta_1$ to minimise the **sum of squared residuals (SSR)**:

$$\min_{\hat\beta_0,\, \hat\beta_1} \sum_{i=1}^{n} e_i^2 = \sum_{i=1}^{n}(y_i - \hat\beta_0 - \hat\beta_1 x_i)^2$$

Setting the first-order conditions (partial derivatives = 0) yields analytical solutions:

$$\hat\beta_1 = \frac{s_{xy}}{s_x^2} = \frac{\sum_i(x_i - \bar{x})(y_i - \bar{y})}{\sum_i(x_i - \bar{x})^2}$$

$$\hat\beta_0 = \bar{y} - \hat\beta_1 \cdot \bar{x}$$

Notice that $\hat\beta_1$ equals the correlation between $x$ and $y$, scaled by the ratio of their standard deviations:
$$\hat\beta_1 = r_{xy} \cdot \frac{s_y}{s_x}$$

This shows that the regression slope is directly linked to correlation — but expressed in the original units.

### The Coefficient of Determination R²

The total variation in $y$ (TSS) can be decomposed into explained variation (ESS) and residual variation (SSR):

$$\underbrace{\sum_i(y_i - \bar{y})^2}_{\text{TSS}} = \underbrace{\sum_i(\hat{y}_i - \bar{y})^2}_{\text{ESS}} + \underbrace{\sum_i e_i^2}_{\text{SSR}}$$

The coefficient of determination measures the share of explained variation:

$$R^2 = \frac{\text{ESS}}{\text{TSS}} = 1 - \frac{\text{SSR}}{\text{TSS}} \in [0, 1]$$

:::{.callout-note}
## Interpreting R² Correctly

$R^2 = 0.15$ means: **15% of the variance in `Sales`** is explained by the model.
The remaining 85% is attributable to other factors not in the model.

A low $R^2$ is **not a failure** — it signals that important variables are missing.
Didactically, this is precisely the point: it motivates the **multiple model** with ShelveLoc and Advertising.

For simple regression, $R^2 = r_{xy}^2$ — the square of the Pearson correlation between $x$ and $y$.
:::

:::{.callout-important}
## Regression as Benchmark — Not as Truth

The regression line is **not** a causal model and **not** an exact prediction.
It is a **market standard**: the sales a store *on average* should achieve
at a given price.

A store with a strongly **negative residual** sells less than the price benchmark would predict —
that is the underperformance signal we are looking for.

> *"Regression does not give us truth. It gives us conditional averages —
> benchmarks against which to judge individual cases."*
> — @taddy_business_2019, p. 32
:::

---

## Step 1: Estimate the Baseline Model

### The Model

Before running `lm()`, let us explicitly state what we are about to estimate:

$$\widehat{\text{Sales}}_i = \hat\beta_0 + \hat\beta_1 \cdot \text{Price}_i$$

We expect $\hat\beta_1 < 0$: stores that charge higher prices should, on average, sell fewer units. The EDA in Lab 2 confirmed this with $r \approx -0.44$.

### Simple Regression: Sales ~ Price

```{r baseline-fit}
#| code-fold: false
library(moderndive)

# Fit model
mod_baseline <- tbl_carseats %>%
  lm(Sales ~ Price, data = .)

# Regression table
mod_baseline %>%
  get_regression_table() %>%
  gt() %>%
  fmt_number(columns = c(estimate, std_error, statistic,
                         p_value, lower_ci, upper_ci),
             decimals = 4) %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = term == "Price")
  ) %>%
  tab_options(table.font.size = "14px") %>%
  tab_footnote("Yellow = slope coefficient for Price | Columns 2–6 relevant only in an inferential (sampling) context — see PVA2")
```

:::{.callout-note}
## Reading the Regression Table — Today's Focus

| Column | Meaning | Today |
|--------|---------|-------|
| `estimate` | Coefficient $\hat\beta$ — the slope/intercept | ✅ Focus here |
| `std_error` | Standard error (sampling concept) | → PVA2 |
| `statistic` | t-value (sampling concept) | → PVA2 |
| `p_value` | p-value (sampling concept) | → PVA2 |
| `lower_ci` / `upper_ci` | Confidence interval (sampling concept) | → PVA2 |

Since we are working with a population (all 400 stores), the coefficients are **parameters**, not estimates. Standard errors, t-values, p-values and confidence intervals only become meaningful in PVA2 when we adopt a sampling perspective.
:::

### Model Quality

```{r baseline-summary}
mod_baseline %>%
  get_regression_summaries() %>%
  select(r_squared, adj_r_squared, rmse, sigma, nobs) %>%
  gt() %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse, sigma),
             decimals = 4) %>%
  tab_options(table.font.size = "14px")
```

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A1: Interpret the Coefficients

Write out the regression equation and interpret the coefficients:

$$\widehat{\text{Sales}} = \hat\beta_0 + \hat\beta_1 \cdot \text{Price}$$

1. **Intercept $\hat\beta_0$:** What does this value mean in context?
   Is it sensibly interpretable (price = 0)?

2. **Slope $\hat\beta_1$:** How does the predicted sales change
   when the price increases by 1 USD? By 10 USD?

3. **$R^2$:** What percentage of the sales variance does the model explain?
   What does this say about the quality of the model?

4. **RMSE:** What does the Root Mean Squared Error tell us about the
   average prediction accuracy (in which units)?
:::

### Visualisation: The Regression Line

```{r scatter-fit}
tbl_carseats %>%
  ggplot(aes(x = Price, y = Sales)) +
  geom_point(alpha = 0.35, color = "#7d0a52", size = 1.8) +
  geom_smooth(method = "lm", color = "#D50006",
              se = TRUE, linewidth = 1.2) +
  annotate("text", x = 70, y = 16,
           label = paste0("Sales = ",
                          round(coef(mod_baseline)[1], 2), " + ",
                          round(coef(mod_baseline)[2], 2), " × Price\n",
                          "R² = ", round(summary(mod_baseline)$r.squared, 3)),
           color = "#D50006", size = 3.5, hjust = 0, fontface = "bold") +
  labs(
    title    = "Baseline Model: Sales ~ Price",
    subtitle = "Regression line = conditional mean E[Sales | Price]",
    x = "Price (USD)", y = "Sales (thousands of units)"
  ) +
  theme_minimal(base_size = 11)
```

---

## Step 2: Residual Analysis

### Compute and Display Residuals

```{r residuals}
#| code-fold: false
reg_points <- mod_baseline %>%
  get_regression_points()

# First 10 observations
reg_points %>%
  head(10) %>%
  gt() %>%
  fmt_number(columns = c(Sales, Price, Sales_hat, residual),
             decimals = 3) %>%
  tab_style(
    style     = cell_fill(color = "#fff0f0"),
    locations = cells_body(rows = residual < -2)
  ) %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = residual > 2)
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("Red = strongly negative residual (underperformer) | Green = strongly positive (outperformer)")
```

:::{.callout-note}
## The Residual as a Measure of Underperformance

$$e_i = y_i - \hat{y}_i = \text{Sales}_i - \widehat{\text{Sales}}_i$$

- **$e_i < 0$:** Store sells **less** than the price benchmark predicts → underperformer
- **$e_i > 0$:** Store sells **more** than expected → outperformer
- **$e_i \approx 0$:** Store is at the market average

The residual quantifies the **deviation from the market standard** —
this is the score for our performance mining (→ Lab 4).
:::

### Residual Plot

```{r residual-plots}
pts <- augment(mod_baseline, data = tbl_carseats)

# Plot 1: Residuals vs. Fitted Values
p1 <- pts %>%
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4, color = "#7d0a52", size = 1.5) +
  geom_hline(yintercept = 0, color = "#D50006",
             linewidth = 1.1, linetype = "dashed") +
  geom_smooth(method = "loess", color = "#502479",
              se = FALSE, linewidth = 0.8) +
  labs(title = "Residuals vs. Fitted Values",
       subtitle = "Ideal: random scatter around 0",
       x = "Predicted value (Ŝales)", y = "Residual") +
  theme_minimal(base_size = 10)

# Plot 2: Residuals coloured by ShelveLoc
p2 <- pts %>%
  ggplot(aes(x = .fitted, y = .resid, color = ShelveLoc)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_hline(yintercept = 0, color = "#686868",
             linewidth = 1, linetype = "dashed") +
  scale_color_manual(values = c("Bad"    = "#D50006",
                                "Medium" = "#b3b2b2",
                                "Good"   = "#502479")) +
  labs(title = "Residuals coloured by shelf location",
       subtitle = "Systematic pattern = missing variable!",
       x = "Predicted value", y = "Residual",
       color = "ShelveLoc") +
  theme_minimal(base_size = 10)

p1 + p2
```

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A2: Interpret the Residual Plot

1. **Left plot:** Does the residual plot show a random pattern or a systematic structure?
   What would an ideal residual plot look like?

2. **Right plot:** Describe the pattern after colouring by ShelveLoc.
   - Where do the red points (Bad) tend to cluster?
   - Where do the purple points (Good) cluster?
   - What does this tell us about a missing variable in the model?

3. **Omitted Variable Bias:** If a confounder (ShelveLoc) correlates with both the regressor
   (Price) and the target variable (Sales) — in which direction is the price coefficient biased?

```r
# Check: Does ShelveLoc correlate with Price?
tbl_carseats %>%
  group_by(ShelveLoc) %>%
  summarise(mean_price = mean(Price),
            mean_sales = mean(Sales))
```
:::

---

## Step 3: OVB Diagnosis and Log Transformation

### Compute R² Manually

```{r r2-manual}
#| code-fold: false
ssq <- reg_points %>%
  summarise(
    TSS = sum((Sales - mean(Sales))^2),
    SSR = sum(residual^2),
    ESS = TSS - SSR,
    R2  = 1 - SSR / TSS
  )
ssq %>%
  gt() %>%
  fmt_number(columns = everything(), decimals = 2) %>%
  tab_options(table.font.size = "14px")
```

### Comparison: Linear vs. Log-Linear Model

The log-linear model modifies the specification as follows:

$$\ln(\widehat{\text{Sales}}_i) = \hat\beta_0 + \hat\beta_1 \cdot \text{Price}_i$$

Here $\hat\beta_1$ is no longer a unit change in Sales per unit of Price, but an approximate **percentage change** in Sales per unit of Price:

$$\hat\beta_1 \approx \frac{\Delta \text{Sales\%}}{1 \text{ USD price increase}}$$

```{r log-comparison}
# Model with log(Sales)
tbl_carseats_log <- tbl_carseats %>%
                      filter(Sales > 0) %>%   # log(0) is undefined
                      mutate(log_sales = log(Sales))

mod_log <- tbl_carseats_log %>%
  lm(log_sales ~ Price, data = .)

# Comparison
tibble(
  Model    = c("Sales ~ Price (linear)", "log(Sales) ~ Price (log-linear)"),
  R2       = c(summary(mod_baseline)$r.squared,
               summary(mod_log)$r.squared),
  `β_Price` = c(coef(mod_baseline)[2],
                coef(mod_log)[2]),
  Interpretation = c(
    "1 USD more → β̂ thousand units more Sales",
    "1 USD more → β̂ × 100% more log(Sales) ≈ % change in Sales"
  )
) %>%
  gt() %>%
  fmt_number(columns = c(R2, `β_Price`), decimals = 4) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-note}
## Interpretation of the Log-Linear Coefficient

In the log-linear model, for small $\hat\beta$:

$$\hat\beta_1 \approx \frac{\Delta \text{Sales\%}}{1 \text{ USD price}}$$

Exactly: $e^{\hat\beta_1} - 1 =$ percentage change in Sales per 1 USD price increase.

```r
# Exact percentage change
(exp(coef(mod_log)[2]) - 1) * 100
```
:::

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A3: Model Comparison and OVB

1. Compare $R^2$ of both models. Does the log transformation improve model fit?

2. **OVB formula:** The bias in the price coefficient can be written as:
$$\hat\beta_{\text{Price, biased}} = \beta_{\text{Price}} + \underbrace{\beta_{\text{ShelveLoc}} \cdot \frac{\text{Cov(Price, ShelveLoc)}}{\text{Var(Price)}}}_{\text{Bias}}$$

   Check empirically: how much does $\hat\beta_{\text{Price}}$ change when ShelveLoc
   is added to the model?

```r
# Model WITHOUT ShelveLoc
mod_without <- lm(Sales ~ Price, data = tbl_carseats)

# Model WITH ShelveLoc
mod_with    <- lm(Sales ~ Price + ShelveLoc, data = tbl_carseats)

# Compare price coefficients
coef(mod_without)["Price"]
coef(mod_with)["Price"]
```

3. In which direction is the coefficient biased? Is this consistent with your
   expectation from A2?
:::

---

## Bonus Tasks

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B1: Visualise the Variance Decomposition

Visualise TSS, ESS, and SSR for the first 20 observations:

```r
first20 <- reg_points %>% slice(1:20) %>%
  mutate(
    ybar   = mean(tbl_carseats$Sales),
    x_rank = row_number()
  )

ggplot(first20, aes(x = x_rank)) +
  # TSS: deviation from the mean
  geom_segment(aes(xend = x_rank, y = ybar, yend = Sales),
               color = "#b3b2b2", linewidth = 1) +
  # SSR: residuals
  geom_segment(aes(xend = x_rank, y = Sales_hat, yend = Sales),
               color = "#D50006", linewidth = 1.2) +
  # ESS: explained deviation
  geom_segment(aes(xend = x_rank, y = ybar, yend = Sales_hat),
               color = "#502479", linewidth = 1.2) +
  geom_point(aes(y = Sales), color = "#7d0a52", size = 2) +
  geom_point(aes(y = Sales_hat), color = "#D50006", shape = 17, size = 2) +
  geom_hline(yintercept = mean(tbl_carseats$Sales),
             linetype = "dashed", color = "#686868") +
  labs(title = "Variance decomposition: TSS = ESS + SSR",
       subtitle = "Grey = TSS | Purple = ESS (explained) | Red = SSR (residual)",
       x = "Observation", y = "Sales") +
  theme_minimal()
```
:::

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B2: Coefficient of Determination and Correlation

Show empirically that in the case of simple regression:
$$R^2 = r_{x,y}^2$$

```r
r_xy <- cor(tbl_carseats$Price, tbl_carseats$Sales)
r_xy^2
summary(mod_baseline)$r.squared
```

Why does this equality hold **only** for simple regression, not for multiple regression?
:::

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B3: Identify Underperformers

Create a first ranking of underperforming stores based on the baseline model:

```r
reg_points %>%
  arrange(residual) %>%
  head(10) %>%
  select(ID, Sales, Price, Sales_hat, residual) %>%
  gt() %>%
  fmt_number(columns = c(Sales, Sales_hat, residual), decimals = 2) %>%
  tab_style(
    style = cell_fill(color = "#fff0f0"),
    locations = cells_body(rows = residual < -3)
  )
```

What do the top underperformers have in common? Look at their ShelveLoc category.
:::

---

## Summary

```{r summary-comparison}
#| echo: false
tibble::tribble(
  ~Criterion,            ~`Baseline (Sales ~ Price)`,   ~`Expected with ShelveLoc`,
  "R²",                  paste0(round(summary(mod_baseline)$r.squared, 3)),
                         "~0.50 (substantially better)",
  "β_Price",             paste0(round(coef(mod_baseline)[2], 3), " USD/thousand"),
                         "Changes due to OVB correction",
  "Residual plot",       "Systematic pattern (ShelveLoc)",
                         "Random scatter around 0",
  "Usability",           "Too inaccurate for decisions",
                         "Better benchmark quality",
  "Main problem",        "Omitted Variable Bias via ShelveLoc",
                         "Resolved by multiple regression"
) %>%
  gt() %>%
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style     = cell_fill(color = "#fff0f0"),
    locations = cells_body(rows = Criterion == "Main problem")
  ) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-important}
## Key Messages of This Lab

1. **Regression = conditional mean** — not a causal model, but a benchmark.

2. **Descriptive perspective:** Since all 400 stores form our population, the coefficients
   are **parameters** (not estimates). Standard errors, p-values, and confidence intervals
   shown by R require a sampling context and become relevant in PVA2.

3. **R² = `r round(summary(mod_baseline)$r.squared, 2)`** — price alone explains only a small share of sales variance. This is not a flaw — it is a signal that important variables are missing.

4. **The residual plot diagnoses OVB** — the systematic ShelveLoc pattern shows that the price coefficient is biased by a missing variable.

5. **Log transformation** improves model assumptions for right-skewed outcomes and makes coefficients interpretable as (approximate) elasticities.
:::

:::{.callout-note}
## Transition to the Live Demo

In the **live demo**, the instructor shows what happens when we add `ShelveLoc` and `Advertising`
to the model:

- $R^2$ rises from ~0.15 to ~0.50
- The price coefficient changes (OVB correction)
- The residual plot loses its systematic pattern

This is the **multiple regression model** — and it leads directly to Lab 4: Performance Mining.
:::

## References {.unnumbered}
