---
title: "Lab 4: Performance Mining & Managerial Memo"
subtitle: "CRISP-DM Phase 5 & 6 — Evaluation & Deployment"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en-US
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Inhalt"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Code anzeigen"
    theme: cosmo
    css: ../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse)
library(moderndive)
library(broom)
library(gt)
library(patchwork)
library(GGally)
library(gapminder)
library(ISLR2)

tbl_carseats <- tibble(Carseats)

# Multiples Modell aus der Live-Demo — direkt verfügbar
mod_multipel <- tbl_carseats %>%
  lm(Sales ~ Price + Advertising + ShelveLoc, data = .)
```

## Learning Objectives and Overview

:::{.callout-note}
## Learning Objectives of This Lab

- Use the multiple regression model as a **scoring tool**
- Calculate and rank **residuals** as a quantitative measure of underperformance
- Communicate results in a **managerial memo** in a way that is suitable for management (translator role)
- Critically reflect on the model’s limitations: R², OVB, **data drought**, make-or-buy
:::

**Scenario:** The live demo showed that the multiple model
`Sales ~ Price + Advertising + ShelveLoc` explains approximately 54% of the variance in sales.
Now we use the **residuals** as a performance score and generate the list of underperformers —
this corresponds to CRISP-DM Phase 5 (Evaluation) and Phase 6 (Deployment).

:::{.callout-tip}
## Schedule
⏱ **15 min.** Required tasks A1–A3 | **15 min.** Group work: Managerial Memo
:::

---

## Background Material: An In-Depth Look at Multiple Regression

This chapter contains supplementary background material that you can read before or after
the lab. It explores the concepts from the live demo in greater depth using
additional datasets.

::: {.panel-tabset}

### Simpson's Paradox

**The problem:** An aggregated trend may be opposite to the trend within
subgroups — if a confounder influences both.

In the car-seat dataset: Stores with good shelf placement (Good) can charge higher prices
**and** have higher sales. This masks the negative price effect in the simple model.

```{r}
#| label: simpson-viz
#| message: false
p_agg <- tbl_carseats %>%
  ggplot(aes(x = Price, y = Sales)) +
  geom_point(alpha = 0.3, color = "#686868", size = 1.5) +
  geom_smooth(method = "lm", color = "#D50006",
              se = FALSE, linewidth = 1.3) +
  annotate("text", x = 70, y = 16,
           label = paste0("Gesamt: r = ",
                          round(cor(tbl_carseats$Price,
                                    tbl_carseats$Sales), 3)),
           color = "#D50006", size = 3.5, hjust = 0) +
  labs(title = "Aggregiert (ohne Gruppen)",
       x = "Preis (USD)", y = "Sales (Tsd.)") +
  theme_minimal(base_size = 10)

p_grp <- tbl_carseats %>%
  ggplot(aes(x = Price, y = Sales, color = ShelveLoc)) +
  geom_point(alpha = 0.35, size = 1.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.1) +
  scale_color_manual(values = c("Bad"    = "#D50006",
                                "Medium" = "#686868",
                                "Good"   = "#502479")) +
  labs(title = "Nach Regalplatz (ShelveLoc)",
       subtitle = "Innerhalb jeder Gruppe: gleicher negativer Trend",
       x = "Preis (USD)", y = "Sales (Tsd.)",
       color = "ShelveLoc") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")

(p_agg + p_grp) +
  plot_annotation(
    title    = "Simpson's Paradoxon: Aggregat vs. Subgruppen",
    subtitle = "Gleiche Daten — aber zwei sehr verschiedene Bilder"
  )
```

**Takeaway:** Always look for confounders and perform subgroup analyses
before interpreting aggregate results. The multiple model solves this problem
by controlling for ShelveLoc.

### Categorical Regression (Gapminder)

**How are dummy variables interpreted?**
The following example using the Gapminder dataset illustrates the logic
using continents—directly applicable to `ShelveLoc`.

```{r gapminder-setup}
gapminder2007 <- gapminder %>%
  filter(year == 2007) %>%
  select(country, continent, lifeExp, gdpPercap)

# Deskriptive Statistics
gapminder2007 %>%
  group_by(continent) %>%
  summarise(
    n            = n(),
    mean_lifeExp = round(mean(lifeExp), 1),
    sd_lifeExp   = round(sd(lifeExp),   1)
  ) %>%
  arrange(mean_lifeExp) %>%
  gt() %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = continent == "Africa")
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("Yellow = Reference group (alphabetical order)")
```

```{r gapminder-boxplot}
gapminder2007 %>%
  ggplot(aes(x = reorder(continent, lifeExp),
             y = lifeExp, fill = continent)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, size = 3, fill = "white") +
  scale_fill_manual(values = c(
    "Africa"   = "#D50006", "Americas" = "#7d0a52",
    "Asia"     = "#502479", "Europe"   = "#686868",
    "Oceania"  = "#b3b2b2"
  )) +
  labs(title   = "Life Expectancy 2007 by continent",
       subtitle = "Diamond = Mean | Sorted by average life expectancy",
       x = NULL, y = "Life Expectancy (Years)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

```{r gapminder-model}
mod_lifeexp <- gapminder2007 %>%
  lm(lifeExp ~ continent, data = .)

mod_lifeexp %>%
  get_regression_table() %>%
  gt() %>%
  fmt_number(columns = c(estimate, std_error,
                         statistic, p_value,
                         lower_ci, upper_ci),
             decimals = 2) %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = term == "(Intercept)")
  ) %>%
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_body(rows = term != "(Intercept)")
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(
    "Regression: lifeExp ~ continent",
    subtitle = "Intercept = Ø Afrika | Alle anderen = Differenz zu Afrika"
  )
```

**Interpretation of coefficients (directly applicable to Carseats):**

| Gapminder | Carseats |
|-----------|---------|
| Intercept = Average life expectancy in Africa (reference) | Intercept = Average sales at Bad (reference) |
| `continentAmericas` = Difference between Americas and Africa | `ShelveLocGood` = Difference between Good and Bad |
| Reference comes first alphabetically | Reference comes first alphabetically |

**Verify numerically:**

```{r gapminder-check}
# Intercept = Ø lifeExp Afrika
gapminder2007 %>%
  filter(continent == "Africa") %>%
  summarise(mean_africa = round(mean(lifeExp), 2))

# Intercept + ShelveLocGood = Ø Americas
coef(mod_lifeexp)["(Intercept)"] +
  coef(mod_lifeexp)["continentAmericas"]

gapminder2007 %>%
  filter(continent == "Americas") %>%
  summarise(mean_americas = round(mean(lifeExp), 2))
```

### Multiple Regression: 2 Numerical Regressors

**Concept:** When there are two numerical regressors, OLS does not estimate a straight line,
but rather a **plane** in three-dimensional space.

The following example uses credit card data from ISLR2.

```{r credit-eda}
tbl_credit <- tibble(Credit) %>%
  select(Balance, Income, Limit, Rating, Age)

# Scatterplot-Matrix
tbl_credit %>%
  select(Balance, Income, Limit) %>%
  ggpairs(
    upper   = list(continuous = wrap("cor", size = 4)),
    lower   = list(continuous = wrap("points",
                                     alpha = 0.3, size = 0.8)),
    diag    = list(continuous = wrap("densityDiag", alpha = 0.6)),
    progress = FALSE
  ) +
  theme_minimal(base_size = 9) +
  labs(title = "Credit card data: Balance, Income, Limit")
```

```{r credit-model}
mod_credit <- tbl_credit %>%
  lm(Balance ~ Income + Limit, data = .)

mod_credit %>%
  get_regression_table() %>%
  gt() %>%
  fmt_number(columns = c(estimate, std_error,
                         statistic, p_value,
                         lower_ci, upper_ci),
             decimals = 4) %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = term != "(Intercept)")
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header(
    "Multiple Regression: Balance ~ Income + Limit",
    subtitle = "Beide Koeffizienten: ceteris paribus"
  )
```

**Interpretation of the coefficients:**

- **Income** (`r round(coef(mod_credit)["Income"], 4)`):
  $1,000 more income → Credit card balance changes by
  `r round(coef(mod_credit)[“Income”], 2)` USD, **ceteris paribus** (limit constant).
- **Limit** (`r round(coef(mod_credit)["Limit"], 4)`):
  $1 more credit limit → Balance changes by
  `r round(coef(mod_credit)["Limit"], 4)` USD, **ceteris paribus** (Income constant).

```{r credit-r2}
mod_credit %>%
  get_regression_summaries() %>%
  select(r_squared, adj_r_squared, rmse) %>%
  gt() %>%
  fmt_number(decimals = 4) %>%
  tab_options(table.font.size = "13px")
```

**Application to Carseats:** The Carseats model combines
*numeric* predictors (Price, Advertising) with a
*categorical* predictor (ShelveLoc) — this is the general case
of multiple regression.

:::

---

## Tasks: Performance Mining

### Step 1 — Review the model

The multiple model from the live demo:

```{r model-recap}
#| code-fold: false
# Modell identisch zur Demo
mod_multipel <- tbl_carseats %>%
  lm(Sales ~ Price + Advertising + ShelveLoc, data = .)

# Modellgüte
mod_multipel %>%
  get_regression_summaries() %>%
  select(r_squared, adj_r_squared, rmse, sigma, nobs) %>%
  gt() %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse, sigma),
             decimals = 4) %>%
  tab_options(table.font.size = "14px")
```

```{r model-coefs}
# Koeffizienten für den Einstieg
mod_multipel %>%
  get_regression_table() %>%
  gt() %>%
  fmt_number(columns = c(estimate, std_error,
                         statistic, p_value,
                         lower_ci, upper_ci),
             decimals = 4) %>%
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_body(rows = str_detect(term, "ShelveLoc"))
  ) %>%
  tab_options(table.font.size = "13px")
```

### Step 2 — Compute Performance Scores

```{r scores}
#| code-fold: false
filial_scored <- mod_multipel %>%
  get_regression_points() %>%
  mutate(
    performance_gap = residual,
    performance_kat = case_when(
      residual < -2 ~ "Underperformer",
      residual >  2 ~ "Outperformer",
      TRUE          ~ "Normal"
    )
  )

# Überblick
filial_scored %>%
  count(performance_kat) %>%
  mutate(anteil = scales::percent(n / sum(n), accuracy = 0.1)) %>%
  arrange(desc(n)) %>%
  gt() %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-note}
## Residual = Measure for Underperformance

$$e_i = \text{Sales}_i - \widehat{\text{Sales}}_i$$

- $e_i \ll 0$: Store sells **significantly less** than the benchmark -- **underperformer**
- $e_i \approx 0$: Store is at the market average
- $e_i \gg 0$: Store sells **more** than expected — **outperformer**

**Important:** The benchmark is already *adjusted* for price, advertising, and shelf space
— so an underperformer is truly worse than
stores with comparable starting conditions.
:::

### Step 3 — Top-10-Ranking of Underperformers

```{r top10}
# Metadaten für Unternehmenskontext ergänzen
filial_scored %>%
  arrange(performance_gap) %>%
  head(10) %>%
  left_join(
    tbl_carseats %>%
      mutate(ID = row_number()) %>%
      select(ID, ShelveLoc, Urban, US, Income, Age),
    by = "ID"
  ) %>%
  select(ID, Sales, Sales_hat, performance_gap,
         ShelveLoc.x, Urban, US, Income, Age) %>%
  rename(ShelveLoc = ShelveLoc.x) %>%
  gt() %>%
  fmt_number(columns = c(Sales, Sales_hat, performance_gap),
             decimals = 2) %>%
  fmt_number(columns = c(Income, Age), decimals = 0) %>%
  tab_style(
    style     = cell_fill(color = "#fff0f0"),
    locations = cells_body(rows = performance_gap < -3)
  ) %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = performance_gap >= -3 &
                             performance_gap < -2)
  ) %>%
  tab_options(table.font.size = "12px") %>%
  tab_header(
    "Top-10-Underperformer",
    subtitle = "Dunkelrot: gap < −3 | Gelb: gap zwischen −3 und −2"
  )
```

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A1 — Analyze Underperformers

**1.** Which **ShelveLoc category** dominates the top 10 underperformers?
Test this empirically:

```r
filial_scored %>%
  left_join(
    tbl_carseats %>% mutate(ID = row_number()) %>%
      select(ID, ShelveLoc),
    by = "ID"
  ) %>%
  filter(performance_kat == "Underperformer") %>%
  count(ShelveLoc) %>%
  mutate(anteil = scales::percent(n / sum(n)))
```

**2.** Is there a **store with prime shelf space** among the top 10 underperformers?
If so, why is this particularly interesting for management?

**3.** What do the **top 3 underperformers** have in common?
Compare income, age, and urban/rural location.
:::

### Step 4 — Visualize the performance gap

```{r performance-plot}
#| fig-height: 5
# Plot 1: Scatter Performance Gap vs. Predicted
p1 <- filial_scored %>%
  ggplot(aes(x = Sales_hat, y = performance_gap,
             color = performance_kat)) +
  geom_point(alpha = 0.6, size = 1.8) +
  geom_hline(yintercept = 0, color = "#333",
             linewidth = 0.8, linetype = "dashed") +
  geom_hline(yintercept = c(-2, 2), color = "#b3b2b2",
             linewidth = 0.5, linetype = "dotted") +
  scale_color_manual(
    values = c("Underperformer" = "#D50006",
               "Normal"         = "#b3b2b2",
               "Outperformer"   = "#502479")
  ) +
  labs(
    title    = "Performance Gap by predicted Sales",
    subtitle = "Threshold: ±2k units",
    x        = "Predicted Ŝales",
    y        = "Performance Gap (Residual)",
    color    = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")

# Plot 2: Top/Bottom 10 as horizontal Bar Chart
p2 <- bind_rows(
  filial_scored %>% arrange(performance_gap) %>%
    head(10) %>% mutate(gruppe = "Underperformer"),
  filial_scored %>% arrange(desc(performance_gap)) %>%
    head(10) %>% mutate(gruppe = "Outperformer")
) %>%
  ggplot(aes(x = reorder(factor(ID), performance_gap),
             y = performance_gap, fill = gruppe)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = 0, color = "#333",
             linewidth = 0.5) +
  scale_fill_manual(values = c("Underperformer" = "#D50006",
                               "Outperformer"   = "#502479")) +
  coord_flip() +
  labs(
    title = "Top-10 Under- and Outperformer",
    x     = "Filial-ID",
    y     = "Performance Gap (in 1.000)",
    fill  = NULL
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "bottom")

p1 + p2
```

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A2 — Interpreting the Visualization

**1.** Is there a **pattern**: Are stores with high predicted sales
more likely to be top performers, or is the distribution even?

**2.** Calculate performance scores **separately by ShelveLoc**:

```r
filial_scored %>%
  left_join(
    tbl_carseats %>% mutate(ID = row_number()) %>%
      select(ID, ShelveLoc),
    by = "ID"
  ) %>%
  group_by(ShelveLoc) %>%
  summarise(
    mean_gap = round(mean(performance_gap), 3),
    sd_gap   = round(sd(performance_gap),   3),
    n_under  = sum(performance_kat == "Underperformer"),
    n_over   = sum(performance_kat == "Outperformer"),
    n_total  = n()
  )
```

**3.** A **stock that’s outperforming and has a prominent shelf position** might be worth considering:
Why? What would management do with this information?
:::

### Step 5 — Model Evaluation

```{r}
#| label: model-eval-plots
#| message: false
pts <- augment(mod_multipel, data = tbl_carseats)

# Residualplot
p_res <- pts %>%
  ggplot(aes(.fitted, .resid)) +
  geom_point(alpha = 0.4, color = "#7d0a52", size = 1.5) +
  geom_hline(yintercept = 0, color = "#D50006",
             linewidth = 1, linetype = "dashed") +
  geom_smooth(method = "loess", color = "#502479",
              se = FALSE, linewidth = 0.8) +
  labs(title = "Residualplot: Multiples Modell",
       subtitle = "Kein klares systematisches Muster — aber noch Reststreuung",
       x = "Vorhergesagter Wert", y = "Residuum") +
  theme_minimal(base_size = 10)

# QQ-Plot
p_qq <- pts %>%
  ggplot(aes(sample = .resid)) +
  stat_qq(color = "#7d0a52", alpha = 0.6) +
  stat_qq_line(color = "#D50006", linewidth = 0.8) +
  labs(title = "QQ-Plot der Residuen",
       subtitle = "Näherungsweise Normalverteilung → für Inferenz in PVA2 wichtig",
       x = "Theoretische Quantile", y = "Beobachtete Quantile") +
  theme_minimal(base_size = 10)

p_res + p_qq
```

```{r r2-vergleich}
# Drei-Modell-Vergleich
mod_basis <- lm(Sales ~ Price, data = tbl_carseats)
mod_adv   <- lm(Sales ~ Price + Advertising, data = tbl_carseats)

bind_rows(
  get_regression_summaries(mod_basis) %>%
    mutate(modell = "① Sales ~ Price"),
  get_regression_summaries(mod_adv) %>%
    mutate(modell = "② Sales ~ Price + Advertising"),
  get_regression_summaries(mod_multipel) %>%
    mutate(modell = "③ Sales ~ Price + Advertising + ShelveLoc")
) %>%
  select(modell, r_squared, adj_r_squared, rmse) %>%
  gt() %>%
  fmt_number(columns = c(r_squared, adj_r_squared, rmse),
             decimals = 4) %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = 2)
  ) %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = 3)
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_header("Modellvergleich: R²-Progression")
```

:::{.callout-caution icon=false}
## `r fontawesome::fa(“pen”)` A3 — Model Evaluation

**1. Residual plot:** Has the ShelveLoc pattern from Lab 2 disappeared?
What residual dispersion do you still see?

**2. R² ≈ 0.54:** For which decisions is this explanatory power sufficient?
Based on this model, would you recommend closing a branch?

**3. Data Drought:** Which variables are missing from the model?
Classify according to data strategy:

| Category | Examples |
|-----------|-----------|
| Present in the dataset, not yet in the model | ? |
| Externally obtainable (commodity data) | ? |
| Can only be collected internally (proprietary data) | ? |
:::

---

## Group Assignment: Managerial Memo

:::{.callout-caution icon=false}
## `r fontawesome::fa("users")` Group assignment (3–4 persons, 15 minutes)

### Task

Write a short **memo to the CEO** (max. 300 words).
You are the data science team—the CEO is not a statistician.

Answer all three questions clearly, concisely, and in a way that is suitable for management.

---

### Question 1 — Model Evaluation: Is R² sufficient?

> *“Our model explains R² ≈ 54% of the variance in sales.
> Would you recommend that management close stores based on this model?
> Explain your recommendation.”*

**What makes for a strong answer?**

<table style="font-size:0.84em; border-collapse:collapse; width:100%;">
<thead>
<tr>
  <th style="padding:5px 10px; background:#f5f5f5; border:1px solid #ddd; width:50%;">Weak answer</th>
  <th style="padding:5px 10px; background:#f5f5f5; border:1px solid #ddd; width:50%;">Strong answer</th>
</tr>
</thead>
<tbody>
<tr>
  <td style="padding:5px 10px; border:1px solid #ddd; vertical-align:top;">
    „R² = 0.54. More than 50% is good enough."
  </td>
  <td style="padding:5px 10px; border:1px solid #ddd; vertical-align:top;">
    „R² = 0.54 means that 46% of the variance in revenue is attributable to unknown factors.
    A decision to close a location based on this information carries a significant risk of error—
    we recommend first supplementing the data in a targeted manner before taking irreversible
    action."
  </td>
</tr>
</tbody>
</table>

---

### Question 2 — OVB: What variables are missing?

> *“What factors are missing from our data that could explain
> why a branch is underperforming?
> Give three specific examples and explain why.”*

**What makes for a strong answer?**

<table style="font-size:0.84em; border-collapse:collapse; width:100%;">
<thead>
<tr>
  <th style="padding:5px 10px; background:#f5f5f5; border:1px solid #ddd; width:50%;">Weak answer</th>
  <th style="padding:5px 10px; background:#f5f5f5; border:1px solid #ddd; width:50%;">Strong answer</th>
</tr>
</thead>
<tbody>
<tr>
  <td style="padding:5px 10px; border:1px solid #ddd; vertical-align:top;">
    „We need more data about our clients."
  </td>
  <td style="padding:5px 10px; border:1px solid #ddd; vertical-align:top;">
    „Population density within a 5-km radius: Stores in sparsely populated regions have
    structurally lower sales—regardless of price, advertising, and shelf space.
    This factor is missing and systematically skews the score for rural stores downward."
  </td>
</tr>
</tbody>
</table>

---

### Frage 3 — Data Strategy: Make or Buy?

> *„Sollen wir externe Geodaten und Demografiedaten von einem Datenhändler einkaufen
> (Commodity Data) oder eine eigene Kundenbefragung aufbauen (Proprietary Data)?
> Argumentieren Sie mit dem AI-Value-Creator-Konzept aus Thomas et al. (2025)."*

**Was macht eine starke Antwort aus?**

<table style="font-size:0.84em; border-collapse:collapse; width:100%;">
<thead>
<tr>
  <th style="padding:5px 10px; background:#f5f5f5; border:1px solid #ddd; width:50%;">Schwache Antwort</th>
  <th style="padding:5px 10px; background:#f5f5f5; border:1px solid #ddd; width:50%;">Starke Antwort</th>
</tr>
</thead>
<tbody>
<tr>
  <td style="padding:5px 10px; border:1px solid #ddd; vertical-align:top;">
    „Wir sollten Daten kaufen — das ist schneller und günstiger."
  </td>
  <td style="padding:5px 10px; border:1px solid #ddd; vertical-align:top;">
    „Geodaten (Commodity Data) sind für alle Wettbewerber beschaffbar —
    kein nachhaltiger Vorteil. Eine eigene Loyalty-Card mit Kaufhistorie
    (Proprietary Data) schafft einen Data Moat.
    Empfehlung: Kurzfristig Commodity kaufen (Pilotprojekt), langfristig
    Loyalty-Programm aufbauen für dauerhaften Wettbewerbsvorteil
    (Thomas et al., 2025: AI Value Creator)."
  </td>
</tr>
</tbody>
</table>
:::

---

## Debrief: Make-or-Buy und Expected Value Framework

### Wann lohnt sich Dateneinkauf?

$$\underbrace{E[\Delta\text{Gewinn}]}_{\text{Mehrwert besserer Entscheidungen}} > \underbrace{C_{\text{Daten}}}_{\text{Beschaffungskosten}}$$

```{r ev-table}
#| echo: false
tibble(
  Parameter = c(
    "Kosten Demografiedaten (Commodity)",
    "Kosten falsche Filialschließung",
    "P(Fehlentscheidung ohne Daten)",
    "P(Fehlentscheidung mit Daten)",
    "E[Δ Gewinn]",
    "Entscheidung"
  ),
  Wert = c(
    "50.000 EUR/Jahr",
    "500.000 EUR",
    "30%",
    "10%",
    "(0.30 − 0.10) × 500.000 = **100.000 EUR**",
    "✅ Datenkauf lohnt sich — 100.000 > 50.000"
  )
) %>%
  gt() %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = Parameter == "Entscheidung")
  ) %>%
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = Parameter == "E[Δ Gewinn]")
  ) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-note}
## Vom AI User zum AI Value Creator

Das Pilotprojekt (heute) hat gezeigt: **Commodity Data** (öffentlicher Carseats-Datensatz)
reicht für einen Proof-of-Concept — aber nicht für nachhaltige Wettbewerbsvorteile.

Der Weg zum **AI Value Creator** (Thomas et al., 2025):

1. **Kurzfristig:** Externe Geodaten einkaufen (Commodity) → Modell verbessern
2. **Mittelfristig:** Loyalty-Programm aufbauen → eigene Transaktionsdaten
3. **Langfristig:** Modell mit Proprietary Data trainieren → nicht kopierbar

:::

---

## CRISP-DM: Abschluss

```{r crisp-summary}
#| echo: false
tibble::tribble(
  ~Phase,                       ~Ergebnis,
  "① Business Understanding",   "Underperformer = Filiale mit stark negativem Residuum",
  "② Data Understanding",       "EDA: Sales leicht rechtsschief, ShelveLoc als Confounder",
  "③ Data Preparation",         "ShelveLoc als Dummy-Variable, Advertising aufgenommen",
  "④ Modeling",                 paste0("Multiples Modell: R² ≈ ",
                                       round(summary(mod_multipel)$r.squared, 2),
                                       " | drei Prädiktoren"),
  "⑤ Evaluation",               "Residualplot: OVB behoben, kein ShelveLoc-Muster mehr",
  "⑥ Deployment",               "Underperformer-Ranking + Managerial Memo + Make-or-Buy"
) %>%
  gt() %>%
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = Phase == "⑥ Deployment")
  ) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-important}
## Kernbotschaften dieses Labs

1. **Residuum = Performance-Score:** Das multiple Modell liefert einen preis-,
   werbe- und regalplatz-bereinigten Benchmark für jede Filiale.

2. **R² ≈ 0.54** ist ein Ausgangspunkt — 46% unerklärte Varianz motiviert
   gezielte Datenergänzung, nicht direkt Filialschließungen.

3. **Translator-Rolle:** Nicht das Modell, sondern die Handlungsempfehlung ist der Wert —
   statistische Befunde in Entscheidungslogik übersetzen.

4. **Make or Buy:** Commodity-Daten für sofortige Verbesserung,
   Proprietary-Daten für dauerhaften Wettbewerbsvorteil.
:::

:::{.callout-note}
## Ausblick: Was kommt in PVA2?

Heute haben wir **deskriptiv** gearbeitet: alle 400 Filialen = Grundgesamtheit (Population).

**Offene Frage:** Was wenn die 400 Filialen nur eine **Stichprobe** wären?
Wie sicher sind unsere Koeffizienten? Hätten wir bei einer anderen Saison
dieselben Ergebnisse? Und wie wahrscheinlich ist es, dass der gemessene Werbeeffekt
nur Zufall ist?

Das sind die Fragen der **statistischen Inferenz** — das Thema von PVA2:
Bootstrap, Konfidenzintervalle, p-Hacking, FDR und das Expected-Value-Framework
für Manager.
:::

## Quellenverzeichnis {.unnumbered}
