---
title: "Lab 2: Data Understanding & Exploratory Data Analysis"
subtitle: "CRISP-DM Phase 2 --- Carseats Case"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en-US
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse)
library(moderndive)
library(skimr)
library(gt)
library(GGally)
library(patchwork)
library(moments)

library(ISLR2)
tbl_carseats <- tibble(Carseats)
```

## Learning Objectives and Overview

:::{.callout-note}
## Learning Objectives

- Establish EDA as a mandatory first *step to understand the data* before modelling (e.g. regression analysis)
- Explore the Carseats dataset: structure, variable types, distributions
- Understand what EDA is (**descriptive** and **inductive**)...
- ...and, equally importantly, what it is *not* (confirmatory, predictive, inferential)
- Therefore know why EDA-derived hypotheses cannot be tested on the same data
- Visualise bivariate relationships and formulate hypotheses for modelling
:::

**Scenario:** You are a data scientist at a retail chain with 400 stores.
Before building a model, you need to understand the data ---
this is CRISP-DM Phase 2: **Data Understanding**.

:::{.callout-tip}
## Schedule
⏱ **20 minutes** --- Mandatory tasks (A1--A4) first, bonus tasks if time permits.
:::

---

## Background: EDA and Its Boundaries

### Focus of This Lab

The primary goal of this lab is **Exploratory Data Analysis (EDA)** of the Carseats
dataset. EDA and the closely related field of **Data Mining (DM)** are frequently
conflated in practice --- and understandably so: the tools overlap, and the boundary is
genuinely blurry in places. Rather than ignoring this, we address it head-on with a compact comparison before diving into the data.

A concrete example of the grey zone: when we fit a regression line to a scatter plot
in EDA, we are already doing something model-like. Is that still EDA or already Data Mining?
The answer depends less on the technique than on the *purpose and epistemic status* of
what we are doing. In EDA, the line is a visual summary to help us see a pattern.
In DM, the model is a deployable artefact evaluated against performance metrics.
Same tool, different frame.

### EDA Is Descriptive --- and Treats the Data as a Population

**Exploratory Data Analysis is purely descriptive.**

When we compute means, standard deviations, or correlations for the 400 Carseats stores,
we are not estimating population parameters from a sample --- we are *describing the data
we have in full*. The 400 stores **are** the population of interest. As a result:

- Summary statistics are **parameters** of this dataset, not estimates subject to
  sampling error.
- There is no sampling variability --- no standard errors, no p-values, no
  confidence intervals are conceptually meaningful at this stage.
- The goal is **understanding and pattern discovery**, not inference or prediction.

We will return to this in Lab 3 when we examine the regression output table critically.

### EDA and Data Mining: What They Share, Where They Differ

Both EDA and Data Mining are **data-driven and inductive**: they do not wait for
a theory to confirm, but search for structure in the data itself.
This distinguishes them from classical confirmatory statistics, which starts from
a hypothesis and asks whether the data are consistent with it.

**What they share:**

- Both use statistical summaries and algorithms to separate signal from noise.
- Both are iterative: findings in one step shape the next question.
- Both are empirical: the data, not prior theory, drive the process.

**Where they diverge** is primarily in *purpose*, *dimensionality*, and *what counts
as success*. The following two tables lay this out systematically.

#### Dimension-by-Dimension Comparison

| Dimension | EDA | Data Mining (DM) |
|-----------|-----|-----------------|
| **Goal** | Data understanding, intuition, uncovering errors and structures | Extraction of stable patterns for commercial use or decision support |
| **Dimensionality** | Low-dimensional: focus on 2D/3D visualisations and descriptive statistics | High-dimensional (Big Data): processes hundreds of variables simultaneously, where visualisation breaks down |
| **Type of insight** | Phenomenological: "What do the data show?" (descriptive) | Pragmatic/functional: "Which patterns are stable enough for profit or prediction?" (correlation > causation) |
| **Methods** | Inspection, plotting, transformation, outlier identification | Algorithmic modelling (associations, clusters, classification) |
| **Human involvement** | Interactive and manual: the analyst is central to the feedback loop | Automated and scalable: algorithms search mass data for significant structures |
| **Validation** | Plausibility checks and visual confirmation | Performance metrics (lift, accuracy, profit); focus on pattern stability |
| **Causality** | Seeks hints at logical relationships --- generates hypotheses | Uses sufficiently stable correlations ("black box" approach often acceptable if profitable) |

#### Focus, Scale, and Success Criteria

| Aspect | EDA (focus of this lab) | Data Mining (DM) |
|--------|------------------------|-----------------|
| **Central actor** | The human: visualisation serves cognitive understanding | The algorithm: detects patterns in dimensions we can no longer see |
| **Dimensionality** | Low: usually 2D to 3D (scatter plots, matrices) | High: linking hundreds of variables ("Big Data") |
| **Goal** | Understanding and hypotheses: "Why do the data behave this way?" | Exploitation: "How do I use this pattern commercially?" |
| **Measure of success** | Insight: are new questions being generated? | Performance: ROI, accuracy, lift, stability of predictions |

:::{.callout-note}
## Where the Boundary Gets Blurry

The grey zone is real and worth naming explicitly:

- **Regression line in a scatter plot:** In EDA, this is a visual aid --- a conditional
  mean drawn to reveal a trend. In DM, the same line is a predictive model evaluated
  by RMSE on held-out data. Same technique, different purpose.
- **Correlation matrix:** In EDA, it flags potentially interesting variable pairs for
  further inspection. In DM, it feeds feature selection algorithms.
- **Outlier detection:** In EDA, an outlier is something the analyst *examines* to
  understand the data better. In DM, it may be automatically flagged and removed or
  modelled separately.

The practical distinction: **EDA ends when the analyst has formed a question or hypothesis
that is worth testing. DM begins when that question is operationalised into a model
evaluated on new data.**
:::

### The Induction Trap: Why EDA-Derived Hypotheses Need New Data

Both EDA and DM are inductive --- they derive conclusions from observed patterns rather
than testing pre-specified hypotheses. This creates a critical constraint.

:::{.callout-warning}
## The Induction Trap

**The critical rule:** A hypothesis generated inductively from a dataset cannot be
validly tested on that same dataset.

Any pattern you discover in a dataset will --- by construction --- appear confirmed when
you apply a statistical test to those same observations. The p-value will be misleadingly
small, the result will look significant, but you have learned nothing new: you are
re-describing what you already saw. This is **circular inference** (or, less charitably,
**data dredging**).

> *Example:* You explore the 400 Carseats stores, observe that `ShelveLoc` appears
> associated with `Sales`, and then run a t-test on the same 400 stores.
> The test will almost certainly be significant --- but only because the same data
> generated both the hypothesis and the "confirmation".

**What to do instead:**

- **Split the data upfront** --- before any exploration --- into an exploration set and a
  confirmation set. This is the logic behind train/test splits and **A/B testing**
  (→ Session 3).
- Or: formulate hypotheses *a priori* from theory and test them on independent data.

In **PVA2** we examine what p-values and confidence intervals actually measure and when
they are valid. For now: **EDA generates questions. It does not answer them.**
:::

### Why Visualisation Is Non-Negotiable: Anscombe's Quartet

The classic demonstration that summary statistics alone are deeply misleading comes from
**Anscombe (1973)**. He hand-constructed four small datasets with *virtually identical*
summary statistics --- yet with fundamentally different structures.

Read the table first. Then examine the plots carefully. Compare panel by panel.

```{r anscombe-stats}
#| code-fold: false
anscombe_long <- anscombe |>
  pivot_longer(
    everything(),
    names_to  = c(".value", "dataset"),
    names_pattern = "(.)(.)"
  ) |>
  mutate(dataset = paste("Dataset", dataset))

anscombe_long |>
  group_by(dataset) |>
  summarise(
    n      = n(),
    mean_x = round(mean(x), 2),
    mean_y = round(mean(y), 2),
    sd_x   = round(sd(x),   2),
    sd_y   = round(sd(y),   2),
    r_xy   = round(cor(x, y), 3)
  ) |>
  gt() |>
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_column_labels()
  ) |>
  tab_options(table.font.size = "13px") |>
  tab_footnote(
    "All four datasets have (almost) identical summary statistics -- the plots below tell four different stories."
  )
```

```{r anscombe-plots}
#| fig-height: 5
#| fig-cap: "Anscombe's Quartet: identical statistics, four fundamentally different structures."
anscombe_long |>
  ggplot(aes(x = x, y = y)) +
  geom_point(color = "#502479", size = 2.5, alpha = 0.85) +
  geom_smooth(method = "lm", color = "#D50006",
              se = TRUE, linewidth = 1) +
  facet_wrap(~dataset, nrow = 2) +
  labs(
    title    = "Anscombe's Quartet",
    subtitle = "Identical n, mean, sd, and correlation -- four entirely different stories",
    x = "x", y = "y"
  ) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))
```

:::{.callout-important}
## What Each Panel Teaches --- and Why It Matters for Regression Diagnostics

| Panel | Structure | What the OLS line misses | Detected by |
|-------|-----------|--------------------------|-------------|
| **I** | Linear, roughly homoscedastic | Nothing --- OLS is appropriate here | Residual plot: random scatter ✓ |
| **II** | Curved (quadratic) relationship | Non-linearity --- a linear model is structurally wrong | Residual plot: curved pattern |
| **III** | One outlier in *y* | Outlier distorts intercept and slope | Residual plot: one extreme point |
| **IV** | One leverage point (*x* = 19) | Single extreme *x*-value pulls entire line | Residual plot: gap in x-values |

All four problems are diagnosed by the **residual plot** --- which you will use in
Lab 3 to detect Omitted Variable Bias.

Note that the regression line in all four panels is an EDA tool here:
it makes the trend visible, but we do not yet evaluate it as a predictive model.
That shift in perspective is what distinguishes EDA from Data Mining.

*The Datasaurus Dozen* (Matejka & Fitzmaurice, 2017) makes the same point with 13
algorithmically generated datasets (including a dinosaur). Anscombe is more instructive
for a regression course because each panel encodes a *different* diagnostic problem.
:::

### Elements of Exploratory Data Analysis

According to @stoetzer_regressionsanalyse_2020, EDA follows four guiding principles:

1. **Substantive meaning** --- The primary goal is understanding variables and their
   plausible relationships, not running formal tests.
2. **Graphical representation** --- Visual inspection is the primary means of gaining insight.
3. **Iterative model building** --- Tentative hypothesis generation in an iterative process
   of model specification and re-specification.
4. **Scepticism and parsimony** --- Scepticism, flexibility, and restraint regarding
   which statistical methods to apply.

:::{.callout-note}
## EDA as CRISP-DM Phase 2

In our Carseats project, EDA (Phase 2: Data Understanding) generates **hypotheses**
for the modelling phase (Phase 4): Which variables explain sales differences?
Which transformations are necessary? Are there outliers or systematic patterns?
:::

---

## Step 1: Getting to Know the Dataset

### Overview: Structure and Variables

```{r glimpse}
#| code-fold: false
glimpse(tbl_carseats)
```

:::{.callout-tip}
## Interpretation
- How many observations (stores) and variables does the dataset contain?
- Which variables are numeric (`dbl`), which are categorical (`chr`/`fct`)?
- What exactly does `Sales` measure? (Pay attention to the unit!)
:::

### Summary Statistics

```{r skim}
tbl_carseats |>
  select(Sales, Price, Advertising, Income, Age, ShelveLoc, Urban, US) |>
  skim() |>
  yank("numeric") |>
  gt() |>
  fmt_number(columns = where(is.numeric), decimals = 2) |>
  tab_options(table.font.size = "13px")
```

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A1: Explore the Dataset

1. How many stores (`n`) does the dataset contain?
2. What is the **range** of `Sales` (Min, Max)? What does this mean in practice?
3. What is the **range** of `Price`? Is the car-seat market price-elastic?
4. What does the variable `ShelveLoc` represent? What categories does it have?

```r
tbl_carseats |> count(ShelveLoc)
tbl_carseats |> count(Urban)
tbl_carseats |> count(US)
```
:::

---

## Step 2: Distribution of the Target Variable Sales

### Check Before You Transform

A common reflex is to log-transform sales data "because sales are always skewed."
**This is wrong as a default.** Whether a transformation is appropriate must be
determined from the data --- not assumed in advance. That is exactly what EDA is for.

```{r sales-distribution}
#| code-fold: false
skew_raw <- round(skewness(tbl_carseats$Sales), 3)
skew_log <- round(skewness(log(tbl_carseats$Sales)), 3)

p1 <- tbl_carseats |>
  ggplot(aes(x = Sales)) +
  geom_histogram(bins = 30, fill = "#D50006", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(tbl_carseats$Sales),
             color = "#502479", linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = median(tbl_carseats$Sales),
             color = "#502479", linewidth = 1.2, linetype = "dotted") +
  annotate("text", x = mean(tbl_carseats$Sales) + 0.4, y = 38,
           label = "Mean", color = "#502479", size = 3.2, hjust = 0) +
  annotate("text", x = median(tbl_carseats$Sales) - 0.4, y = 33,
           label = "Median", color = "#502479", size = 3.2, hjust = 1) +
  labs(
    title    = "Sales -- raw data",
    subtitle = paste0("Skewness = ", skew_raw,
                      "  |  Mean = ", round(mean(tbl_carseats$Sales), 2),
                      "  |  Median = ", round(median(tbl_carseats$Sales), 2)),
    x = "Sales (thousands of units)", y = "Count"
  ) +
  theme_minimal(base_size = 11)

p2 <- tbl_carseats |>
  ggplot(aes(x = log(Sales))) +
  geom_histogram(bins = 30, fill = "#686868", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(log(tbl_carseats$Sales)),
             color = "#D50006", linewidth = 1.2, linetype = "dashed") +
  labs(
    title    = "log(Sales) -- after transformation",
    subtitle = paste0("Skewness = ", skew_log,
                      "  |  Mean = ", round(mean(log(tbl_carseats$Sales)), 3)),
    x = "log(Sales)", y = "Count"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(color = "#686868"))

p1 + p2
```

:::{.callout-important}
## Result: The Log Transformation Makes Things Worse Here

The raw `Sales` data are already **approximately symmetric**:

- Skewness of `Sales`: `r skew_raw` --- close to zero.
- Mean (`r round(mean(tbl_carseats$Sales), 2)`) and Median
  (`r round(median(tbl_carseats$Sales), 2)`) are nearly identical.

After the log transformation, skewness becomes `r skew_log` --- **more negative**,
meaning the transformation *introduces* left-skewness rather than removing right-skewness.
This is the opposite of what log transformation is designed to achieve.

**Lesson: EDA before transformation --- always.**
We will work with **untransformed Sales** throughout this course.
:::

:::{.callout-note}
## When Is a Log Transformation Appropriate?

The log transformation is motivated by a **multiplicative generating process**:

$$\text{Sales}_i = \alpha \cdot \text{Price}_i^{\beta_1} \cdot \text{Adv}_i^{\beta_2} \cdot \varepsilon_i$$

Taking logs converts this into an additive, OLS-compatible model whose coefficients
become **elasticities**. But the precondition --- a right-skewed raw distribution ---
must first be verified empirically.

Contrast: `Advertising` *is* right-skewed (many stores spend nothing; a few spend a lot).
A log transformation of `Advertising` would be more defensible. Check this in task A2.
:::

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A2: Assess the Distribution

1. Describe the distribution of `Sales` in your own words.
   Is it symmetric, right-skewed, or left-skewed? How close are mean and median?

2. Compute skewness for `Sales`, `log(Sales)`, and `Advertising` for comparison:

```r
library(moments)
skewness(tbl_carseats$Sales)
skewness(log(tbl_carseats$Sales))
skewness(tbl_carseats$Advertising)   # contrast: this one IS right-skewed
```

3. For which variable is a log transformation more justified: `Sales` or `Advertising`?
   Justify your answer based on skewness values and the histograms.
:::

---

## Step 3: Bivariate Relationships

### Price vs. Sales

```{r scatter-price-sales}
#| code-fold: false
tbl_carseats |>
  ggplot(aes(x = Price, y = Sales)) +
  geom_point(alpha = 0.4, color = "#7d0a52", size = 1.8) +
  geom_smooth(method = "lm", color = "#D50006",
              se = TRUE, linewidth = 1.2) +
  geom_smooth(method = "loess", color = "#502479",
              se = FALSE, linewidth = 1, linetype = "dashed") +
  labs(
    title    = "Scatterplot: Price vs. Sales",
    subtitle = paste0("Correlation r = ",
                      round(cor(tbl_carseats$Price, tbl_carseats$Sales), 3),
                      "  |  Red = OLS line (visual summary)  |  Purple dashed = LOESS"),
    x = "Price (USD)", y = "Sales (thousands of units)"
  ) +
  theme_minimal(base_size = 11)
```

:::{.callout-tip}
## Interpretation
- What does the correlation say? Is it strong or weak?
- The OLS line here is an **EDA tool** (visual summary of the trend), not a
  deployed predictive model. Does it agree with the LOESS curve? What does that indicate?
- Why is the scatter around the line so high? What might be missing?
:::

### Scatterplot Matrix: All Variables at Once

```{r ggpairs}
#| fig-height: 8
tbl_carseats |>
  select(Sales, Price, Advertising, Income, Age, ShelveLoc) |>
  ggpairs(
    aes(color = ShelveLoc, alpha = 0.5),
    upper = list(continuous = wrap("cor", size = 3)),
    lower = list(continuous = wrap("points", alpha = 0.3, size = 0.8)),
    diag  = list(continuous = wrap("densityDiag", alpha = 0.6))
  ) +
  scale_color_manual(values = c(
    "Bad"    = "#D50006",
    "Medium" = "#686868",
    "Good"   = "#502479"
  )) +
  scale_fill_manual(values = c(
    "Bad"    = "#D50006",
    "Medium" = "#686868",
    "Good"   = "#502479"
  )) +
  theme_minimal(base_size = 9) +
  labs(title    = "Scatterplot matrix: Carseats variables",
       subtitle = "Coloured by shelf location (ShelveLoc)")
```

:::{.callout-note}
## GGally::ggpairs() --- EDA in One Line of Code

`ggpairs()` from the `GGally` package produces in a single call:

- **Diagonal:** Distributions of individual variables
- **Upper triangle:** Correlation coefficients
- **Lower triangle:** Scatterplots of all variable pairs

The correlation numbers in the upper triangle are summary statistics --- and as
Anscombe showed, summary statistics alone are insufficient. Always cross-check with
the scatter plots in the lower triangle, which can reveal non-linearities, outliers,
or subgroup structure that a single number cannot capture.
:::

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A3: Bivariate Analysis

1. Which variables correlate **most strongly** with `Sales`? Create a ranked list:

```r
tbl_carseats |>
  select(where(is.numeric)) |>
  cor() |>
  as_tibble(rownames = "variable") |>
  select(variable, Sales) |>
  arrange(desc(abs(Sales)))
```

2. Are there variables strongly correlated **with each other** (multicollinearity risk)?
   What could this mean for a multiple regression?

3. Describe the relationship between `ShelveLoc` and `Sales` in words.
   Which category has the highest average sales?

```r
tbl_carseats |>
  group_by(ShelveLoc) |>
  summarise(mean_sales = mean(Sales),
            sd_sales   = sd(Sales),
            n          = n())
```
:::

---

## Step 4: Formulating Hypotheses

```{r hypotheses-viz}
#| fig-height: 4
p_shelf <- tbl_carseats |>
  ggplot(aes(x = ShelveLoc, y = Sales, fill = ShelveLoc)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
  scale_fill_manual(values = c("Bad"    = "#D50006",
                               "Medium" = "#686868",
                               "Good"   = "#502479")) +
  labs(title = "Sales by shelf location",
       x = "Shelf location", y = "Sales (thousands)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

p_adv <- tbl_carseats |>
  ggplot(aes(x = Advertising, y = Sales)) +
  geom_point(alpha = 0.3, color = "#7d0a52", size = 1.5) +
  geom_smooth(method = "lm", color = "#D50006", se = FALSE) +
  labs(title = "Advertising vs. Sales",
       subtitle = paste0("r = ",
                         round(cor(tbl_carseats$Advertising,
                                   tbl_carseats$Sales), 3)),
       x = "Advertising budget (thousands USD)", y = "Sales (thousands)") +
  theme_minimal(base_size = 11)

p_shelf + p_adv
```

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A4: Hypotheses for Modelling

Based on your EDA, formulate **two hypotheses** for the goal
"Identification of underperforming stores".

**Example structure:**
> *"I expect that [variable X] has a [positive/negative] relationship with Sales,
> because [substantive justification]."*

Additionally:

1. For which variable might a **non-linear effect** be plausible?

2. Which variables should **definitely** be included in the regression model?
   Which probably not --- and why?

3. Which variables are **missing** that would be important for explaining sales differences?
   (Keyword: *Data Drought*, @thomasAIValueCreators2025)

4. **Critical reflection (induction trap):** You have now used all 400 stores to
   generate these hypotheses inductively. What would need to happen *before exploration*
   if you wanted to run valid statistical tests afterwards?
:::

---

## Bonus Tasks

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B1: Estimate Price Elasticity

Fit a simple log-log regression for exploratory purposes. This is an EDA exercise ---
the regression line is a visual-analytic tool, not a deployed model. Since the hypothesis
was generated from the same data, the result is descriptive only, not a valid
confirmatory test.

```r
mod_loglog <- tbl_carseats |>
  mutate(log_sales = log(Sales),
         log_price = log(Price)) |>
  lm(log_sales ~ log_price, data = _)

get_regression_table(mod_loglog)
```

- What does the coefficient for `log_price` mean as a price elasticity?
- Compare $R^2$ with the linear model `Sales ~ Price`. What changes?
- Given that `Sales` is not right-skewed: does the log-log specification add value here?
:::

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B2: Outlier Analysis

```r
tbl_carseats |>
  mutate(rank = dense_rank(desc(Sales))) |>
  filter(rank <= 5 | rank >= (n() - 4)) |>
  select(Sales, Price, Advertising, ShelveLoc, Income, rank) |>
  arrange(rank)
```

Are there stores with extreme price-sales combinations (high price, high sales --- or vice versa)?
:::

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B3: Segmented EDA

```r
tbl_carseats |>
  ggplot(aes(x = Price, y = Sales, color = Urban)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("Yes" = "#502479", "No" = "#D50006")) +
  facet_wrap(~US) +
  labs(title = "Price vs. Sales by location (Urban) and region (US)",
       x = "Price (USD)", y = "Sales (thousands)") +
  theme_minimal()
```

Any indications of interaction effects between Urban status and US region?
:::

---

## Summary

```{r summary-table}
#| echo: false
tibble::tribble(
  ~`EDA Step`,           ~`Finding`,                                         ~`Consequence for Modelling`,
  "Structure",           "400 stores, 11 variables, no NAs",                 "Clean dataset, ready to use",
  "Sales distribution",  "Approximately symmetric (skewness ~ 0.18)",        "No log transformation needed",
  "Price effect",        "r ~ -0.44, negative linear trend",                 "Include Price as regressor",
  "ShelveLoc",           "Good > Medium > Bad (marked difference)",          "Include as categorical variable",
  "Advertising",         "r ~ +0.58, strongest positive predictor",          "Include Advertising in the model",
  "Missing variables",   "No competition, no demographics, no location",     "Data Drought -> pilot project"
) |>
  gt() |>
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = `EDA Step` == "Sales distribution")
  ) |>
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = `EDA Step` == "Missing variables")
  ) |>
  tab_options(table.font.size = "13px")
```

:::{.callout-important}
## Key Messages of This Lab

1. **EDA first --- always.** Summary statistics without visualisation are insufficient.
   Anscombe's quartet: four datasets, identical numbers, four completely different
   structures and regression diagnostics.

2. **EDA and DM are both inductive.** Neither starts from a pre-specified hypothesis.
   They differ in purpose (understanding vs. exploitation), dimensionality (low vs. high),
   and success criteria (insight vs. performance metrics).

3. **The EDA/DM boundary is real but blurry.** A regression line in a scatter plot is
   still EDA if the purpose is visual understanding; it becomes DM when it is a deployed
   model evaluated on held-out data. Purpose, not technique, draws the line.

4. **The induction trap.** Hypotheses discovered by looking at data cannot be tested on
   those same data. If confirmatory testing is intended, split the data *before*
   exploration (-> A/B testing, Session 3).

5. **Check before transforming.** `Sales` is already approximately symmetric.
   The log transformation introduces left-skewness here --- the opposite of its intended
   effect. Always verify empirically.

6. **ShelveLoc and Advertising** are the strongest predictors --- both absent from
   the baseline model (-> Lab 3).

7. **Data Drought.** Important explanatory variables (location, competition, demographics)
   are missing --- motivating the Make-or-Buy question in Lab 4.
:::

:::{.callout-note}
## Transition to Lab 3

In Lab 3 we fit the **baseline model** `Sales ~ Price` and examine it critically.
A low $R^2$ and a systematic pattern in the residual plot point to
**Omitted Variable Bias** caused by ShelveLoc.

We also revisit the regression output table: since we are working with a population,
what do "estimate", "std_error", and "p_value" actually mean?
:::

## References {.unnumbered}
