---
title: "Lab 2: Data Understanding & Exploratory Data Analysis"
subtitle: "CRISP-DM Phase 2 --- Carseats Case"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en-US
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse)
library(skimr)
library(gt)
library(GGally)
library(patchwork)
library(moments)
library(datasauRus)
library(janitor)
library(DataExplorer)
library(psych)

library(ISLR2)
tbl_carseats <- tibble(Carseats)
```

## Learning Objectives and Overview

:::{.callout-note}
## Learning Objectives

- Establish EDA as a mandatory first *step to understand the data* before modelling (e.g. regression analysis)
- Explore the Carseats dataset: structure, variable types, distributions
- Understand what EDA is (**descriptive** and **inductive**)...
- ...and, equally importantly, what it is *not* (confirmatory, predictive, inferential)
- Recognise why EDA-derived hypotheses cannot be tested on the same data --- and know what to do instead
- Visualise bivariate relationships and formulate hypotheses for modelling
:::

**Scenario:** You are the data scientist of a retail chain with 400 stores.
The business goal (CRISP-DM Phase 1) is already defined: identify underperforming stores
so that management can take targeted action.
Before any model is built, the data must be understood.
That is CRISP-DM Phase 2: **Data Understanding** --- the focus of this lab.

## 🧭 How to Navigate This Lab

| Symbol | Box colour | Meaning | When to read |
|--------|-----------|---------|-------------|
| ⚠️ | <span style="color:#e02424">**Red — Action Required**</span> | Run code, interpret output, answer questions | **Do this in the live session** |
| 💡 | <span style="color:#2f9e44">**Green — Best Practice / Tip**</span> | Coding tips, R functions, shortcuts | Skim in the lab |
| 📚 | <span style="color:#1c7ed6">**Blue, collapsed — Background Theory**</span> | Concepts, theory, context | Skip in the lab — read at home |
| ℹ️ | <span style="color:#1c7ed6">**Blue, open — Note**</span> | Short clarifications and reminders | Read when you reach them |

**Normal text and code chunks are the main content** --- work through them in order.
**Background sections** are marked with a *collapsed blue box* and **can be expanded** at any time.

⏱ **20 minutes** for mandatory tasks (A1--A5). Bonus tasks if time permits.

---

## Icebreaker: Never Trust a Statistic You Haven't Plotted

Before we turn to the Carseats data, a quick thought experiment.
The table below shows descriptive statistics for 13 different datasets:
means, standard deviations, and the correlation between x and y.

```{r datasaurus-stats}
#| code-fold: true
#| code-summary: "Show code: summary statistics"
library(datasauRus)

datasaurus_dozen |>
  group_by(dataset) |>
  summarise(
    n      = n(),
    mean_x = mean(x),
    mean_y = mean(y),
    sd_x   = sd(x),
    sd_y   = sd(y),
    cor_xy = cor(x, y)
  ) |>
  gt() |>
  fmt_number(columns = where(is.numeric), decimals = 2) |>
  tab_header(
    title    = "Summary Statistics: Datasaurus Dozen",
    subtitle = "Means, standard deviations, and correlations look virtually identical across all 13 datasets."
  ) |>
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = dataset == "dino")
  ) |>
  tab_options(table.font.size = "13px")
```

Statistically, these 13 datasets are almost indistinguishable.
An algorithm working from summary statistics alone would treat them identically.
Now look at the actual data:

```{r datasaurus-viz}
#| code-fold: true
#| code-summary: "Show code: scatter plots"
#| fig-height: 8
#| fig-width: 10
#| out-width: "100%"
datasaurus_dozen |>
  ggplot(aes(x = x, y = y, color = dataset)) +
  geom_point(alpha = 0.8, size = 1.0) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "none",
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  ) +
  facet_wrap(~dataset, ncol = 4) +
  labs(
    title    = "The Datasaurus Dozen",
    subtitle = "Same summary statistics --- completely different structures.",
    x = "x", y = "y"
  )
```

:::{.callout-important}
## The Central Lesson

The numbers in the table are almost identical. The plots reveal a dinosaur, a star,
a circle, an X, and several other completely different shapes --- all with the same mean,
standard deviation, and correlation.

**Conclusion: summary statistics are a necessary but not sufficient description of data.
Visualisation is not optional --- it is the primary EDA tool.**

This motivates everything we do in this lab.
:::

---

## Background: Where We Are in CRISP-DM

In the lecture, **Phase 1 (Business Understanding)** was introduced:
the business goal is to identify underperforming stores. The question is now:
do our data support this goal? What do they contain? What is their quality?
These are Phase 2 questions.

```{r crisp-overview}
#| echo: false
tibble::tribble(
  ~Phase, ~Name,                  ~Content,                                           ~Status,
  "1",    "Business Understanding","Goal: identify underperforming stores",            "Done (lecture)",
  "2",    "Data Understanding",   "EDA: structure, quality, distributions, patterns", "← This lab",
  "3",    "Data Preparation",     "Cleaning, encoding, feature engineering",          "Lab 3",
  "4",    "Modelling",            "Regression model: Sales ~ Price + ...",            "Lab 3 + Demo",
  "5",    "Evaluation",           "R², residuals, OVB check",                        "Demo + Lab 4",
  "6",    "Deployment",           "Performance ranking, managerial memo",             "Lab 4"
) |>
  gt() |>
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = Phase == "2")
  ) |>
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_column_labels()
  ) |>
  tab_options(table.font.size = "13px")
```

:::{.callout-note collapse="true"}
## 📚 Background: EDA and Data Mining --- Orientation

EDA and Data Mining (DM) are frequently used interchangeably in practice ---
and there are genuine overlaps. Both are **data-driven and inductive**:
they do not wait for a pre-specified hypothesis but search for structure in the data.
This distinguishes both from classical confirmatory statistics, which starts from
a hypothesis and asks whether the data are consistent with it.

The differences lie in *purpose*, *scale*, and *success criteria*:

**What they share:**
Both use statistical summaries and algorithms to separate signal from noise.
Both are iterative. Both treat the data as the source of new questions.

**Where they diverge:**

| Aspect | EDA | Data Mining (DM) |
|--------|-----|-----------------|
| **Central actor** | The human analyst: visualisation serves cognitive understanding | The algorithm: detects patterns at scales humans cannot see |
| **Dimensionality** | Low: 2D/3D visualisations and descriptive statistics | High: hundreds of variables simultaneously ("Big Data") |
| **Goal** | Understanding, intuition, error detection, hypothesis generation | Extraction of stable, commercially exploitable patterns |
| **Type of insight** | Phenomenological --- "What do these data show?" | Pragmatic --- "Which patterns are stable enough for profit or prediction?" |
| **Methods** | Plotting, transformation, outlier inspection, correlation | Association rules, clustering, classification, regression models |
| **Human involvement** | Interactive and manual: analyst is in the feedback loop | Automated and scalable |
| **Validation** | Plausibility and visual confirmation | Performance metrics: lift, accuracy, ROI |
| **Causality** | Seeks logical-conceptual connections; generates hypotheses | Stable correlations are sufficient ("black box" acceptable if profitable) |
| **Success** | New question generated; anomaly noticed; pattern understood | Prediction accurate; model deployable; ROI positive |

:::{.callout-note}
## The Grey Zone

The boundary is genuinely blurry. When we fit a regression line to a scatter plot,
are we doing EDA or Data Mining?

- **In EDA:** the line is a *visual summary* --- a conditional mean drawn to make a trend
  perceptible. We look at it, think about it, and decide what to ask next.
- **In DM:** the same line is a *deployed model*, evaluated on held-out data by RMSE
  and compared against alternatives.

Same technique, different epistemic frame. In this lab, all models are EDA tools ---
they serve understanding, not deployment. The shift to DM begins in →Lab 3.
:::
:::

### The Induction Trap

Both EDA and DM are inductive. This creates a constraint that is easy to overlook.

:::{.callout-warning}
## Hypotheses from EDA Cannot Be Tested on the Same Data

**The rule:** A hypothesis generated by looking at a dataset cannot be validly tested
on that same dataset.

Any pattern you discover inductively will appear confirmed by a statistical test on
the same observations --- by construction. The p-value will be small, the result will
look significant, but you have learned nothing new. This is **circular inference**
(or: data dredging, p-hacking).

> *In our case:* After this lab you will have noticed that `ShelveLoc` seems associated
> with `Sales`. Running a t-test on the same 400 stores to "confirm" this is circular.
> The data have already been interrogated; they cannot serve as their own judge.

**What to do:**
Split the data *before* any exploration: one part for EDA, one part for testing.
This is the logic behind train/test splits and **A/B testing** (→ Session 3).
In PVA2 we examine what p-values and confidence intervals actually mean.
**For now: EDA generates questions --- it does not answer them.**
:::

---

## Step 1: Getting to Know the Dataset

### Structural Overview

```{r glimpse}
#| code-fold: false
glimpse(tbl_carseats)
```

:::{.callout-tip}
## 💡 What to read from `glimpse()`
- Number of rows (stores) and columns (variables)
- Data types: `<dbl>` = numeric, `<chr>` = character/text
- First values: do they look plausible?
:::

### Comprehensive Summary with `skim()`

`skim()` from the `skimr` package gives a structured, type-aware overview in one call:
missing values, quantiles, mean, standard deviation, and a small inline histogram.

```{r skim}
#| code-fold: false
tbl_carseats |> skim()
```

:::{.callout-note collapse="true"}
## 📚 Background: Reading a `skim()` output

The output is split by data type. For numeric variables, pay attention to:

- `n_missing`: missing values --- here 0 for all variables, so no imputation needed.
- `mean` and `p50` (median): if they diverge noticeably, the distribution is skewed.
- `hist`: the inline sparkline gives a first impression of shape.
- `sd`, `p0`, `p100`: range and spread --- are there implausible extremes?
:::

### Frequency Tables for Categorical Variables

For categorical variables, `count()` and `tabyl()` (from `janitor`) give
counts and proportions in one step.

```{r freq-tables}
#| code-fold: false
library(janitor)

# Simple count
tbl_carseats |> count(ShelveLoc, sort = TRUE)

# With proportions
tbl_carseats |>
  tabyl(ShelveLoc) |>
  adorn_pct_formatting(digits = 1) |>
  gt() |>
  tab_options(table.font.size = "13px")
```

```{r freq-urban-us}
# Cross-tabulation: Urban × US
tbl_carseats |>
  tabyl(Urban, US) |>
  adorn_totals(c("row", "col")) |>
  adorn_percentages("all") |>
  adorn_pct_formatting(digits = 1) |>
  adorn_ns() |>
  gt() |>
  tab_header(title = "Urban × US: counts (%) of all 400 stores") |>
  tab_options(table.font.size = "13px")
```

### Extended Descriptives with `psych::describe()`

`describe()` from the `psych` package adds skewness and kurtosis to the standard
summary --- useful for assessing distributional shape.

```{r psych-describe}
library(psych)

tbl_carseats |>
  select(Sales, Price, Advertising, Income, Age, Population) |>
  describe() |>
  as_tibble(rownames = "variable") |>
  select(variable, n, mean, sd, median, min, max, skew, kurtosis) |>
  gt() |>
  fmt_number(columns = where(is.numeric), decimals = 2) |>
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = variable == "Sales")
  ) |>
  tab_options(table.font.size = "13px") |>
  tab_footnote("Skew: |value| < 0.5 = approximately symmetric; > 1 = notably skewed")
```

### Quick Data Quality Overview with `DataExplorer`

```{r dataexplorer-intro}
library(DataExplorer)

# Types and missings at a glance
plot_intro(tbl_carseats)
```

:::{.callout-important icon=false}
## ⚠️ A1: Explore the Dataset

Using `glimpse()`, `skim()`, `tabyl()`, and `describe()`:

1. How many stores and variables does the dataset contain?
2. What is the **range** of `Sales`? What does one unit represent?
3. What proportion of stores have "Good" shelf placement?
4. Which numeric variable has the highest skewness? Does this surprise you?
5. Are there any missing values?
:::

---

## Step 2: Univariate Distributions

### Target Variable: Sales

A natural first question for any outcome variable: is the distribution symmetric,
or does it have a long tail that might cause problems for OLS? And --- given the
standard textbook advice to log-transform business variables --- should we apply a
log transformation here?

We look at both versions side by side and let the data answer.


:::{.callout-important icon=false}
## ⚠️ A2a: First Impression

Look at both histograms carefully before reading on.

1. Describe the distribution of raw `Sales` in your own words.
   Is it symmetric, right-skewed, or left-skewed? How close are mean and median?

2. The standard textbook advice for business variables is to apply a log transformation.
   Does `log1p(Sales)` follow this expectation --- or does the result surprise you?
   What does the comparison of the two skewness values tell you?
:::

```{r sales-hist}
#| code-fold: true
skew_sales <- round(skewness(tbl_carseats$Sales), 3)
skew_log   <- round(skewness(log1p(tbl_carseats$Sales)), 3)

p_raw <- tbl_carseats |>
  ggplot(aes(x = Sales)) +
  geom_histogram(bins = 30, fill = "#D50006", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(tbl_carseats$Sales),
             color = "#502479", linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = median(tbl_carseats$Sales),
             color = "#502479", linewidth = 1.2, linetype = "dotted") +
  annotate("text", x = mean(tbl_carseats$Sales) + 0.4, y = 37,
           label = "Mean", color = "#502479", size = 3.2, hjust = 0) +
  annotate("text", x = median(tbl_carseats$Sales) - 0.4, y = 32,
           label = "Median", color = "#502479", size = 3.2, hjust = 1) +
  labs(
    title    = "Sales --- raw",
    subtitle = paste0("Skewness = ", skew_sales,
                      "  |  Mean = ", round(mean(tbl_carseats$Sales), 2),
                      "  |  Median = ", round(median(tbl_carseats$Sales), 2)),
    x = "Sales (thousands of units)", y = "Count"
  ) +
  theme_minimal(base_size = 10)

p_log <- tbl_carseats |>
  ggplot(aes(x = log1p(Sales))) +
  geom_histogram(bins = 30, fill = "#686868", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(log1p(tbl_carseats$Sales)),
             color = "#D50006", linewidth = 1.2, linetype = "dashed") +
  labs(
    title    = "log1p(Sales) --- after transformation",
    subtitle = paste0("Skewness = ", skew_log,
                      "  |  Mean = ", round(mean(log1p(tbl_carseats$Sales)), 3)),
    x = "log1p(Sales) = log(Sales + 1)", y = "Count"
  ) +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(color = "#686868"))

p_raw + p_log +
  plot_annotation(
    title    = "Sales: raw vs. log-transformed",
    subtitle = "Does the log transformation improve distributional shape here?"
  )
```

Raw `Sales` (skewness = `r skew_sales`) is only slightly positive, hence already
approximately symmetric: this is also indicated by the fact that mean and median are
nearly identical and the distribution is bell-shaped.
After the `log1p` transformation, skewness becomes `r skew_log` --- considerably more
negative, meaning the transformation introduces left-skewness rather than removing
right-skewness --- the opposite of its intended effect.
**EDA before transformation --- always.**

This result is not accidental. It follows from how store-level unit sales are generated,
which is worth understanding. The background box below explains why.

:::{.callout-note collapse="true"}
## 📚 Background: When Is a Log Transformation Appropriate?

Many econometrics textbooks (Wooldridge, Stock & Watson) present log transformation
as a standard remedy for two problems:

- **Right-skewness:** Income, house prices, and firm sizes are almost always right-skewed
  (a few very large values, many small ones). Logging compresses the upper tail.
- **Interpretability:** Log-transformed models yield coefficients interpretable as
  elasticities (percentage changes).

Because students typically learn with wage or house price data, a heuristic forms:
*"business variable → log-transform."* This heuristic is often justified --- but it
conflates two fundamentally different types of business variable.

| Variable type | Typical distribution | Log appropriate? |
|---------------|---------------------|-----------------|
| **Revenue, wealth, asset prices** | Right-skewed: driven by a few very large values (power-law dynamics) | Usually yes |
| **Unit sales at store level** | Often approximately symmetric | Check empirically --- often no |

Why the difference? Unit sales of a stable consumer good across many similar stores
are the sum of many independent purchase decisions. The **Central Limit Theorem**
tells us that such sums converge to a symmetric distribution regardless of the
distribution of individual purchases. Revenue, by contrast, is frequently dominated
by a small number of very large transactions --- the power-law dynamics that make log
transformation genuinely necessary.

**The Carseats data illustrate this perfectly.** `Sales` measures unit sales across
400 stores. The EDA confirms the theoretical expectation: near-symmetry, no need for
transformation. A reflexive log-transform would have made the distribution *worse* ---
and we would only know this if we had looked at the data first.
:::

:::{.callout-important icon=false}
## ⚠️ A2b: Understanding the Transformation

1. The histogram above uses `log1p(Sales)` rather than `log(Sales)`.
   Find out why by running the naive version yourself:

```{r}
library(moments)
skewness(tbl_carseats$Sales)
skewness(log(tbl_carseats$Sales))   # Does this work as expected?
```

   - What does R return? Why?
   - Investigate the root cause:

```{r}
# Any zeros in Sales?
tbl_carseats |> filter(Sales == 0) |> nrow()

# What does log(0) return?
log(0)

# And how does skewness() handle that?
skewness(c(1, 2, 3, -Inf))
```

   - Is `Sales = 0` a plausible business observation, or a data quality issue?

2. Now compute skewness for `Advertising` and compare:

```{r}
skewness(log1p(tbl_carseats$Sales))
skewness(tbl_carseats$Advertising)
```

   For which variable is a log transformation more defensible: `Sales` or `Advertising`?
   Justify using the skewness values and the theoretical argument from the background box.

3. Look at the histogram of `Population` (coming up in the next plot).
   Is its shape surprising? What might explain it?
:::

---

## Step 3: Group-Level Descriptives with `group_by()` and `summarise()`

Having described each variable in isolation, we now condition on grouping variables
--- the first step towards multivariate thinking. `group_by()` combined with
`summarise()` is the most direct tidyverse tool for this: it computes descriptive
statistics *separately for each group*, making differences that are invisible in
marginal summaries immediately apparent. In a full multivariate analysis (regression,
Lab 3+), these group differences will reappear as coefficients --- but here we
describe them first without imposing a model.

### Sales by Shelf Location

```{r group-shelveloc}
#| code-fold: false
tbl_carseats |>
  group_by(ShelveLoc) |>
  summarise(
    n          = n(),
    mean_sales = round(mean(Sales), 2),
    median_sales = round(median(Sales), 2),
    sd_sales   = round(sd(Sales),   2),
    min_sales  = round(min(Sales),  2),
    max_sales  = round(max(Sales),  2)
  ) |>
  arrange(desc(mean_sales)) |>
  gt() |>
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = ShelveLoc == "Good")
  ) |>
  tab_style(
    style     = cell_fill(color = "#fff0f0"),
    locations = cells_body(rows = ShelveLoc == "Bad")
  ) |>
  tab_options(table.font.size = "13px") |>
  tab_header("Sales by shelf location (ShelveLoc)")
```

### Sales by Urban/US Segment

```{r group-urban-us}
tbl_carseats |>
  group_by(Urban, US) |>
  summarise(
    n          = n(),
    mean_sales = round(mean(Sales), 2),
    sd_sales   = round(sd(Sales),   2),
    .groups = "drop"
  ) |>
  gt() |>
  tab_options(table.font.size = "13px") |>
  tab_header("Sales by Urban status and US region")
```

### Price Distribution by Shelf Location

```{r price-by-shelveloc}
tbl_carseats |>
  group_by(ShelveLoc) |>
  summarise(
    mean_price = round(mean(Price), 2),
    sd_price   = round(sd(Price),   2),
    n          = n()
  ) |>
  gt() |>
  tab_options(table.font.size = "13px") |>
  tab_header("Price by ShelveLoc",
             subtitle = "Do stores with better shelf space charge higher prices?")
```

:::{.callout-important icon=false}
## ⚠️ A3: Group Comparisons

1. What is the difference in average sales between "Good" and "Bad" shelf locations?
   Is this difference large relative to the overall standard deviation of `Sales`?

2. Do stores with better shelf locations also charge higher prices on average?
   What would this imply for a simple regression of `Sales` on `Price` alone?
   (This question foreshadows **Omitted Variable Bias** --- discussed in →Lab 3.)

3. Extend the group-by analysis: compute mean `Sales` and `Advertising` for the
   four combinations of Urban × US:

```r
tbl_carseats |>
  group_by(Urban, US) |>
  summarise(
    mean_sales = mean(Sales),
    mean_adv   = mean(Advertising),
    n          = n(),
    .groups = "drop"
  )
```

   What pattern do you see? Are there systematic differences in advertising spend?
:::

---

## Step 4: Bivariate Relationships

### Price and Sales --- A First Look

```{r scatter-price-sales}
#| code-fold: true
tbl_carseats |>
  ggplot(aes(x = Price, y = Sales)) +
  geom_point(alpha = 0.4, color = "#7d0a52", size = 1.8) +
  geom_smooth(method = "lm",    color = "#D50006",
              se = FALSE, linewidth = 1.2) +
  geom_smooth(method = "loess", color = "#502479",
              se = FALSE, linewidth = 1, linetype = "dashed") +
  labs(
    title    = "Price vs. Sales",
    subtitle = paste0("Correlation r = ",
                      round(cor(tbl_carseats$Price, tbl_carseats$Sales), 3),
                      "  |  Red = OLS line  |  Purple dashed = LOESS"),
    x = "Price (USD)", y = "Sales (thousands of units)"
  ) +
  theme_minimal(base_size = 11)
```

The OLS line (red) and the LOESS curve (purple, see note below) are drawn here as **descriptive visual summaries**, not as inferential models. Here, they both tell the same story: the relationship between Price and Sales is approximately linear and negative.
This is useful prior information for Lab 3 --- a linear model is a reasonable
starting point.


:::{.callout-tip}
## 💡 ggplot default: `se = TRUE`
`geom_smooth()` adds a grey confidence band around the line by default.
This band represents inferential uncertainty around the fitted line and only makes
sense when treating the data as a sample. In EDA on a population, set `se = FALSE`
to avoid a misleading visual.
:::



:::{.callout-note collapse="true"}
## 📚 What is LOESS?

LOESS (*locally estimated scatterplot smoothing*) fits a smooth curve through the
data without assuming a specific functional form --- no linearity, no fixed slope.
It does this by fitting many small local regressions across the range of x, then
connecting them into a continuous curve.

In EDA, LOESS serves as a **linearity check**: if OLS line and LOESS curve run
closely together (as here), the linear assumption is supported. If they diverge,
a non-linear model or transformation may be needed.
:::


### Scatterplot Matrix: All Variables at Once

```{r ggpairs-raw}
#| code-fold: true
#| code-summary: "Show code: scatterplot matrix"
#| fig-height: 8
#| fig-width: 10
tbl_carseats |>
  select(Sales, Price, Advertising, Income, Age, ShelveLoc) |>
  ggpairs(
    aes(color = ShelveLoc, alpha = 0.4),
    upper = list(continuous = wrap("cor", size = 2.8)),
    lower = list(continuous = wrap("points", alpha = 0.25, size = 0.7)),
    diag  = list(continuous = wrap("densityDiag", alpha = 0.5))
  ) +
  scale_color_manual(values = c(
    "Bad"    = "#D50006",
    "Medium" = "#686868",
    "Good"   = "#502479"
  )) +
  scale_fill_manual(values = c(
    "Bad"    = "#D50006",
    "Medium" = "#686868",
    "Good"   = "#502479"
  )) +
  theme_minimal(base_size = 8) +
  labs(title    = "Scatterplot matrix: Carseats variables",
       subtitle = "Coloured by shelf location (ShelveLoc)")
```

Look at the rightmost column carefully. The shelf location categories appear in the
order **Bad --- Good --- Medium** rather than the natural rank Bad < Medium < Good.
In practice, unexpected category orderings like this are often a signal worth
investigating: they can indicate a coding error that leads to wrong statistical
results --- for example, if a model interprets the levels in the wrong order.

:::{.callout-important icon=false}
## ⚠️ A4a: Spot the Problem

1. Check the class and levels of `ShelveLoc`:
```r
class(tbl_carseats$ShelveLoc)
levels(tbl_carseats$ShelveLoc)
```

   Is `ShelveLoc` already a factor? If so --- why is the order still wrong?

2. Here, the wrong ordering does not affect statistical results: `ggpairs` 
   displays the existing factor levels as-is, but regression dummy coding and
   group comparisons are unaffected by alphabetical vs. natural ordering.
   The problem is purely visual --- but visual clarity matters in EDA.
   Fix it for the plot by relevelling:
```r
tbl_carseats <- tbl_carseats |>
  mutate(ShelveLoc = factor(ShelveLoc,
                            levels = c("Bad", "Medium", "Good")))
```

   Re-run the `ggpairs` plot. What changes in the rightmost column?
:::

:::{.callout-note collapse="true"}
## 📚 character vs. factor vs. ordered factor

| Class | Ordering | Typical use |
|-------|----------|-------------|
| `character` | None --- alphabetical by default | Raw text, free-form labels |
| `factor` | Fixed levels, alphabetical by default if not specified | Nominal categories (Urban, US) |
| `ordered factor` | Levels have a meaningful rank | Ordinal categories (education level, satisfaction) |

The general principle: **scale level determines variable class.**
Nominal → `factor`, ordinal → `ordered factor`, metric → `numeric`.

For `ShelveLoc`, a plain `factor` with correctly specified levels is sufficient
for EDA and regression. An `ordered factor` would additionally signal the rank
Bad < Medium < Good to R's modelling functions --- relevant if you want
monotonicity constraints or ordinal contrasts, but not required here.

Mismatches between content and class are a common source of subtle errors and
are exactly the kind of thing EDA is designed to catch.
:::
```{r ggpairs-fixed}
#| code-fold: true
#| code-summary: "Show code: scatterplot matrix (corrected level order)"
#| fig-height: 8
#| fig-width: 10
tbl_carseats |>
  mutate(ShelveLoc = factor(ShelveLoc,
                            levels = c("Bad", "Medium", "Good"))) |>
  select(Sales, Price, Advertising, Income, Age, ShelveLoc) |>
  ggpairs(
    aes(color = ShelveLoc, alpha = 0.4),
    upper = list(continuous = wrap("cor", size = 2.8)),
    lower = list(continuous = wrap("points", alpha = 0.25, size = 0.7)),
    diag  = list(continuous = wrap("densityDiag", alpha = 0.5))
  ) +
  scale_color_manual(values = c(
    "Bad"    = "#D50006",
    "Medium" = "#686868",
    "Good"   = "#502479"
  )) +
  scale_fill_manual(values = c(
    "Bad"    = "#D50006",
    "Medium" = "#686868",
    "Good"   = "#502479"
  )) +
  theme_minimal(base_size = 8) +
  labs(title    = "Scatterplot matrix: Carseats variables (corrected)",
       subtitle = "ShelveLoc levels: Bad < Medium < Good")
```

### Correlation Matrix

```{r correlation-matrix}
cor_mat <- tbl_carseats |>
  select(Sales, Price, Advertising, Income, Age, Population) |>
  cor() |>
  round(3)

cor_mat |>
  as_tibble(rownames = "variable") |>
  gt() |>
  fmt_number(columns = where(is.numeric), decimals = 3) |>
  data_color(
    columns   = where(is.numeric),
    rows      = everything(),
    palette   = c("#D50006", "white", "#502479"),
    domain    = c(-1, 1)
  ) |>
  tab_options(table.font.size = "12px") |>
  tab_header("Correlation matrix (numeric variables)")
```

```{r correlations-ranked}
#| code-fold: false
# Ranked correlations with Sales
cor_mat |>
  as_tibble(rownames = "variable") |>
  select(variable, Sales) |>
  filter(variable != "Sales") |>
  arrange(desc(abs(Sales))) |>
  gt() |>
  fmt_number(columns = Sales, decimals = 3) |>
  tab_options(table.font.size = "13px") |>
  tab_header("Correlation with Sales --- ranked by absolute value")
```

:::{.callout-important icon=false}
## ⚠️ A4b: Bivariate Analysis

1. Which numeric variable has the **strongest** linear correlation with `Sales`?
   Which has the weakest?

2. Are there pairs of *predictors* that are strongly correlated with each other
   (potential multicollinearity)? Why would this matter in a regression model?

3. Inspect the `ggpairs` matrix carefully. The `ShelveLoc` colouring reveals
   subgroup differences invisible in aggregated correlations:
   - Does the slope of the Price--Sales relationship look different across ShelveLoc groups?
   - What does this imply for modelling?
:::

---

## Step 5: Boxplots and Conditional Distributions

Boxplots add a second perspective to the group-level summaries from Step 3:
instead of a single mean, they show the full distributional shape --- median,
interquartile range, whiskers, and outliers --- *per group*.
The two key questions from A3 reappear here visually:
how much do Sales differ across shelf locations, and do stores with better
shelf space also charge higher prices?

Both plots use `ShelveLoc` with corrected level order (Bad < Medium < Good),
carried forward from the recodification in Step 4.

```{r boxplots-core}
#| code-fold: false
#| fig-height: 3.5
# ── Data: reorder factor levels ───────────────────────────────────────────────
tbl_cs <- tbl_carseats |>
  mutate(ShelveLoc = factor(ShelveLoc,
                            levels = c("Bad", "Medium", "Good")))

# ── Core plots ────────────────────────────────────────────────────────────────
p1 <- tbl_cs |>
  ggplot(aes(x = ShelveLoc, y = Sales, fill = ShelveLoc)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white")

p2 <- tbl_cs |>
  ggplot(aes(x = ShelveLoc, y = Price, fill = ShelveLoc)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white")

p1 + p2
```

```{r boxplots-layout}
#| code-fold: true
#| code-summary: "Show code: colours, labels & theme"
#| fig-height: 3.5
col_shelf <- c("Bad" = "#D50006", "Medium" = "#686868", "Good" = "#502479")

p1 <- tbl_cs |>
  ggplot(aes(x = ShelveLoc, y = Sales, fill = ShelveLoc)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +
  scale_fill_manual(values = col_shelf) +
  labs(title    = "Sales by shelf location",
       subtitle = "Diamond = mean | line = median",
       x = NULL, y = "Sales (thousands)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")

p2 <- tbl_cs |>
  ggplot(aes(x = ShelveLoc, y = Price, fill = ShelveLoc)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +
  scale_fill_manual(values = col_shelf) +
  labs(title    = "Price by shelf location",
       subtitle = "Do premium shelves command higher prices?",
       x = NULL, y = "Price (USD)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")

p1 + p2
```

The left panel confirms the group-level finding from Step 3: Good > Medium > Bad,
with a substantial gap. The right panel raises a flag: stores with better shelf
placement tend to charge *higher* prices --- not lower. This is the pattern that
will reappear in Lab 3 as **Omitted Variable Bias**.

```{r adv-sales-scatter}
#| code-fold: false
#| fig-height: 3.5
# Advertising vs. Sales coloured by ShelveLoc
# (Option 3: layout comments inline)
tbl_cs |>
  ggplot(aes(x = Advertising, y = Sales, color = ShelveLoc)) +
  geom_point(alpha = 0.35, size = 1.4) +                       # <- data
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.9) +    # <- trend per group
  scale_color_manual(values = col_shelf) +                      # --- layout ---
  labs(title    = "Advertising vs. Sales by shelf location",
       subtitle = paste0("Overall r = ",
                         round(cor(tbl_cs$Advertising, tbl_cs$Sales), 3)),
       x = "Advertising (thousands USD)", y = "Sales (thousands)",
       color = "ShelveLoc") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

All three groups show a positive advertising--sales relationship, and the slopes
are similar across shelf locations. The vertical separation between the three trend
lines reflects the ShelveLoc main effect already seen in the boxplots.

---

## Step 6: Formulating Hypotheses

The EDA has surfaced several patterns. The next step is to translate observations
into testable hypotheses for the modelling phase (→Labs 3 and 4).

:::{.callout-important icon=false}
## ⚠️ A5: Hypotheses for Modelling

Based on your EDA, formulate **three hypotheses** relevant to the goal
"Identify underperforming stores".

**Structure:**
> *"I expect [variable X] to have a [positive/negative] association with Sales because
> [substantive reasoning]. This is visible in [specific EDA finding: table/plot]."*

Additionally:

1. The `ggpairs` matrix shows that `ShelveLoc` is associated with both `Price` and
   `Sales`. What does this imply for a simple regression of `Sales` on `Price` alone?
   (One sentence is enough --- we will formalise this in →Lab 3 as Omitted Variable Bias.)

2. Which variables should be included in the regression model?
   Which are probably redundant --- and why?

3. What variables are **missing** from the dataset that could explain sales differences?
   (Keyword: *Data Drought*, @thomasAIValueCreators2025)

4. **The induction trap:** You have now explored all 400 stores. What would you need to
   do *before* this exploration if you wanted to run valid statistical tests on these
   hypotheses?
:::

---

## Bonus Tasks

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B1: Automated EDA Report

`DataExplorer` can generate a full HTML EDA report in one line:

```r
# Generates a self-contained HTML report
create_report(tbl_carseats,
              output_file  = "EDA_Carseats_report.html",
              output_dir   = ".",
              report_title = "Carseats EDA Report")
```

Open the report. Which visualisations are most informative?
Which would you include in a presentation to management?
:::

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B2: Missing Data Visualisation

The Carseats data have no missing values --- but real datasets rarely do.
Explore a dataset with missings using `visdat` and `naniar`:

```r
library(visdat)
library(naniar)

# Visual overview of data types and missings
vis_dat(airquality)

# Summary of missing values
miss_var_summary(airquality)

# Missingness as a pattern
gg_miss_upset(airquality)
```

What patterns do you see? Are certain variables always missing together?
:::

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B3: Segmented EDA

Conduct the EDA separately for urban and non-urban stores.
Do the price and advertising effects differ?

```r
tbl_carseats |>
  ggplot(aes(x = Price, y = Sales, color = Urban)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("Yes" = "#502479", "No" = "#D50006")) +
  facet_wrap(~US) +
  labs(title = "Price vs. Sales by location (Urban) and region (US)",
       x = "Price (USD)", y = "Sales (thousands)") +
  theme_minimal()
```

Any indications of interaction effects between Urban status and US region?
:::

---

## Summary

```{r summary-table}
#| echo: false
tibble::tribble(
  ~`EDA Finding`,          ~`Key result`,                                     ~`Implication for Lab 3+`,
  "Structure",             "400 stores, 11 variables, 0 missing values",      "Dataset ready to use directly",
  "Sales distribution",    "Approximately symmetric (skewness ~ 0.18)",       "No log transformation needed",
  "Advertising",           "Right-skewed (many zeros; skewness ~ 2)",         "Log(Advertising+1) worth considering",
  "ShelveLoc effect",      "Good: ~8 vs. Bad: ~5 thousand units (mean)",      "Large categorical effect -> must include",
  "Price--Sales corr.",    "r ~ -0.44, negative linear trend",                "Include Price; but OVB risk via ShelveLoc",
  "Advertising--Sales",    "r ~ +0.58, strongest numeric predictor",          "Include Advertising in model",
  "Missing variables",     "No competition, demographics, or location data",  "Data Drought -> motivates Lab 4"
) |>
  gt() |>
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = `EDA Finding` == "Sales distribution")
  ) |>
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = `EDA Finding` == "Missing variables")
  ) |>
  tab_options(table.font.size = "13px")
```

:::{.callout-important}
## Key Messages of This Lab

1. **EDA is CRISP-DM Phase 2.** It follows Business Understanding and precedes all
   modelling. Its purpose is data comprehension, not prediction.

2. **Never trust a statistic you haven't plotted.** The Datasaurus Dozen illustrates
   this dramatically: identical summary statistics, completely different realities.

3. **EDA and Data Mining are both inductive.** They differ in scale, purpose, and
   success criteria --- not in their logical structure. The OLS line in a scatter plot
   is EDA when it aids understanding; it becomes a DM artefact when evaluated on new data.

4. **The induction trap.** Hypotheses generated from data cannot be tested on those same
   data. Plan the data split before exploration.

5. **EDA before transformation.** `Sales` is symmetric (CLT applies to store-level unit
   sales); the log transformation makes it *worse*. `Advertising` is right-skewed and
   *is* a candidate for transformation. Always check empirically.

6. **`group_by()` + `summarise()` is the most powerful EDA tool in the tidyverse.**
   Conditional means reveal patterns that global statistics hide.
:::

:::{.callout-note}
## Transition to Lab 3

In →Lab 3 we fit the **baseline model** `Sales ~ Price`.
The EDA findings predict exactly what we will see:
a low $R^2$ (Price alone is not enough) and a systematic ShelveLoc pattern in the
residual plot --- the signature of **Omitted Variable Bias**.

We will also examine the regression output table critically:
since we treat the 400 stores as a population, what do "estimate", "std_error",
and "p_value" actually mean?
:::

## References {.unnumbered}
