---
title: "Lab 2: Data Understanding & Exploratory Data Analysis"
subtitle: "CRISP-DM Phase 2 — Carseats Case"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en-US
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse)
library(moderndive)
library(skimr)
library(gt)
library(GGally)
library(patchwork)
library(moments)

library(ISLR2)
tbl_carseats <- tibble(Carseats)
```

## Learning Objectives and Overview

:::{.callout-note}
## Learning Objectives

- Explore the Carseats dataset: structure, variable types, distributions
- Understand EDA as **descriptive** and **inductive** — and know what follows from this
- Visualise bivariate relationships and formulate hypotheses for modelling
- Establish EDA as a mandatory first step before any regression analysis
:::

**Scenario:** You are a data scientist at a retail chain with 400 stores.
Before building a model, you need to understand the data ---
this is CRISP-DM Phase 2: **Data Understanding**.

:::{.callout-tip}
## Schedule
⏱ **20 minutes** --- Mandatory tasks (A1--A4) first, bonus tasks if time permits.
:::

---

## Background: Why EDA First?

### EDA Is Descriptive --- and Treats the Data as a Population

A crucial distinction before we begin: **Exploratory Data Analysis (EDA) is purely descriptive.**

When we summarise the 400 Carseats stores using means, standard deviations, or correlations,
we are not estimating population parameters from a sample --- we are *describing the data we have*.
The 400 stores **are** the population of interest (or at least the complete dataset at hand).
As a result:

- Summary statistics (means, correlations, etc.) are **parameters**, not estimates.
- There is **no sampling variability** to account for --- no standard errors, no p-values,
  no confidence intervals.
- The goal is **understanding and pattern discovery**, not inference.

This has an important practical implication: the familiar table of inferential statistics
(t-values, p-values, confidence intervals) that appears in regression output makes no conceptual
sense in a purely descriptive, population-level analysis. We will return to this in Lab 3.

### EDA vs. Data Mining --- and the Induction--Deduction Distinction

:::{.callout-note}
## EDA vs. Data Mining

Both EDA and Data Mining are empirical, data-driven approaches --- but they differ
fundamentally in purpose, logic, and method:

| Dimension | EDA | Data Mining |
|-----------|-----|-------------|
| **Goal** | Understand the data; generate hypotheses | Extract actionable patterns; build predictive models |
| **Reasoning** | **Inductive** --- patterns in data suggest hypotheses | **Deductive** --- hypotheses are tested against data |
| **Orientation** | Descriptive --- "What is in the data?" | Predictive/prescriptive --- "What will happen?" |
| **Methods** | Summary statistics, visualisation, correlation | Regression, classification, clustering, association rules |
| **Output** | Insights, hypotheses, data quality assessment | Scores, rankings, predictions, rules |
| **Role in CRISP-DM** | Phase 2: Data Understanding | Phases 4--6: Modelling, Evaluation, Deployment |
| **Population treatment** | Treats available data as population | Typically assumes a sample; uses inferential statistics |

**In this course:**\
Lab 2 (this lab) is EDA --- descriptive, inductive exploration of the Carseats data.\
Lab 4 transitions to Data Mining: using the regression model as a scoring tool.
:::

:::{.callout-warning}
## The Induction Trap: Why You Cannot Test What You Discovered in the Same Data

EDA is **inductive**: you observe patterns in the data and let them suggest hypotheses.
This is a legitimate and valuable first step. The danger arises immediately afterwards.

**The critical rule:** A hypothesis generated inductively from a dataset **cannot be
validly tested on that same dataset.**

Why not? Because any pattern you "discover" in a dataset will --- by construction --- appear
to be confirmed when you apply a statistical test to those very same observations.
The p-value will be misleadingly small; the result will look significant; but you have
learned nothing new. You are simply re-describing what you already saw.
This is called **circular inference** or, less charitably, **data dredging**.

> *Example in our case:* You explore the 400 Carseats stores, notice that `ShelveLoc`
> appears associated with `Sales`, and then run a t-test on the same 400 stores to
> "confirm" it. The test will almost certainly be significant --- but only because you used
> the same data to generate and to test the hypothesis. The data cannot be their own judge.

**What to do instead:**

- **Split the data upfront** --- before any exploration --- into an *exploration set* and a
  *confirmation set*. This is the logic behind train/test splits and, more formally,
  **A/B testing** (→ Session 3).
- Or: formulate hypotheses *a priori* on theoretical grounds and test them on fresh data.

In **PVA2** we will examine what p-values and confidence intervals actually measure and
when they are valid. For now: **EDA generates questions --- it does not answer them.**
:::

### Why Visualisation Is Non-Negotiable: Anscombe's Quartet

The classic demonstration that summary statistics alone are deeply misleading comes from
**Anscombe (1973)**. He hand-constructed four small datasets with *virtually identical*
summary statistics --- yet with fundamentally different structures.

Read the table first. Then look at the scatter plots. Compare row by row.

```{r anscombe-stats}
#| code-fold: false
anscombe_long <- anscombe |>
  pivot_longer(
    everything(),
    names_to  = c(".value", "dataset"),
    names_pattern = "(.)(.)"
  ) |>
  mutate(dataset = paste("Dataset", dataset))

anscombe_long |>
  group_by(dataset) |>
  summarise(
    n      = n(),
    mean_x = round(mean(x), 2),
    mean_y = round(mean(y), 2),
    sd_x   = round(sd(x),   2),
    sd_y   = round(sd(y),   2),
    r_xy   = round(cor(x, y), 3)
  ) |>
  gt() |>
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_column_labels()
  ) |>
  tab_options(table.font.size = "13px") |>
  tab_footnote(
    "All four datasets have (almost) identical summary statistics -- yet the plots below tell four different stories."
  )
```

```{r anscombe-plots}
#| fig-height: 5
#| fig-cap: "Anscombe's Quartet: same n, mean, sd, and correlation --- four entirely different structures."
anscombe_long |>
  ggplot(aes(x = x, y = y)) +
  geom_point(color = "#502479", size = 2.5, alpha = 0.85) +
  geom_smooth(method = "lm", color = "#D50006",
              se = TRUE, linewidth = 1) +
  facet_wrap(~dataset, nrow = 2) +
  labs(
    title    = "Anscombe's Quartet",
    subtitle = "Identical summary statistics -- four different stories",
    x = "x", y = "y"
  ) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))
```

:::{.callout-important}
## What Each Panel Teaches --- and Why It Matters for Regression Diagnostics

| Panel | Structure | What the OLS line misses | Detected by |
|-------|-----------|--------------------------|-------------|
| **I** | Linear, roughly homoscedastic | Nothing --- OLS is appropriate | Residual plot: random scatter ✓ |
| **II** | Curved (quadratic) relationship | The non-linearity --- linear model is structurally wrong | Residual plot: curved pattern |
| **III** | One outlier in *y* | Outlier distorts intercept and slope | Residual plot: one extreme point |
| **IV** | One leverage point (*x* = 19) | Single extreme *x*-value pulls entire line | Residual plot: gap in x-values |

All four problems are diagnosed using the **residual plot** --- which you will apply
in Lab 3 to detect Omitted Variable Bias.

*The Datasaurus Dozen* (Matejka & Fitzmaurice, 2017) extends the same argument to 13
algorithmically generated datasets with 142 points each (including a dinosaur shape).
Anscombe's quartet is more instructive for a regression course because each panel
encodes a *different* diagnostic problem --- not just a different visual shape.
:::

### Elements of Exploratory Data Analysis

According to @stoetzer_regressionsanalyse_2020, EDA follows four guiding principles:

1. **Substantive meaning** --- The primary goal is understanding variables and their
   plausible relationships, not running formal tests.

2. **Graphical representation** --- Visual inspection of variables and distributions
   is the primary means of gaining insight.

3. **Iterative model building** --- The focus is on tentative hypothesis generation in
   an iterative process of model specification and re-specification.

4. **Scepticism and parsimony** --- The starting points are scepticism, flexibility,
   and restraint regarding which statistical methods to apply.

:::{.callout-note}
## EDA as CRISP-DM Phase 2
In our Carseats project, EDA serves to generate **hypotheses** for the modelling phase:
Which variables explain sales differences across stores?
Which transformations are necessary? Are there outliers or systematic patterns?
:::

---

## Step 1: Getting to Know the Dataset

### Overview: Structure and Variables

```{r glimpse}
#| code-fold: false
glimpse(tbl_carseats)
```

:::{.callout-tip}
## Interpretation
- How many observations (stores) and variables does the dataset contain?
- Which variables are numeric (`dbl`), which are categorical (`chr`/`fct`)?
- What exactly does `Sales` measure? (Pay attention to the unit!)
:::

### Summary Statistics

```{r skim}
tbl_carseats |>
  select(Sales, Price, Advertising, Income, Age, ShelveLoc, Urban, US) |>
  skim() |>
  yank("numeric") |>
  gt() |>
  fmt_number(columns = where(is.numeric), decimals = 2) |>
  tab_options(table.font.size = "13px")
```

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A1: Explore the Dataset

1. How many stores (`n`) does the dataset contain?
2. What is the **range** of `Sales` (Min, Max)? What does this mean in practice?
3. What is the **range** of `Price`? Is the car-seat market price-elastic?
4. What does the variable `ShelveLoc` represent? What categories does it have?

```r
tbl_carseats |> count(ShelveLoc)
tbl_carseats |> count(Urban)
tbl_carseats |> count(US)
```
:::

---

## Step 2: Distribution of the Target Variable Sales

### Check Before You Transform

A common reflex in applied work is to log-transform sales data "because sales are
always skewed." **This is wrong as a default.** Whether a transformation is appropriate
must be determined from the data --- not assumed in advance. That is exactly what EDA is for.

```{r sales-distribution}
#| code-fold: false
skew_raw <- round(skewness(tbl_carseats$Sales), 3)
skew_log <- round(skewness(log(tbl_carseats$Sales)), 3)

p1 <- tbl_carseats |>
  ggplot(aes(x = Sales)) +
  geom_histogram(bins = 30, fill = "#D50006", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(tbl_carseats$Sales),
             color = "#502479", linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = median(tbl_carseats$Sales),
             color = "#502479", linewidth = 1.2, linetype = "dotted") +
  annotate("text", x = mean(tbl_carseats$Sales) + 0.4, y = 38,
           label = "Mean", color = "#502479", size = 3.2, hjust = 0) +
  annotate("text", x = median(tbl_carseats$Sales) - 0.4, y = 33,
           label = "Median", color = "#502479", size = 3.2, hjust = 1) +
  labs(
    title    = "Sales -- raw data",
    subtitle = paste0("Skewness = ", skew_raw,
                      "  |  Mean = ", round(mean(tbl_carseats$Sales), 2),
                      "  |  Median = ", round(median(tbl_carseats$Sales), 2)),
    x = "Sales (thousands of units)", y = "Count"
  ) +
  theme_minimal(base_size = 11)

p2 <- tbl_carseats |>
  ggplot(aes(x = log(Sales))) +
  geom_histogram(bins = 30, fill = "#686868", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(log(tbl_carseats$Sales)),
             color = "#D50006", linewidth = 1.2, linetype = "dashed") +
  labs(
    title    = "log(Sales) -- after transformation",
    subtitle = paste0("Skewness = ", skew_log,
                      "  |  Mean log(Sales) = ",
                      round(mean(log(tbl_carseats$Sales)), 3)),
    x = "log(Sales)", y = "Count"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(color = "#686868"))

p1 + p2
```

:::{.callout-important}
## Result: The Log Transformation Makes Things Worse Here

The raw `Sales` data are already **approximately symmetric**:

- Skewness of `Sales`: `r skew_raw` --- close to zero.
- Mean and Median barely differ.

After the log transformation, skewness becomes `r skew_log` --- **more negative**,
meaning the transformation *introduces* left-skewness rather than removing right-skewness.
This is the opposite of what log transformation is designed to achieve.

**Lesson: EDA before transformation --- always.**
We will work with **untransformed Sales** throughout this course.
:::

:::{.callout-note}
## When Is a Log Transformation Appropriate?

The log transformation is theoretically motivated by a **multiplicative generating process**:

$$\text{Sales}_i = \alpha \cdot \text{Price}_i^{\beta_1} \cdot \text{Adv}_i^{\beta_2} \cdot \varepsilon_i$$

Taking logs converts this into an additive, OLS-compatible model whose coefficients are
**elasticities**. But the practical precondition --- a right-skewed raw distribution ---
must first be verified empirically.

Contrast: the variable `Advertising` *is* right-skewed (many stores spend nothing; a few
spend a lot). A log transformation of `Advertising` would be more defensible. Check
this in task A2 below.
:::

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A2: Assess the Distribution

1. Describe the distribution of `Sales` in your own words.
   Is it symmetric, right-skewed, or left-skewed? How close are mean and median?

2. Compute skewness for `Sales`, `log(Sales)`, and `Advertising` for comparison:

```r
library(moments)
skewness(tbl_carseats$Sales)
skewness(log(tbl_carseats$Sales))
skewness(tbl_carseats$Advertising)   # this one IS right-skewed -- compare!
```

3. For which variable is a log transformation more justified: `Sales` or `Advertising`?
   Justify based on the skewness values and histograms.
:::

---

## Step 3: Bivariate Relationships

### Price vs. Sales

```{r scatter-price-sales}
#| code-fold: false
tbl_carseats |>
  ggplot(aes(x = Price, y = Sales)) +
  geom_point(alpha = 0.4, color = "#7d0a52", size = 1.8) +
  geom_smooth(method = "lm", color = "#D50006",
              se = TRUE, linewidth = 1.2) +
  geom_smooth(method = "loess", color = "#502479",
              se = FALSE, linewidth = 1, linetype = "dashed") +
  labs(
    title    = "Scatterplot: Price vs. Sales",
    subtitle = paste0("Correlation r = ",
                      round(cor(tbl_carseats$Price, tbl_carseats$Sales), 3),
                      "  |  Red = OLS line  |  Purple dashed = LOESS"),
    x = "Price (USD)", y = "Sales (thousands of units)"
  ) +
  theme_minimal(base_size = 11)
```

:::{.callout-tip}
## Interpretation
- What does the correlation say? Is it strong or weak?
- Does the OLS line agree with the LOESS trend? What does that indicate?
- Why is the scatter around the line so high? What is missing from the model?
:::

### Scatterplot Matrix: All Variables at Once

```{r ggpairs}
#| fig-height: 8
tbl_carseats |>
  select(Sales, Price, Advertising, Income, Age, ShelveLoc) |>
  ggpairs(
    aes(color = ShelveLoc, alpha = 0.5),
    upper = list(continuous = wrap("cor", size = 3)),
    lower = list(continuous = wrap("points", alpha = 0.3, size = 0.8)),
    diag  = list(continuous = wrap("densityDiag", alpha = 0.6))
  ) +
  scale_color_manual(values = c(
    "Bad"    = "#D50006",
    "Medium" = "#686868",
    "Good"   = "#502479"
  )) +
  scale_fill_manual(values = c(
    "Bad"    = "#D50006",
    "Medium" = "#686868",
    "Good"   = "#502479"
  )) +
  theme_minimal(base_size = 9) +
  labs(title    = "Scatterplot matrix: Carseats variables",
       subtitle = "Coloured by shelf location (ShelveLoc)")
```

:::{.callout-note}
## GGally::ggpairs() --- EDA in One Line of Code

`ggpairs()` from the `GGally` package produces in a single call:

- **Diagonal:** Distributions of individual variables
- **Upper triangle:** Correlation coefficients
- **Lower triangle:** Scatterplots of all variable pairs

The correlation coefficients in the upper triangle are summary statistics --- and as
Anscombe showed, summary statistics alone are not enough. Always cross-check with the
scatter plots in the lower triangle, which can reveal non-linearities, outliers, or
subgroup structure that a single number cannot capture.
:::

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A3: Bivariate Analysis

1. Which variables correlate **most strongly** with `Sales`? Create a ranked list:

```r
tbl_carseats |>
  select(where(is.numeric)) |>
  cor() |>
  as_tibble(rownames = "variable") |>
  select(variable, Sales) |>
  arrange(desc(abs(Sales)))
```

2. Are there variables strongly correlated **with each other** (multicollinearity risk)?
   What could this mean for a multiple regression?

3. Describe the relationship between `ShelveLoc` and `Sales` in words.
   Which category has the highest average sales?

```r
tbl_carseats |>
  group_by(ShelveLoc) |>
  summarise(mean_sales = mean(Sales),
            sd_sales   = sd(Sales),
            n          = n())
```
:::

---

## Step 4: Formulating Hypotheses

```{r hypotheses-viz}
#| fig-height: 4
p_shelf <- tbl_carseats |>
  ggplot(aes(x = ShelveLoc, y = Sales, fill = ShelveLoc)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
  scale_fill_manual(values = c("Bad"    = "#D50006",
                               "Medium" = "#686868",
                               "Good"   = "#502479")) +
  labs(title = "Sales by shelf location",
       x = "Shelf location", y = "Sales (thousands)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

p_adv <- tbl_carseats |>
  ggplot(aes(x = Advertising, y = Sales)) +
  geom_point(alpha = 0.3, color = "#7d0a52", size = 1.5) +
  geom_smooth(method = "lm", color = "#D50006", se = FALSE) +
  labs(title = "Advertising vs. Sales",
       subtitle = paste0("r = ",
                         round(cor(tbl_carseats$Advertising,
                                   tbl_carseats$Sales), 3)),
       x = "Advertising budget (thousands USD)", y = "Sales (thousands)") +
  theme_minimal(base_size = 11)

p_shelf + p_adv
```

:::{.callout-caution icon=false}
## `r fontawesome::fa("pen")` A4: Hypotheses for Modelling

Based on your EDA, formulate **two hypotheses** for the goal
"Identification of underperforming stores".

**Example structure:**
> *"I expect that [variable X] has a [positive/negative] relationship with Sales,
> because [substantive justification]."*

Additionally:

1. For which variable might a **non-linear effect** be plausible?

2. Which variables should **definitely** be included in the regression model?
   Which probably not --- and why?

3. Which variables are **missing** that would be important for explaining sales differences?
   (Keyword: *Data Drought*, @thomasAIValueCreators2025)

4. **Critical reflection (induction trap):** You have now used all 400 stores to
   generate these hypotheses. What would need to happen *before exploration*
   if you wanted to run valid statistical tests afterwards?
:::

---

## Bonus Tasks

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B1: Estimate Price Elasticity

Fit a simple log-log regression for exploratory purposes. Note: this is a
descriptive exercise --- since the hypothesis was generated from the same data,
the result is not a valid confirmatory test.

```r
mod_loglog <- tbl_carseats |>
  mutate(log_sales = log(Sales),
         log_price = log(Price)) |>
  lm(log_sales ~ log_price, data = _)

get_regression_table(mod_loglog)
```

- What does the coefficient for `log_price` mean as a price elasticity?
- Compare $R^2$ with the linear model `Sales ~ Price`. What changes?
- Given that `Sales` is not right-skewed: does the log-log specification add value here?
:::

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B2: Outlier Analysis

```r
tbl_carseats |>
  mutate(rank = dense_rank(desc(Sales))) |>
  filter(rank <= 5 | rank >= (n() - 4)) |>
  select(Sales, Price, Advertising, ShelveLoc, Income, rank) |>
  arrange(rank)
```

Are there stores with extreme price-sales combinations (high price, high sales --- or vice versa)?
:::

:::{.callout-tip icon=false}
## `r fontawesome::fa("star")` B3: Segmented EDA

```r
tbl_carseats |>
  ggplot(aes(x = Price, y = Sales, color = Urban)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("Yes" = "#502479", "No" = "#D50006")) +
  facet_wrap(~US) +
  labs(title = "Price vs. Sales by location (Urban) and region (US)",
       x = "Price (USD)", y = "Sales (thousands)") +
  theme_minimal()
```

Any indications of interaction effects?
:::

---

## Summary

```{r summary-table}
#| echo: false
tibble::tribble(
  ~`EDA Step`,           ~`Finding`,                                       ~`Consequence for Modelling`,
  "Structure",           "400 stores, 11 variables, no NAs",               "Clean dataset, ready to use",
  "Sales distribution",  "Approximately symmetric (skewness ~ 0.18)",      "No log transformation needed",
  "Price effect",        "r ~ -0.44, negative linear trend",               "Include Price as regressor",
  "ShelveLoc",           "Good > Medium > Bad (marked difference)",        "Include as categorical variable",
  "Advertising",         "r ~ +0.58, strongest positive predictor",        "Include Advertising in the model",
  "Missing variables",   "No competition, no demographics, no location",   "Data Drought -> pilot project"
) |>
  gt() |>
  tab_style(
    style     = cell_fill(color = "#f5f0fa"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = `EDA Step` == "Sales distribution")
  ) |>
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = `EDA Step` == "Missing variables")
  ) |>
  tab_options(table.font.size = "13px")
```

:::{.callout-important}
## Key Messages of This Lab

1. **EDA first --- always.** Summary statistics without visualisation are insufficient.
   Anscombe's quartet proves it: four datasets, identical statistics, four completely
   different structures and regression diagnostics.

2. **EDA is descriptive and inductive.** It treats the available data as a population
   and generates hypotheses from observed patterns. No p-values or confidence intervals
   are needed --- or valid --- at this stage.

3. **The induction trap.** Hypotheses discovered by looking at data cannot be tested on
   those same data. If confirmatory testing is intended, split the data *before* any
   exploration (-> A/B testing, Session 3).

4. **Check before transforming.** `Sales` in the Carseats data is already approximately
   symmetric. The log transformation introduces left-skewness --- the opposite of its
   intended effect. Transformation decisions must always be grounded in EDA.

5. **ShelveLoc and Advertising** are the strongest predictors --- both absent from the
   baseline model (-> Lab 3).

6. **Data Drought.** Important explanatory variables (location, competition, demographics)
   are missing --- motivating the Make-or-Buy question in Lab 4.
:::

:::{.callout-note}
## Transition to Lab 3

In Lab 3 we fit the **baseline model** `Sales ~ Price` and examine it critically.
A low $R^2$ and a systematic pattern in the residual plot will point to
**Omitted Variable Bias** caused by ShelveLoc.

We will also revisit the regression output table and ask: since we are working with
a population, what do "estimate", "std_error", and "p_value" actually mean here?
:::

## References {.unnumbered}
