---
title: "Session 1 · Lab 1: R Projects & Quarto"
subtitle: "Project Structure, Data Import, and First Presentation"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: false
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(fontawesome)
library(openintro)
library(gt)
data("ames", package = "openintro", envir = .GlobalEnv)
```

## Lab 1 · Overview

### Learning Objectives

- Create an `r fa('r-project')` project with a clean folder structure

- Import data from packages and local files

- Learn initial exploratory commands (`dim()`, `head()`, `glimpse()`)

- Create and render a Quarto presentation

- Use inline `r fa('r-project')` code in Quarto

### Today's Roadmap

| Phase      | Content                                        |
|------------|------------------------------------------------|
| 🔵 Concept | Project structure & data import                |
| 🔵 Demo    | Explore `ames`, Quarto & inline code           |
| 🟡 Task    | Import your own datasets                       |
| 🟢 Task    | Create Quarto slides                           |

<br> ⏱️ **Approx. 30 minutes** · 👤 **Individual work**

::: {.callout-note appearance="minimal"}
**Prerequisites:** `r fa('r-project')`, RStudio, and the packages `tidyverse`, `openintro`, and `gt` are installed.
:::

## Step 1 — Unzip the Template & Rename

### Instructions

1.  Download `R-Project_Template.zip`\
    ([Moodle link](https://moodle.ffhs.ch/pluginfile.php/5752746/mod_assign/introattachment/0/R-Project_Template.zip?forcedownload=1))
2.  Unzip the archive
3.  Rename the folder to `ER014-PVA1_XY`\
    **X** = first letter of your last name\
    **Y** = first letter of your first name\
    *(Example: Maria Müller → `ER014-PVA1_MM`)*
4.  Open the folder and check the structure

::: {.callout-warning appearance="minimal"}
No spaces or special characters in the folder name — this avoids path problems across all operating systems!
:::

### Expected Folder Structure

``` text
ER014-PVA1_XY/
├── data/
│   ├── raw/        ← Raw data (never modify!)
│   └── tidy/       ← Cleaned data (tidy format)
├── R/              ← R scripts
├── qmd/            ← Quarto files
├── figs/           ← Figures (created with R)
├── img/            ← Image files (from web or graphics software)
├── lit/            ← Literature (PDFs etc.) and bibliography (.bib, .json etc.)
└── README.md       ← Project documentation (in the root directory)
```

💡 **Core principle:** Raw data always remains unchanged.\
Everything else can be regenerated from raw data at any time.

## Step 2 — Create an RStudio Project

::: panel-tabset
### Instructions

1.  Open RStudio
2.  **File → New Project → Existing Directory**
3.  Navigate to your folder `ER014-PVA1_XY`
4.  Click **Create Project**
5.  RStudio restarts — the project name appears in the top right

✅ **Success criterion:** The project name appears in the top right of RStudio.

### Screenshot Note

``` text
┌─────────────────────────────────────────┐
│ File  Edit  Code  View  ...             │
├─────────────────────────────────────────┤
│                    [ER014-PVA1_MM ▼]    │  ← Project name
│                                         │
│  Environment  History  ...              │
│  ─────────────────────────              │
│                                         │
└─────────────────────────────────────────┘
```

### What Happens in the Background?

RStudio creates an `.Rproj` file in the folder:

``` text
ER014-PVA1_XY/
├── ER014-PVA1_XY.Rproj   ← NEW
├── data/
│   ├── raw/
│   └── tidy/
...
```

**Why does this matter?**

When opening the `.Rproj` file, RStudio automatically sets the working directory to this folder. Relative paths like `"data/raw/file.csv"` therefore work on **any computer** without `setwd()` — this is the foundation of reproducible analyses.
:::

## Step 3 — Importing Data: The Concept

`r fa('r-project')` can load data from many sources. Here is an overview of the most important approaches:

### Way 1: Data from a Package

Many `r fa('r-project')` packages include datasets directly — **no file needed**.

```{r}
#| label: demo-import-package
# Load package
library(openintro)
# Load dataset into the global environment
data(ames)
```

::: {.callout-note appearance="minimal"}
**Installing ≠ Loading**\
`install.packages()` downloads the package **once** from the internet.\
`library()` activates it for the **current `r fa('r-project')` session** — repeat after every restart.\
`install.packages()` does **not** belong in a script — only run it once in the console.
:::

### Way 2: Local File

```{r}
#| eval: false
# CSV (semicolon-separated, e.g. German Excel export)
tbl_autos <- read_csv2(xfun::from_root(
  "data/raw/autos.csv"))

# R native binary format
tbl_buli <- read_rds(xfun::from_root(
  "data/raw/buli.rds"))
```

::: {.callout-note appearance="minimal"}
**`read_csv` vs. `read_csv2`**\
`,` as separator → `read_csv()` · `;` as separator → `read_csv2()`\
If in doubt: open the file in a text editor and check.\
Wrong separator → all values land in **one** column!
:::

### Way 3: File Directly from the Web

```{r}
#| eval: false
# Pass URL directly
url <- "https://example.org/data.csv"
tbl <- read_csv(url)

# Download and unzip an archive
tmp <- tempfile(fileext = ".zip")
download.file("https://example.org/data.zip", tmp)
tbl <- read_csv(unzip(tmp, exdir = tempdir()))
```

Useful for publicly available datasets that are regularly updated.

### Way 4: API (Application Programming Interface)

Many institutions provide APIs — `r fa('r-project')` fetches data directly without manual downloading:

```{r}
#| eval: false
# Example: Data from the Swiss Federal Statistical Office
library(BFS)

# Search catalogue and retrieve dataset
bfs_get_data(number_bfs = "px-x-0102010000_101",
             language = "de")
```

::: {.callout-note appearance="minimal"}
**Outlook:** Web scraping and APIs will be used practically in **ER015** and **ER018**.\
Curious? → [`BFS` package on CRAN](https://cran.r-project.org/package=BFS) · [Documentation](https://felixluginbuhl.com/BFS/)
:::

## Step 3 — Importing Data: Reproducibility

### Why in a Script or Quarto Chunk?

Importing data by clicking (e.g. via *Import Dataset* in RStudio) is tempting — but **not reproducible**: no one else can trace what steps were taken.

✅ **Import code always belongs in an `r fa('r-project')` script or a Quarto chunk.**\
That way, anyone — including your future self — can rerun the analysis from scratch.

::: {.callout-tip appearance="minimal"}
**Rule of thumb:** If you click something, document it as code afterwards.\
If you have to make the same click twice, write it as a function.
:::

### Paths: The Most Common Stumbling Block

```{r}
#| eval: false
# ❌ Absolute path — only works on THIS machine
read_csv("C:/Users/Maria/Desktop/autos.csv")

# ❌ Relative path without a project — depends on
#    the current working directory
read_csv("data/raw/autos.csv")   # only works if
                                  # wd is set correctly

# ✅ With xfun::from_root() — always relative to
#    the .Rproj file, regardless of the subfolder
library(xfun)
read_csv2(from_root("data/raw/autos.csv"))
```

::: {.callout-note appearance="minimal"}
`xfun::from_root()` searches for the `.Rproj` file in the current or parent directory and builds the path from there. This also works when the script is run from `R/` or the `.qmd` from `qmd/`.
:::

## Step 3 — Download Files

Download these two files and save them in `data/raw/`:

### Files

1.  `autos_(StockerUIBK)_20240414.csv`\
    ([Moodle link](https://moodle.ffhs.ch/pluginfile.php/5752753/mod_assign/introattachment/0/autos_%28StockerUIBK%29_20240414.csv?forcedownload=1))

2.  `buli_raw_2023-09-06.rds`\
    ([Moodle link](https://moodle.ffhs.ch/pluginfile.php/5752753/mod_assign/introattachment/0/buli_raw_2023-09-06.rds?forcedownload=1))

::: {.callout-warning appearance="minimal"}
Do **not** rename the files — the code references the original names exactly.
:::

### Verification

``` text
data/raw/
├── autos_(StockerUIBK)_20240414.csv
└── buli_raw_2023-09-06.rds
```

### Three Datasets — Three Import Methods

| Dataset     | Import method       | Function                      |
|-------------|---------------------|-------------------------------|
| `ames`      | Package `openintro` | `data(..., envir=.GlobalEnv)` |
| `tbl_autos` | File `.csv`         | `read_csv2()`                 |
| `tbl_buli`  | File `.rds`         | `read_rds()`                  |

<br>

| Format | Separator        | Note                          |
|--------|------------------|-------------------------------|
| `.csv` | `;` (semicolon)  | Text format, readable anywhere|
| `.rds` | —                | Compressed, `r fa('r-project')` only |

## Step 4 — Create an `r fa('r-project')` Script

**File → New File → R Script** → save as `R/EDA_IMS.R`

```{r}
#| eval: false
# ============================================
# EDA_IMS.R  —  Exploratory Data Analysis
# PVA1 · Lab 1
# ============================================

# 1. Load packages --------------------------------
#    (Packages must be installed once first:
#     install.packages(c("tidyverse","openintro","gt","xfun")))
library(tidyverse)
library(openintro)
library(gt)

# 2. Load package data ----------------------------
data("ames", package = "openintro", envir = .GlobalEnv)

# 3. Import external files -----------------------
my_in_file <- "data/raw/autos_(StockerUIBK)_20240414.csv"
tbl_autos  <- read_csv2(xfun::from_root(my_in_file))

my_in_file <- "data/raw/buli_raw_2023-09-06.rds"
tbl_buli   <- read_rds(xfun::from_root(my_in_file))
```

::: {.callout-tip appearance="minimal"}
**Running code:** Cursor on a line → **Ctrl+Enter** (Win) / **Cmd+Enter** (Mac) runs that line.\
Select multiple lines → **Ctrl+Enter** runs all selected lines.\
After import, the objects appear in the **Environment** panel in the top right.
:::

## 🔵 Demo — Exploring `ames`

:::: panel-tabset
### dim() / nrow() / ncol()

```{r}
dim(ames)    # Result: [rows, columns]
nrow(ames)   # rows only
ncol(ames)   # columns only
```

❓ What do these numbers mean in context?\
→ *2930 rows = 2930 houses sold; 82 columns = 82 variables (price, area, location …)*

### head()

```{r}
head(ames)   # first 6 rows (default)
```

❓ `head(ames, n = 10)` — what changes?

### glimpse()

```{r}
glimpse(ames)
```

::: {.callout-note appearance="minimal"}
**`head()` vs. `glimpse()`**\
`head()` shows the first rows as a table — practical with few columns.\
`glimpse()` flips the perspective: each **variable** becomes a row, including its data type (`<fct>`, `<int>`, `<dbl>`). Ideal for many columns or for a quick check of data types.
:::
::::

## 🔵 Demo — Understanding Quarto

### Why Quarto?

Quarto combines **`r fa('r-project')` code, results, and text** in a single document. Instead of manually copying tables and figures into a Word document, they are generated directly during rendering — automatically, reproducibly, and always up to date. The same `.qmd` source document can be output in various formats:

| Format                   | Use case                              |
|--------------------------|---------------------------------------|
| HTML                     | Interactive reports, websites         |
| PDF                      | Submissions, publications             |
| Word (`.docx`)           | Collaborative editing                 |
| Presentation (Reveal.js) | Talks, these labs                     |



### The Three Building Blocks of a Quarto File

A `.qmd` file always consists of the same three elements:

* YAML header (metadata & output format)
* Code (`r fa('r-project')`, Python, Julia …)
* Text and formatting (Markdown)


#### YAML Header
```         
---                        ← YAML header starts with ---
title: "My Report"
format: html
---                        ← and ends with ---
```

#### Code

`r fa('r-project')` code can be embedded as **code chunks** or as **inline code**.

* Code chunks: Insert with **Ctrl+Alt+I** (Win) or via *Insert → R*:

[` ```{r} `]{.verbatim} R code [` ``` `]{.verbatim}


::: {.callout-note appearance="minimal"}
**Code chunk options:**
Whether code chunks appear **in the output document** (usually not desired!) can be controlled with chunk options. They are typically set with `#|` (hash-pipe) directly above the code.
:::

To show both the result and the code block in the final document, use `#| echo: true` at the beginning of the chunk:

```{r}
#| echo: true # (= default — does not need to be stated explicitly)
4 + 3
```

With `#| echo: false`, only the result is shown, the code is hidden — useful for presentations:

```{r}
#| echo: false
4 + 3
```


With `#| eval: false`, the code is displayed but not executed — useful for example code:

```{r}
#| eval: false
4 + 3
```

* Inline code (→ see [Demo section](@inline)).


#### Text and Formatting

Quarto uses **Markdown** for text formatting: `**bold**`, `*italic*`, `# Heading`, `- List item`. A compact reference: [Quarto Markdown Basics](https://quarto.org/docs/authoring/markdown-basics.html).

### Excursus: Tables in Quarto Documents

For structured data tables in the output document, the `gt` package works well:

```{r}
ames %>%
  select(area, price, Neighborhood, Bedroom.AbvGr, Lot.Area) %>%
  head(5) %>%
  gt() %>%
  tab_header(
    title    = "Ames Housing — first 5 rows (selection)",
    subtitle = "Source: openintro package"
  )
```

::: {.callout-note appearance="minimal"}
**Review from the preparatory material: tidyverse and the pipe**\
The **magrittr** package introduced the idea of "piping" to `r fa('r-project')` in 2013. Since 2014, the **magrittr** operator `%>%` has been implemented in the **dplyr** package and can be used directly by loading the **tidyverse** meta-package.

The pipe operator `%>%` makes it easy to chain a series of operations:

`x %>% f()` means: take `x` and pass it as the **first** argument to `f()`.

Due to the enormous success of the tidyverse, a native pipe was also added to base `r fa('r-project')` in version 4.1.0. This **native pipe** `|>` works identically for most use cases.

`r fa('exclamation-circle')` We use `%>%` from `magrittr` because it offers a wider range of functionality, for example when used with the `lm()` function.
:::

## 🔵 Demo — Understanding Inline `r fa('r-project')` Code  {#inline} 

### Syntax

Inline `r fa('r-project')` code is introduced with a backtick (&#96;), the letter `r`, and a space — immediately before the `r fa('r-project')` expression — and closed with a backtick (&#96;). Quarto evaluates the expression during rendering and inserts the result directly into the running text.

### Before: Hard-Coded Numbers

In the source, numbers are typed in manually:

``` markdown
The Ames dataset contains 2930 rows and 82 columns.
```

Rendered result — numbers are fixed/immutable text in the document:

> The Ames dataset contains 2930 rows and 82 columns.

❌ If the data changes, the text is no longer correct — and you often don't notice. Also prone to typos.

### After: Inline `r fa('r-project')` Code

In the source, `r fa('r-project')` expressions replace the hard-coded numbers:

The Ames dataset contains &#96;r nrow(ames)&#96; rows and &#96;r ncol(ames)&#96; columns.

Rendered result — Quarto evaluates the expressions and inserts the values:

> The Ames dataset contains **`r nrow(ames)`** rows and **`r ncol(ames)`** columns.

✅ The numbers update automatically — computed from the data at every render. No more typos either!

### Further Examples

| Source                               | Rendered result                |
|--------------------------------------|--------------------------------|
| `` `r nrow(ames)` ``                 | `r nrow(ames)`                 |
| `` `r ncol(ames)` ``                 | `r ncol(ames)`                 |
| `` `r round(mean(ames$price), 0)` `` | `r round(mean(ames$price), 0)` |
| `` `r Sys.Date()` ``                 | `r Sys.Date()`                 |

::: {.callout-tip appearance="minimal"}
**When inline code, when a chunk?**\
Inline code is suited for **single values** in running text (count, mean, date).\
For tables and figures → code chunk.
:::

## 🟡 Task — Import & Explore Datasets

Run the import commands for `tbl_autos` and `tbl_buli` in `R/EDA_IMS.R` and explore the datasets **analogously to the `ames` demo**:

### A: Check Dimensions

```{r}
#| eval: false
dim(tbl_autos)
nrow(tbl_autos)
ncol(tbl_autos)

dim(tbl_buli)
nrow(tbl_buli)
ncol(tbl_buli)
```

❓ What does one row represent in each case?\
*(One car? One match? One matchday?)*

### B: Explore Structure

```{r}
#| eval: false
glimpse(tbl_autos)
glimpse(tbl_buli)
```

❓ What variable types appear?\
*(`<chr>`, `<dbl>`, `<date>`, `<fct>` …)*

### C: Display First Rows

```{r}
#| eval: false
head(tbl_autos)
head(tbl_buli)
```

### D: Document Findings

Record your observations as comments in the script:

```{r}
#| eval: false
# tbl_autos: ___ rows, ___ columns
# One row = one ________________
# Variable types: ___________________

# tbl_buli:  ___ rows, ___ columns
# One row = one ________________
# Variable types: ___________________
```

::: {.callout-tip appearance="minimal"}
These notes form the basis for the Quarto slides in the next step.
:::

## 🟢 Task — Create a Quarto Presentation

::::::: panel-tabset
### 1. Create File

1.  **File → New File → Quarto Presentation → Reveal.js**
2.  Save as `qmd/DataOverview.qmd`
3.  Delete all example content
4.  Insert YAML header:

``` yaml
---
title: "Data Overview"
subtitle: "PVA1 · Lab 1"
author: "Your Name"
date: today
format:
  revealjs:
    theme: default
    slide-number: true
execute:
  echo: true
  warning: false
---
```

### 2. Setup Chunk

Directly after the YAML header — loads all objects **invisibly**:

```` markdown
```{{r}}
#| label: setup
#| include: false
library(tidyverse)
library(openintro)
library(gt)
data("ames", package = "openintro", envir = .GlobalEnv)

my_in_file <- "data/raw/autos_(StockerUIBK)_20240414.csv"
tbl_autos  <- read_csv2(xfun::from_root(my_in_file))

my_in_file <- "data/raw/buli_raw_2023-09-06.rds"
tbl_buli   <- read_rds(xfun::from_root(my_in_file))
```
````

::: {.callout-important appearance="minimal"}
`#| include: false` — code runs during rendering but does **not** appear in the rendered document.\
All three objects are then available on every slide.
:::

### 3. Slide: Cars

Refer to the `ames` demo:

```` markdown
## Cars Dataset

The dataset contains **`nrow(tbl_autos)`** observations
and **`ncol(tbl_autos)`** variables.

```{{r}}
tbl_autos %>%
  head(5) %>%
  gt() %>%
  tab_header(title = "First 5 rows: Cars")
```
````

::: {.callout-note appearance="minimal"}
`` `{{r}} nrow(tbl_autos)` `` works here because `tbl_autos` was loaded in the setup chunk — just like `ames` in the demo.
:::

### 4. Slide: Bundesliga

```` markdown
## Bundesliga Dataset

The dataset contains **`nrow(tbl_buli)`** observations
and **`ncol(tbl_buli)`** variables.

```{{r}}
tbl_buli %>%
  head(5) %>%
  select(1:6) %>%
  gt() %>%
  tab_header(title = "First 5 rows: Bundesliga (selection)")
```
````

::: {.callout-note appearance="minimal"}
`select(1:6)` selects the first 6 columns — for wide datasets this prevents overfull slides.
:::

### 5. Slide: Comparison

``` markdown
## Dataset Overview

| Dataset    | Rows                | Columns             |
|------------|---------------------|---------------------|
| ames       | `nrow(ames)`        | `ncol(ames)`        |
| Cars       | `nrow(tbl_autos)` | `ncol(tbl_autos)` |
| Bundesliga | `nrow(tbl_buli)`  | `ncol(tbl_buli)`  |

Values are inserted automatically during rendering.
```

### 6. Render & Check

1.  **Ctrl+S** — save file
2.  **Render** (or **Ctrl+Shift+K**)
3.  Presentation opens in the browser

✅ **Success criterion:** All three slides show correct dimensions and tables.

::: {.callout-tip appearance="minimal"}
Render error? → Read the line number in the error message, isolate the affected chunk in the console and test it.
:::
:::::::

## Bonus Tasks

### B1: Variable Types as a Table

Create a slide with variable names and data types:

```{r}
#| eval: false
tibble(
  Variable = names(tbl_autos),
  Type     = map_chr(tbl_autos, class)
) %>%
  gt() %>%
  tab_header(title = "Variables: Cars")
```

### B2: Missing Values

Show only columns with at least one NA:

```{r}
#| eval: false
tbl_autos %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(),
               names_to  = "Variable",
               values_to = "Missing") %>%
  filter(Missing > 0) %>%
  gt() %>%
  tab_header(title = "Missing values: Cars")
```

### B3: Visualisation with `ames`

```{r}
#| eval: false
ames %>%
  ggplot(aes(x = area, y = price)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  scale_y_continuous(labels = scales::dollar) +
  labs(
    title    = "Living Area vs. Sale Price",
    subtitle = "Ames, Iowa 2006–2010",
    x        = "Living area (sqft)",
    y        = "Sale price (USD)"
  ) +
  theme_minimal()
```

### B4: Change the Presentation Theme

Experiment with the `theme` parameter in the YAML of `DataOverview.qmd`:

``` yaml
format:
  revealjs:
    theme: moon
```

Options: `sky` · `beige` · `serif` · `simple` · `night` · `moon`

❓ Which theme looks appropriate for an academic presentation?

## Common Errors & Solutions

::: callout-warning
#### ❌ File Not Found

```         
Error: cannot open file 'data/raw/...'
```

**Cause:** No `r fa('r-project')` project open, or a typo in the path.

**Solution:**\
1. Is the project name visible in the top right of RStudio?\
2. Is the file really in `data/raw/`?\
3. Use `xfun::from_root()`
:::

::: callout-warning
#### ❌ All Data in One Column

Dataset has only one column, although several are expected.

**Cause:** Wrong separator — `read_csv()` instead of `read_csv2()` (or vice versa).

**Solution:** Open the file in a text editor and check: comma or semicolon as separator?
:::

::: callout-warning
#### ❌ Package Not Installed

```         
Error: there is no package called 'gt'
```

**Solution:** Install once, then load:

``` r
install.packages("gt")
library(gt)
```

Never write `install.packages()` in a script — only run it in the console!
:::

::: callout-warning
#### ❌ Object Not Found

```         
Error: object 'tbl_autos' not found
```

**Cause:** Import code was not executed.

**Solution:**\
- In the script: run import lines first\
- In Quarto: is the setup chunk with `#| include: false` present?\
- Did you forget `data("ames", package = "openintro", envir = .GlobalEnv)` after `library(openintro)`?
:::

## Lab 1 · Summary

::: callout-tip
#### What We Have Learned

✅ R projects guarantee reproducible, machine-independent paths

✅ **Two import methods:** `data(..., envir=.GlobalEnv)` for package data · `read_csv2()`/`read_rds()` for files

✅ `xfun::from_root()` — paths always relative to the `.Rproj` file

✅ `dim()`, `head()`, `glimpse()` for a first overview of any dataset

✅ `%>%` — the magrittr pipe chains analysis steps readably

✅ Inline R code keeps documents automatically consistent
:::

::: callout-important
#### Checklist Before Submission

☐ Folder named correctly (`ER014-PVA1_XY`)

☐ `.Rproj` file present and open

☐ Both files in `data/raw/` (unchanged)

☐ Script `R/EDA_IMS.R` runs **without errors** from top to bottom

☐ `qmd/DataOverview.qmd` renders without errors

☐ All three datasets shown in the presentation with correct dimensions

☐ Comparison slide: inline code shows correct values
:::

## Appendix: Integrating Reference Management Software

Various reference management tools are available, including free options. This brief introduction focuses on Zotero, a free application available for Windows, macOS, and Linux. Zotero integrates with web browsers (as an extension), Microsoft Word, and RStudio (in combination with Quarto/RMarkdown).

If you already use other software (e.g. Mendeley, Citavi), you have two options:

* Integrate that software into RStudio (guides are available online), or
* Migrate your library to Zotero.

### Installing Zotero and Integrating with RStudio

#### Installing Zotero and Useful Add-ons

Steps to install Zotero:

* Download the installer from the [Zotero website](www.zotero.org)
* Create a free account

Useful add-ons:

* Install **Better BibTeX for Zotero** in the Zotero desktop app. Follow the instructions on the [BetterBibTeX website](https://retorque.re/zotero-better-bibtex/installation/).
* Install the [Zotero Connector](http://www.zotero.org/download/connectors) for your browser (Chrome, Edge, Safari, or Firefox) to import references directly from the browser into your library.

* Note: Zotero also integrates directly with MS Word. After installing Zotero, a "Zotero" tab appears automatically in Word.

#### Integration with RStudio

To embed Zotero in RStudio:

1. Run the following command in the console:

> `remotes::install_github("paleolimbot/rbbt")`

2. After installation, close and restart RStudio.

3. Zotero should now appear in the RStudio Addins menu.

Always make sure all files (.R, .qmd, etc.) are created with UTF-8 encoding to avoid compatibility issues.


### Creating a Bibliography in Quarto/RMarkdown {#zotero}

* First, create a bibliography file. In Zotero:

> `File > Export Library`

* Choose the format CSL JSON (other formats such as BibLaTeX are also supported, but CSL JSON is increasingly standard)

* Save the file in the `lit/` folder of your R project, e.g. as `my_bib.json`

* Add the following two lines to the YAML header to automatically generate the bibliography:

> `bibliography: my_bib.json`\
> `reference-section-title: References`

Note: `reference-section-title` specifies the heading for the bibliography section (here: "References"). Alternatives include "Literature" or "Sources".

If the output is in MS Word format, a Word template can be created to apply the correct formatting automatically. Guidance is available on the [RStudio website](https://rmarkdown.rstudio.com/articles_docx.html).

## Resources

📖 [r4ds.hadley.nz](https://r4ds.hadley.nz) — Ch. 6 & 7\
*(Workflow: scripts & projects)*

📖 [quarto.org/docs/output-formats](https://quarto.org/docs/output-formats/html-basics.html)\
*(HTML documents in Quarto)*

📖 [xfun documentation](https://yihui.org/xfun/)\
*(from_root())*

📺 RStudio: *"Project-oriented workflow"* (Webinar)

<br> **Questions? → Lab 1 Debrief**
