<!-- ============================================================
INSTRUCTOR CUSTOMISATION — update before rendering for each cohort
Search for the marker text rather than relying on exact line numbers.

  "buli_raw_{date}.rds"   Dataset filename — update date if new scrape
  xfun::from_root(...)    Adjust path if project structure differs
  set.seed(42)            Change seed for different cohort results
============================================================ -->
---
title: "Session 2 · Lab 3: p-Hacking & False Discovery Rate"
subtitle: "Understanding and Controlling the Multiplicity Problem"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse)    # data wrangling & plotting
library(moderndive)   # get_regression_table()
library(infer)        # resampling framework
library(gt)           # formatted tables

date <- '2023-09-06'
tbl_buli_raw <- read_rds(
  file = xfun::from_root("data","raw",
                         glue::glue('buli_raw_{date}.rds')))

tbl_buli <- tbl_buli_raw %>%
              mutate(
                age       = lubridate::year(Sys.Date()) -
                            lubridate::year(player_dob),
                log_value = log(player_market_value_euro),
                pos_bin   = if_else(player_position == "Goalkeeper",
                                    "Goalkeeper", "Outfield")
              ) %>%
              rename(name = player_name,
                     mv   = player_market_value_euro) %>%
              filter(!is.na(mv), !is.na(age))

options(scipen = 999)
```

## Scenario & Orientation

In Labs 1 and 2 we tested **one** hypothesis — "does age predict market value?" — and found a significant result. But what happens when a data scientist tests **hundreds** of variables against the same outcome? Some will appear significant purely by chance. This lab makes that problem viscerally real through simulation.

```{r crisp-overview}
#| echo: false
tibble::tribble(
  ~Phase, ~Name,                    ~Content,                                ~Status,
  "1",    "Business Understanding", "Identify under-/overvalued players",    "Done (Session 1)",
  "2",    "Data Understanding",     "EDA of Bundesliga market values",       "Done (Session 1)",
  "3",    "Data Preparation",       "log-transform, position dummy",         "Done (Session 1)",
  "4",    "Modelling",              "OLS: log_value ~ age",                  "Done (Session 1)",
  "5",    "Evaluation",             "Multiple testing & false discoveries",  "← This lab",
  "6",    "Deployment",             "Transfer decision framework",           "Session 2 lecture"
) %>%
  gt() %>%
  tab_style(cell_fill("#f0fff0"), cells_body(rows = Phase == "5")) %>%
  tab_style(cell_fill("#f5f0fa"), cells_column_labels()) %>%
  tab_options(table.font.size = "13px")
```

**Navigation guide**

| Symbol | Box colour | Meaning | When to read |
|--------|-----------|---------|--------------|
| ⚠️ | Red | Mandatory task | Live session |
| 💡 | Green | Tip / best practice | Anytime |
| 📚 | Blue (collapsed) | Background theory | At home |
| ℹ️ | Blue (open) | Short clarification | When needed |

:::{.callout-tip}
## ⏱ Time Budget
**25 minutes** — complete mandatory tasks (A1–A3) first; bonus tasks if time permits.
:::

---

## Icebreaker: The Jellybean Problem

Imagine a lab tests 20 jellybean colours for a link to acne. At α = 0.05, how many colours do you **expect** to appear significant — even if jellybeans have nothing to do with acne?

$$20 \times 0.05 = 1$$

One colour will cross the significance threshold by pure chance. If the lab publishes only *that* result — "Green jellybeans linked to acne!" — the finding looks real. But it is noise.

:::{.callout-important}
## The Core Problem
At α = 0.05, **every 20th test** produces a false positive. With 100 tests, you get ~5 false discoveries. With 1,000 tests, you get ~50. This is not a bug — it is the mathematical consequence of the α-level. **This motivates everything we do in this lab.**
:::

We will now simulate this exact problem with the Bundesliga data.

---

## Step 1: Simulating p-Hacking

We generate 100 completely random variables — with no real relationship to market value — and test each one individually against `log_value`.

```{r phacking}
#| code-fold: false
#| cache: true
set.seed(42)
n_tests <- 100
n_obs   <- nrow(tbl_buli)

p_values_random <- map_dbl(1:n_tests, function(i) {
  tbl_sim <- tbl_buli %>%
               mutate(rand_var = rnorm(n_obs))
  lm(log_value ~ rand_var, data = tbl_sim) %>%
    get_regression_table() %>%
    filter(term == "rand_var") %>%
    pull(p_value)
})

# result
n_significant <- sum(p_values_random < 0.05)
cat("Significant at α = 0.05:", n_significant, "of", n_tests, "tests\n")
cat("Expected (theoretical):  ", n_tests * 0.05)
```

`r n_significant` tests are significant — despite **no real effect** in any of them.

```{r phacking-viz}
tibble(
  test_nr     = 1:n_tests,
  p_value     = p_values_random,
  significant = p_values_random < 0.05
) %>%
  ggplot(aes(x = p_value, fill = significant)) +
  geom_histogram(bins = 20, color = "white", boundary = 0) +
  geom_vline(xintercept = 0.05, color = "#D50006",
             linetype = "dashed", linewidth = 1.2) +
  annotate("text", x = 0.07, y = Inf, label = "α = 0.05",
           hjust = 0, vjust = 1.5, color = "#D50006", size = 3.5) +
  scale_fill_manual(
    values = c("TRUE"  = "#D50006", "FALSE" = "#b3b2b2"),
    labels = c("TRUE"  = glue::glue("p < 0.05 ({n_significant} falsely significant!)"),
               "FALSE" = "p ≥ 0.05 (correctly not significant)")
  ) +
  labs(
    x        = "p-value",
    y        = "Number of tests",
    fill     = NULL,
    title    = glue::glue("100 tests with random variables — {n_significant} are p < 0.05"),
    subtitle = "All variables are completely random — no real relationship exists"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

:::{.callout-note collapse="true"}
## 📚 Background: Why Is the p-Value Distribution Uniform?
Under $H_0$ (no real effect), p-values are **uniformly distributed** on [0, 1]. This means: in any interval of width 0.05, about 5% of all p-values fall — including the interval [0, 0.05]. That is why α = 0.05 produces exactly 5% false positives by design.
:::

---

## Step 2: The Real Problem in the Bundesliga Dataset

Now we test all numeric variables in the actual dataset against `log_value` — without any content-driven hypotheses. This is data mining in its rawest form.

```{r many-tests}
#| cache: true
# prepare dataset for testing
tbl_test <- tbl_buli_raw %>%
              mutate(
                log_value = log(player_market_value_euro),
                age       = lubridate::year(Sys.Date()) -
                            lubridate::year(player_dob)
              ) %>%
              select(log_value, where(is.numeric),
                     -player_market_value_euro)

# test all variables against log_value
p_values_real <- tbl_test %>%
  select(-log_value) %>%
  names() %>%
  map_dfr(function(var) {
    tryCatch({
      p <- lm(as.formula(glue::glue("log_value ~ `{var}`")),
              data = tbl_test) %>%
            broom::tidy() %>%
            filter(term != "(Intercept)") %>%
            pull(p.value)
      tibble(variable = var, p_value = p[1])
    }, error = function(e) {
      tibble(variable = var, p_value = NA_real_)
    })
  }) %>%
  filter(!is.na(p_value)) %>%
  arrange(p_value)

# top 10 — apparently significant variables
p_values_real %>%
  head(10) %>%
  gt() %>%
  fmt_number(columns = p_value, decimals = 4) %>%
  tab_style(
    style     = cell_fill(color = "#fff0f0"),
    locations = cells_body(rows = p_value < 0.05)
  ) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("Red = p < 0.05 without correction")
```

`r sum(p_values_real$p_value < 0.05, na.rm=TRUE)` of `r nrow(p_values_real)` variables are p < 0.05 without correction. But how many are real effects versus noise?

---

## Step 3: Benjamini-Hochberg — Controlling the FDR

Instead of controlling each individual α (**FWER**, e.g. Bonferroni), the **False Discovery Rate** controls the *proportion* of false discoveries among all discoveries:

$$\text{FDR} = \frac{\text{False Discoveries}}{\text{All Discoveries}}$$

The **Benjamini-Hochberg procedure** works as follows:

1. Sort p-values in ascending order: $p_{(1)} \leq p_{(2)} \leq \cdots \leq p_{(m)}$
2. For desired FDR level $q$: find the largest $k$ with $p_{(k)} \leq \frac{k}{m} \cdot q$
3. Declare tests $1, \ldots, k$ as significant

In R: `p.adjust(p_values, method = "BH")`

```{r bh}
#| code-fold: false
p_values_adjusted <- p_values_real %>%
  mutate(
    p_bonferroni   = p.adjust(p_value, method = "bonferroni"),
    p_BH           = p.adjust(p_value, method = "BH"),
    sig_raw        = p_value      < 0.05,
    sig_bonferroni = p_bonferroni < 0.05,
    sig_BH         = p_BH         < 0.05
  )

# comparison of methods
p_values_adjusted %>%
  summarise(
    `No correction`   = sum(sig_raw,        na.rm = TRUE),
    `Bonferroni`       = sum(sig_bonferroni, na.rm = TRUE),
    `BH (FDR)`         = sum(sig_BH,         na.rm = TRUE)
  ) %>%
  pivot_longer(everything(),
               names_to  = "Method",
               values_to = "Significant") %>%
  gt() %>%
  tab_style(
    style     = cell_fill(color = "#f0fff0"),
    locations = cells_body(rows = Method == "BH (FDR)")
  ) %>%
  tab_options(table.font.size = "14px") %>%
  tab_footnote("Green = recommended method for many simultaneous tests")
```

```{r bh-viz}
p_values_adjusted %>%
  slice_head(n = 20) %>%
  mutate(variable = fct_reorder(variable, p_value)) %>%
  ggplot(aes(y = variable)) +
  geom_point(aes(x = p_value,      color = "Raw"),
             size = 3, alpha = 0.8) +
  geom_point(aes(x = p_BH,         color = "BH-adjusted"),
             size = 3, alpha = 0.8, shape = 17) +
  geom_point(aes(x = p_bonferroni, color = "Bonferroni"),
             size = 3, alpha = 0.8, shape = 15) +
  geom_vline(xintercept = 0.05, linetype = "dashed",
             color = "#686868", linewidth = 0.8) +
  scale_color_manual(
    values = c("Raw"         = "#D50006",
               "BH-adjusted" = "#502479",
               "Bonferroni"  = "#686868")
  ) +
  labs(
    x        = "p-value",
    y        = "Variable (Top 20)",
    color    = NULL,
    title    = "Raw vs. corrected significance levels",
    subtitle = "Dashed line: α = 0.05 | Triangle = BH | Square = Bonferroni"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

:::{.callout-note}
## When to Use Which Method?

| Situation | Method | Rationale |
|-----------|--------|-----------|
| 1 test, clear hypothesis | α = 0.05 classical | No multiplicity problem |
| Few tests (5–20) | **Bonferroni** | Conservative, controls FWER |
| Many tests (>50) | **BH (FDR)** | Less conservative, more power |
| Data mining, exploratory | **BH (FDR)** | Standard in data science |
:::

---

## Mandatory Tasks

:::{.callout-important icon=false}
## ⚠️ A1: Scaling the Simulation

Run the simulation with `n_tests = 1000`.

1. How many of 1,000 random tests are significant?
2. Does this match the theoretical expectation ($1000 \times 0.05 = 50$)?
3. What would happen if a researcher ran these 1,000 tests and published only the significant ones?
:::

:::{.callout-important icon=false}
## ⚠️ A2: Bonferroni vs. BH

Interpret the comparison of the three methods in the table above.

1. Which variables remain significant after BH correction?
2. Do these variables make substantive sense as market-value predictors?
3. Why is Bonferroni more conservative than BH here?
:::

:::{.callout-important icon=false}
## ⚠️ A3: Break-Even Formula

In the lecture we showed:

$$p_{\text{A}} > \frac{c(\text{FP})}{c(\text{FP}) + c(\text{FN})}$$

Suppose the FDR correction increases your confidence that a significant finding is real. How does this change your estimated $p_{\text{A}}$ for a Bundesliga transfer? Is the purchase more worthwhile with a BH-corrected result?
:::

---

## Bonus Tasks

:::{.callout-tip icon=false}
## 💡 B1: Reproduce p-Hacking

Simulate the classic p-hacking scenario:

```r
# researcher tests 20 variables and reports only significant ones
set.seed(1)
results <- map_dfr(1:20, function(i) {
  tbl_sim <- tbl_buli %>% mutate(x = rnorm(n()))
  p <- lm(log_value ~ x, data = tbl_sim) %>%
         broom::tidy() %>%
         filter(term == "x") %>%
         pull(p.value)
  tibble(test = i, p_value = p, significant = p < 0.05)
})
filter(results, significant)
```

How many "discoveries" get published? What would be the correct conclusion?
:::

:::{.callout-tip icon=false}
## 💡 B2: xkcd Significant

Read: [xkcd: Significant](https://xkcd.com/882/)

Explain the joke with statistical precision — and name which p-hacking practice from this lab is depicted.
:::

:::{.callout-tip icon=false}
## 💡 B3: FDR vs. FWER

Research the difference between **FDR** (False Discovery Rate) and **FWER** (Family-Wise Error Rate).

- In which context is FWER more important than FDR?
- Give one medical and one data-science example where the choice of method would lead to a different decision.
:::

---

## Summary

```{r summary-table}
#| echo: false
tibble::tribble(
  ~`Finding`,                  ~`Key Result`,                              ~`Implication for Session 2 Lecture`,
  "Multiplicity problem",      "5% of random tests are falsely significant", "Cannot trust raw p-values in data mining",
  "p-Hacking simulation",      paste0(n_significant, " of 100 random tests p < 0.05"),
                                                                            "Selective reporting inflates false discovery rate",
  "Bonferroni vs. BH",         "BH retains more true discoveries",          "BH is the standard for exploratory analysis",
  "Business connection",       "FDR correction increases p_A confidence",   "Feeds into Expected Value Framework"
) %>%
  gt() %>%
  tab_style(cell_fill("#f5f0fa"), cells_column_labels()) %>%
  tab_style(cell_fill("#fff8e1"), cells_body(rows = c(1, 2))) %>%
  tab_style(cell_fill("#f0fff0"), cells_body(rows = c(3, 4))) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-note}
## What Comes Next

Even when a finding is **truly significant** (after FDR correction) — it does not mean it is **practically relevant**. That is the topic of the next block: effect sizes and the difference between statistical and practical significance.

**Dataset used:** `buli_raw_2023-09-06.rds` (Bundesliga players, season 2022/23)
:::
