<!-- ============================================================
INSTRUCTOR CUSTOMISATION — update before rendering for each cohort
Search for the marker text rather than relying on exact line numbers.

  "buli_raw_{date}.rds"   Dataset filename — update date if new scrape
  xfun::from_root(...)    Adjust path if project structure differs
  set.seed(2024)          Change seed for different cohort results
============================================================ -->
---
title: "Session 2 · Lab 2: Stochastic Regression Analysis"
subtitle: "From Descriptive Fit to Inferential Reasoning — Classical and Resampling Approaches"
author: "Prof. Dr. Jörg Schoder"
date: today
lang: en
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    css: ../../../css/styles.css
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse)    # data wrangling & plotting
library(moderndive)   # get_regression_table/summaries/points
library(infer)        # specify-generate-fit
library(gt)           # formatted tables

date <- '2023-09-06'
my_in_file <- glue::glue('buli_raw_{date}.rds')
tbl_buli_raw <- read_rds(file = xfun::from_root("data","raw", my_in_file))

tbl_buli <- tbl_buli_raw %>%
              mutate(
                age       = lubridate::year(Sys.Date()) -
                            lubridate::year(player_dob),
                log_value = log(player_market_value_euro),
                pos_bin   = if_else(player_position == "Goalkeeper",
                                    "Goalkeeper", "Outfield")
              ) %>%
              rename(name = player_name,
                     mv   = player_market_value_euro) %>%
              filter(!is.na(mv), !is.na(age))

options(scipen = 999)
```

## Scenario & Orientation

In Lab 1 we established *that* the age effect is significant — via bootstrap, permutation, and t-test. But we treated `get_regression_table()` as a black box. This lab opens that box.

We start with a fundamental distinction: in Session 1, regression was **descriptive** — we computed the best-fitting line through the data we had. Now we treat that line as an **estimate** of an unknown population relationship. This shift — from conditional mean to conditional expected value — is the conceptual foundation of all statistical inference in regression.

```{r crisp-overview}
#| echo: false
tibble::tribble(
  ~Phase, ~Name,                    ~Content,                                        ~Status,
  "1",    "Business Understanding", "Identify under-/overvalued players",            "Done (Session 1)",
  "2",    "Data Understanding",     "EDA of Bundesliga market values",               "Done (Session 1)",
  "3",    "Data Preparation",       "log-transform, position dummy",                 "Done (Session 1)",
  "4",    "Modelling",              "OLS: simple + interaction model",               "Done (Session 1)",
  "5",    "Evaluation",             "Stochastic regression: classical + resampling", "← This lab",
  "6",    "Deployment",             "Transfer decision framework",                   "Session 2 lecture"
) %>%
  gt() %>%
  tab_style(cell_fill("#f0fff0"), cells_body(rows = Phase == "5")) %>%
  tab_style(cell_fill("#f5f0fa"), cells_column_labels()) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-note}
## Learning Objectives

- Distinguish between the **population model** (Greek letters) and the **sample model** (Latin letters)
- Interpret all columns of `get_regression_table()`, `get_regression_summaries()`, and `get_regression_points()` — connecting each to its resampling counterpart from Lab 1
- Use `infer::fit()` to compute bootstrap CIs and permutation p-values for **multiple regression**
- Assess regression assumptions via **residual diagnostics** and understand when resampling is preferable
- Recognise that regression estimates the **conditional mean** — and that this is only one way to model a distribution
:::

**Navigation guide**

| Symbol | Box colour | Meaning | When to read |
|--------|-----------|---------|--------------|
| ⚠️ | Red | Mandatory task | Live session |
| 💡 | Green | Tip / best practice | Anytime |
| 📚 | Blue (collapsed) | Background theory | At home |
| ℹ️ | Blue (open) | Short clarification | When needed |

:::{.callout-tip}
## ⏱ Time Budget
**50 minutes** — complete mandatory tasks (A1–A6) first; bonus tasks if time permits.
:::

---

## Step 1: Population Model vs. Sample Model — The Stocker Notation

The key conceptual shift in this lab is understanding that every fitted regression is a **sample-based estimate** of an unknown **population relationship**.

### 1.1 Two Models, Two Notations

```{r notation-table}
#| echo: false
tibble::tribble(
  ~Element,                ~`Population (unknown parameters)`,                      ~`Sample (statistics as estimators)`,
  "Model equation",        "$$Y = \\beta_0 + \\beta_1 X + \\varepsilon$$",
                                                                         "$$\\hat{y} = \\hat{\\beta}_0 + \\hat{\\beta}_1 x + e$$",
  "Intercept",             "$$\\beta_0$$",                   "$$\\hat{\\beta}_0$$",
  "Slope",                 "$$\\beta_1$$",                   "$$\\hat{\\beta}_1$$",
  "Error term",            "$$\\varepsilon$$ (epsilon — unobservable)",   "$$\\hat{\\varepsilon} = y - \\hat{y}$$ ",
  "Error variance",        "$$\\sigma^2$$ (unknown)",                     "$$s^2 = \\frac{\\sum \\hat{\\varepsilon}_i^2}{n-k-1}$$ (estimated)",
  "Meaning of the line",   "Conditional expected value: $$E[Y|X]$$",     "Conditional mean: $$\\bar{y}|x$$",
  "Coefficients are...",   "Fixed parameters (constants of nature)",      "Random variables (vary across samples)"
) %>%
  gt() %>%
  fmt_markdown(columns = everything()) %>%
  tab_style(cell_fill("#f5f0fa"), cells_column_labels()) %>%
  tab_style(cell_text(weight = "bold"), cells_body(columns = Element)) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("Greek letters = population (unknown); Latin letters = sample (estimated)")
```

The population model is what we want to know but can never observe directly. The sample model is what we compute from data. **Every number in a regression output is a sample quantity** — an estimate of an unknown population parameter.

:::{.callout-important}
## The Fundamental Shift
In Session 1, the regression line was a **description** of the data: the conditional mean $\bar{y}|x$. Now it becomes an **estimate** of the conditional expected value $E[Y|X]$. Same line, different interpretation — and the difference is precisely the uncertainty we quantified via bootstrap in Lab 1.
:::

### 1.2 The Conditional Mean Is Only One Parameter

OLS regression estimates **one feature** of the conditional distribution of $Y|X$: the mean. But a distribution has many other features — spread, shape, tails.

```{r quantile-teaser}
#| echo: false
#| fig-height: 4
tbl_buli %>%
  ggplot(aes(x = age, y = log_value)) +
  geom_point(alpha = 0.15, color = "#b3b2b2", size = 0.7) +
  geom_smooth(method = "lm", se = FALSE,
              color = "#D50006", linewidth = 1.2) +
  geom_quantile(quantiles = c(0.1, 0.9),
                color = "#502479", linewidth = 0.8,
                linetype = "dashed") +
  annotate("text", x = 35, y = 17.5, label = "90th percentile",
           color = "#502479", size = 3, hjust = 0) +
  annotate("text", x = 35, y = 13.5, label = "10th percentile",
           color = "#502479", size = 3, hjust = 0) +
  annotate("text", x = 35, y = 15.2, label = "OLS mean",
           color = "#D50006", size = 3, hjust = 0, fontface = "bold") +
  labs(x     = "Age",
       y     = "log(Market Value)",
       title = "OLS estimates the mean — but the spread matters too") +
  theme_minimal(base_size = 11)
```

The red line is OLS — the conditional *mean*. The dashed purple lines show the 10th and 90th percentiles of the conditional distribution (via quantile regression). The spread around the mean is not constant — a pattern we will revisit in the residual diagnostics.

:::{.callout-note collapse="true"}
## 📚 Background: Beyond the Mean — Quantile Regression and Beyond
OLS answers: "What is the *average* market value for a player of age $x$?" But a scout might ask: "What is the market value that only 10% of players of that age exceed?" — that is a question about the *upper tail*, not the mean.

**Quantile regression** (Koenker & Bassett, 1978) estimates conditional *quantiles* instead of the conditional mean. Other approaches model the entire conditional distribution: distributional regression, GAMLSS, or Bayesian methods. These are beyond the scope of ER014, but the key takeaway is: **the mean is a useful summary, but it is not the whole story.** Keep this in mind when interpreting $R^2$ — a low $R^2$ does not mean the model is useless; it means the *mean* does not capture all the variation.
:::

---

## Step 2: The Classical Toolkit — `lm()` and Three moderndive Helpers

The `moderndive` package provides three functions that extract everything from an `lm()` object in tidy, readable form:

| Function | Returns | Population counterpart |
|----------|---------|----------------------|
| `get_regression_table()` | Coefficients, SE, t, p, CI | Inference on $\beta_j$ |
| `get_regression_summaries()` | $R^2$, adj. $R^2$, RMSE, F-stat | Model-level diagnostics |
| `get_regression_points()` | Fitted values $\hat{y}$ and residuals $e$ | Estimates of $\varepsilon$ |

### 2.1 Coefficients: `get_regression_table()`

```{r reg-table}
#| code-fold: false
lm_age <- lm(log_value ~ age, data = tbl_buli)
get_regression_table(lm_age)
```

```{r reg-table-fmt}
#| echo: false
get_regression_table(lm_age) %>%
  gt() %>%
  fmt_number(columns = c(estimate, std_error, statistic,
                         p_value, lower_ci, upper_ci), decimals = 4) %>%
  tab_style(cell_fill("#fff8e1"),
            cells_body(rows = term == "age")) %>%
  tab_options(table.font.size = "14px") %>%
  tab_footnote("Yellow: age coefficient — each column is explained below")
```

Each column maps to a concept from Lab 1:

| Column | What it is | Lab 1 counterpart |
|--------|-----------|-------------------|
| `estimate` | $b_1$ — the sample slope | `obs_slope` from `calculate(stat = "slope")` |
| `std_error` | $\text{SE}(b_1)$ — estimated standard deviation of the sampling distribution | Width of the bootstrap distribution |
| `statistic` | $t = b_1 / \text{SE}(b_1)$ — how many SEs away from zero | Position of `obs_slope` relative to the null distribution |
| `p_value` | Probability of seeing $|t|$ this large if $\beta_1 = 0$ | Permutation p-value |
| `lower_ci`, `upper_ci` | 95% CI assuming normality: $b_1 \pm 1.96 \cdot \text{SE}$ | Bootstrap percentile CI |

:::{.callout-note collapse="true"}
## 📚 Background: Where Does the Standard Error Come From?
The classical SE of the slope coefficient is:

$$\text{SE}(b_1) = \frac{s}{\sqrt{\sum(x_i - \bar{x})^2}}$$

where $s = \sqrt{\frac{\sum e_i^2}{n - 2}}$ is the residual standard error. This formula **assumes** that the errors $\varepsilon$ are i.i.d. with constant variance (homoskedasticity). If that assumption fails, the formula gives the wrong SE — and the bootstrap from Lab 1 is more reliable.
:::

### 2.2 Model-Level Summary: `get_regression_summaries()`

```{r reg-summaries}
#| code-fold: false
get_regression_summaries(lm_age)
```

```{r reg-summaries-fmt}
#| echo: false
get_regression_summaries(lm_age) %>%
  gt() %>%
  fmt_number(decimals = 3) %>%
  tab_style(cell_fill("#fff8e1"),
            cells_body(columns = c(r_squared, statistic))) %>%
  tab_options(table.font.size = "14px") %>%
  tab_footnote("Yellow: R² (how much variation explained) and F-statistic (global significance)")
```

| Column | What it is | Interpretation |
|--------|-----------|----------------|
| `r_squared` | $R^2 = 1 - \text{SSR}/\text{SST}$ | Proportion of variance in $y$ explained by the model |
| `adj_r_squared` | Adjusted for number of predictors | Penalises adding useless predictors |
| `mse` | Mean squared error | Average squared residual |
| `rmse` | $\sqrt{\text{MSE}}$ | Residual standard error $s$ — in the units of $y$ |
| `sigma` | Same as RMSE for simple models | Estimate of $\sigma$ (population error SD) |
| `statistic` | F-statistic | Global test: are *all* slopes jointly zero? (Lab 1, Step 5.2) |
| `p_value` | p-value for F-test | Significance of the overall model |
| `df` | Degrees of freedom | Number of estimated slope parameters |
| `nobs` | Number of observations | Sample size |

### 2.3 Fitted Values and Residuals: `get_regression_points()`

```{r reg-points}
#| code-fold: false
reg_points <- get_regression_points(lm_age)
reg_points %>% head(6)
```

```{r reg-points-fmt}
#| echo: false
reg_points %>%
  head(6) %>%
  gt() %>%
  fmt_number(columns = c(log_value, log_value_hat, residual), decimals = 3) %>%
  tab_style(cell_fill("#fff8e1"),
            cells_body(columns = c(log_value_hat, residual))) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("Yellow: ŷ (fitted value) and e = y − ŷ (residual)")
```

Every row is a player. The `residual` column is the **observable counterpart** of the unobservable population error $\varepsilon$ — and it is the basis for all diagnostics in Step 5.

---

## Step 3: Interpreting SE, t, p, CI — The Bridge to Lab 1

Having seen the three tables, we now connect every classical quantity to its resampling counterpart. The point is not that one is "better" — it is that **they measure the same thing**.

```{r bridge-table}
#| echo: false
tibble::tribble(
  ~`Classical (parametric)`,                       ~`Resampling (Lab 1)`,                          ~`What it measures`,
  "SE from formula",                                "SD of bootstrap distribution",                 "Precision of the estimate",
  "t = estimate / SE",                              "Position of obs_stat in null distribution",    "Signal-to-noise ratio",
  "p-value from t-distribution",                    "Permutation p-value",                          "Extremity under H₀",
  "CI = estimate ± 1.96 · SE",                      "Bootstrap percentile CI",                      "Plausible range for β",
  "F-statistic (get_regression_summaries)",          "Permutation F-stat (Lab 1, Step 5.3)",         "Global model significance"
) %>%
  gt() %>%
  tab_style(cell_fill("#f5f0fa"), cells_column_labels()) %>%
  tab_style(cell_fill("#f0fff0"), cells_body()) %>%
  tab_options(table.font.size = "13px")
```

When do the two approaches **diverge**? When the classical assumptions (normality, homoskedasticity) are violated. We check those assumptions in Step 5.

---

## Step 4: Multiple Regression with `infer::fit()`

In Lab 1 we used `calculate(stat = "slope")` for a single predictor. For the interaction model `log_value ~ age * pos_bin`, we need `fit()` — which returns bootstrap or permutation results for **all coefficients at once**.

### 4.1 Classical Baseline

```{r lm-interaction}
#| code-fold: false
lm_inter <- lm(log_value ~ age * pos_bin, data = tbl_buli)
get_regression_table(lm_inter)
```

```{r lm-interaction-fmt}
#| echo: false
get_regression_table(lm_inter) %>%
  gt() %>%
  fmt_number(columns = c(estimate, std_error, statistic,
                         p_value, lower_ci, upper_ci), decimals = 4) %>%
  tab_style(cell_fill("#fff8e1"),
            cells_body(rows = p_value < 0.05)) %>%
  tab_options(table.font.size = "13px") %>%
  tab_footnote("Yellow = significant at α = 0.05")
```

### 4.2 Bootstrap CIs via `fit()`

```{r boot-fit}
#| code-fold: false
#| cache: true
set.seed(2024)

# --- observed coefficients ---
obs_fit <- tbl_buli %>%
             specify(log_value ~ age * pos_bin) %>%
             fit()

# --- bootstrap ---
boot_fit <- tbl_buli %>%
              specify(log_value ~ age * pos_bin) %>%
              generate(reps = 1000, type = "bootstrap") %>%
              fit()

# --- bootstrap CIs ---
ci_fit <- boot_fit %>%
            get_confidence_interval(
              point_estimate = obs_fit,
              type           = "percentile",
              level          = 0.95
            )
ci_fit
```

```{r boot-fit-viz}
#| echo: false
#| fig-height: 6
boot_fit %>%
  visualize() +
  shade_confidence_interval(endpoints = ci_fit) +
  facet_wrap(~ term, scales = "free_x") +
  labs(title = "Bootstrap CIs for all coefficients (interaction model)",
       x     = "Coefficient value",
       y     = "Count") +
  theme_minimal(base_size = 10)
```

### 4.3 Permutation Test via `fit()`

```{r perm-fit}
#| code-fold: false
#| cache: true
set.seed(2024)

null_fit <- tbl_buli %>%
              specify(log_value ~ age * pos_bin) %>%
              hypothesize(null = "independence") %>%
              generate(reps = 1000, type = "permute") %>%
              fit()

p_fit <- null_fit %>%
           get_p_value(obs_stat = obs_fit, direction = "two-sided")
p_fit
```

```{r perm-fit-viz}
#| echo: false
#| fig-height: 6
null_fit %>%
  visualize() +
  shade_p_value(obs_stat = obs_fit, direction = "two-sided") +
  facet_wrap(~ term, scales = "free_x") +
  labs(title = "Null distributions — permutation test (interaction model)",
       x     = "Coefficient under H₀",
       y     = "Count") +
  theme_minimal(base_size = 10)
```

### 4.4 Classical vs. Resampling — Side by Side

```{r compare-ci}
#| echo: false
trad_ci <- get_regression_table(lm_inter) %>%
             select(term, lower_ci_trad = lower_ci, upper_ci_trad = upper_ci,
                    p_trad = p_value)

boot_ci <- ci_fit %>%
             rename(lower_ci_boot = lower_ci, upper_ci_boot = upper_ci)

p_boot <- p_fit %>% rename(p_boot = p_value)

trad_ci %>%
  left_join(boot_ci, by = "term") %>%
  left_join(p_boot, by = "term") %>%
  gt() %>%
  fmt_number(decimals = 4) %>%
  tab_spanner("Classical (t-test)", columns = c(lower_ci_trad, upper_ci_trad, p_trad)) %>%
  tab_spanner("Bootstrap / Permutation", columns = c(lower_ci_boot, upper_ci_boot, p_boot)) %>%
  tab_style(cell_fill("#f5f0fa"), cells_column_labels()) %>%
  tab_style(cell_fill("#f0fff0"),
            cells_body(rows = p_trad < 0.05 & p_boot < 0.05)) %>%
  tab_style(cell_fill("#fff8e1"),
            cells_body(rows = (p_trad < 0.05) != (p_boot < 0.05))) %>%
  tab_options(table.font.size = "12px") %>%
  tab_footnote("Green = both methods agree; Yellow = methods disagree")
```

When both methods agree (green rows), we can be confident in the result. If they ever disagree (yellow), the classical assumptions may be violated — and the resampling result is more trustworthy.

---

## Step 5: Residual Diagnostics — When Do Classical Assumptions Matter?

The classical formulas for SE, t, p, and CI rest on assumptions about the residuals. If these assumptions fail, the formulas may give wrong answers — while bootstrap/permutation remain valid. We check three assumptions:

### 5.1 Linearity: Residuals vs. Fitted

```{r resid-fitted}
#| code-fold: false
reg_points_inter <- get_regression_points(lm_inter)
```

```{r resid-fitted-plot}
#| echo: false
reg_points_inter %>%
  ggplot(aes(x = log_value_hat, y = residual)) +
  geom_point(alpha = 0.3, color = "#686868", size = 0.8) +
  geom_hline(yintercept = 0, color = "#D50006", linewidth = 0.8) +
  geom_smooth(method = "loess", se = FALSE,
              color = "#502479", linewidth = 0.9, linetype = "dashed") +
  labs(x     = expression(hat(y)~"(fitted value)"),
       y     = expression(e~"(residual)"),
       title = "Residuals vs. Fitted Values",
       subtitle = "Purple dashed = LOESS smoother; should be flat if linearity holds") +
  theme_minimal(base_size = 11)
```

If the LOESS smoother deviates strongly from the horizontal red line, the linearity assumption is questionable.

### 5.2 Normality: QQ-Plot

```{r qq-plot}
#| echo: false
reg_points_inter %>%
  ggplot(aes(sample = residual)) +
  stat_qq(alpha = 0.3, color = "#686868", size = 0.8) +
  stat_qq_line(color = "#D50006", linewidth = 0.8) +
  labs(x     = "Theoretical quantiles",
       y     = "Sample quantiles",
       title = "Normal Q-Q Plot of Residuals",
       subtitle = "Points should follow the red line if normality holds") +
  theme_minimal(base_size = 11)
```

Deviations in the tails are common with real data. Moderate deviations are acceptable for large samples (CLT protects the t-test); severe deviations favour the bootstrap.

### 5.3 Homoskedasticity: Scale-Location Plot

```{r scale-location}
#| echo: false
reg_points_inter %>%
  mutate(sqrt_abs_resid = sqrt(abs(residual))) %>%
  ggplot(aes(x = log_value_hat, y = sqrt_abs_resid)) +
  geom_point(alpha = 0.3, color = "#686868", size = 0.8) +
  geom_smooth(method = "loess", se = FALSE,
              color = "#502479", linewidth = 0.9, linetype = "dashed") +
  labs(x     = expression(hat(y)~"(fitted value)"),
       y     = expression(sqrt("|e|")),
       title = "Scale-Location Plot",
       subtitle = "If spread increases with fitted values → heteroskedasticity") +
  theme_minimal(base_size = 11)
```

If the purple smoother slopes upward, the error variance is not constant — the classical SEs are biased. Bootstrap SEs remain valid.

:::{.callout-note collapse="true"}
## 📚 Background: Parametric vs. Nonparametric — and BLUE
The classical OLS estimator is **BLUE** — Best Linear Unbiased Estimator — under the Gauss-Markov assumptions (linearity, exogeneity, homoskedasticity, no multicollinearity). "Best" means smallest variance among all linear unbiased estimators.

When these assumptions hold, the parametric SE is more *efficient* (smaller) than the bootstrap SE. When they fail, the parametric SE is *wrong* — and bootstrap is more *honest*.

This is why we always check residual diagnostics: not to decide whether to use regression, but to decide **which standard errors to trust**. We return to BLUE and its implications in Session 3 when we discuss multicollinearity and regularisation.
:::

:::{.callout-note}
## Practical Rule of Thumb
| Diagnostics look clean | Use classical SEs — they are more efficient |
|---|---|
| Diagnostics look problematic | Use bootstrap CIs — they are more robust |
| Not sure | Report both — if they agree, the result is solid |
:::

---

## Mandatory Tasks

:::{.callout-important icon=false}
## ⚠️ A1: Population vs. Sample Notation

Look at the notation table in Step 1.1.

1. What is the key difference between $\beta_1$ and $b_1$?
2. Why is $\varepsilon$ "unobservable" while $e$ is "observable"?
3. A colleague writes: "The slope of our model is 0.067." Is this a statement about $\beta_1$ or $b_1$? Why does the distinction matter?
:::

:::{.callout-important icon=false}
## ⚠️ A2: Reading `get_regression_table()`

Look at the regression table for `lm_age` in Step 2.1.

1. What does the `std_error` for age tell you? Relate it to the bootstrap distribution from Lab 1.
2. The `statistic` column shows a t-value. What is it — conceptually — and when is it "large enough"?
3. The CI is [`r round(get_regression_table(lm_age) %>% filter(term == "age") %>% pull(lower_ci), 4)`, `r round(get_regression_table(lm_age) %>% filter(term == "age") %>% pull(upper_ci), 4)`]. Does it match the bootstrap CI from Lab 1? Why might they differ slightly?
:::

:::{.callout-important icon=false}
## ⚠️ A3: Reading `get_regression_summaries()`

Look at the model summary in Step 2.2.

1. The $R^2$ is `r round(get_regression_summaries(lm_age) %>% pull(r_squared), 3)`. What does this mean in words?
2. Is a low $R^2$ a reason to discard the model? Consider the quantile regression plot in Step 1.2.
3. What does the F-statistic test? How does it relate to the permutation F-test from Lab 1, Step 5.3?
:::

:::{.callout-important icon=false}
## ⚠️ A4: Interaction Model — Classical vs. Resampling

Look at the comparison table in Step 4.4.

1. For which coefficients do both methods agree on significance?
2. The interaction term `age:pos_binOutfield` — what does its sign tell you about the age effect for outfield players vs. goalkeepers?
3. If the two methods disagreed for a coefficient, which would you trust more? Why?
:::

:::{.callout-important icon=false}
## ⚠️ A5: Residual Diagnostics

Examine the three diagnostic plots in Step 5.

1. Does the residuals-vs-fitted plot suggest a linearity problem?
2. Does the QQ-plot suggest serious non-normality?
3. Does the scale-location plot suggest heteroskedasticity?
4. Given your answers: are the classical SEs from `get_regression_table()` trustworthy for this model?
:::

:::{.callout-important icon=false}
## ⚠️ A6: The Conditional Mean Is Not Everything

Look at the quantile regression plot in Step 1.2.

1. What does the gap between the 10th and 90th percentile lines tell you?
2. For a 30-year-old player, roughly how large is the spread between the 10th and 90th percentile of log market value?
3. Why might a scout care more about the upper quantile than the mean?
:::

---

## Bonus Tasks

:::{.callout-tip icon=false}
## 💡 B1: Robust Standard Errors

The `estimatr` package provides `lm_robust()` which computes heteroskedasticity-consistent (HC) standard errors. Compare:

```r
# install.packages("estimatr")  # if needed
library(estimatr)
lm_robust(log_value ~ age, data = tbl_buli, se_type = "HC2") %>%
  tidy() %>%
  select(term, estimate, std.error, p.value, conf.low, conf.high)
```

How do the HC2 standard errors compare to the classical SEs and the bootstrap SEs?
:::

:::{.callout-tip icon=false}
## 💡 B2: Bootstrap for `get_regression_summaries()` Quantities

Bootstrap the $R^2$ of the simple age model:

```r
set.seed(2024)
boot_r2 <- map_dbl(1:1000, function(i) {
  tbl_buli %>%
    slice_sample(n = nrow(tbl_buli), replace = TRUE) %>%
    lm(log_value ~ age, data = .) %>%
    summary() %>%
    pluck("r.squared")
})
quantile(boot_r2, c(0.025, 0.975))
```

How variable is $R^2$ across samples? Does it ever reach zero?
:::

:::{.callout-tip icon=false}
## 💡 B3: Residual Diagnostics for the Simple Model

Repeat the three diagnostic plots from Step 5 for the simple model `lm_age` (without interaction). Do the diagnostics look better or worse? What does this tell you about the interaction term's role?
:::

:::{.callout-tip icon=false}
## 💡 B4: Influence Points

Use `get_regression_points()` to identify the 5 players with the largest absolute residuals. Who are they? Are they potential outliers — or just players that the model cannot explain well?

```r
get_regression_points(lm_inter) %>%
  mutate(abs_resid = abs(residual)) %>%
  arrange(desc(abs_resid)) %>%
  head(5)
```
:::

---

## Summary

```{r summary-table}
#| echo: false
tibble::tribble(
  ~`Finding`,                            ~`Key Result`,                                      ~`Implication`,
  "Population vs. sample notation",      "Greek = unknown; Latin = estimated",                "Every regression output is an estimate, not truth",
  "get_regression_table()",              "SE, t, p, CI — all have resampling counterparts",   "Classical and resampling measure the same things",
  "get_regression_summaries()",          paste0("R² = ", round(get_regression_summaries(lm_age) %>% pull(r_squared), 3)),
                                                                                              "Low R² ≠ useless model; mean ≠ whole distribution",
  "fit() for multiple regression",       "Bootstrap CIs for all coefficients at once",         "Scales resampling to any model complexity",
  "Classical ≈ Resampling (when valid)", "CIs agree when assumptions hold",                   "Diagnostics determine which SEs to trust",
  "Residual diagnostics",               "Check linearity, normality, homoskedasticity",       "→ BLUE assumptions revisited in Session 3"
) %>%
  gt() %>%
  tab_style(cell_fill("#f5f0fa"), cells_column_labels()) %>%
  tab_style(cell_fill("#f0fff0"), cells_body(rows = c(1, 2, 4))) %>%
  tab_style(cell_fill("#fff8e1"), cells_body(rows = c(3, 5, 6))) %>%
  tab_options(table.font.size = "13px")
```

:::{.callout-note}
## What Comes Next

In **Lab 3** we confront the **multiplicity problem**: when you test many hypotheses at once, false discoveries are inevitable. We simulate p-hacking, apply the Benjamini-Hochberg procedure, and connect everything to the **Expected Value Framework** — because the business decision depends on the *cost* of errors, not just their probability.

**Dataset used:** `buli_raw_2023-09-06.rds` (Bundesliga players, season 2022/23)
:::
